### 2023-09-22 paper/paper.qmd
Update with docs and quarto template for paper

---
title: "Preparing your manuscript"
format:
    agu-pdf:
        keep-tex: true
    agu-html: default
author:
  - name: Florian Börgel
    affiliations: 
      - ref: IOW
    orcid: 0000-0003-2402-304X
    email: florian.boergel@io-warnemuende.de
    corresponding: true

  - name: Sven Karsten
    affiliations:
      - ref: IOW
    orcid: 0000-0001-8715-9476
    email: sven.karsten@io-warnemuende.de

affiliations:
  - id: IOW
    name: Leibniz-Institute for Baltic Sea Research Warnemünde

abstract: |
  The abstract (1) states the nature of the investigation and (2) summarizes the important conclusions. The abstract should be suitable for indexing. Your abstract should:

    - Be set as a single paragraph.
    - Be less than 250 words for all journals except GRL, for which the limit is 150 words.
    - Not include table or figure mentions.
    - Avoid reference citations unless dependent on or directly related to another paper (e.g., companion, comment, reply, or commentary on another paper(s)). AGU’s Style Guide discusses formatting citations in abstracts.
    - Define all abbreviations.
plain-language-summary: |
  A Plain Language Summary (PLS) can be an incredibly effective science communication tool. By summarizing your paper in non-technical terms, you can explain your research and its relevance to a much broader audience. A PLS is required for submissions to AGU Advances, G-Cubed, GeoHealth, GRL, JAMES, JGR: Biogeosciences, JGR: Oceans, JGR: Planets, JGR: Solid Earth, JGR: Atmospheres, Space Weather, and Reviews of Geophysics, but optional for other journals.  A PLS should be no longer than 200 words and should be free of jargon, acronyms, equations, and any technical information that would be unknown to people from outside your scientific discipline. Read our tips for creating an effective PLS.
keywords: []
key-points:
  - Key Points convey the main points and conclusions of the article. 
  - Up to three key point statements are allowed, and each is limited to at most 140 characters with no abbreviations.
  - Key Points are not included in the word count.
bibliography: bibliography.bib  
citation:
  container-title: Geophysical Research Letters
keep-tex: true
date: last-modified
---

## Section Heading

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum hendrerit facilisis velit sit amet malesuada. Phasellus ornare nibh augue, maximus sodales ex tristique vitae. Vivamus non sollicitudin orci, aliquam placerat metus. Maecenas volutpat orci felis, vel finibus urna consectetur sed. Integer in dui ac dui mollis imperdiet. Quisque sed dapibus nibh. Aenean non luctus leo. Phasellus luctus mauris id aliquet dictum. Aliquam fermentum semper massa, vel dignissim nibh dictum et. See @Hubbard2021.

Phasellus interdum tincidunt ex, a euismod massa pulvinar at. Ut fringilla ut nisi nec volutpat. Morbi imperdiet congue tincidunt. Vivamus eget rutrum purus. Etiam et pretium justo. Donec et egestas sem. Donec molestie ex sit amet viverra egestas. Nullam justo nulla, fringilla at iaculis in, posuere non mauris. Ut eget imperdiet elit.

In luctus mauris vitae imperdiet luctus. Morbi volutpat ligula ut tortor fermentum, eu ornare felis luctus. Donec semper diam vitae mattis posuere. Suspendisse facilisis purus nisi, sit amet egestas ex tempor ut. Cras tortor nulla, euismod at fermentum vel, dictum vel justo. Aenean commodo interdum diam nec placerat. Nunc vestibulum felis at est tincidunt, at euismod dui vestibulum. Nulla venenatis tortor at auctor iaculis. Donec consectetur neque ut sagittis ornare. Nullam pharetra felis tempor suscipit efficitur. Curabitur nibh ex, euismod at congue hendrerit, egestas id mi. Duis porttitor neque in commodo elementum. Fusce vitae fermentum nisi, euismod viverra augue. Curabitur at mi pretium, accumsan purus nec, tempus turpis.

Donec non semper dui, quis aliquet est. Quisque quis sapien at massa ultricies egestas. Duis consequat ultricies erat, a pulvinar nisl vestibulum id. Sed tristique turpis ligula, et tempor lectus iaculis at. Vivamus commodo sapien ac turpis vestibulum dapibus. Morbi tristique arcu metus, et laoreet nisi varius nec. Pellentesque habitant morbi tristique senectus et netus et malesuada fames ac turpis egestas. Fusce sit amet nisl at mauris suscipit aliquet. Nulla vitae dignissim urna. Suspendisse sit amet arcu vitae magna blandit mattis. Vivamus convallis efficitur pulvinar. Sed cursus elit nulla. Sed porta, arcu a euismod pretium, odio dui lacinia lacus, ac vulputate nulla augue eget ex. Nullam consequat ligula sit amet mattis aliquam. Nulla risus urna, ultrices vel ullamcorper id, ornare viverra nunc.

Nunc in lobortis lacus. Duis maximus urna leo, varius sodales arcu interdum nec. Pellentesque imperdiet dolor in leo eleifend dapibus. Ut dapibus, lectus non viverra gravida, ipsum ex faucibus tellus, quis iaculis risus tellus eget augue. Nullam a viverra est. Cras velit nisi, interdum in lacus at, vehicula mattis elit. Curabitur eu viverra purus. Proin pellentesque, metus vitae congue convallis, lorem metus feugiat mi, sit amet auctor purus ligula bibendum ante. Nam id justo scelerisque, rhoncus lectus in, fermentum libero. Donec tincidunt egestas ex ac eleifend. Cras faucibus ipsum a nunc faucibus fermentum. Integer et maximus lacus. Nam dictum nibh id viverra convallis.

## Acknowledgments

Phasellus interdum tincidunt ex, a euismod massa pulvinar at. Ut fringilla ut nisi nec volutpat. Morbi imperdiet congue tincidunt. Vivamus eget rutrum purus. Etiam et pretium justo. Donec et egestas sem. Donec molestie ex sit amet viverra egestas. Nullam justo nulla, fringilla at iaculis in, posuere non mauris. Ut eget imperdiet elit.

## Open research

Phasellus interdum tincidunt ex, a euismod massa pulvinar at. Ut fringilla ut nisi nec volutpat. Morbi imperdiet congue tincidunt. Vivamus eget rutrum purus. Etiam et pretium justo. Donec et egestas sem. Donec molestie ex sit amet viverra egestas. Nullam justo nulla, fringilla at iaculis in, posuere non mauris. Ut eget imperdiet elit.

## References {.unnumbered}

:::{#refs}

:::

### 2023-09-22 paper/paper.tex
Update with docs and quarto template for paper

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{agujournal2019}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}
\linenumbers
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Preparing your manuscript},
  pdfauthor={Florian Börgel; Sven Karsten},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\journalname{Geophysical Research Letters}

\draftfalse

\begin{document}
\title{Preparing your manuscript}

\authors{Florian Börgel\affil{1}, Sven Karsten\affil{1}}
\affiliation{1}{Leibniz-Institute for Baltic Sea Research Warnemünde, }
\correspondingauthor{Florian Börgel}{florian.boergel@io-warnemuende.de}

\begin{abstract}
The abstract (1) states the nature of the investigation and (2)
summarizes the important conclusions. The abstract should be suitable
for indexing. Your abstract should:

\begin{itemize}
\tightlist
\item
  Be set as a single paragraph.
\item
  Be less than 250 words for all journals except GRL, for which the
  limit is 150 words.
\item
  Not include table or figure mentions.
\item
  Avoid reference citations unless dependent on or directly related to
  another paper (e.g., companion, comment, reply, or commentary on
  another paper(s)). AGU's Style Guide discusses formatting citations in
  abstracts.
\item
  Define all abbreviations.
\end{itemize}
\end{abstract}

\section*{Plain Language Summary}
A Plain Language Summary (PLS) can be an incredibly effective science
communication tool. By summarizing your paper in non-technical terms,
you can explain your research and its relevance to a much broader
audience. A PLS is required for submissions to AGU Advances, G-Cubed,
GeoHealth, GRL, JAMES, JGR: Biogeosciences, JGR: Oceans, JGR: Planets,
JGR: Solid Earth, JGR: Atmospheres, Space Weather, and Reviews of
Geophysics, but optional for other journals. A PLS should be no longer
than 200 words and should be free of jargon, acronyms, equations, and
any technical information that would be unknown to people from outside
your scientific discipline. Read our tips for creating an effective PLS.

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden, sharp corners]}{\end{tcolorbox}}\fi

\hypertarget{section-heading}{%
\section{Section Heading}\label{section-heading}}

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum
hendrerit facilisis velit sit amet malesuada. Phasellus ornare nibh
augue, maximus sodales ex tristique vitae. Vivamus non sollicitudin
orci, aliquam placerat metus. Maecenas volutpat orci felis, vel finibus
urna consectetur sed. Integer in dui ac dui mollis imperdiet. Quisque
sed dapibus nibh. Aenean non luctus leo. Phasellus luctus mauris id
aliquet dictum. Aliquam fermentum semper massa, vel dignissim nibh
dictum et. See Hubbard et al. (2021).

Phasellus interdum tincidunt ex, a euismod massa pulvinar at. Ut
fringilla ut nisi nec volutpat. Morbi imperdiet congue tincidunt.
Vivamus eget rutrum purus. Etiam et pretium justo. Donec et egestas sem.
Donec molestie ex sit amet viverra egestas. Nullam justo nulla,
fringilla at iaculis in, posuere non mauris. Ut eget imperdiet elit.

In luctus mauris vitae imperdiet luctus. Morbi volutpat ligula ut tortor
fermentum, eu ornare felis luctus. Donec semper diam vitae mattis
posuere. Suspendisse facilisis purus nisi, sit amet egestas ex tempor
ut. Cras tortor nulla, euismod at fermentum vel, dictum vel justo.
Aenean commodo interdum diam nec placerat. Nunc vestibulum felis at est
tincidunt, at euismod dui vestibulum. Nulla venenatis tortor at auctor
iaculis. Donec consectetur neque ut sagittis ornare. Nullam pharetra
felis tempor suscipit efficitur. Curabitur nibh ex, euismod at congue
hendrerit, egestas id mi. Duis porttitor neque in commodo elementum.
Fusce vitae fermentum nisi, euismod viverra augue. Curabitur at mi
pretium, accumsan purus nec, tempus turpis.

Donec non semper dui, quis aliquet est. Quisque quis sapien at massa
ultricies egestas. Duis consequat ultricies erat, a pulvinar nisl
vestibulum id. Sed tristique turpis ligula, et tempor lectus iaculis at.
Vivamus commodo sapien ac turpis vestibulum dapibus. Morbi tristique
arcu metus, et laoreet nisi varius nec. Pellentesque habitant morbi
tristique senectus et netus et malesuada fames ac turpis egestas. Fusce
sit amet nisl at mauris suscipit aliquet. Nulla vitae dignissim urna.
Suspendisse sit amet arcu vitae magna blandit mattis. Vivamus convallis
efficitur pulvinar. Sed cursus elit nulla. Sed porta, arcu a euismod
pretium, odio dui lacinia lacus, ac vulputate nulla augue eget ex.
Nullam consequat ligula sit amet mattis aliquam. Nulla risus urna,
ultrices vel ullamcorper id, ornare viverra nunc.

Nunc in lobortis lacus. Duis maximus urna leo, varius sodales arcu
interdum nec. Pellentesque imperdiet dolor in leo eleifend dapibus. Ut
dapibus, lectus non viverra gravida, ipsum ex faucibus tellus, quis
iaculis risus tellus eget augue. Nullam a viverra est. Cras velit nisi,
interdum in lacus at, vehicula mattis elit. Curabitur eu viverra purus.
Proin pellentesque, metus vitae congue convallis, lorem metus feugiat
mi, sit amet auctor purus ligula bibendum ante. Nam id justo
scelerisque, rhoncus lectus in, fermentum libero. Donec tincidunt
egestas ex ac eleifend. Cras faucibus ipsum a nunc faucibus fermentum.
Integer et maximus lacus. Nam dictum nibh id viverra convallis.

\hypertarget{acknowledgments}{%
\section{Acknowledgments}\label{acknowledgments}}

Phasellus interdum tincidunt ex, a euismod massa pulvinar at. Ut
fringilla ut nisi nec volutpat. Morbi imperdiet congue tincidunt.
Vivamus eget rutrum purus. Etiam et pretium justo. Donec et egestas sem.
Donec molestie ex sit amet viverra egestas. Nullam justo nulla,
fringilla at iaculis in, posuere non mauris. Ut eget imperdiet elit.

\hypertarget{open-research}{%
\section{Open research}\label{open-research}}

Phasellus interdum tincidunt ex, a euismod massa pulvinar at. Ut
fringilla ut nisi nec volutpat. Morbi imperdiet congue tincidunt.
Vivamus eget rutrum purus. Etiam et pretium justo. Donec et egestas sem.
Donec molestie ex sit amet viverra egestas. Nullam justo nulla,
fringilla at iaculis in, posuere non mauris. Ut eget imperdiet elit.

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\vspace{1em}

\leavevmode\vadjust pre{\hypertarget{ref-Hubbard2021}{}}%
Hubbard, B., Christoffersen, P., Doyle, S. H., Chudley, T. R.,
Schoonman, C. M., Law, R., \& Bougamont, M. (2021). Borehole-based
characterization of deep mixed-mode crevasses at a greenlandic outlet
glacier. \emph{{AGU} Advances}, \emph{2}(2).
\url{https://doi.org/10.1029/2020av000291}

\end{CSLReferences}

\end{document}

### 2023-09-25 BalticRiverPrediction/BaltNet.py
small errors, checkpointing and proofofconcept

@@ -3,7 +3,7 @@
 # %% auto 0
 __all__ = ['BaltNet', 'BaseLineModel', 'LightningModel', 'AtmosphericDataset', 'AtmosphereDataModule']
 
-# %% ../nbs/02_BaltNet.ipynb 1
+# %% ../nbs/02_BaltNet.ipynb 2
 import torch
 import torch.nn as nn
 import lightning as L
@@ -23,7 +23,7 @@ from tqdm import tqdm
 from .convLSTM import ConvLSTM
 from .sharedUtilities import read_netcdfs, preprocess, plot_loss_and_acc
 
-# %% ../nbs/02_BaltNet.ipynb 2
+# %% ../nbs/02_BaltNet.ipynb 3
 class BaltNet(nn.Module):
     def __init__(self, modelPar):
         super(BaltNet, self).__init__()
@@ -71,7 +71,7 @@ class BaltNet(nn.Module):
         return x
 
 
-# %% ../nbs/02_BaltNet.ipynb 3
+# %% ../nbs/02_BaltNet.ipynb 4
 class BaseLineModel(nn.Module):
     def __init__(self, modelPar):
         super(BaseLineModel, self).__init__()
@@ -93,7 +93,7 @@ class BaseLineModel(nn.Module):
         x = self.fc_layers(x).squeeze()
         return x
 
-# %% ../nbs/02_BaltNet.ipynb 4
+# %% ../nbs/02_BaltNet.ipynb 5
 class LightningModel(L.LightningModule):
     def __init__(self, model, learning_rate, cosine_t_max):
         super().__init__()
@@ -153,7 +153,7 @@ class LightningModel(L.LightningModule):
         return [opt], [sch]
 
 
-# %% ../nbs/02_BaltNet.ipynb 5
+# %% ../nbs/02_BaltNet.ipynb 6
 class AtmosphericDataset(Dataset):
     def __init__(self, datapath, transform=None):
 
@@ -183,8 +183,9 @@ class AtmosphericDataset(Dataset):
 
         self.rainData = (rainDataMean, rainDataSTD)
         self.runoffData = (runoffDataMean, runoffDataSTD)
+        self.timeRange = (str(rainData.time.min().data), str(rainData.time.max().data))
 
-        np.savetext(
+        np.savetxt(
             "/silor/boergel/paper/runoff_prediction/data/runoffMeanStd.txt",
             [runoffDataMean, runoffDataSTD]
         )
[…]

### 2023-09-25 BalticRiverPrediction/sharedUtilities.py

@@ -3,7 +3,7 @@
 # %% auto 0
 __all__ = ['preprocess', 'read_netcdfs', 'transform_calendar', 'plot_loss_and_acc']
 
-# %% ../nbs/03_shared_utilities.ipynb 1
+# %% ../nbs/03_shared_utilities.ipynb 2
 import xarray as xr
 from glob import glob
 from tqdm import tqdm 
@@ -12,12 +12,12 @@ import numpy as np
 import pandas as pd
 import matplotlib.pyplot as plt
 
-# %% ../nbs/03_shared_utilities.ipynb 3
+# %% ../nbs/03_shared_utilities.ipynb 4
 def preprocess(ds):
     return ds.sel(time=str(ds.time[len(ds.time)//2].dt.year.data)).resample(time="1D").mean()
 
 
-# %% ../nbs/03_shared_utilities.ipynb 4
+# %% ../nbs/03_shared_utilities.ipynb 5
 def read_netcdfs(files, dim, transform_func, transform_calendar=None, cftime = True):
     """Reads multiples netcdfs files"""
     def process_one_path(path):
@@ -44,7 +44,7 @@ def read_netcdfs(files, dim, transform_func, transform_calendar=None, cftime = T
     combined = xr.concat(datasets, dim)
     return combined
 
-# %% ../nbs/03_shared_utilities.ipynb 5
+# %% ../nbs/03_shared_utilities.ipynb 6
 def transform_calendar(ds,
                        timedim="time",
                        calendarname="proleptic_gregorin"):
@@ -52,7 +52,7 @@ def transform_calendar(ds,
     ds[timedim].attrs['calendar'] = calendarname
     return ds
 
-# %% ../nbs/03_shared_utilities.ipynb 6
+# %% ../nbs/03_shared_utilities.ipynb 7
 def plot_loss_and_acc(
     log_dir, loss_ylim=(0.0, 0.9), acc_ylim=(0.7, 1.0), save_loss=None, save_acc=None
 ):

### 2023-09-25 src/runMultipleTest.py

@@ -58,7 +58,14 @@ if __name__ == "__main__":
 
     # save best model 
     callbacks = [
-        ModelCheckpoint(save_top_k=1, mode="max", monitor="val_mse", save_last=True)
+        ModelCheckpoint(
+            dirpath="/silor/boergel/paper/runoff_prediction/data/modelWeights/",
+            filename="BaltNetTopOne",
+            save_top_k=1,
+            mode="max",
+            monitor="val_mse",
+            save_last=True
+        )
     ]
 
     trainer = L.Trainer(
@@ -71,7 +78,6 @@ if __name__ == "__main__":
             name="BaltNet1"
         ),
         deterministic=True,
-        default_root_dir="/silor/boergel/paper/runoff_prediction/data/modelWeights/"
     )
 
     trainer.fit(model=LighningBaltNet, datamodule=dataLoader)

### 2023-09-27 BalticRiverPrediction/BaltNet.py
Implement ray hyperparameter tuner

@@ -7,6 +7,7 @@ __all__ = ['BaltNet', 'BaseLineModel', 'LightningModel', 'AtmosphericDataset', '
 import torch
 import torch.nn as nn
 import lightning as L
+import pytorch_lightning as pl
 from lightning.pytorch.callbacks import ModelCheckpoint
 import numpy as np
 
@@ -94,7 +95,7 @@ class BaseLineModel(nn.Module):
         return x
 
 # %% ../nbs/02_BaltNet.ipynb 5
-class LightningModel(L.LightningModule):
+class LightningModel(pl.LightningModule):
     def __init__(self, model, learning_rate, cosine_t_max):
         super().__init__()
 
@@ -155,35 +156,25 @@ class LightningModel(L.LightningModule):
 
 # %% ../nbs/02_BaltNet.ipynb 6
 class AtmosphericDataset(Dataset):
-    def __init__(self, datapath, transform=None):
+    def __init__(self, input_size, data, runoff, transform=None):
 
-        data = read_netcdfs(
-            files=f"{datapath}/atmosphericForcing/????/rain.mom.dta.nc",
-            dim="time",
-            transform_func=lambda ds:preprocess(ds)
-            )       
 
-        runoff = read_netcdfs(
-            f"{datapath}/runoffData/combined_fastriver_*.nc",
-            dim="river",
-            transform_func= lambda ds:ds.roflux.resample(time="1D").mean(),
-            cftime=False
-            )
-        
-        runoff = runoff.transpose("time", "river")
-        runoff = runoff.sel(time=slice(None, str(2017)))
+        # Following the technical description of the river data (Germo et al.)
+        # the original river data is limited to 1979 to 2011
 
-        rainData = data["RAIN"]
-        runoffData = runoff.sel(time=slice(str(rainData.time.min().data), str(rainData.time.max().data)))
+        # runoffData = runoff.sel(time=slice(str(1979), str(2011)))
+        runoffData = runoff.transpose("time", "river")
+        runoffDataMean = runoff.mean()
+        runoffDataSTD = runoff.std()
 
+        rainData = data["RAIN"].sel(time=slice(str(1979), str(2011)))
         rainDataMean = rainData.mean()
         rainDataSTD = rainData.std()
-        runoffDataMean = runoffData.mean("time")
-        runoffDataSTD = runoffData.std("time")
 
         self.rainData = (rainDataMean, rainDataSTD)
         self.runoffData = (runoffDataMean, runoffDataSTD)
[…]

### 2023-09-27 src/runMultipleTest.py

@@ -34,7 +34,7 @@ if __name__ == "__main__":
     modelParameters = {
     "input_dim":30, # timesteps
     "hidden_dim":1, # Channels -> right now only precipitation
-    "kernel_size":(3,3), # applied for spatial convolutions
+    "kernel_size":(4,4), # applied for spatial convolutions
     "num_layers":2, # number of convLSTM layers
     "batch_first":True, # first index is batch
     "bias":True, 

### 2023-09-27 src/runTuning.py

@@ -1,0 +1,132 @@
+import torch
+import torch.nn as nn
+import pytorch_lightning as pl
+
+from BalticRiverPrediction.BaltNet import BaltNet
+from BalticRiverPrediction.BaltNet import LightningModel
+from BalticRiverPrediction.BaltNet import AtmosphereDataModule
+from BalticRiverPrediction.sharedUtilities import read_netcdfs, preprocess
+
+import ray
+from ray import tune
+import ray.train.lightning
+from ray.train import RunConfig, ScalingConfig, CheckpointConfig
+from ray.train.torch import TorchTrainer
+from ray.tune.schedulers import ASHAScheduler
+
+
+def train_func(config):
+    torch.set_float32_matmul_precision("medium")
+  
+    # Loads the atmospheric data in batches
+    dataLoader = AtmosphereDataModule(
+    data=data,
+    runoff=runoff,
+    batch_size=64,
+    input_size=config["input_dim"]
+    )
+
+    num_epochs = 50
+
+    # initalize model
+    pyTorchBaltNet = BaltNet(modelPar=config)
+
+    # Lightning model wrapper
+    LighningBaltNet = LightningModel(
+        pyTorchBaltNet,
+        learning_rate=1e-3,
+        cosine_t_max=num_epochs
+    )
+
+    trainer = pl.Trainer(
+        accelerator="auto",
+        devices="auto",
+        strategy=ray.train.lightning.RayDDPStrategy(),
+        callbacks=[ray.train.lightning.RayTrainReportCallback()],
+        plugins=[ray.train.lightning.RayLightningEnvironment()],
+        max_epochs=num_epochs
+    )
+
+    trainer = ray.train.lightning.prepare_trainer(trainer)
+    trainer.fit(model=LighningBaltNet, datamodule=dataLoader)
+
+def tune_BaltNet_asha(ray_trainer, config, num_samples=1):
+    scheduler = ASHAScheduler(max_t=50, grace_period=1, reduction_factor=2)
+
+    tuner = tune.Tuner(
+        ray_trainer,
+        param_space={"train_loop_config": config},
+        tune_config=tune.TuneConfig(
+            metric="val_mse",
+            mode="max",
+            num_samples=num_samples,
+            scheduler=scheduler,
+            max_concurrent_trials=4
+        ),
+    )
+    return tuner.fit()
+
+if __name__ == "__main__":
+
+    torch.set_float32_matmul_precision("medium")
+
[…]

### 2023-09-27 BalticRiverPrediction/BaltNet.py
Update model, update tuning, introduce tensorboard

@@ -57,8 +57,10 @@ class BaltNet(nn.Module):
 
         self.fc_layers = torch.nn.Sequential(
             torch.nn.Linear(self.linear_dim, 512),
+            torch.nn.BatchNorm1d(512),  
             torch.nn.ReLU(),
             torch.nn.Linear(512, 256),
+            torch.nn.BatchNorm1d(256),  
             torch.nn.ReLU(),
             torch.nn.Linear(256, 97)
             )
@@ -129,17 +131,15 @@ class LightningModel(pl.LightningModule):
     
     def validation_step(self, batch, batch_idx):
         loss, true_labes, predicted_labels = self._shared_step(batch)
+        mse = self.val_mse(predicted_labels, true_labes)
         self.log("val_loss", loss, sync_dist=True)
-        self.val_mse(predicted_labels, true_labes)
-        self.log(
-            "val_mse", self.val_mse, prog_bar=True, sync_dist=True
-        )
+        self.log("val_mse", mse, prog_bar=True, sync_dist=True)
     
     def test_step(self, batch, _):
         loss, true_labels, predicted_labels = self._shared_step(batch)
-        self.test_mse(predicted_labels, true_labels)
+        mse = self.test_mse(predicted_labels, true_labels)
         self.log("test_loss", loss, rank_zero_only=True)
-        self.log("test_mse", self.test_mse, sync_dist=True)
+        self.log("test_mse", mse, sync_dist=True)
         return loss
     
     def predict_step(self, batch, batch_idx: int, dataloader_idx: int = 0):
@@ -148,7 +148,7 @@ class LightningModel(pl.LightningModule):
 
     
     def configure_optimizers(self):
-        opt = torch.optim.Adam(self.parameters(), lr=self.learning_rate)
+        opt = torch.optim.AdamW(self.parameters(), lr=self.learning_rate)
         sch = torch.optim.lr_scheduler.CosineAnnealingLR(opt, T_max=self.cosine_t_max)
 
         return [opt], [sch]
@@ -162,37 +162,39 @@ class AtmosphericDataset(Dataset):
         # Following the technical description of the river data (Germo et al.)
         # the original river data is limited to 1979 to 2011
 
-        # runoffData = runoff.sel(time=slice(str(1979), str(2011)))
-        runoffData = runoff.transpose("time", "river")
-        runoffDataMean = runoff.mean()
-        runoffDataSTD = runoff.std()
[…]

### 2023-09-27 src/runMultipleTest.py

@@ -31,6 +31,8 @@ if __name__ == "__main__":
     batch_size=64
     )
 
+    # Note that this set of parameters will be defined by runTuning.py
+
     modelParameters = {
     "input_dim":30, # timesteps
     "hidden_dim":1, # Channels -> right now only precipitation

### 2023-09-27 src/runTuning.py

@@ -12,7 +12,7 @@ from ray import tune
 import ray.train.lightning
 from ray.train import RunConfig, ScalingConfig, CheckpointConfig
 from ray.train.torch import TorchTrainer
-from ray.tune.schedulers import ASHAScheduler
+from ray.tune.schedulers import FIFOScheduler
 
 
 def train_func(config):
@@ -50,18 +50,18 @@ def train_func(config):
     trainer = ray.train.lightning.prepare_trainer(trainer)
     trainer.fit(model=LighningBaltNet, datamodule=dataLoader)
 
-def tune_BaltNet_asha(ray_trainer, config, num_samples=1):
-    scheduler = ASHAScheduler(max_t=50, grace_period=1, reduction_factor=2)
+def tune_BaltNet_FIFO(ray_trainer, config, num_samples=1):
+    scheduler = FIFOScheduler()
 
     tuner = tune.Tuner(
         ray_trainer,
         param_space={"train_loop_config": config},
         tune_config=tune.TuneConfig(
             metric="val_mse",
-            mode="max",
+            mode="min",
             num_samples=num_samples,
             scheduler=scheduler,
-            max_concurrent_trials=4
+            max_concurrent_trials=2
         ),
     )
     return tuner.fit()
@@ -84,7 +84,9 @@ if __name__ == "__main__":
         transform_func= lambda ds:ds.sel(time=slice(str(1979), str(2011))).roflux.resample(time="1D").mean(),
         cftime=False
         )  
-    ray.init()
+    
+    #TODO: May be not needed to call ray.init()
+    ray.init(num_gpus=2)
 
     modelParameters = {
     "input_dim": tune.choice([32,64,128]), # timesteps
@@ -97,36 +99,30 @@ if __name__ == "__main__":
     "dimensions": (191, 206) # dimensions of atmospheric forcing
     }
 
+    # Scaling config provides the resource allocated for the trainer
     scaling_config = ScalingConfig(
         num_workers=1,
         use_gpu=True,
         resources_per_worker={"GPU":1, "CPU":8}
         )
 
+    # Runtime configuration for training and tuning runs
     run_config = RunConfig(
     checkpoint_config=CheckpointConfig(
         num_to_keep=2,
         checkpoint_score_attribute="val_mse",
         checkpoint_score_order="min",
         ),
+        local_dir="/silor/boergel/paper/runoff_prediction/data/",
[…]

### 2023-09-28 BalticRiverPrediction/BaltNet.py
Tuning procedure finished

@@ -57,10 +57,8 @@ class BaltNet(nn.Module):
 
         self.fc_layers = torch.nn.Sequential(
             torch.nn.Linear(self.linear_dim, 512),
-            torch.nn.BatchNorm1d(512),  
             torch.nn.ReLU(),
             torch.nn.Linear(512, 256),
-            torch.nn.BatchNorm1d(256),  
             torch.nn.ReLU(),
             torch.nn.Linear(256, 97)
             )
@@ -148,7 +146,7 @@ class LightningModel(pl.LightningModule):
 
     
     def configure_optimizers(self):
-        opt = torch.optim.AdamW(self.parameters(), lr=self.learning_rate)
+        opt = torch.optim.Adam(self.parameters(), lr=self.learning_rate)
         sch = torch.optim.lr_scheduler.CosineAnnealingLR(opt, T_max=self.cosine_t_max)
 
         return [opt], [sch]
@@ -207,13 +205,14 @@ class AtmosphericDataset(Dataset):
 class AtmosphereDataModule(pl.LightningDataModule):
     def __init__(self, data, runoff, batch_size=64, num_workers=8, add_first_dim=True, input_size=30):
         super().__init__()
+
+        self.data = data
+        self.runoff = runoff
         self.batch_size = batch_size
         self.num_workers = num_workers
         self.add_first_dim = add_first_dim
         self.input_size = input_size
-        self.data = data
-        self.runoff = runoff
-
+    
     def setup(self, stage:str):
         UserWarning("Loading atmospheric data ...")
         dataset = AtmosphericDataset(data=self.data, runoff=self.runoff, input_size=self.input_size)

### 2023-09-28 src/runMultipleTest.py

@@ -1,52 +1,64 @@
 
 import torch
-import torch.nn as nn
 import lightning as L
 
 from lightning.pytorch.callbacks import ModelCheckpoint
 from lightning.pytorch.loggers import CSVLogger
 
-from torch.utils.data import DataLoader, Dataset
-from torch.utils.data.dataset import random_split
-
 from BalticRiverPrediction.BaltNet import BaltNet
 from BalticRiverPrediction.BaltNet import LightningModel
 from BalticRiverPrediction.BaltNet import AtmosphereDataModule
+from BalticRiverPrediction.sharedUtilities import read_netcdfs, preprocess
 
 if __name__ == "__main__":
 
-    # Our GPU has tensor cores, hence mixed precision training is enabled
-    # see https://sebastianraschka.com/blog/2023/llm-mixed-precision-copy.html
-    # for more
-
-    torch.set_float32_matmul_precision("medium")
-    
-    # set random seed for reproducibility
+    # set seed for reproducibility   
     L.pytorch.seed_everything(123)
 
+    # Our GPU has tensor cores, hence mixed precision training is enabled
+    # see https://sebastianraschka.com/blog/2023/llm-mixed-precision-copy.html
 
-    # Loads the atmospheric data in batches
-    dataLoader = AtmosphereDataModule(
-    datapath="/silor/boergel/paper/runoff_prediction/data",
-    batch_size=64
-    )
+    torch.set_float32_matmul_precision("medium")
+
+    datapath="/silor/boergel/paper/runoff_prediction/data"
+
+    data = read_netcdfs(
+        files=f"{datapath}/atmosphericForcing/????/rain.mom.dta.nc",
+        dim="time",
+        transform_func=lambda ds:preprocess(ds)
+        )       
+
+    runoff = read_netcdfs(
+        f"{datapath}/runoffData/combined_fastriver_*.nc",
+        dim="river",
+        transform_func= lambda ds:ds.sel(time=slice(str(1979), str(2011))).roflux.resample(time="1D").mean(),
+        cftime=False
+        )   
 
     # Note that this set of parameters will be defined by runTuning.py
 
     modelParameters = {
-    "input_dim":30, # timesteps
+    "input_dim":16, # timesteps
     "hidden_dim":1, # Channels -> right now only precipitation
-    "kernel_size":(4,4), # applied for spatial convolutions
-    "num_layers":2, # number of convLSTM layers
+    "kernel_size":(5,5), # applied for spatial convolutions
[…]

### 2023-09-28 src/runTuning.py

@@ -22,7 +22,7 @@ def train_func(config):
     dataLoader = AtmosphereDataModule(
     data=data,
     runoff=runoff,
-    batch_size=64,
+    batch_size=16,
     input_size=config["input_dim"]
     )
 
@@ -89,10 +89,10 @@ if __name__ == "__main__":
     ray.init(num_gpus=2)
 
     modelParameters = {
-    "input_dim": tune.choice([32,64,128]), # timesteps
+    "input_dim": tune.choice([16,32]), # timesteps
     "hidden_dim":1, # Channels -> right now only precipitation
-    "kernel_size": tune.choice([(3,3),(5,5),(7,7)]), # applied for spatial convolutions, only uneven numbers
-    "num_layers": tune.choice([2,3,4]), # number of convLSTM layers
+    "kernel_size": tune.choice([(5,5),(7,7)]), # applied for spatial convolutions, only uneven numbers
+    "num_layers": tune.choice([3,4,7]), # number of convLSTM layers
     "batch_first":True, # first index is batch
     "bias":True, 
     "return_all_layers": False, 
@@ -103,7 +103,7 @@ if __name__ == "__main__":
     scaling_config = ScalingConfig(
         num_workers=1,
         use_gpu=True,
-        resources_per_worker={"GPU":1, "CPU":8}
+        resources_per_worker={"GPU":1, "CPU":16}
         )
 
     # Runtime configuration for training and tuning runs
@@ -125,4 +125,4 @@ if __name__ == "__main__":
     run_config=run_config,
     )
 
-    results = tune_BaltNet_FIFO(ray_trainer, modelParameters, num_samples=27)
+    results = tune_BaltNet_FIFO(ray_trainer, modelParameters, num_samples=8)

### 2023-09-28 paper/paper.qmd
Force myself to write text

@@
 ---
+## Introduction
 
-## Section Heading
+River runoff is an important component of the global water cycle as it comprises about one third of the precipitation over land areas (Hagemann, 2020).
+Moreover, accurate runoff forecasting, especially over extended periods, is pivotal for effective water resources management, as highlighted by studies such as Yang et al.
+(2018), Tan et al.
+(2018), and Fang et al.
+(2019).
 
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Vestibulum hendrerit facilisis velit sit amet malesuada.
-Phasellus ornare nibh augue, maximus sodales ex tristique vitae.
-Vivamus non sollicitudin orci, aliquam placerat metus.
-Maecenas volutpat orci felis, vel finibus urna consectetur sed.
-Integer in dui ac dui mollis imperdiet.
-Quisque sed dapibus nibh.
-Aenean non luctus leo.
-Phasellus luctus mauris id aliquet dictum.
-Aliquam fermentum semper massa, vel dignissim nibh dictum et.
-See @Hubbard2021.
+Over the past decades, models for long-term runoff forecasting have been primarily bifurcated into physically based models and data-based models.
+While the former attempts to emulate intricate and nonlinear physical hydrological processes, the latter hinges on establishing statistical models that delineate the relationship between large-scale climate patterns and catchment runoff.
 
-Phasellus interdum tincidunt ex, a euismod massa pulvinar at.
-Ut fringilla ut nisi nec volutpat.
-Morbi imperdiet congue tincidunt.
-Vivamus eget rutrum purus.
-Etiam et pretium justo.
-Donec et egestas sem.
-Donec molestie ex sit amet viverra egestas.
-Nullam justo nulla, fringilla at iaculis in, posuere non mauris.
-Ut eget imperdiet elit.
+Machine Learning (ML) models, such as those employing artificial neural networks, support vector machines, adaptive neuro-fuzzy inference systems, and notably, Long Short-Term Memory (LSTM) neural networks, have gained traction for long-term hydrological forecasting due to their commendable performance (Humphrey et al 2016, Huang et al 2014, Ashrafi et al 2017, Yuan et al 2018, Xu et al 2021).
 
-In luctus mauris vitae imperdiet luctus.
-Morbi volutpat ligula ut tortor fermentum, eu ornare felis luctus.
-Donec semper diam vitae mattis posuere.
-Suspendisse facilisis purus nisi, sit amet egestas ex tempor ut.
-Cras tortor nulla, euismod at fermentum vel, dictum vel justo.
-Aenean commodo interdum diam nec placerat.
-Nunc vestibulum felis at est tincidunt, at euismod dui vestibulum.
-Nulla venenatis tortor at auctor iaculis.
-Donec consectetur neque ut sagittis ornare.
-Nullam pharetra felis tempor suscipit efficitur.
-Curabitur nibh ex, euismod at congue hendrerit, egestas id mi.
-Duis porttitor neque in commodo elementum.
-Fusce vitae fermentum nisi, euismod viverra augue.
-Curabitur at mi pretium, accumsan purus nec, tempus turpis.
+LSTM networks, an evolution of the classical Recurrent Neural Networks (RNNs), have shown stability and efficacy in sequence-to-sequence predictions, such as using climatic indices for rainfall estimation or long-term hydrological forecasting.
+However, a limitation of LSTMs is their inability to effectively capture two-dimensional structures, an area where Convolutional Neural Networks (CNNs) excel.
+Recognizing this, we introduce the ConvLSTM, which integrates the strengths of both LSTM and CNN, to extract spatiotemporal features from precipitation fields for predicting river runoff in the Baltic Sea catchment, summarized by 97 inidivual rivers.
 
-Donec non semper dui, quis aliquet est.
-Quisque quis sapien at massa ultricies egestas.
-Duis consequat ultricies erat, a pulvinar nisl vestibulum id.
-Sed tristique turpis ligula, et tempor lectus iaculis at.
-Vivamus commodo sapien ac turpis vestibulum dapibus.
-Morbi tristique arcu metus, et laoreet nisi varius nec.
-Pellentesque habitant morbi tristique senectus et netus et malesuada fames ac turpis egestas.
-Fusce sit amet nisl at mauris suscipit aliquet.
-Nulla vitae dignissim urna.
-Suspendisse sit amet arcu vitae magna blandit mattis.
-Vivamus convallis efficitur pulvinar.
-Sed cursus elit nulla.
-Sed porta, arcu a euismod pretium, odio dui lacinia lacus, ac vulputate nulla augue eget ex.
-Nullam consequat ligula sit amet mattis aliquam.
-Nulla risus urna, ultrices vel ullamcorper id, ornare viverra nunc.
+Modeling the Baltic Sea is to a large part the result of the quality of the freshwater input, that is used for the simulation.
+Meier and Kauker(2003) showed that decadal salinity variations of about 1 $g$ $kg^{-1}$ are caused, inter alia, by annual runoff variations.
+Further, Meier and Kauker (2003) showed that about 50 \% of the decadal salinity variability can be explained by variations in freshwater input into the Baltic Sea.
 
-Nunc in lobortis lacus.
-Duis maximus urna leo, varius sodales arcu interdum nec.
-Pellentesque imperdiet dolor in leo eleifend dapibus.
-Ut dapibus, lectus non viverra gravida, ipsum ex faucibus tellus, quis iaculis risus tellus eget augue.
-Nullam a viverra est.
-Cras velit nisi, interdum in lacus at, vehicula mattis elit.
-Curabitur eu viverra purus.
-Proin pellentesque, metus vitae congue convallis, lorem metus feugiat mi, sit amet auctor purus ligula bibendum ante.
-Nam id justo scelerisque, rhoncus lectus in, fermentum libero.
-Donec tincidunt egestas ex ac eleifend.
-Cras faucibus ipsum a nunc faucibus fermentum.
-Integer et maximus lacus.
-Nam dictum nibh id viverra convallis.
+This paper delves into the application of deep learning, particularly ConvLSTM, to the challenging task of precipitation nowcasting, a domain yet to fully harness the potential of advanced machine learning techniques.
+We present ConvLSTM as a novel solution to this spatiotemporal sequence forecasting challenge, highlighting its advantages and potential future applications.
+
+## Methods
+
+### LSTM network
+
+The Long Short-Term Memory (LSTM), a specialized form of Recurrent Neural Networks (RNNs), is specifically tailored for modeling temporal sequences.
+Its unique design allows it to adeptly handle long-range dependencies, setting it apart from traditional RNNs in terms of accuracy.
+This prowess in modeling long-range dependencies has been validated in various studies [12, 11, 17, 23].
+The cornerstone of LSTM's innovation is its memory cell, $c_t$t, which functions as a repository of state information, also refered to as long-term memory.
+This cell is accessed, modified, and reset through several self-parameterized gates.
+As new inputs arrive, their information is accumulated into the cell if the input gate is activated.
+The forget gate $f_t$ defines the percentage of the previous cell status that is stored $c_{t-1}$​.
+The decision to propagate the latest cell output, $c_t$, to the final state, $h_t$, is governed by the output gate, $o_t$.
+A significant advantage of this architecture is the memory cell's ability to retain gradients.
+This mechanism addresses the vanishing gradient problem, where, as input sequences elongate, the influence of initial stages becomes harder to capture, causing gradients of early input points to approach zero.
+The LSTM's activation function, inherently recurrent, mirrors the identity function with a consistent derivative of 1.0, ensuring the gradient remains stable throughout backpropagation.
+
+One LSTM cell hence maybe expressed as:
+
+$$
+\begin{aligned}
+i_t &= \sigma(W_{xi} x_t + W_{hi} h_{t-1} + W_{ci} \circ c_{t-1} + b_i) \\
+f_t &= \sigma(W_{xf} x_t + W_{hf} h_{t-1} + W_{cf} \circ c_{t-1} + b_f) \\
+c_t &= f_t \circ c_{t-1} + i_t \circ \tanh(W_{xc} x_t + W_{hc} h_{t-1} + b_c) \\
+o_t &= \sigma(W_{xo} x_t + W_{ho} h_{t-1} + W_{co} \circ c_t + b_o) \\
+h_t &= o_t \circ \tanh(c_t)
+\end{aligned}
+$$
+
+### ConvLSTM network
+
+The FC-LSTM, while adept in many scenarios, falters in encoding spatial information when handling spatiotemporal data due to its reliance on full connections in both input-to-state and state-to-state transitions.
+Addressing this limitation, our design ensures that all inputs $X_1, \ldots, X_t$, cell outputs $C_1, \ldots, C_t$, hidden states $H_1, \ldots, H_t$, and gates $i_t, f_t, o_t$ of the ConvLSTM are 3D tensors.
+The last two dimensions of these tensors represent spatial dimensions, specifically rows and columns.
+Conceptually, these inputs and states can be visualized as vectors positioned on a spatial grid.
+
+In the ConvLSTM, the future state of a specific cell on this grid is determined by the inputs and past states of its neighboring cells.
+This spatial consideration is integrated by employing a convolution operator in both state-to-state and input-to-state transitions, as illustrated in Fig.
+2.
+The foundational equations for ConvLSTM are:
+
+$$
+\begin{aligned}
+i_t &= \sigma(W_{xi} \ast X_t + W_{hi} \ast H_{t-1} + W_{ci} \circ C_{t-1} + b_i) \\
+f_t &= \sigma(W_{xf} \ast X_t + W_{hf} \ast H_{t-1} + W_{cf} \circ C_{t-1} + b_f) \\
+C_t &= f_t \circ C_{t-1} + i_t \circ \tanh(W_{xc} \ast X_t + W_{hc} \ast H_{t-1} + b_c) \\
+o_t &= \sigma(W_{xo} \ast X_t + W_{ho} \ast H_{t-1} + W_{co} \circ C_t + b_o) \\
+H_t &= o_t \circ \tanh(C_t)
+\end{aligned}
+$$
+
+Considering the states as representations of moving entities, a ConvLSTM with a larger transitional kernel can capture rapid movements, while a smaller kernel is apt for slower motions.
+Drawing parallels with [16], the traditional FC-LSTM's inputs, cell outputs, and hidden states can also be envisioned as 3D tensors, with the last two dimensions being unitary.
+Thus, the FC-LSTM can be seen as a specific instance of ConvLSTM where all features are concentrated on a singular cell.
+
+To maintain consistency in the dimensions of states and inputs, padding is essential before convolution operations.
+This padding, especially at boundary points, can be interpreted as leveraging the state of external surroundings for computations.
+Typically, all LSTM states are initialized to zero before the first input, symbolizing a lack of knowledge about future events.
+Similarly, zero-padding (as adopted in this study) on hidden states sets the external world's state to zero, implying no prior information about external conditions.
+This padding approach allows for differential treatment of boundary points, proving advantageous in many scenarios.
+For instance, in a system with a ball moving and bouncing off unseen walls, the existence of these walls can be inferred by observing the ball's repeated bounces, a deduction challenging if boundary points share the same state transition dynamics as inner points.
 

### 2023-09-28 paper/paper.tex
Force myself to write text

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt, interior hidden, enhanced]}{\end{tcolorbox}}\fi
 
-\hypertarget{section-heading}{%
-\section{Section Heading}\label{section-heading}}
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
-Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum
-hendrerit facilisis velit sit amet malesuada. Phasellus ornare nibh
-augue, maximus sodales ex tristique vitae. Vivamus non sollicitudin
-orci, aliquam placerat metus. Maecenas volutpat orci felis, vel finibus
-urna consectetur sed. Integer in dui ac dui mollis imperdiet. Quisque
-sed dapibus nibh. Aenean non luctus leo. Phasellus luctus mauris id
-aliquet dictum. Aliquam fermentum semper massa, vel dignissim nibh
-dictum et. See Hubbard et al. (2021).
+River runoff is an important component of the global water cycle as it
+comprises about one third of the precipitation over land areas
+(Hagemann, 2020). Moreover, accurate runoff forecasting, especially over
+extended periods, is pivotal for effective water resources management,
+as highlighted by studies such as Yang et al.~(2018), Tan et al.~(2018),
+and Fang et al.~(2019).
 
-Phasellus interdum tincidunt ex, a euismod massa pulvinar at. Ut
-fringilla ut nisi nec volutpat. Morbi imperdiet congue tincidunt.
-Vivamus eget rutrum purus. Etiam et pretium justo. Donec et egestas sem.
-Donec molestie ex sit amet viverra egestas. Nullam justo nulla,
-fringilla at iaculis in, posuere non mauris. Ut eget imperdiet elit.
+Over the past decades, models for long-term runoff forecasting have been
+primarily bifurcated into physically based models and data-based models.
+While the former attempts to emulate intricate and nonlinear physical
+hydrological processes, the latter hinges on establishing statistical
+models that delineate the relationship between large-scale climate
+patterns and catchment runoff.
 
-In luctus mauris vitae imperdiet luctus. Morbi volutpat ligula ut tortor
-fermentum, eu ornare felis luctus. Donec semper diam vitae mattis
-posuere. Suspendisse facilisis purus nisi, sit amet egestas ex tempor
-ut. Cras tortor nulla, euismod at fermentum vel, dictum vel justo.
-Aenean commodo interdum diam nec placerat. Nunc vestibulum felis at est
-tincidunt, at euismod dui vestibulum. Nulla venenatis tortor at auctor
-iaculis. Donec consectetur neque ut sagittis ornare. Nullam pharetra
-felis tempor suscipit efficitur. Curabitur nibh ex, euismod at congue
-hendrerit, egestas id mi. Duis porttitor neque in commodo elementum.
-Fusce vitae fermentum nisi, euismod viverra augue. Curabitur at mi
-pretium, accumsan purus nec, tempus turpis.
+Machine Learning (ML) models, such as those employing artificial neural
+networks, support vector machines, adaptive neuro-fuzzy inference
+systems, and notably, Long Short-Term Memory (LSTM) neural networks,
+have gained traction for long-term hydrological forecasting due to their
+commendable performance (Humphrey et al 2016, Huang et al 2014, Ashrafi
+et al 2017, Yuan et al 2018, Xu et al 2021).
 
-Donec non semper dui, quis aliquet est. Quisque quis sapien at massa
-ultricies egestas. Duis consequat ultricies erat, a pulvinar nisl
-vestibulum id. Sed tristique turpis ligula, et tempor lectus iaculis at.
-Vivamus commodo sapien ac turpis vestibulum dapibus. Morbi tristique
-arcu metus, et laoreet nisi varius nec. Pellentesque habitant morbi
-tristique senectus et netus et malesuada fames ac turpis egestas. Fusce
-sit amet nisl at mauris suscipit aliquet. Nulla vitae dignissim urna.
-Suspendisse sit amet arcu vitae magna blandit mattis. Vivamus convallis
-efficitur pulvinar. Sed cursus elit nulla. Sed porta, arcu a euismod
-pretium, odio dui lacinia lacus, ac vulputate nulla augue eget ex.
-Nullam consequat ligula sit amet mattis aliquam. Nulla risus urna,
-ultrices vel ullamcorper id, ornare viverra nunc.
+LSTM networks, an evolution of the classical Recurrent Neural Networks
+(RNNs), have shown stability and efficacy in sequence-to-sequence
+predictions, such as using climatic indices for rainfall estimation or
+long-term hydrological forecasting. However, a limitation of LSTMs is
+their inability to effectively capture two-dimensional structures, an
+area where Convolutional Neural Networks (CNNs) excel. Recognizing this,
+we introduce the ConvLSTM, which integrates the strengths of both LSTM
+and CNN, to extract spatiotemporal features from precipitation fields
+for predicting river runoff in the Baltic Sea catchment, summarized by
+97 inidivual rivers.
 
-Nunc in lobortis lacus. Duis maximus urna leo, varius sodales arcu
-interdum nec. Pellentesque imperdiet dolor in leo eleifend dapibus. Ut
-dapibus, lectus non viverra gravida, ipsum ex faucibus tellus, quis
-iaculis risus tellus eget augue. Nullam a viverra est. Cras velit nisi,
-interdum in lacus at, vehicula mattis elit. Curabitur eu viverra purus.
-Proin pellentesque, metus vitae congue convallis, lorem metus feugiat
-mi, sit amet auctor purus ligula bibendum ante. Nam id justo
-scelerisque, rhoncus lectus in, fermentum libero. Donec tincidunt
-egestas ex ac eleifend. Cras faucibus ipsum a nunc faucibus fermentum.
-Integer et maximus lacus. Nam dictum nibh id viverra convallis.
+Modeling the Baltic Sea is to a large part the result of the quality of
+the freshwater input, that is used for the simulation. Meier and
+Kauker(2003) showed that decadal salinity variations of about 1 \(g\)
+\(kg^{-1}\) are caused, inter alia, by annual runoff variations.
+Further, Meier and Kauker (2003) showed that about 50 \% of the decadal
+salinity variability can be explained by variations in freshwater input
+into the Baltic Sea.
+
+This paper delves into the application of deep learning, particularly
+ConvLSTM, to the challenging task of precipitation nowcasting, a domain
+yet to fully harness the potential of advanced machine learning
+techniques. We present ConvLSTM as a novel solution to this
+spatiotemporal sequence forecasting challenge, highlighting its
+advantages and potential future applications.
+
+\hypertarget{methods}{%
+\section{Methods}\label{methods}}
+
+\hypertarget{lstm-network}{%
+\subsection{LSTM network}\label{lstm-network}}
+
+The Long Short-Term Memory (LSTM), a specialized form of Recurrent
+Neural Networks (RNNs), is specifically tailored for modeling temporal
+sequences. Its unique design allows it to adeptly handle long-range
+dependencies, setting it apart from traditional RNNs in terms of
+accuracy. This prowess in modeling long-range dependencies has been
+validated in various studies {[}12, 11, 17, 23{]}. The cornerstone of
+LSTM's innovation is its memory cell, \(c_t\)t, which functions as a
+repository of state information, also refered to as long-term memory.
+This cell is accessed, modified, and reset through several
+self-parameterized gates. As new inputs arrive, their information is
+accumulated into the cell if the input gate is activated. The forget
+gate \(f_t\) defines the percentage of the previous cell status that is
+stored \(c_{t-1}\)\hspace{0pt}. The decision to propagate the latest
+cell output, \(c_t\), to the final state, \(h_t\), is governed by the
+output gate, \(o_t\). A significant advantage of this architecture is
+the memory cell's ability to retain gradients. This mechanism addresses
+the vanishing gradient problem, where, as input sequences elongate, the
+influence of initial stages becomes harder to capture, causing gradients
+of early input points to approach zero. The LSTM's activation function,
+inherently recurrent, mirrors the identity function with a consistent
+derivative of 1.0, ensuring the gradient remains stable throughout
+backpropagation.
+
+One LSTM cell hence maybe expressed as:
+
+\[
+\begin{aligned}
+i_t &= \sigma(W_{xi} x_t + W_{hi} h_{t-1} + W_{ci} \circ c_{t-1} + b_i) \\
+f_t &= \sigma(W_{xf} x_t + W_{hf} h_{t-1} + W_{cf} \circ c_{t-1} + b_f) \\
+c_t &= f_t \circ c_{t-1} + i_t \circ \tanh(W_{xc} x_t + W_{hc} h_{t-1} + b_c) \\
+o_t &= \sigma(W_{xo} x_t + W_{ho} h_{t-1} + W_{co} \circ c_t + b_o) \\
+h_t &= o_t \circ \tanh(c_t)
+\end{aligned}
+\]
+
+\hypertarget{convlstm-network}{%
+\subsection{ConvLSTM network}\label{convlstm-network}}
+
+The FC-LSTM, while adept in many scenarios, falters in encoding spatial
+information when handling spatiotemporal data due to its reliance on
+full connections in both input-to-state and state-to-state transitions.
+Addressing this limitation, our design ensures that all inputs
+\(X_1, \ldots, X_t\), cell outputs \(C_1, \ldots, C_t\), hidden states
+\(H_1, \ldots, H_t\), and gates \(i_t, f_t, o_t\) of the ConvLSTM are 3D
+tensors. The last two dimensions of these tensors represent spatial
+dimensions, specifically rows and columns. Conceptually, these inputs
+and states can be visualized as vectors positioned on a spatial grid.
+
+In the ConvLSTM, the future state of a specific cell on this grid is
+determined by the inputs and past states of its neighboring cells. This
+spatial consideration is integrated by employing a convolution operator
+in both state-to-state and input-to-state transitions, as illustrated in
+Fig. 2. The foundational equations for ConvLSTM are:
+
+\[
+\begin{aligned}
+i_t &= \sigma(W_{xi} \ast X_t + W_{hi} \ast H_{t-1} + W_{ci} \circ C_{t-1} + b_i) \\
+f_t &= \sigma(W_{xf} \ast X_t + W_{hf} \ast H_{t-1} + W_{cf} \circ C_{t-1} + b_f) \\
+C_t &= f_t \circ C_{t-1} + i_t \circ \tanh(W_{xc} \ast X_t + W_{hc} \ast H_{t-1} + b_c) \\
+o_t &= \sigma(W_{xo} \ast X_t + W_{ho} \ast H_{t-1} + W_{co} \circ C_t + b_o) \\
+H_t &= o_t \circ \tanh(C_t)
+\end{aligned}
+\]
+
+Considering the states as representations of moving entities, a ConvLSTM
+with a larger transitional kernel can capture rapid movements, while a
+smaller kernel is apt for slower motions. Drawing parallels with
+{[}16{]}, the traditional FC-LSTM's inputs, cell outputs, and hidden
+states can also be envisioned as 3D tensors, with the last two
+dimensions being unitary. Thus, the FC-LSTM can be seen as a specific
+instance of ConvLSTM where all features are concentrated on a singular
+cell.
+
+To maintain consistency in the dimensions of states and inputs, padding
+is essential before convolution operations. This padding, especially at
+boundary points, can be interpreted as leveraging the state of external
+surroundings for computations. Typically, all LSTM states are
+initialized to zero before the first input, symbolizing a lack of
+knowledge about future events. Similarly, zero-padding (as adopted in
+this study) on hidden states sets the external world's state to zero,
+implying no prior information about external conditions. This padding
+approach allows for differential treatment of boundary points, proving
+advantageous in many scenarios. For instance, in a system with a ball
+moving and bouncing off unseen walls, the existence of these walls can
+be inferred by observing the ball's repeated bounces, a deduction
+challenging if boundary points share the same state transition dynamics
+as inner points.
 
@@
 \hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
+\begin{CSLReferences}{0}{0}
 \vspace{1em}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Hubbard2021}{}}%
-Hubbard, B., Christoffersen, P., Doyle, S. H., Chudley, T. R.,
-Schoonman, C. M., Law, R., \& Bougamont, M. (2021). Borehole-based
-characterization of deep mixed-mode crevasses at a greenlandic outlet
-glacier. \emph{{AGU} Advances}, \emph{2}(2).
-\url{https://doi.org/10.1029/2020av000291}
 

### 2023-09-29 paper/paper.qmd
Update paper

@@
 
-    - Be set as a single paragraph.
-    - Be less than 250 words for all journals except GRL, for which the limit is 150 words.
-    - Not include table or figure mentions.
-    - Avoid reference citations unless dependent on or directly related to another paper (e.g., companion, comment, reply, or commentary on another paper(s)).
-AGU’s Style Guide discusses formatting citations in abstracts.
-    - Define all abbreviations.
 plain-language-summary: |
@@
 The Long Short-Term Memory (LSTM), a specialized form of Recurrent Neural Networks (RNNs), is specifically tailored for modeling temporal sequences.
-Its unique design allows it to adeptly handle long-range dependencies, setting it apart from traditional RNNs in terms of accuracy.
-This prowess in modeling long-range dependencies has been validated in various studies [12, 11, 17, 23].
-The cornerstone of LSTM's innovation is its memory cell, $c_t$t, which functions as a repository of state information, also refered to as long-term memory.
+Its unique design allows it to adeptly handle long-range dependencies, setting it apart from traditional RNNs in terms of accuracy (see @fig-lstm).
+
+![Inner structure of a Long Short-Term Memory Cell](draw_lstm.pdf){#fig-lstm}
+
+This performance in modeling long-range dependencies has been validated in various studies.
+The key component of LSTM's innovation is its memory cell, $c_t$t, which stores state information, also refered to as long-term memory.
 This cell is accessed, modified, and reset through several self-parameterized gates.
-As new inputs arrive, their information is accumulated into the cell if the input gate is activated.
-The forget gate $f_t$ defines the percentage of the previous cell status that is stored $c_{t-1}$​.
-The decision to propagate the latest cell output, $c_t$, to the final state, $h_t$, is governed by the output gate, $o_t$.
+For the input of the sequence  $x_t$ input, the forget gate $f_t$ defines the percentage of the previous long-term memory status $c_{t-1}$ that should be retained stored​.
+Next the input gate $i_t$ decides how much of the input is added to the the long-term memory, forming the updated cell state $c_{t}$.
+The decision to propagate the latest cell output, $c_t$, to the final state, $h_t$, is governed by the output gate, $o_t$, representing the updated short-term memory of the hidden state $h_t$.
 A significant advantage of this architecture is the memory cell's ability to retain gradients.
@@
 
+with
+
+- $x_t$: Input vector at time step $t$.
+- $h_{t-1}$: Hidden state from the previous time step.
+- $C_{t-1}$: Cell state from the previous time step.
+- $W$ and $b$: Weight matrices and bias vectors, respectively, associated with the gates of the LSTM.
+The subscripts denote the specific gate or operation they are associated with (e.g., $W_f$ and $b_f$ are the weight matrix and bias for the forget gate, respectively).
+- $\sigma$: Sigmoid activation function, which squashes values between 0 and 1.
+- $\tanh$: Hyperbolic tangent activation function, which squashes values between -1 and 1
+
 ### ConvLSTM network
 
-The FC-LSTM, while adept in many scenarios, falters in encoding spatial information when handling spatiotemporal data due to its reliance on full connections in both input-to-state and state-to-state transitions.
-Addressing this limitation, our design ensures that all inputs $X_1, \ldots, X_t$, cell outputs $C_1, \ldots, C_t$, hidden states $H_1, \ldots, H_t$, and gates $i_t, f_t, o_t$ of the ConvLSTM are 3D tensors.
+The FC-LSTM fails to handle information when handling spatiotemporal data due to its reliance on full connections in both input-to-state and state-to-state transitions.
+To adress this limitation we use a convLSTM architecture.
+convLSTM replaces the fully connected operations in the LSTM with convolutional operations.
+Hence, all inputs $X_1, \ldots, X_t$, cell outputs $C_1, \ldots, C_t$, hidden states $H_1, \ldots, H_t$, and gates $i_t, f_t, o_t$ of the ConvLSTM are 3D tensors.
 The last two dimensions of these tensors represent spatial dimensions, specifically rows and columns.
@@
 
-Considering the states as representations of moving entities, a ConvLSTM with a larger transitional kernel can capture rapid movements, while a smaller kernel is apt for slower motions.
-Drawing parallels with [16], the traditional FC-LSTM's inputs, cell outputs, and hidden states can also be envisioned as 3D tensors, with the last two dimensions being unitary.
-Thus, the FC-LSTM can be seen as a specific instance of ConvLSTM where all features are concentrated on a singular cell.
+In summary, the ConvLSTM excels at processing tasks that demand a combined understanding of spatial patterns and temporal sequences in data.
+It merges the image-processing capabilities of Convolutional Neural Networks (CNNs) with the time-series modeling of Long Short-Term Memory (LSTM) networks.
 
-To maintain consistency in the dimensions of states and inputs, padding is essential before convolution operations.
-This padding, especially at boundary points, can be interpreted as leveraging the state of external surroundings for computations.
-Typically, all LSTM states are initialized to zero before the first input, symbolizing a lack of knowledge about future events.
-Similarly, zero-padding (as adopted in this study) on hidden states sets the external world's state to zero, implying no prior information about external conditions.
-This padding approach allows for differential treatment of boundary points, proving advantageous in many scenarios.
-For instance, in a system with a ball moving and bouncing off unseen walls, the existence of these walls can be inferred by observing the ball's repeated bounces, a deduction challenging if boundary points share the same state transition dynamics as inner points.
+### Implemented model architecture
+
+We first implemented the ConvLSTM using and encoder/decoder structure as discussed in ...
+.
+To predict all 97 rivers entering the Baltic Sea at once, we flatten the output and use fully connected layers resulting in 97 outputs.
+
+For the computation we use the following set of hyper parameters:
+
+| Parameter name | Parameter size | 
+|------|------|
+| Num. timesteps     | 32    |
+| Conv. Kernelsize    | (7,7)    |
+| Num. ConvLSTM layers    | 4    |
+| Batch size| 32 |
+| Learning Rate | 1e-4 with CosineAnnealing |
+
+: Hyperparameters {#tbl-letters}
 

### 2023-09-29 paper/paper.tex
Update paper

@@
 \documentclass[
-]{agujournal2019}
+  letterpaper,
+  DIV=11,
+  numbers=noendperiod]{scrartcl}
 
@@
 \setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{5}
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 % Make \paragraph and \subparagraph free-standing
@@
 
-\usepackage{url} %this package should fix any errors with URLs in refs.
-\usepackage{lineno}
-\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
-\usepackage{soul}
-\linenumbers
+\KOMAoption{captions}{tableheading}
 \makeatletter
@@
 
-\journalname{Geophysical Research Letters}
-
-\draftfalse
+\title{Preparing your manuscript}
+\author{Florian Börgel \and Sven Karsten}
+\date{9/29/23}
 
 \begin{document}
-\title{Preparing your manuscript}
-
-\authors{Florian Börgel\affil{1}, Sven Karsten\affil{1}}
-\affiliation{1}{Leibniz-Institute for Baltic Sea Research Warnemünde, }
-\correspondingauthor{Florian Börgel}{florian.boergel@io-warnemuende.de}
-
+\maketitle
 \begin{abstract}
@@
 for indexing. Your abstract should:
-
-\begin{itemize}
-\tightlist
-\item
-  Be set as a single paragraph.
-\item
-  Be less than 250 words for all journals except GRL, for which the
-  limit is 150 words.
-\item
-  Not include table or figure mentions.
-\item
-  Avoid reference citations unless dependent on or directly related to
-  another paper (e.g., companion, comment, reply, or commentary on
-  another paper(s)). AGU's Style Guide discusses formatting citations in
-  abstracts.
-\item
-  Define all abbreviations.
-\end{itemize}
 \end{abstract}
-
-\section*{Plain Language Summary}
-A Plain Language Summary (PLS) can be an incredibly effective science
-communication tool. By summarizing your paper in non-technical terms,
-you can explain your research and its relevance to a much broader
-audience. A PLS is required for submissions to AGU Advances, G-Cubed,
-GeoHealth, GRL, JAMES, JGR: Biogeosciences, JGR: Oceans, JGR: Planets,
-JGR: Solid Earth, JGR: Atmospheres, Space Weather, and Reviews of
-Geophysics, but optional for other journals. A PLS should be no longer
-than 200 words and should be free of jargon, acronyms, equations, and
-any technical information that would be unknown to people from outside
-your scientific discipline. Read our tips for creating an effective PLS.
-
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt, interior hidden, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, enhanced, frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden, boxrule=0pt, sharp corners]}{\end{tcolorbox}}\fi
 
 \hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+\subsection{Introduction}\label{introduction}}
 
@@
 \hypertarget{methods}{%
-\section{Methods}\label{methods}}
+\subsection{Methods}\label{methods}}
 
 \hypertarget{lstm-network}{%
-\subsection{LSTM network}\label{lstm-network}}
+\subsubsection{LSTM network}\label{lstm-network}}
 
@@
 dependencies, setting it apart from traditional RNNs in terms of
-accuracy. This prowess in modeling long-range dependencies has been
-validated in various studies {[}12, 11, 17, 23{]}. The cornerstone of
-LSTM's innovation is its memory cell, \(c_t\)t, which functions as a
-repository of state information, also refered to as long-term memory.
-This cell is accessed, modified, and reset through several
-self-parameterized gates. As new inputs arrive, their information is
-accumulated into the cell if the input gate is activated. The forget
-gate \(f_t\) defines the percentage of the previous cell status that is
-stored \(c_{t-1}\)\hspace{0pt}. The decision to propagate the latest
-cell output, \(c_t\), to the final state, \(h_t\), is governed by the
-output gate, \(o_t\). A significant advantage of this architecture is
-the memory cell's ability to retain gradients. This mechanism addresses
-the vanishing gradient problem, where, as input sequences elongate, the
+accuracy (see Figure~\ref{fig-lstm}).
+
+\begin{figure}
+
+{\centering \includegraphics{draw_lstm.pdf}
+
+}
+
+\caption{\label{fig-lstm}Inner structure of a Long Short-Term Memory
+Cell}
+
+\end{figure}
+
+This performance in modeling long-range dependencies has been validated
+in various studies. The key component of LSTM's innovation is its memory
+cell, \(c_t\)t, which stores state information, also refered to as
+long-term memory. This cell is accessed, modified, and reset through
+several self-parameterized gates. For the input of the sequence \(x_t\)
+input, the forget gate \(f_t\) defines the percentage of the previous
+long-term memory status \(c_{t-1}\) that should be retained
+stored\hspace{0pt}. Next the input gate \(i_t\) decides how much of the
+input is added to the the long-term memory, forming the updated cell
+state \(c_{t}\). The decision to propagate the latest cell output,
+\(c_t\), to the final state, \(h_t\), is governed by the output gate,
+\(o_t\), representing the updated short-term memory of the hidden state
+\(h_t\). A significant advantage of this architecture is the memory
+cell's ability to retain gradients. This mechanism addresses the
+vanishing gradient problem, where, as input sequences elongate, the
 influence of initial stages becomes harder to capture, causing gradients
@@
 
+with
+
+\begin{itemize}
+\tightlist
+\item
+  \(x_t\): Input vector at time step \(t\).
+\item
+  \(h_{t-1}\): Hidden state from the previous time step.
+\item
+  \(C_{t-1}\): Cell state from the previous time step.
+\item
+  \(W\) and \(b\): Weight matrices and bias vectors, respectively,
+  associated with the gates of the LSTM. The subscripts denote the
+  specific gate or operation they are associated with (e.g., \(W_f\) and
+  \(b_f\) are the weight matrix and bias for the forget gate,
+  respectively).
+\item
+  \(\sigma\): Sigmoid activation function, which squashes values between
+  0 and 1.
+\item
+  \(\tanh\): Hyperbolic tangent activation function, which squashes
+  values between -1 and 1
+\end{itemize}
+
 \hypertarget{convlstm-network}{%
-\subsection{ConvLSTM network}\label{convlstm-network}}
+\subsubsection{ConvLSTM network}\label{convlstm-network}}
 
-The FC-LSTM, while adept in many scenarios, falters in encoding spatial
-information when handling spatiotemporal data due to its reliance on
-full connections in both input-to-state and state-to-state transitions.
-Addressing this limitation, our design ensures that all inputs
+The FC-LSTM fails to handle information when handling spatiotemporal
+data due to its reliance on full connections in both input-to-state and
+state-to-state transitions. To adress this limitation we use a convLSTM
+architecture. convLSTM replaces the fully connected operations in the
+LSTM with convolutional operations. Hence, all inputs
 \(X_1, \ldots, X_t\), cell outputs \(C_1, \ldots, C_t\), hidden states
@@
 
-Considering the states as representations of moving entities, a ConvLSTM
-with a larger transitional kernel can capture rapid movements, while a
-smaller kernel is apt for slower motions. Drawing parallels with
-{[}16{]}, the traditional FC-LSTM's inputs, cell outputs, and hidden
-states can also be envisioned as 3D tensors, with the last two
-dimensions being unitary. Thus, the FC-LSTM can be seen as a specific
-instance of ConvLSTM where all features are concentrated on a singular
-cell.
+In summary, the ConvLSTM excels at processing tasks that demand a
+combined understanding of spatial patterns and temporal sequences in
+data. It merges the image-processing capabilities of Convolutional
+Neural Networks (CNNs) with the time-series modeling of Long Short-Term
+Memory (LSTM) networks.
 
-To maintain consistency in the dimensions of states and inputs, padding
-is essential before convolution operations. This padding, especially at
-boundary points, can be interpreted as leveraging the state of external
-surroundings for computations. Typically, all LSTM states are
-initialized to zero before the first input, symbolizing a lack of
-knowledge about future events. Similarly, zero-padding (as adopted in
-this study) on hidden states sets the external world's state to zero,
-implying no prior information about external conditions. This padding
-approach allows for differential treatment of boundary points, proving
-advantageous in many scenarios. For instance, in a system with a ball
-moving and bouncing off unseen walls, the existence of these walls can
-be inferred by observing the ball's repeated bounces, a deduction
-challenging if boundary points share the same state transition dynamics
-as inner points.
+\hypertarget{implemented-model-architecture}{%
+\subsubsection{Implemented model
+architecture}\label{implemented-model-architecture}}
+
+We first implemented the ConvLSTM using and encoder/decoder structure as
+discussed in \ldots{} . To predict all 97 rivers entering the Baltic Sea
+at once, we flatten the output and use fully connected layers resulting
+in 97 outputs.
+
+For the computation we use the following set of hyper parameters:
+
+\hypertarget{tbl-letters}{}
+\begin{longtable}[]{@{}ll@{}}
+\caption{\label{tbl-letters}Hyperparameters}\tabularnewline
+\toprule()
+Parameter name & Parameter size \\
+\midrule()
+\endfirsthead
+\toprule()
+Parameter name & Parameter size \\
+\midrule()
+\endhead
+Num. timesteps & 32 \\
+Conv. Kernelsize & (7,7) \\
+Num. ConvLSTM layers & 4 \\
+Batch size & 32 \\
+~Learning Rate & 1e-4 with CosineAnnealing \\
+\bottomrule()
+\end{longtable}
 
 \hypertarget{acknowledgments}{%
-\section{Acknowledgments}\label{acknowledgments}}
+\subsection{Acknowledgments}\label{acknowledgments}}
 
@@
 \hypertarget{open-research}{%
-\section{Open research}\label{open-research}}
+\subsection{Open research}\label{open-research}}
 
@@
 \hypertarget{references}{%
-\section*{References}\label{references}}
-\addcontentsline{toc}{section}{References}
+\subsection*{References}\label{references}}
+\addcontentsline{toc}{subsection}{References}
 
@@
 \begin{CSLReferences}{0}{0}
-\vspace{1em}
-
 \end{CSLReferences}

### 2023-09-29 BalticRiverPrediction/BaltNet.py
prog

@@ -95,7 +95,7 @@ class BaseLineModel(nn.Module):
         return x
 
 # %% ../nbs/02_BaltNet.ipynb 5
-class LightningModel(pl.LightningModule):
+class LightningModel(L.LightningModule):
     def __init__(self, model, learning_rate, cosine_t_max):
         super().__init__()
 
@@ -165,12 +165,12 @@ class AtmosphericDataset(Dataset):
         self.input_size = input_size
 
         runoffData = runoff.transpose("time", "river")
-        runoffDataMean = runoffData.mean()
-        runoffDataSTD = runoffData.std()
+        runoffDataMean = runoffData.mean("time")
+        runoffDataSTD = runoffData.std("time")
 
         rainData = data["RAIN"].sel(time=self.timeRange)
-        rainDataMean = rainData.mean()
-        rainDataSTD = rainData.std()
+        rainDataMean = rainData.mean("time")
+        rainDataSTD = rainData.std("time")
 
         self.rainData = (rainDataMean, rainDataSTD)
         self.runoffData = (runoffDataMean, runoffDataSTD)
@@ -202,7 +202,7 @@ class AtmosphericDataset(Dataset):
 
 
 # %% ../nbs/02_BaltNet.ipynb 7
-class AtmosphereDataModule(pl.LightningDataModule):
+class AtmosphereDataModule(L.LightningDataModule):
     def __init__(self, data, runoff, batch_size=64, num_workers=8, add_first_dim=True, input_size=30):
         super().__init__()
 

### 2023-09-29 src/runMultipleTest.py

@@ -13,7 +13,7 @@ from BalticRiverPrediction.sharedUtilities import read_netcdfs, preprocess
 if __name__ == "__main__":
 
     # set seed for reproducibility   
-    L.pytorch.seed_everything(123)
+    L.seed_everything(123)
 
     # Our GPU has tensor cores, hence mixed precision training is enabled
     # see https://sebastianraschka.com/blog/2023/llm-mixed-precision-copy.html
@@ -38,9 +38,9 @@ if __name__ == "__main__":
     # Note that this set of parameters will be defined by runTuning.py
 
     modelParameters = {
-    "input_dim":16, # timesteps
+    "input_dim":32, # timesteps
     "hidden_dim":1, # Channels -> right now only precipitation
-    "kernel_size":(5,5), # applied for spatial convolutions
+    "kernel_size":(7,7), # applied for spatial convolutions
     "num_layers":4, # number of convLSTM layers
     "batch_first":True, # first index is batch
     "bias":True, 
@@ -52,7 +52,7 @@ if __name__ == "__main__":
     dataLoader = AtmosphereDataModule(
     data=data,
     runoff=runoff,
-    batch_size=16,
+    batch_size=32,
     input_size=modelParameters["input_dim"]
     )
 
@@ -66,7 +66,7 @@ if __name__ == "__main__":
     # Lightning model wrapper
     LighningBaltNet = LightningModel(
         pyTorchBaltNet,
-        learning_rate=1e-3,
+        learning_rate=1e-4,
         cosine_t_max=num_epochs
     )
 
@@ -86,7 +86,7 @@ if __name__ == "__main__":
         callbacks=callbacks,
         max_epochs=num_epochs,
         accelerator="cuda",
-        devices=2,
+        devices=1,
         logger=CSVLogger(
             save_dir="/silor/boergel/paper/runoff_prediction/logs",
             name="BaltNet1"
@@ -95,4 +95,3 @@ if __name__ == "__main__":
     )
 
     trainer.fit(model=LighningBaltNet, datamodule=dataLoader)
-    # trainer.test(model=LighningBaltNet, datamodule=dataLoader)

### 2023-09-29 src/runTuning.py

@@ -14,7 +14,8 @@ from ray.train import RunConfig, ScalingConfig, CheckpointConfig
 from ray.train.torch import TorchTrainer
 from ray.tune.schedulers import FIFOScheduler
 
-
+#IMPORTANT: Note that the model needs to be build with pl as well instead of lighning
+#ERROR IN LIGHTNING
 def train_func(config):
     torch.set_float32_matmul_precision("medium")
   

### 2024-03-14 paper/paper.qmd
Update paper

@@
 
-River runoff is an important component of the global water cycle as it comprises about one third of the precipitation over land areas (Hagemann, 2020).
-Moreover, accurate runoff forecasting, especially over extended periods, is pivotal for effective water resources management, as highlighted by studies such as Yang et al.
-(2018), Tan et al.
-(2018), and Fang et al.
-(2019).
+River runoff is an important component of the global water cycle as it comprises about one third of the precipitation over land areas [@hagemannHighResolutionDischarge2020].
+For the Baltic Sea river runoff is even more important, as the Baltic Sea is nearly decoupled from the open ocean (see Figure).
+As a consequence, the salinity of the Baltic Sea is driven to a large part by freshwater supply from rivers.
+More generally, the freshwater input into the Baltic Sea comes either as river runoff or a positive net precipitation (precipitation minus evaporation) over the sea surface.
+The net precipitation accounts for 11 \% and the river input for 89 \% of the total freshwater input (Meier and Doescher, 2002).
+Modeling the Baltic Sea is therefore to a large part the result of the quality of the river input, that is used for the simulation.
 
-Over the past decades, models for long-term runoff forecasting have been primarily bifurcated into physically based models and data-based models.
-While the former attempts to emulate intricate and nonlinear physical hydrological processes, the latter hinges on establishing statistical models that delineate the relationship between large-scale climate patterns and catchment runoff.
+In the context of climate change studies, river runoff is usually generated in two ways.
+First, river runoff as input for ocean models can be created using hyrdological models such as the Hydrological Discharge (HD) model [@hagemannHighResolutionDischarge2020].
+HD calculates the water balance using hydrological processes (e.g.
+snow, glaciers, soil moisture, groundwater contribution).
+It represents a complex forecast tool that uses underlying physical processes.
+A different approach would use data-based models that intergrate statistical correction, using the land surface schemes of global or regional climate models.
 
-Machine Learning (ML) models, such as those employing artificial neural networks, support vector machines, adaptive neuro-fuzzy inference systems, and notably, Long Short-Term Memory (LSTM) neural networks, have gained traction for long-term hydrological forecasting due to their commendable performance (Humphrey et al 2016, Huang et al 2014, Ashrafi et al 2017, Yuan et al 2018, Xu et al 2021).
+The releatively recent rise of machine learning (ML) models has been mostly explored for river runoff forcasting, as accurate runoff forecasting, especially over extended periods, is pivotal for effective water resources management [@fangExaminingApplicabilityDifferent2019;@tanAdaptiveMiddleLongterm2018;@yangDerivingOperatingRules2018].
+Common approach us employ artificial neural networks, support vector machines, adaptive neuro-fuzzy inference systems, and notably, Long Short-Term Memory (LSTM) neural networks and have gained traction for long-term hydrological forecasting due to their commendable performance (Humphrey et al 2016, Huang et al 2014, Ashrafi et al 2017, Yuan et al 2018, Xu et al 2021).
 
@@
 However, a limitation of LSTMs is their inability to effectively capture two-dimensional structures, an area where Convolutional Neural Networks (CNNs) excel.
-Recognizing this, we introduce the ConvLSTM, which integrates the strengths of both LSTM and CNN, to extract spatiotemporal features from precipitation fields for predicting river runoff in the Baltic Sea catchment, summarized by 97 inidivual rivers.
+
+Recognizing this, we use the ConvLSTM, which integrates the strengths of both LSTM and CNN, to extract spatiotemporal features from atmospheric fields for predicting river runoff in the Baltic Sea catchment, summarized by 97 inidivual rivers.
+
+Following the previous description of the complex hyrdological model E-Hype - when present E-Hype should be used.
+However, in absence of a fully functioning hydrogolocial model, that also uses a rather complex parametrization, RNNs represent a robust way to predict river runoff for any give period of time using atmospheric forcing.
 
@@
 - $W$ and $b$: Weight matrices and bias vectors, respectively, associated with the gates of the LSTM.
-The subscripts denote the specific gate or operation they are associated with (e.g., $W_f$ and $b_f$ are the weight matrix and bias for the forget gate, respectively).
-- $\sigma$: Sigmoid activation function, which squashes values between 0 and 1.
-- $\tanh$: Hyperbolic tangent activation function, which squashes values between -1 and 1
+The subscripts denote the specific gate or operation they are associated with (e.g., $W_f$ and $b_f$ are the weight matrix and bias for the forget gate).
+- $\sigma$: Sigmoid activation function ($0,\ldots 1$)
+- $\tanh$: Hyperbolic tangent activation function ($ -1 \ldots 1$)
 
@@
 
+An overview of the model structure is given below
+
+![BaltConvLSTM](model_structure.png){#fig-baltNet}
+
 For the computation we use the following set of hyper parameters:
@@
 |------|------|
-| Num. timesteps     | 32    |
-| Conv. Kernelsize    | (7,7)    |
-| Num. ConvLSTM layers    | 4    |
-| Batch size| 32 |
-| Learning Rate | 1e-4 with CosineAnnealing |
+| Channel size | 4 |
+| Num. hidden layer |10 |
+| Num. timesteps     | 30    |
+| Conv. Kernelsize    | (9,9)    |
+| Num. ConvLSTM layers    | 1    |
+| Batch size| 64 |
+| Learning Rate | 1e-3 with CosineAnnealing |
 

### 2024-03-14 paper/paper.tex
Update paper

@@
 \documentclass[
-  letterpaper,
-  DIV=11,
-  numbers=noendperiod]{scrartcl}
+]{agujournal2019}
 
@@
 \setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+\setcounter{secnumdepth}{5}
 % Make \paragraph and \subparagraph free-standing
@@
 
-\KOMAoption{captions}{tableheading}
+\usepackage{url} %this package should fix any errors with URLs in refs.
+\usepackage{lineno}
+\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
+\usepackage{soul}
+\linenumbers
 \makeatletter
@@
 
-\title{Preparing your manuscript}
-\author{Florian Börgel \and Sven Karsten}
-\date{9/29/23}
+\journalname{Geophysical Research Letters}
+
+\draftfalse
 
 \begin{document}
-\maketitle
+\title{Preparing your manuscript}
+
+\authors{Florian Börgel\affil{1}, Sven Karsten\affil{1}}
+\affiliation{1}{Leibniz-Institute for Baltic Sea Research Warnemünde, }
+\correspondingauthor{Florian Börgel}{florian.boergel@io-warnemuende.de}
+
 \begin{abstract}
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, enhanced, frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden, boxrule=0pt, sharp corners]}{\end{tcolorbox}}\fi
+
+\section*{Plain Language Summary}
+A Plain Language Summary (PLS) can be an incredibly effective science
+communication tool. By summarizing your paper in non-technical terms,
+you can explain your research and its relevance to a much broader
+audience. A PLS is required for submissions to AGU Advances, G-Cubed,
+GeoHealth, GRL, JAMES, JGR: Biogeosciences, JGR: Oceans, JGR: Planets,
+JGR: Solid Earth, JGR: Atmospheres, Space Weather, and Reviews of
+Geophysics, but optional for other journals. A PLS should be no longer
+than 200 words and should be free of jargon, acronyms, equations, and
+any technical information that would be unknown to people from outside
+your scientific discipline. Read our tips for creating an effective PLS.
+
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, sharp corners, interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable]}{\end{tcolorbox}}\fi
 
 \hypertarget{introduction}{%
-\subsection{Introduction}\label{introduction}}
+\section{Introduction}\label{introduction}}
 
 River runoff is an important component of the global water cycle as it
-comprises about one third of the precipitation over land areas
-(Hagemann, 2020). Moreover, accurate runoff forecasting, especially over
-extended periods, is pivotal for effective water resources management,
-as highlighted by studies such as Yang et al.~(2018), Tan et al.~(2018),
-and Fang et al.~(2019).
+comprises about one third of the precipitation over land areas (Hagemann
+et al., 2020). For the Baltic Sea river runoff is even more important,
+as the Baltic Sea is nearly decoupled from the open ocean (see Figure).
+As a consequence, the salinity of the Baltic Sea is driven to a large
+part by freshwater supply from rivers. More generally, the freshwater
+input into the Baltic Sea comes either as river runoff or a positive net
+precipitation (precipitation minus evaporation) over the sea surface.
+The net precipitation accounts for 11 \% and the river input for 89 \%
+of the total freshwater input (Meier and Doescher, 2002). Modeling the
+Baltic Sea is therefore to a large part the result of the quality of the
+river input, that is used for the simulation.
 
-Over the past decades, models for long-term runoff forecasting have been
-primarily bifurcated into physically based models and data-based models.
-While the former attempts to emulate intricate and nonlinear physical
-hydrological processes, the latter hinges on establishing statistical
-models that delineate the relationship between large-scale climate
-patterns and catchment runoff.
+In the context of climate change studies, river runoff is usually
+generated in two ways. First, river runoff as input for ocean models can
+be created using hyrdological models such as the Hydrological Discharge
+(HD) model (Hagemann et al., 2020). HD calculates the water balance
+using hydrological processes (e.g.~snow, glaciers, soil moisture,
+groundwater contribution). It represents a complex forecast tool that
+uses underlying physical processes. A different approach would use
+data-based models that intergrate statistical correction, using the land
+surface schemes of global or regional climate models.
 
-Machine Learning (ML) models, such as those employing artificial neural
-networks, support vector machines, adaptive neuro-fuzzy inference
-systems, and notably, Long Short-Term Memory (LSTM) neural networks,
-have gained traction for long-term hydrological forecasting due to their
-commendable performance (Humphrey et al 2016, Huang et al 2014, Ashrafi
-et al 2017, Yuan et al 2018, Xu et al 2021).
+The releatively recent rise of machine learning (ML) models has been
+mostly explored for river runoff forcasting, as accurate runoff
+forecasting, especially over extended periods, is pivotal for effective
+water resources management (Fang et al., 2019; Tan et al., 2018; Yang et
+al., 2018). Common approach us employ artificial neural networks,
+support vector machines, adaptive neuro-fuzzy inference systems, and
+notably, Long Short-Term Memory (LSTM) neural networks and have gained
+traction for long-term hydrological forecasting due to their commendable
+performance (Humphrey et al 2016, Huang et al 2014, Ashrafi et al 2017,
+Yuan et al 2018, Xu et al 2021).
 
@@
 their inability to effectively capture two-dimensional structures, an
-area where Convolutional Neural Networks (CNNs) excel. Recognizing this,
-we introduce the ConvLSTM, which integrates the strengths of both LSTM
-and CNN, to extract spatiotemporal features from precipitation fields
-for predicting river runoff in the Baltic Sea catchment, summarized by
-97 inidivual rivers.
+area where Convolutional Neural Networks (CNNs) excel.
+
+Recognizing this, we use the ConvLSTM, which integrates the strengths of
+both LSTM and CNN, to extract spatiotemporal features from atmospheric
+fields for predicting river runoff in the Baltic Sea catchment,
+summarized by 97 inidivual rivers.
+
+Following the previous description of the complex hyrdological model
+E-Hype - when present E-Hype should be used. However, in absence of a
+fully functioning hydrogolocial model, that also uses a rather complex
+parametrization, RNNs represent a robust way to predict river runoff for
+any give period of time using atmospheric forcing.
 
@@
 \hypertarget{methods}{%
-\subsection{Methods}\label{methods}}
+\section{Methods}\label{methods}}
 
 \hypertarget{lstm-network}{%
-\subsubsection{LSTM network}\label{lstm-network}}
+\subsection{LSTM network}\label{lstm-network}}
 
@@
   specific gate or operation they are associated with (e.g., \(W_f\) and
-  \(b_f\) are the weight matrix and bias for the forget gate,
-  respectively).
+  \(b_f\) are the weight matrix and bias for the forget gate).
 \item
-  \(\sigma\): Sigmoid activation function, which squashes values between
-  0 and 1.
+  \(\sigma\): Sigmoid activation function (\(0,\ldots 1\))
 \item
-  \(\tanh\): Hyperbolic tangent activation function, which squashes
-  values between -1 and 1
+  \(\tanh\): Hyperbolic tangent activation function (\$-1 \ldots1)
 \end{itemize}
@@
 \hypertarget{convlstm-network}{%
-\subsubsection{ConvLSTM network}\label{convlstm-network}}
+\subsection{ConvLSTM network}\label{convlstm-network}}
 
@@
 \hypertarget{implemented-model-architecture}{%
-\subsubsection{Implemented model
+\subsection{Implemented model
 architecture}\label{implemented-model-architecture}}
@@
 in 97 outputs.
+
+An overview of the model structure is given below
+
+\begin{figure}
+
+{\centering \includegraphics{model_structure.png}
+
+}
+
+\caption{\label{fig-baltNet}BaltConvLSTM}
+
+\end{figure}
 
@@
 \endhead
-Num. timesteps & 32 \\
-Conv. Kernelsize & (7,7) \\
-Num. ConvLSTM layers & 4 \\
-Batch size & 32 \\
-~Learning Rate & 1e-4 with CosineAnnealing \\
+Channel size & 4 \\
+Num. hidden layer & 10 \\
+Num. timesteps & 30 \\
+Conv. Kernelsize & (9,9) \\
+Num. ConvLSTM layers & 1 \\
+Batch size & 64 \\
+~Learning Rate & 1e-3 with CosineAnnealing \\
 \bottomrule()
@@
 \hypertarget{acknowledgments}{%
-\subsection{Acknowledgments}\label{acknowledgments}}
+\section{Acknowledgments}\label{acknowledgments}}
 
@@
 \hypertarget{open-research}{%
-\subsection{Open research}\label{open-research}}
+\section{Open research}\label{open-research}}
 
@@
 \hypertarget{references}{%
-\subsection*{References}\label{references}}
-\addcontentsline{toc}{subsection}{References}
+\section*{References}\label{references}}
+\addcontentsline{toc}{section}{References}
 
 \hypertarget{refs}{}
-\begin{CSLReferences}{0}{0}
+\begin{CSLReferences}{1}{0}
+\vspace{1em}
+
+\leavevmode\vadjust pre{\hypertarget{ref-fangExaminingApplicabilityDifferent2019}{}}%
+Fang, W., Huang, S., Ren, K., Huang, Q., Huang, G., Cheng, G., \& Li, K.
+(2019). Examining the applicability of different sampling techniques in
+the development of decomposition-based streamflow forecasting models.
+\emph{Journal of Hydrology}, \emph{568}, 534--550.
+\url{https://doi.org/10.1016/j.jhydrol.2018.11.020}
+
+\leavevmode\vadjust pre{\hypertarget{ref-hagemannHighResolutionDischarge2020}{}}%
+Hagemann, S., Stacke, T., \& Ho-Hagemann, H. T. M. (2020). High
+{Resolution Discharge Simulations Over Europe} and the {Baltic Sea
+Catchment}. \emph{Frontiers in Earth Science}, \emph{8}.
+\url{https://doi.org/10.3389/feart.2020.00012}
+
+\leavevmode\vadjust pre{\hypertarget{ref-tanAdaptiveMiddleLongterm2018}{}}%
+Tan, Q.-F., Lei, X.-H., Wang, X., Wang, H., Wen, X., Ji, Y., \& Kang,
+A.-Q. (2018). An adaptive middle and long-term runoff forecast model
+using {EEMD-ANN} hybrid approach. \emph{Journal of Hydrology},
+\emph{567}, 767--780.
+\url{https://doi.org/10.1016/j.jhydrol.2018.01.015}
+
+\leavevmode\vadjust pre{\hypertarget{ref-yangDerivingOperatingRules2018}{}}%
+Yang, Z., Liu, P., Cheng, L., Wang, H., Ming, B., \& Gong, W. (2018).
+Deriving operating rules for a large-scale hydro-photovoltaic power
+system using implicit stochastic optimization. \emph{Journal of Cleaner
+Production}, \emph{195}, 562--572.
+\url{https://doi.org/10.1016/j.jclepro.2018.05.154}
+
 \end{CSLReferences}

### 2024-03-14 paper/graph.py

@@ -1,0 +1,38 @@
+import graphviz
+
+def draw_balt_net():
+    dot = graphviz.Digraph(format='png')
+
+    # Add nodes
+    dot.node('Input', 'Input Layer')
+    dot.node('Encoder', 'ConvLSTM Encoder')
+    dot.node('Decoder', 'ConvLSTM Decoder')
+    dot.node('Attention', 'Spatial Attention')
+    dot.node('Flatten', 'Flatten Output')
+    dot.node('FC1', 'Fully Connected Layer 1')
+    dot.node('ReLU1', 'ReLU Activation 1')
+    dot.node('FC2', 'Fully Connected Layer 2')
+    dot.node('ReLU2', 'ReLU Activation 2')
+    dot.node('FC3', 'Fully Connected Layer 3')
+    dot.node('Output', 'Output Layer')
+
+    # Add edges
+    dot.edge('Input', 'Encoder')
+    dot.edge('Encoder', 'Decoder')
+    dot.edge('Decoder', 'Attention')
+    dot.edge('Attention', 'Flatten')
+    dot.edge('Flatten', 'FC1')
+    dot.edge('FC1', 'ReLU1')
+    dot.edge('ReLU1', 'FC2')
+    dot.edge('FC2', 'ReLU2')
+    dot.edge('ReLU2', 'FC3')
+    dot.edge('FC3', 'Output')
+
+    return dot
+
+# Draw the network
+balt_net_graph = draw_balt_net()
+balt_net_graph.render('BaltNet_Structure')
+
+# The network structure is saved as 'BaltNet_Structure.png' in the current directory
+

### 2024-03-15 paper/paper.qmd
update paper draft

@@
 ---
+
 ## Introduction
@@
 River runoff is an important component of the global water cycle as it comprises about one third of the precipitation over land areas [@hagemannHighResolutionDischarge2020].
-For the Baltic Sea river runoff is even more important, as the Baltic Sea is nearly decoupled from the open ocean (see Figure).
-As a consequence, the salinity of the Baltic Sea is driven to a large part by freshwater supply from rivers.
-More generally, the freshwater input into the Baltic Sea comes either as river runoff or a positive net precipitation (precipitation minus evaporation) over the sea surface.
-The net precipitation accounts for 11 \% and the river input for 89 \% of the total freshwater input (Meier and Doescher, 2002).
-Modeling the Baltic Sea is therefore to a large part the result of the quality of the river input, that is used for the simulation.
-
 In the context of climate change studies, river runoff is usually generated in two ways.
 First, river runoff as input for ocean models can be created using hyrdological models such as the Hydrological Discharge (HD) model [@hagemannHighResolutionDischarge2020].
-HD calculates the water balance using hydrological processes (e.g.
+HD models calculates the water balance using hydrological processes (e.g.
 snow, glaciers, soil moisture, groundwater contribution).
-It represents a complex forecast tool that uses underlying physical processes.
+It represents a complex forecasting tool that uses underlying physical processes.
 A different approach would use data-based models that intergrate statistical correction, using the land surface schemes of global or regional climate models.
 
-The releatively recent rise of machine learning (ML) models has been mostly explored for river runoff forcasting, as accurate runoff forecasting, especially over extended periods, is pivotal for effective water resources management [@fangExaminingApplicabilityDifferent2019;@tanAdaptiveMiddleLongterm2018;@yangDerivingOperatingRules2018].
-Common approach us employ artificial neural networks, support vector machines, adaptive neuro-fuzzy inference systems, and notably, Long Short-Term Memory (LSTM) neural networks and have gained traction for long-term hydrological forecasting due to their commendable performance (Humphrey et al 2016, Huang et al 2014, Ashrafi et al 2017, Yuan et al 2018, Xu et al 2021).
+The releatively recent rise of machine learning (ML) models has been mostly explored for river runoff forecasting, as accurate runoff forecasting, especially over extended periods, is pivotal for effective water resources management [@fangExaminingApplicabilityDifferent2019; @tanAdaptiveMiddleLongterm2018; @yangDerivingOperatingRules2018].
+Common approaches employ artificial neural networks, support vector machines, adaptive neuro-fuzzy inference systems, and notably, Long Short-Term Memory (LSTM) neural networks that have gained traction for long-term hydrological forecasting due to their excellent performance (Humphrey et al 2016, Huang et al 2014, Ashrafi et al 2017, Yuan et al 2018, Xu et al 2021).
 
@@
 However, a limitation of LSTMs is their inability to effectively capture two-dimensional structures, an area where Convolutional Neural Networks (CNNs) excel.
+Recognizing this, we use the ConvLSTM, which integrates the strengths of both LSTM and CNN, to extract spatiotemporal features from atmospheric fields for predicting river runoff.
 
-Recognizing this, we use the ConvLSTM, which integrates the strengths of both LSTM and CNN, to extract spatiotemporal features from atmospheric fields for predicting river runoff in the Baltic Sea catchment, summarized by 97 inidivual rivers.
+TODO: More about ConvLSTM
 
-Following the previous description of the complex hyrdological model E-Hype - when present E-Hype should be used.
-However, in absence of a fully functioning hydrogolocial model, that also uses a rather complex parametrization, RNNs represent a robust way to predict river runoff for any give period of time using atmospheric forcing.
+In the following we will use the Baltic Sea catchment to illustrate our approach, while in principal the methodology we propose is universally applicable across various geographic regions.
+The Baltic Sea serves as a challenging example due to its unique hydrological characteristics, being nearly decoupled from the open ocean (see Figure).
 
-Modeling the Baltic Sea is to a large part the result of the quality of the freshwater input, that is used for the simulation.
-Meier and Kauker(2003) showed that decadal salinity variations of about 1 $g$ $kg^{-1}$ are caused, inter alia, by annual runoff variations.
-Further, Meier and Kauker (2003) showed that about 50 \% of the decadal salinity variability can be explained by variations in freshwater input into the Baltic Sea.
+As a consequence, the salinity of the Baltic Sea is driven to a large part by freshwater supply from rivers.
+More generally, the freshwater input into the Baltic Sea comes either as river runoff or a positive net precipitation (precipitation minus evaporation) over the sea surface.
+The net precipitation accounts for 11 % and the river input for 89 % of the total freshwater input (Meier and Doescher, 2002).
+Modeling the Baltic Sea is therefore to a large part the result of the quality of the river input, that is used for the simulation.
+This makes the accurate modeling of river runoff especially critical for simulations pertaining to the Baltic Sea.
+In the following we show that in absence of a fully functioning hydrogolocial model, that also uses a rather complex parametrization, ConvLSTM represent a robust way to predict river runoff for any given period of time using only atmospheric forcing.
 
-This paper delves into the application of deep learning, particularly ConvLSTM, to the challenging task of precipitation nowcasting, a domain yet to fully harness the potential of advanced machine learning techniques.
-We present ConvLSTM as a novel solution to this spatiotemporal sequence forecasting challenge, highlighting its advantages and potential future applications.
+In this work, we present a ConvLSTM architecture as a novel solution to this spatiotemporal sequence forecasting challenge, highlighting its advantages and potential future applications.
 
@@
 
-![Inner structure of a Long Short-Term Memory Cell](draw_lstm.pdf){#fig-lstm}
+![Inner structure of a Long Short-Term Memory Cell](draw_lstm.png){#fig-lstm fig-align="center"}
 
@@
 -   $\sigma$: Sigmoid activation function ($0,\ldots 1$)
-- $\tanh$: Hyperbolic tangent activation function ($ -1 \ldots 1$)
+-   $\tanh$: Hyperbolic tangent activation function (\$ -1 \ldots 1\$)
 
@@
 
-We first implemented the ConvLSTM using and encoder/decoder structure as discussed in ...
-.
-To predict all 97 rivers entering the Baltic Sea at once, we flatten the output and use fully connected layers resulting in 97 outputs.
+The ConvLSTM architectures uses and encoder/decoder structure as discussed in TODO.
+To predict all 97 rivers entering the Baltic Sea, we flatten the output and use fully connected layers to map onto the individual rivers outputs.
 
@@
 | Parameter name       | Parameter size            |
-|------|------|
+|----------------------|---------------------------|
 | Channel size         | 4                         |
-| Num. hidden layer |10 |
+| Num. hidden layer    | 10                        |
 | Num. timesteps       | 30                        |
@@
 | Num. ConvLSTM layers | 1                         |
-| Batch size| 64 |
+| Batch size           | 64                        |
 |  Learning Rate       | 1e-3 with CosineAnnealing |
@@
 : Hyperparameters {#tbl-letters}
+
+As input the model receives 30 days of atmospheric surface fields temperature $T$, precipitation $P$, specific humidity $Q$ and wind speed $W$, with a daily resolution to predict the river runoff $R$ at the time step $\Delta t+1$ , which can be summarized as
+
+$R_{\Delta t+1} = f\left(T_{t-30:t}, P_{t-30:t}, Q_{t-30:t}, W_{t-30:t}\right)$
+
+with f being a function maps the 30 days of daily atmospheric surface fields data to the predicted river runoff.
+
+The choice of atmospheric fields was based on the implemented river runoff calculation in the atmospheric model COSMO-CLM which uses these four fields to calculate an river runoff estimate.
+
+## Results
+
+For the evaluation of the model performance we consider the period 1979 to 2011.
+For this period no bias correction was applied to the orignal E-HYPE dataset.
+We chose a split of 80% training data, 10% validation data to evaluate the performance of the model during training, and 10% training data that is finally used to evaluate the performance of the model after training.
+
+@fig-PerformanceNeuralNetworkRunoff shows the perfomance of the model using the test dataset.
+The predicted total river runoff for the Baltic Sea is closely matching the original data.
+Zooming in on the largest individual rivers (lower panels) it can be seen that that also the prediction of the inidivual rivers is close to the original data.
+
+![Figure ..](images/paste-1.png){#fig-PerformanceNeuralNetworkRunoff}
+
+The accuracy of the model is further displayed in @fig-statistical-evaluationNN . The correlation is close to 1 and the residuals show the error of the model is below 1% of the original data.
+
+![Figure ...](images/paste-4.png){#fig-statistical-evaluationNN}
+
+Lastly, we evaluated the performance of the runoff model by incorporating the predicted river runoff as forcing into the ocean model MOM5.
+This provides a robust validation of the runoff model against more complex real world conditions.
+This allows us to ensure that the predictions accurately reflect the impact of the river discharge on the ocean dynamics, validating the temporal and spatial variability of the the river discharge.
+@fig-by15 shows the salinity comparison between the original E-HYPE river runoff and the predicted river runoff at BY15 - a central stations in the Baltic Sea.
+
+![](images/paste-5.png){#fig-by15}
 
@@
 
-:::{#refs}
-
+::: {#refs}
 :::

### 2024-03-15 paper/paper.tex
update paper draft

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, sharp corners, interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, interior hidden, boxrule=0pt, sharp corners, enhanced]}{\end{tcolorbox}}\fi
 
@@
 
-{\centering \includegraphics{draw_lstm.pdf}
+{\centering \includegraphics{draw_lstm.png}
 
@@
 \item
-  \(\tanh\): Hyperbolic tangent activation function (\$-1 \ldots1)
+  \(\tanh\): Hyperbolic tangent activation function (\$ -1 \ldots 1\$)
 \end{itemize}
@@
 
-We first implemented the ConvLSTM using and encoder/decoder structure as
-discussed in \ldots{} . To predict all 97 rivers entering the Baltic Sea
-at once, we flatten the output and use fully connected layers resulting
-in 97 outputs.
+The ConvLSTM architectures uses and encoder/decoder structure as
+discussed in TODO. To predict all 97 rivers entering the Baltic Sea, we
+flatten the output and use fully connected layers to map onto the
+individual rivers outputs.
 
@@
 \end{longtable}
+
+As input the model receives 30 days of atmospheric surface fields
+temperature \(T\), precipitation \(P\), specific humidity \(Q\) and wind
+speed \(W\), with a daily resolution to predict the river runoff \(R\)
+at the time step \(\Delta t+1\) , which can be summarized as
+
+\(R_{\Delta t+1} = f\left(T_{t-30:t}, P_{t-30:t}, Q_{t-30:t}, W_{t-30:t}\right)\)
+
+with f being a function maps the 30 days of daily atmospheric surface
+fields data to the predicted river runoff.
+
+The choice of atmospheric fields was based on the implemented river
+runoff calculation in the atmospheric model COSMO-CLM which uses these
+four fields to calculate an river runoff estimate.
+
+\hypertarget{results}{%
+\section{Results}\label{results}}
+
+For the evaluation of the model performance we consider the period 1979
+to 2011. For this period no bias correction was applied to the orignal
+E-HYPE dataset. We chose a split of 80\% training data, 10\% validation
+data to evaluate the performance of the model during training, and 10\%
+training data that is finally used to evaluate the performance of the
+model after training.
+
+Figure~\ref{fig-PerformanceNeuralNetworkRunoff} shows the perfomance of
+the model using the test dataset. The predicted total river runoff for
+the Baltic Sea is closely matching the original data. Zooming in on the
+largest individual rivers (lower panels) it can be seen that that also
+the prediction of the inidivual rivers is close to the original data.
+
+\begin{figure}
+
+{\centering \includegraphics{images/paste-1.png}
+
+}
+
+\caption{\label{fig-PerformanceNeuralNetworkRunoff}Figure ..}
+
+\end{figure}
+
+The accuracy of the model is further displayed in
+Figure~\ref{fig-statistical-evaluationNN} . The correlation is close to
+1 and the residuals show the error of the model is below 1\% of the
+original data.
+
+\begin{figure}
+
+{\centering \includegraphics{images/paste-4.png}
+
+}
+
+\caption{\label{fig-statistical-evaluationNN}Figure \ldots{}}
+
+\end{figure}
+
+Lastly, we evaluated the performance of the runoff model by
+incorporating the predicted river runoff as forcing into the ocean model
+MOM5. This provides a robust validation of the runoff model against more
+complex real world conditions. This allows us to ensure that the
+predictions accurately reflect the impact of the river discharge on the
+ocean dynamics, validating the temporal and spatial variability of the
+the river discharge. \textbf{?@fig-by15} shows the salinity comparison
+between the original E-HYPE river runoff and the predicted river runoff
+at BY15 - a central stations in the Baltic Sea.
+
+\includegraphics{images/paste-5.png}
 

### 2024-03-15 paper/paper.qmd
Update paper

@@
 ---
-title: "Preparing your manuscript"
+title: "From Precipitation to Prediction: Integrating ConvLSTM Models for Comprehensive River Runoff Forecasting"
 format:
@@
 abstract: |
-  The abstract (1) states the nature of the investigation and (2) summarizes the important conclusions. The abstract should be suitable for indexing. Your abstract should:
-
+    a
 plain-language-summary: |
-  A Plain Language Summary (PLS) can be an incredibly effective science communication tool.
-By summarizing your paper in non-technical terms, you can explain your research and its relevance to a much broader audience.
-A PLS is required for submissions to AGU Advances, G-Cubed, GeoHealth, GRL, JAMES, JGR: Biogeosciences, JGR: Oceans, JGR: Planets, JGR: Solid Earth, JGR: Atmospheres, Space Weather, and Reviews of Geophysics, but optional for other journals.
-A PLS should be no longer than 200 words and should be free of jargon, acronyms, equations, and any technical information that would be unknown to people from outside your scientific discipline.
-Read our tips for creating an effective PLS.
+    b
 keywords: []
@@
 citation:
-  container-title: Geophysical Research Letters
+  container-title: Climate Dynamics
 keep-tex: true
@@
 However, a limitation of LSTMs is their inability to effectively capture two-dimensional structures, an area where Convolutional Neural Networks (CNNs) excel.
-Recognizing this, we use the ConvLSTM, which integrates the strengths of both LSTM and CNN, to extract spatiotemporal features from atmospheric fields for predicting river runoff.
+Reconizing this limitation [@shi2015] proposed a convolutional LSTM (ConvLSTM) architectures, which combines the strengths of both LSTM and CNN.
+The ConvLSTM network has been proven useful for precipitation nowcasting [@shi2015] or for river runoff forecasting [@ha2021; \@niu2020].
 
-TODO: More about ConvLSTM
-
-In the following we will use the Baltic Sea catchment to illustrate our approach, while in principal the methodology we propose is universally applicable across various geographic regions.
+In the following we will show that in absence of a fully functioning hydrogolocial model, that also uses a rather complex parametrization, ConvLSTM also represent a robust way to predict multiple rivers at once for any given period using only atmospheric forcing.
+In this work we use the Baltic Sea catchment to illustrate our approach, while in principal the methodology we propose is universally applicable across various geographic regions.
 The Baltic Sea serves as a challenging example due to its unique hydrological characteristics, being nearly decoupled from the open ocean (see Figure).
-
 As a consequence, the salinity of the Baltic Sea is driven to a large part by freshwater supply from rivers.
@@
 This makes the accurate modeling of river runoff especially critical for simulations pertaining to the Baltic Sea.
-In the following we show that in absence of a fully functioning hydrogolocial model, that also uses a rather complex parametrization, ConvLSTM represent a robust way to predict river runoff for any given period of time using only atmospheric forcing.
 
-In this work, we present a ConvLSTM architecture as a novel solution to this spatiotemporal sequence forecasting challenge, highlighting its advantages and potential future applications.
+In this work we will, we present a ConvLSTM architecture that is able to predict daily river runoff for 97 rivers across the Baltic Sea catchment.
 
 ## Methods
+
+### Runoff data used for training
+
+### Atmospheric Forcing
 

### 2024-03-15 paper/paper.tex
Update paper

@@
 \hypersetup{
-  pdftitle={Preparing your manuscript},
+  pdftitle={From Precipitation to Prediction: Integrating ConvLSTM Models for Comprehensive River Runoff Forecasting},
   pdfauthor={Florian Börgel; Sven Karsten},
@@
 
-\journalname{Geophysical Research Letters}
+\journalname{Climate Dynamics}
 
@@
 \begin{document}
-\title{Preparing your manuscript}
+\title{From Precipitation to Prediction: Integrating ConvLSTM Models for
+Comprehensive River Runoff Forecasting}
 
@@
 \begin{abstract}
-The abstract (1) states the nature of the investigation and (2)
-summarizes the important conclusions. The abstract should be suitable
-for indexing. Your abstract should:
+a
 \end{abstract}
@@
 \section*{Plain Language Summary}
-A Plain Language Summary (PLS) can be an incredibly effective science
-communication tool. By summarizing your paper in non-technical terms,
-you can explain your research and its relevance to a much broader
-audience. A PLS is required for submissions to AGU Advances, G-Cubed,
-GeoHealth, GRL, JAMES, JGR: Biogeosciences, JGR: Oceans, JGR: Planets,
-JGR: Solid Earth, JGR: Atmospheres, Space Weather, and Reviews of
-Geophysics, but optional for other journals. A PLS should be no longer
-than 200 words and should be free of jargon, acronyms, equations, and
-any technical information that would be unknown to people from outside
-your scientific discipline. Read our tips for creating an effective PLS.
+b
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, interior hidden, boxrule=0pt, sharp corners, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 comprises about one third of the precipitation over land areas (Hagemann
-et al., 2020). For the Baltic Sea river runoff is even more important,
-as the Baltic Sea is nearly decoupled from the open ocean (see Figure).
-As a consequence, the salinity of the Baltic Sea is driven to a large
-part by freshwater supply from rivers. More generally, the freshwater
-input into the Baltic Sea comes either as river runoff or a positive net
-precipitation (precipitation minus evaporation) over the sea surface.
-The net precipitation accounts for 11 \% and the river input for 89 \%
-of the total freshwater input (Meier and Doescher, 2002). Modeling the
-Baltic Sea is therefore to a large part the result of the quality of the
-river input, that is used for the simulation.
-
-In the context of climate change studies, river runoff is usually
-generated in two ways. First, river runoff as input for ocean models can
-be created using hyrdological models such as the Hydrological Discharge
-(HD) model (Hagemann et al., 2020). HD calculates the water balance
-using hydrological processes (e.g.~snow, glaciers, soil moisture,
-groundwater contribution). It represents a complex forecast tool that
-uses underlying physical processes. A different approach would use
-data-based models that intergrate statistical correction, using the land
-surface schemes of global or regional climate models.
+et al., 2020). In the context of climate change studies, river runoff is
+usually generated in two ways. First, river runoff as input for ocean
+models can be created using hyrdological models such as the Hydrological
+Discharge (HD) model (Hagemann et al., 2020). HD models calculates the
+water balance using hydrological processes (e.g.~snow, glaciers, soil
+moisture, groundwater contribution). It represents a complex forecasting
+tool that uses underlying physical processes. A different approach would
+use data-based models that intergrate statistical correction, using the
+land surface schemes of global or regional climate models.
 
 The releatively recent rise of machine learning (ML) models has been
-mostly explored for river runoff forcasting, as accurate runoff
+mostly explored for river runoff forecasting, as accurate runoff
 forecasting, especially over extended periods, is pivotal for effective
 water resources management (Fang et al., 2019; Tan et al., 2018; Yang et
-al., 2018). Common approach us employ artificial neural networks,
-support vector machines, adaptive neuro-fuzzy inference systems, and
-notably, Long Short-Term Memory (LSTM) neural networks and have gained
-traction for long-term hydrological forecasting due to their commendable
+al., 2018). Common approaches employ artificial neural networks, support
+vector machines, adaptive neuro-fuzzy inference systems, and notably,
+Long Short-Term Memory (LSTM) neural networks that have gained traction
+for long-term hydrological forecasting due to their excellent
 performance (Humphrey et al 2016, Huang et al 2014, Ashrafi et al 2017,
@@
 their inability to effectively capture two-dimensional structures, an
-area where Convolutional Neural Networks (CNNs) excel.
+area where Convolutional Neural Networks (CNNs) excel. Reconizing this
+limitation (SHI et al., 2015) proposed a convolutional LSTM (ConvLSTM)
+architectures, which combines the strengths of both LSTM and CNN. The
+ConvLSTM network has been proven useful for precipitation nowcasting
+(SHI et al., 2015) or for river runoff forecasting {[}Ha et al. (2021);
+@niu2020{]}.
 
-Recognizing this, we use the ConvLSTM, which integrates the strengths of
-both LSTM and CNN, to extract spatiotemporal features from atmospheric
-fields for predicting river runoff in the Baltic Sea catchment,
-summarized by 97 inidivual rivers.
+In the following we will show that in absence of a fully functioning
+hydrogolocial model, that also uses a rather complex parametrization,
+ConvLSTM also represent a robust way to predict multiple rivers at once
+for any given period using only atmospheric forcing. In this work we use
+the Baltic Sea catchment to illustrate our approach, while in principal
+the methodology we propose is universally applicable across various
+geographic regions. The Baltic Sea serves as a challenging example due
+to its unique hydrological characteristics, being nearly decoupled from
+the open ocean (see Figure). As a consequence, the salinity of the
+Baltic Sea is driven to a large part by freshwater supply from rivers.
+More generally, the freshwater input into the Baltic Sea comes either as
+river runoff or a positive net precipitation (precipitation minus
+evaporation) over the sea surface. The net precipitation accounts for 11
+\% and the river input for 89 \% of the total freshwater input (Meier
+and Doescher, 2002). Modeling the Baltic Sea is therefore to a large
+part the result of the quality of the river input, that is used for the
+simulation. This makes the accurate modeling of river runoff especially
+critical for simulations pertaining to the Baltic Sea.
 
-Following the previous description of the complex hyrdological model
-E-Hype - when present E-Hype should be used. However, in absence of a
-fully functioning hydrogolocial model, that also uses a rather complex
-parametrization, RNNs represent a robust way to predict river runoff for
-any give period of time using atmospheric forcing.
-
-Modeling the Baltic Sea is to a large part the result of the quality of
-the freshwater input, that is used for the simulation. Meier and
-Kauker(2003) showed that decadal salinity variations of about 1 \(g\)
-\(kg^{-1}\) are caused, inter alia, by annual runoff variations.
-Further, Meier and Kauker (2003) showed that about 50 \% of the decadal
-salinity variability can be explained by variations in freshwater input
-into the Baltic Sea.
-
-This paper delves into the application of deep learning, particularly
-ConvLSTM, to the challenging task of precipitation nowcasting, a domain
-yet to fully harness the potential of advanced machine learning
-techniques. We present ConvLSTM as a novel solution to this
-spatiotemporal sequence forecasting challenge, highlighting its
-advantages and potential future applications.
+In this work we will, we present a ConvLSTM architecture that is able to
+predict daily river runoff for 97 rivers across the Baltic Sea
+catchment.
 
@@
 \section{Methods}\label{methods}}
+
+\hypertarget{runoff-data-used-for-training}{%
+\subsection{Runoff data used for
+training}\label{runoff-data-used-for-training}}
+
+\hypertarget{atmospheric-forcing}{%
+\subsection{Atmospheric Forcing}\label{atmospheric-forcing}}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-ha2021}{}}%
+Ha, S., Liu, D., \& Mu, L. (2021). Prediction of Yangtze River
+streamflow based on deep learning neural network with El
+Niño{\textendash}Southern Oscillation. \emph{Scientific Reports},
+\emph{11}(1), 11738. \url{https://doi.org/10.1038/s41598-021-90964-3}
+
 \leavevmode\vadjust pre{\hypertarget{ref-hagemannHighResolutionDischarge2020}{}}%
@@
 \url{https://doi.org/10.3389/feart.2020.00012}
+
+\leavevmode\vadjust pre{\hypertarget{ref-shi2015}{}}%
+SHI, X., Chen, Z., Wang, H., Yeung, D.-Y., Wong, W., \& WOO, W. (2015).
+Convolutional LSTM network: A machine learning approach for
+precipitation nowcasting. In (Vol. 28). Curran Associates, Inc.
+Retrieved from
+\url{https://proceedings.neurips.cc/paper/2015/hash/07563a3fe3bbe7e3ba84431ad9d055af-Abstract.html}
 

### 2024-03-18 paper/paper.qmd
paper

@@
 
+The runoff data covering the period 1979 to 2011 is based on an E-HYPE hindcast simulation that was forced by a regional downscaling of ERA-Interim [@dee2011] with RCA3 [@theross] and implemented into NEMO-Nordic [@hordoir2019] as a mass flux.
+For the periods before (1961 to 1978) and after (2012 to 2018) additional spatial temporal corrections have been applied to the runoff data, and have therefore been ignored.
+For more information see @gröger2022 and references therein.
+
 ### Atmospheric Forcing
+
+The UERRA-HARMONIE regional reanalysis dataset was developed as part of the FP7 UERRA project (Uncertainties in Ensembles of Regional Re-Analyses, [http://www.uerra.eu/)](http://www.uerra.eu/),).
+The UERRA-HARMONIE system represents a comprehensive, high-resolution reanalysis covering a wide range of essential climate variables.
+This dataset encompasses data on air temperature, pressure, humidity, wind speed and direction, cloud cover, precipitation, albedo, surface heat fluxes, and radiation fluxes from January 1961 to July 2019.
+With a horizontal resolution of 11 km and analyses conducted at 00 UTC, 06 UTC, 12 UTC, and 18 UTC, it also provides hourly resolution forecast model data.
+UERRA-HARMONIE is accessible through the Copernicus Climate Data Store (CDS, <https://cds.climate.copernicus.eu/#!/home),> initially produced during the UERRA project and later transitioned to the Copernicus Climate Change Service (C3S, <https://climate.copernicus.eu/copernicus-regionalreanalysis-europe).>
+
+### Ocean Model
+
+To simulate the Baltic Sea, we use a coupled 3-dimensional ocean model Baltic Sea, called the Modular Ocean Model (MOM).
+This model uses a finite-difference method to solve the full set of primitive equations to calculate the motion of water and the transport of heat and salt.
+The K-profile parameterization (KPP) was used as turbulence closure scheme.
+The model's western boundary opens into the Skagerrak and connects the Baltic Sea to the North Sea.
+The maximum depth was set at 264 meters.
+A more detailed description of the setup can be found in [@gröger2022].
 
@@
 
+![](images/paste-6.png)
+
 ## Acknowledgments

### 2024-03-18 paper/paper.tex
paper

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced, interior hidden, sharp corners, frame hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 
+The runoff data covering the period 1979 to 2011 is based on an E-HYPE
+hindcast simulation that was forced by a regional downscaling of
+ERA-Interim (Dee et al., 2011) with RCA3 ({``The rossby centre regional
+climate model RCA3,''} n.d.) and implemented into NEMO-Nordic (Hordoir
+et al., 2019) as a mass flux. For the periods before (1961 to 1978) and
+after (2012 to 2018) additional spatial temporal corrections have been
+applied to the runoff data, and have therefore been ignored. For more
+information see Gröger et al. (2022) and references therein.
+
 \hypertarget{atmospheric-forcing}{%
 \subsection{Atmospheric Forcing}\label{atmospheric-forcing}}
+
+The UERRA-HARMONIE regional reanalysis dataset was developed as part of
+the FP7 UERRA project (Uncertainties in Ensembles of Regional
+Re-Analyses, \href{http://www.uerra.eu/}{http://www.uerra.eu/)},). The
+UERRA-HARMONIE system represents a comprehensive, high-resolution
+reanalysis covering a wide range of essential climate variables. This
+dataset encompasses data on air temperature, pressure, humidity, wind
+speed and direction, cloud cover, precipitation, albedo, surface heat
+fluxes, and radiation fluxes from January 1961 to July 2019. With a
+horizontal resolution of 11 km and analyses conducted at 00 UTC, 06 UTC,
+12 UTC, and 18 UTC, it also provides hourly resolution forecast model
+data. UERRA-HARMONIE is accessible through the Copernicus Climate Data
+Store (CDS, \url{https://cds.climate.copernicus.eu/\#!/home),} initially
+produced during the UERRA project and later transitioned to the
+Copernicus Climate Change Service (C3S,
+\url{https://climate.copernicus.eu/copernicus-regionalreanalysis-europe).}
+
+\hypertarget{ocean-model}{%
+\subsection{Ocean Model}\label{ocean-model}}
+
+To simulate the Baltic Sea, we use a coupled 3-dimensional ocean model
+Baltic Sea, called the Modular Ocean Model (MOM). This model uses a
+finite-difference method to solve the full set of primitive equations to
+calculate the motion of water and the transport of heat and salt. The
+K-profile parameterization (KPP) was used as turbulence closure scheme.
+The model's western boundary opens into the Skagerrak and connects the
+Baltic Sea to the North Sea. The maximum depth was set at 264 meters. A
+more detailed description of the setup can be found in (Gröger et al.,
+2022).
 
@@
 
+\includegraphics{images/paste-6.png}
+
 \hypertarget{acknowledgments}{%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-dee2011}{}}%
+Dee, D. P., Uppala, S. M., Simmons, A. J., Berrisford, P., Poli, P.,
+Kobayashi, S., et al. (2011). The ERA-Interim reanalysis: configuration
+and performance of the data assimilation system. \emph{Quarterly Journal
+of the Royal Meteorological Society}, \emph{137}(656), 553--597.
+\url{https://doi.org/10.1002/qj.828}
+
 \leavevmode\vadjust pre{\hypertarget{ref-fangExaminingApplicabilityDifferent2019}{}}%
@@
 \url{https://doi.org/10.1016/j.jhydrol.2018.11.020}
+
+\leavevmode\vadjust pre{\hypertarget{ref-gruxf6ger2022}{}}%
+Gröger, M., Placke, M., Meier, M., Börgel, F., Brunnabend, S.-E.,
+Dutheil, C., et al. (2022). \emph{The Baltic Sea model inter-comparison
+project BMIP {\textendash} a platform for model development, evaluation,
+and uncertainty assessment}. Retrieved from
+\url{https://gmd.copernicus.org/preprints/gmd-2022-160/}
 
@@
 \url{https://doi.org/10.3389/feart.2020.00012}
+
+\leavevmode\vadjust pre{\hypertarget{ref-hordoir2019}{}}%
+Hordoir, R., Axell, L., Höglund, A., Dieterich, C., Fransner, F.,
+Gröger, M., et al. (2019). Nemo-nordic 1.0: A NEMO-based ocean model for
+the baltic and north seas {\textendash} research and operational
+applications. \emph{Geoscientific Model Development}, \emph{12}(1),
+363--386. \url{https://doi.org/10.5194/gmd-12-363-2019}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-theross}{}}%
+The rossby centre regional climate model RCA3: Model description and
+performance - tellus a: Dynamic meteorology and oceanography. (n.d.).
+Retrieved from
+\url{https://a.tellusjournals.se/articles/10.1111/j.1600-0870.2010.00478.x}
+
 \leavevmode\vadjust pre{\hypertarget{ref-yangDerivingOperatingRules2018}{}}%

### 2024-03-18 paper/paper.tex
update newest

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced, interior hidden, sharp corners, frame hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, interior hidden, breakable, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 
-\includegraphics{images/paste-5.png}
-
-\includegraphics{images/paste-6.png}
+\includegraphics{images/paste-7.png}
 
 \hypertarget{acknowledgments}{%
-\section{Acknowledgments}\label{acknowledgments}}
+\section[Acknowledgments]{\texorpdfstring{\protect\includegraphics{images/paste-8.png}Acknowledgments}{Acknowledgments}}\label{acknowledgments}}
 

### 2024-03-20 paper/paper.qmd
update newest
most recent images added

@@
 ---
+git pul---
 title: "From Precipitation to Prediction: Integrating ConvLSTM Models for Comprehensive River Runoff Forecasting"
@@
 
-@fig-PerformanceNeuralNetworkRunoff shows the perfomance of the model using the test dataset.
+The accuracy of the model is further displayed in @fig-statistical-evaluationNN . The correlation is close to 1 and the residuals show the error of the model is below 1% of the original data.
+
+![](images/paste-13.png)
+
+@fig-PerformanceNeuralNetworkRunoff shows the performance of the model using the test dataset.
 The predicted total river runoff for the Baltic Sea is closely matching the original data.
@@
 
-![Figure ..](images/paste-1.png){#fig-PerformanceNeuralNetworkRunoff}
-
-The accuracy of the model is further displayed in @fig-statistical-evaluationNN . The correlation is close to 1 and the residuals show the error of the model is below 1% of the original data.
-
-![Figure ...](images/paste-4.png){#fig-statistical-evaluationNN}
+![](images/paper_total_river_runoff.png)
 
@@
 
-![](images/paste-5.png){#fig-by15}
+![](images/paste-9.png)
 
-![](images/paste-6.png)
+![](images/paste-12.png)
 

### 2024-04-16 BalticRiverPrediction/BaltNet.py
Removed all unnecessary stuff form BaltNet architecture

@@ -35,21 +35,11 @@ class BaltNet(nn.Module):
         for k, v in modelPar.items():
             setattr(self, k, v)
 
-        self.encoder = ConvLSTM(
+        self.convLSTM = ConvLSTM(
             input_dim=self.input_dim,
             hidden_dim=self.hidden_dim,
             kernel_size=self.kernel_size,
             num_layers=self.num_layers,
-            batch_first=self.batch_first,
-            bias=self.bias,
-            return_all_layers=False
-        )
-
-        self.decoder = ConvLSTM(
-            input_dim=self.hidden_dim,
-            hidden_dim=self.hidden_dim,
-            kernel_size=self.kernel_size,
-            num_layers=1,
             batch_first=self.batch_first,
             bias=self.bias,
             return_all_layers=False
@@ -67,48 +57,21 @@ class BaltNet(nn.Module):
             nn.Linear(256, 97)
         )
 
-        # Creating separate attention weights for each river
-        # self.attention_weights = nn.Parameter(torch.randn(self.hidden_dim, 1, 1), requires_grad=True)  # 97 rivers
-
-    # def spatial_attention(self, x):
-    #     """Spatial attention mechanism."""
-    #     B, T, C, H, W = x.size()
-
-    #     x = x.view(B * T, C, H, W)
-        
-    #     # Apply attention weights for all rivers
-    #     self.attention_map = torch.sigmoid(F.conv2d(x, self.attention_weights.unsqueeze(0), bias=None, stride=1, padding=0))
-        
-    #     # Weighted sum
-    #     output = x * self.attention_map  # B*T, C, H, W
-    #     output = output.view(B, T, C, H, W)  # B, T, C
-
-    #     return output
-
     def forward(self, x):
         B, _, _, _, _ = x.size()
 
-        # Pass through encoder
-        encoder_outputs, encoder_hidden = self.encoder(x)
+        # Pass through convLSTM 
+        convLSTM_outputs, _ = self.convLSTM(x)
 
-        # Use the entire encoder output as input to the decoder
-        decoder_input = encoder_outputs[0][:,-1,:,:,:].unsqueeze(1)
-
-        # Pass through decoder using the final hidden state of the encoder
-        decoder_outputs, _ = self.decoder(decoder_input, encoder_hidden)
-
-        # Apply spatial attention
[…]

### 2024-04-16 paper/paper.qmd
More text in results

@@
 
-For the evaluation of the model performance we consider the period 1979 to 2011.
-For this period no bias correction was applied to the orignal E-HYPE dataset.
-We chose a split of 80% training data, 10% validation data to evaluate the performance of the model during training, and 10% training data that is finally used to evaluate the performance of the model after training.
+The model was trained with daily data for the period 1979 to 2011, as this period represents the mapping between the original atmospheric forcing and E-HYPE without further bias correction to match observations.
+The data was then divided into randomly chosen splits of 80% training data, 10% validation data to evaluate the performance of the model during training, and 10% training data that is finally used to evaluate the performance of the model after training.
+The model was trained for 400 epochs and the model weights with the lowest mean squared error during training have been stored.
 
-The accuracy of the model is further displayed in @fig-statistical-evaluationNN . The correlation is close to 1 and the residuals show the error of the model is below 1% of the original data.
+The accuracy of the model is displayed in Figure @fig-statistical-evaluationNN.
+For this evaluation, the test dataset was utilized.
+The left panel presents a scatter plot comparing the original E-HYPE data with the predicted river runoff, revealing a high correlation of 0.997.
+The middle panel illustrates the relative prediction error in relation to the original E-HYPE data, indicating that, on daily timescales, the model can predict river runoff with an accuracy of $\pm$ 5%.
+Partly, this may be attributed to the use of a different atmospheric dataset than that what was used to force the original E-HYPE model.
+However, by applying a rolling-mean with a 5-day window, the prediction error is reduced to less than 1%.
+Lastly, the right panel demonstrates that the distribution of residuals follows a Gaussian shape, suggesting that our model does not exhibit bias by systematically over or underestimating river runoff values.
 
-![](images/paste-13.png)
+![](../src/figures/paper_error_metrics.png)
 
-@fig-PerformanceNeuralNetworkRunoff shows the performance of the model using the test dataset.
+@fig-PerformanceNeuralNetworkRunoff shows the predicted and the original river runoff using the test dataset.
 The predicted total river runoff for the Baltic Sea is closely matching the original data.
-Zooming in on the largest individual rivers (lower panels) it can be seen that that also the prediction of the inidivual rivers is close to the original data.
+Zooming in on the largest individual rivers (lower panels) it can be seen that that also the prediction of the individual rivers is close to the original data.
 
@@
 
-Lastly, we evaluated the performance of the runoff model by incorporating the predicted river runoff as forcing into the ocean model MOM5.
+Lastly, we evaluated the performance of the runoff model by incorporating the predicted river runoff as runoff forcing into the ocean model MOM5.
 This provides a robust validation of the runoff model against more complex real world conditions.
@@
 @fig-by15 shows the salinity comparison between the original E-HYPE river runoff and the predicted river runoff at BY15 - a central stations in the Baltic Sea.
+It can be seen that the model simulation using the predicted river runoff by the ConvLSTM is closely mirroring the control simulation.
+The left panel shows the surface salinity, representing the high-frequency variations in the salinity, which is heavily affected by river runoff.
+The right panel shows the bottom salinity which can be viewed as a low-pass filter in the Baltic Sea, which is also closely mirrored by the ConvLSTM predictions.
 

### 2024-04-18 paper/paper.qmd
WIP on the methods part in paper.qmd

--- 2024-04-16 paper/paper.qmd
@@
     email: sven.karsten@io-warnemuende.de
+
+  - name: Karoline Rummel
+    affiliations:
+      - ref: IOW
+    email: karoline.rummel@io-warnemuende.de
 
@@
 
-The Long Short-Term Memory (LSTM), a specialized form of Recurrent Neural Networks (RNNs), is specifically tailored for modeling temporal sequences.
+Before turning to the convolutional LSTM, the simpler architecture of the plain LSTM is examined and should serve as a role model for the later consideration of the full ConvLSTM.
+
+The Long Short-Term Memory (LSTM), a specialized form of Recurrent Neural Networks (RNNs), is specifically tailored for modeling temporal sequences $\vec{X}^t[1], \ldots \vec{X}^t[\tau],\ldots \vec{X}^t[N_\tau]$ of $N_\tau$ input quantities $\vec{X}^t[\tau] = (X^t_k[\tau]) \in \mathbb{R}^{N_k}$.
+The sequence is taken from a dataset given in form of a time series $\{\vec{X}^t\}$ and the endpoint should coincide with the specific element in the time series connected to time $t$, i.e.
+$\vec{X}^t[N_\tau] \equiv \vec{X}^t$.
+The $N_k$ is the number of used input "channels" and can, for example, represent different measurable quantities.
 Its unique design allows it to adeptly handle long-range dependencies, setting it apart from traditional RNNs in terms of accuracy (see @fig-lstm).
@@
 This performance in modeling long-range dependencies has been validated in various studies.
-The key component of LSTM's innovation is its memory cell, $c_t$t, which stores state information, also refered to as long-term memory.
-This cell is accessed, modified, and reset through several self-parameterized gates.
-For the input of the sequence $x_t$ input, the forget gate $f_t$ defines the percentage of the previous long-term memory status $c_{t-1}$ that should be retained stored​.
-Next the input gate $i_t$ decides how much of the input is added to the the long-term memory, forming the updated cell state $c_{t}$.
-The decision to propagate the latest cell output, $c_t$, to the final state, $h_t$, is governed by the output gate, $o_t$, representing the updated short-term memory of the hidden state $h_t$.
+The key component of LSTM's innovation is its memory cell, $\vec{C}^t[\tau] = (C^t_h[\tau]) \in \mathbb{R}^{N_h}$, which stores state information, also referred to as long-term memory.
+The index $\tau \in [1, N_\tau]$ corresponds to the $\tau$-th element in the input sequence.
+The cell state is a vector consists of elements, each associated with one of the $N_h$ hidden layers, labeled by $h$.
+This vector is determined through several self-parameterized gates, all in the same vector space as $\vec{C}^t[\tau]$.
+The forget gate $\vec{F}^t[\tau]$ defines the percentage of the previous long-term memory status $\vec{C}^t[\tau-1]$ that should be retained stored​.
+The input gate $\vec{I}^t[\tau]$ decides how much of the input is added to the the long-term memory, forming the updated cell state.
+The so-called output gate $\vec{O}^t[\tau]$ then determines how much of the latest cell output, $\vec{C}^t[\tau]$, is propagated to the final state, $\vec{H}^t[\tau]$ representing the updated short-term memory of the hidden state.
+
+For a fixed point $\tau$ in the sequence, the action of a LSTM cell, i.e.
+the connection between the input $\vec{X}^t[\tau]$, the various gates and the state vectors, is given as follows.
+First, the elements of the input sequence together with the hidden state are mapped onto auxiliary gate vectors, collectively denoted by $\vec{g}^t[\tau] = (g^t_h[\tau]) \in \mathbb{R}^{N_h}$, via $$
+\begin{aligned}
+g_h^t[\tau] =  \mathcal{M}^{g}_{hk} \, X^t_k[\tau] +  \mathcal{N}^{g}_{hh'} \, H^t_{h'}[\tau-1] + \mathcal{B}^g_{h} \ ,
+\end{aligned}
+$$
+
+where $g=i,f,o,c$ stands for the input, forget, output and cell-state gate, respectively and Eistein's summation convention is employed, i.e.
+indices that appear twice are summed over.
+The calligraphic symbols $\mathcal{M}^{g}_{hk}, \mathcal{N}^{g}_{hh'}$ and $\mathcal{B}^g_{h}$ are the free parameters of the network that are optimized for the given problem during the training, which is at the heart of the any machine learning approach.
+The matrix $\pmb{\mathcal{M}}^{g} = (\mathcal{M}^{g}_{hk}) \in \mathbb{R}^{N_h \times N_k}$ can be interpreted as a Markovian-like contribution of the current input $\vec{X}^t[\tau]$ to the gates, whereas the $\pmb{\mathcal{N}}^{g} = (\mathcal{N}^{g}_{hh'}) \in \mathbb{R}^{N_h \times N_h}$ scales a non-Markovian part determined by the hidden state of the last sequence point $\tau-1$.
+The vector $\vec{\mathcal{B}}^g = (\mathcal{B}^g_{h}) \in \mathbb{R}^{N_h}$ is a learnable bias.
+It should be stressed that these parameters do neither depend on $t$ nor on $\tau$ and are thus optimized once for the complete dataset $\{\vec{X}^t\}$.
+
+Note that this mapping is sometimes extended by a contribution to the $g_h^t[\tau]$ from the past cell state $\vec{C}^t[\tau-1]$ [@XXX].
+Nevertheless, this mechanism called "peeping" is not further considered in this work.
+
+For the sake of brevity, we write the mapping more compactly in matrix-vector form as $$
+\begin{aligned}
+\vec{g}^t[\tau] = \pmb{\mathcal{M}}^{g}\vec{X}^t[\tau] + \pmb{\mathcal{N}}^{g}  \vec{H}^t[ \tau-1]+ \vec{\mathcal{B}}^g \,
+\end{aligned}
+$$ {#eq-LSTM-gates}
+
+Second, the actual gate vectors are computed by the core equations of the LSTM as proposed by [@XXX]:
+
+$$
+\begin{aligned}
+\vec{I}^t[\tau] &= \sigma(\vec{i}^t[\tau]) \\
+\vec{F}^t[\tau] &= \sigma(\vec{f}^t[\tau]) \\
+\vec{O}^t[\tau] &= \sigma(\vec{o}^t[\tau]) \\
+\vec{C}^t[\tau] &= \vec{F}^t[\tau] \circ \vec{C}^t[\tau-1]  + \vec{I}^t[\tau] \circ \tanh(\vec{c}^t[\tau]) \\
+\vec{H}^t[\tau]  &= \vec{O}^t[\tau] \circ \tanh(\vec{C}^t[\tau]) \ ,
+\end{aligned}
+$$ {#eq-LSTM}
+
+where $\sigma$ denotes the logisitc sigmoid function, $\tanh$ is the hyperbolic tangent and the $\circ$ stands for the Hadamard product (all applied in an element-wise fashion to the vectors).
+In the last two equations the role of the input, forget and output gates as described above become apparent.
+
+The third step in a single layer LSTM (as employed for the work presented here) is then to provide the output of the current LSTM cell, i.e.
+$\vec{H}^t[\tau]$ and $\vec{C}^t[\tau]$, to the subsequent LSTM cell that processes the next element $\vec{X}^t[\tau+1]$ of the input sequence.
+
+The full action of the LSTM network up to the end of the sequence can therefore be written as a nested function call
+
+$$
+\left(\vec{H}^t[N_\tau], \vec{C}^t[N_\tau] \right) = L \left( \vec{X}^t[N_\tau], L \left( \vec{X}^t[N_\tau-1], \ldots L \left( \vec{X}^t[1], (\vec{H}^t[0], \vec{C}^t[0] \right) \ldots \right) \right) \ ,
+$$
+
+where $L$ represents Eqs. (@eq-LSTM-gates, @eq-LSTM) and the initial conditions are usually chosen as $\vec{H}^t[0]=\vec{C}^t[0]=0$.
+
+The final output $\vec{H}^t[N_\tau]$ and $\vec{C}^t[N_\tau]$ encode information on the full input sequence ending at time $t$.
+This information has to be decoded to obtain usueful information via an appropriate subsequent layer, as it is described in the Sec.
+@XXX.
+
 A significant advantage of this architecture is the memory cell's ability to retain gradients.
@@
 
-One LSTM cell hence maybe expressed as:
+### ConvLSTM network
 
-$$
+Although the plain LSTM has high performance in handling temporal sequences of point-like quantities it is not designed to recognize spatial features in seuqences of two-dimensional maps.
+To adress this limitation we use a convLSTM architecture.
+
+In this kind of network the elements of the input sequence are given as (kind of) vector fields $\vec{X}^t[\tau] = (X^t_k[x,y,\tau]) \in \mathbb{R}^{N_k \times (N_x \times N_y)}$, where $x\in[1, N_x]$ and $y \in [1, N_y]$ run over the horizontal and vertical dimensions of the map, respectively.
+In order to enable the "learning" of spatial patterns, the free parameters of the network are replaced by two-dimenmsional convolutional kernels $\pmb{\mathcal{M}}^{g} = (\mathcal{M}^{g}_{hk}[\xi, \eta]) \in \mathbb{R}^{(N_h \times N_k)\times (N_\xi \times N_\eta)}$ and $\pmb{\mathcal{N}}^{g} = (\mathcal{N}^{g}_{hh'}[\xi, \eta]) \in \mathbb{R}^{(N_h \times N_h)\times (N_\xi \times N_\eta)}$.
+The width and the height of the kernels are given by $N_\xi$ and $N_\eta$, respectively and $\xi\in [-(N_\xi-1)/2,(N_\xi-1)/2], \eta\in [-(N_\eta-1)/2,(N_\eta-1)/2]$, where we assume odd numbers for the kernel sizes.
+
+The mapping from the input quantities to the gates is then given by a convolution with these kernels $$
 \begin{aligned}
-i_t &= \sigma(W_{xi} x_t + W_{hi} h_{t-1} + W_{ci} \circ c_{t-1} + b_i) \\
-f_t &= \sigma(W_{xf} x_t + W_{hf} h_{t-1} + W_{cf} \circ c_{t-1} + b_f) \\
-c_t &= f_t \circ c_{t-1} + i_t \circ \tanh(W_{xc} x_t + W_{hc} h_{t-1} + b_c) \\
-o_t &= \sigma(W_{xo} x_t + W_{ho} h_{t-1} + W_{co} \circ c_t + b_o) \\
-h_t &= o_t \circ \tanh(c_t)
+g^t_h[x,y,\tau] & =  \mathcal{M}^{g}_{hk} [\xi,\eta]\, X^t_k[x-\xi, y-\eta, \tau]  +  \mathcal{N}^{g}_{hh'}[\xi,\eta] \, H^t_{h'}[x-\xi, y-\eta, \tau-1] + \mathcal{B}^g_{h}[x,y] \ .
 \end{aligned}
@@
 
-with
-
--   $x_t$: Input vector at time step $t$.
--   $h_{t-1}$: Hidden state from the previous time step.
--   $C_{t-1}$: Cell state from the previous time step.
--   $W$ and $b$: Weight matrices and bias vectors, respectively, associated with the gates of the LSTM.
-The subscripts denote the specific gate or operation they are associated with (e.g., $W_f$ and $b_f$ are the weight matrix and bias for the forget gate).
--   $\sigma$: Sigmoid activation function ($0,\ldots 1$)
--   $\tanh$: Hyperbolic tangent activation function (\$ -1 \ldots 1\$)
-
-### ConvLSTM network
-
-The FC-LSTM fails to handle information when handling spatiotemporal data due to its reliance on full connections in both input-to-state and state-to-state transitions.
-To adress this limitation we use a convLSTM architecture.
-convLSTM replaces the fully connected operations in the LSTM with convolutional operations.
-Hence, all inputs $X_1, \ldots, X_t$, cell outputs $C_1, \ldots, C_t$, hidden states $H_1, \ldots, H_t$, and gates $i_t, f_t, o_t$ of the ConvLSTM are 3D tensors.
-The last two dimensions of these tensors represent spatial dimensions, specifically rows and columns.
-Conceptually, these inputs and states can be visualized as vectors positioned on a spatial grid.
-
-In the ConvLSTM, the future state of a specific cell on this grid is determined by the inputs and past states of its neighboring cells.
-This spatial consideration is integrated by employing a convolution operator in both state-to-state and input-to-state transitions, as illustrated in Fig.
-2.
-The foundational equations for ConvLSTM are:
-
+It becomes immediatly apparent that in case of the convLSTM, the bias, gate and state vectors must become vector fields ($\in \mathbb{R}^{N_h \times (N_x \times N_y)}$) as well.
+We can write this mapping in the same fashion as Eq.
+(@eq-LSTM-gates) but with replacing the normal matrix-vector multiplication by a convolution (denoted with $\ast$), i.e.\
 $$
 \begin{aligned}
-i_t &= \sigma(W_{xi} \ast X_t + W_{hi} \ast H_{t-1} + W_{ci} \circ C_{t-1} + b_i) \\
-f_t &= \sigma(W_{xf} \ast X_t + W_{hf} \ast H_{t-1} + W_{cf} \circ C_{t-1} + b_f) \\
-C_t &= f_t \circ C_{t-1} + i_t \circ \tanh(W_{xc} \ast X_t + W_{hc} \ast H_{t-1} + b_c) \\
-o_t &= \sigma(W_{xo} \ast X_t + W_{ho} \ast H_{t-1} + W_{co} \circ C_t + b_o) \\
-H_t &= o_t \circ \tanh(C_t)
+\vec{g}^t[\tau] = \pmb{\mathcal{M}}^{g} \ast \vec{X}^t[\tau] + \pmb{\mathcal{N}}^{g} \ast  \vec{H}^t[ \tau-1]+ \vec{\mathcal{B}}^g \,
 \end{aligned}
 $$
+
+The susebsequent processing of the $\vec{g}^t[\tau]$ remains symbolically the same as presented in Eq.
+(@eq-LSTM) but with all appearing quantities meaning standing now for vector fields instead of simple vectors.
 

### 2024-04-18 equations/equations.tex

\documentclass[10pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{fourier}
\usepackage{amsbsy}
\usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}
\usepackage[most]{tcolorbox}
\tcbset{colback=yellow!10!white, colframe=red!50!black, 
        highlight math style= {enhanced, %<-- needed for the ’remember’ options
            colframe=red,colback=red!10!white,boxsep=0pt}
        }

\author{Sven Karsten}

\newcommand{\del}[2]{\frac{\partial #1}{\partial #2}}

\begin{document}

\section{Runoff prediction architecture}

\subsection{Encoder}

Input quantity $ X^t_k[x, y, \tau]$ consists of a $N_\tau$ long sequence (starting at $t-N_\tau+1$ and ending at time $t$)  of $N_k$ 2D maps with $N_x \times N_y$ pixels, i.e.
\begin{align}
x \in [1,N_x] \\
y \in [1,N_y] \\
\tau \in [1,N_\tau] \\
k \in [1,N_k]
\end{align}

For any gate $g \in [i, f, o, c] $ define the gate input as a spatial convolution of the instantaneous input with a learnable "Markovian" weighting kernel plus a "non-Markovian" part that consists of the hidden state from the previous time instance convoluted with another learnable kernel  
\begin{align}
g^t_h[x,y,\tau] & =  \mathcal{M}^{g}_{hk} [\xi,\eta]\, X^t_k[x-\xi, y-\eta, \tau] \\
& +  \mathcal{N}^{g}_{hh'}[\xi,\eta] \, H^t_{h'}[x-\xi, y-\eta, \tau-1] + \mathcal{B}^g_{h}[x,y]
\end{align}
%
The auxiliary dimensions are
\begin{align}
h,h' \in [1,N_h] \\
\xi\in [-(N_\xi-1)/2,(N_\xi-1)/2]\\
\eta\in [-(N_\eta-1)/2,(N_\eta-1)/2]
\end{align}
%
where the convolution kernel sizes $N_\xi, N_\eta$ are assumed to be odd.
%
Thus the adjustable objects have dimensions
\begin{align}
\mathcal{M}^{g}_{hk}[\xi,\eta] \in \mathbb{R}^{4\times(N_h\times N_c) \times(N_\xi \times N_\eta)} \\
\mathcal{N}^{g}_{hh'}[\xi,\eta] \in \mathbb{R}^{4\times(N_h\times N_h) \times (N_\xi \times N_\eta)} \\
\mathcal{B}^g_{h}[x,y] \in \mathbb{R}^{4 \times(N_h) \times (N_x\times N_y) }
\end{align}

%Concatenate Markovian and non-Markovian parts
%\begin{align}
%G^g_h[x,y,\tau,t]  = \left(\mathcal{M}^{g}_{hc} [\xi,\eta], \mathcal{N}^{g}_{hh'}[\xi,\eta]\right) \cdot \left(X_c[x-\xi, y-\eta, t-N_\tau+\tau], H_{h'}[x-\xi, y-\eta, \tau-1,t]\right)+ \mathcal{B}^g_{h}[x,y]
%\end{align}
%
We can now introduce compact matrix-vector and spatial convolution notation
\begin{align}
\vec{g}^t[\tau] = \left(\pmb{\mathcal{M}}^{g}, \pmb{\mathcal{N}}^{g}\right) \ast \left(\vec{X}^t[\tau], \vec{H}^t[ \tau-1]\right)+ \vec{\mathcal{B}}^g
\end{align}

The hidden state evolves according to the following iteration formula for LSTM networks
%
\begin{align}
I^t_h[x,y,\tau] & = \sigma (i^t_h[x,y,\tau]) \\
F^t_h[x,y,\tau] & = \sigma (f^t_h[x,y,\tau]) \\
O^t_h[x,y,\tau] & = \sigma (o^t_h[x,y,\tau]) \\
C^t_h[x,y,\tau] & = F^t_h[x,y,\tau] \, C^t_h[x,y,\tau-1] + I^t_h[x,y,\tau] \, \tanh (c^t_h[x,y,\tau]) \\
H^t_h[x,y,\tau] & = O^t_h[x,y,\tau]\, \tanh(C^t_h[x,y,\tau])
\end{align}
%
going over $\tau \in [1,N_\tau]$ with initial conditions $H_h[x,y,0,t]=0$ and $C_h[x,y,0,t]=0$.
%
The stated LSTM connections of the gates can be abbreviated by a functions $L$ that returns the hidden network state for the next iteration step
%
\begin{align}
\left(\vec{H}^t[\tau], \vec{C}^t[\tau] \right) = L \left( \left\{ \vec{g}^t[\tau] \right\}, \vec{C}^t[\tau-1] \right) 
\end{align}

\subsection{Decoder}

Take last hidden state of encoder $H^t_h[x,y,N_\tau]$ and use it as input for the decoder.
%
Perform a single step (no iteration anymore!) with initial conditions $H^t_h[x,y,N_\tau]$ and $C^t_h[x,y,N_\tau]$, thus
%
\begin{align}
\tilde{g}^t_h[x,y] & =  \tilde{\mathcal{M}}^{g}_{hh'} [\xi,\eta]\, H^t_{h'}[x-\xi, y-\eta,N_\tau] \\
& +  \tilde{\mathcal{N}}^{g}_{hh'}[\xi,\eta] \, H^t_{h'}[x-\xi, y-\eta, N_\tau] + \tilde{\mathcal{B}}^g_{h}[x,y] \\
& = \tilde{\mathcal{Q}}^{g}_{hh'} [\xi,\eta]\, H^t_{h'}[x-\xi, y-\eta,N_\tau] + \tilde{\mathcal{B}}^g_{h}[x,y]
\end{align}
%
\begin{align}
\tilde{I}^t_h[x,y] & = \sigma (\tilde{i}^t_h[x,y]) \\
\tilde{F}^t_h[x,y] & = \sigma (\tilde{f}^t_h[x,y]) \\
\tilde{O}^t_h[x,y] & = \sigma (\tilde{o}^t_h[x,y]) \\
\tilde{C}^t_h[x,y] & = \tilde{F}^t_h[x,y,\tau] \, C^t_h[x,y,N_\tau] + \tilde{I}^t_h[x,y] \, \tanh (\tilde{c}^t_h[x,y]) \\
\tilde{H}^t_h[x,y] & = \tilde{O}^t_h[x,y]\, \tanh(\tilde{C}^t_h[x,y])
\end{align}
%
The action of the decoder can be compactly written as
\begin{align}
\left(\tilde{\vec{H}}^t, \tilde{\vec{C}}^t \right) = L \left( \left\{ \tilde{\vec{g}}^t \right\}, \vec{C}^t[N_\tau] \right) 
\end{align}

\subsection{Fully connected layer}

Choose $a \in [1, 512]$, $b \in [1, 256]$ then
%
\begin{align}
R_r[t] = \mathcal{W}^{3}_{rb}\mathrm{ReLU}\left(\mathcal{W}^{2}_{ba}\mathrm{ReLU}\left(\mathcal{W}^{1}_{ah} [x,y] \tilde{H}^t_h[x,y]\right)\right)
\end{align}
%
with $r \in [1, 97]$.

\section{Interpretation of a linearized model}

We can write our final result $R_r[t]$ for the $r$-th river as a function $M_r$ (for model) which itself is a nested function call of the $r$-th component of the fully connected layer $F$, the decoder $D$ and the encoder $E$ on a given input time sequence  $\{X^t_k[x,y,\tau]\}$, i.e.
\begin{tcolorbox}[ams align]
R_r[t] = M_r(\{X^t_k[x,y,\tau]\}) = F_r(D(E(\{X^t_k[x,y,\tau]\})))
\end{tcolorbox}
%
The idea is to approximate each $R_r[t]$ like
%
\begin{tcolorbox}[ams align]
R_r[t] \approx A_r[t] = a_r + \omega_{rk}[x,y,\tau] X^t_k[x,y,\tau]
\end{tcolorbox}
%
or expressed in continuous variables
\begin{align}
A_r(t) = a_r + \intop_{0}^{\tau_{\text{max}}} \text{d}\tau \intop_{\mathbb{R}^2} \text{d}x \text{d}y \, \omega_{rk}(x,y,\tau) X^t_k(x,y,\tau)
\end{align}
%
which means that the runoff is approximately determined by the spatially dependent memory kernel that weights the influence of each time instance of each channel back to the time $t-\tau_{\text{max}}$.

The offset $a_r$ and the spatiotemporal weighting $\omega_{rk}[x,y,\tau]$ should be given by the first order Taylor expansion around a reference input sequence $\{X_k[x,y,\tau]\}$ that is not connected to any specific time $t$ (thus no superscript $t$) i.e.
%
\begin{align}
A_r[t] = M_r(\{X_k[x,y,\tau]\}) + \del{ M_r(\{X_k[x,y,\tau]\})}{X_k[x,y,\tau]}(X^t_k[x,y,\tau]-X_k[x,y,\tau])
\end{align}
thus
\begin{tcolorbox}[ams align]
a_r = M_r(\{X_k[x,y,\tau]\})) - \del{ M_r(\{X_k[x,y,\tau]\})}{X_k[x,y,\tau]}X_k[x,y,\tau], \quad \omega_{rk}[x,y,\tau] = \del{ M_r(\{X_k[x,y,\tau]\})}{X_k[x,y,\tau]}
\end{tcolorbox}
%
The upcoming task is hence to calculate the derivative of the model network with respect to the input.

In order to keep the overview it is beneficial to write the output of each function in the nest as individual variables, i.e.
\begin{tcolorbox}[ams align]
\{\epsilon_{sh}[x,y]\} = E(\{X_k[x,y,\tau]\}) \\
\{\tilde{H}_h[x,y]\} = D(\{\epsilon^s_h[x,y]\}) \\
\phi_r = F_r(\{\tilde{H}_h[x,y]\})
\end{tcolorbox}
where the additional index $s$ in $\epsilon_{sh}[x,y]$ goes over the internal states of the encoder's output, i.e. 
\begin{tcolorbox}[ams align]
\epsilon_{1h}[x,y] = C_h[x,y,N_\tau], \quad \epsilon_{2h}[x,y] = H_h[x,y,N_\tau].
\end{tcolorbox}
%
Note that the superscript $t$ is now omitted in the internal states since we calculate these quantities with respect to the reference sequence $\{X_k[x,y,\tau]\}$ that should not be connected to any specific time.
%
Note further that the decoder's output state $\tilde{H}_h[x,y]$ has already been defined above.

As a first step we can now further evaluate the derivative via the chain rule
\begin{align}
\omega_{rk}[x,y,\tau] = \del{ M_r(\{X_k[x,y,\tau]\})}{X_k[x,y,\tau]} = \del{\phi_r}{\tilde{H}_{h''}[x'',y'']} \del{\tilde{H}_{h''}[x'',y'']}{\epsilon_{s'h'}[x',y']} \del{\epsilon_{s'h'}[x',y']}{X_k[x,y,\tau]}
\end{align}
%
The next task is to unfold the dependence of $\epsilon_{sh}[x,y]$ on $\{X_k[x,y,\tau]\}$.
%
For that purpose we define the action of one ConvLSTM cell as function $K$. 
Then $\epsilon_{sh}[x,y]$ is the $N_\tau$-fold nested function
\begin{align}
\epsilon_{sh}[x,y] = E(\{X_k[x,y,\tau]\}) = K(\{X_k[x,y,N_\tau]\},K(\{X_k[x,y,N_\tau-1]\},\ldots, \{X_k[x,y,2]\}, K(\{X_k[x,y,1]\},0)\ldots ))
\end{align}
%
We can now define the output of the $K$-calls recursively for a fixed $\tau$ as 
%
\begin{align}
\{\kappa_{sh}[x,y,\tau]\} = K\left(\{X_k[x,y,\tau]\}, \{\kappa_{sh}[x,y,\tau-1]\} \right)
\end{align}
with $\kappa_{sh}[x,y,0]=0$ and $\kappa_{sh}[x,y,N_\tau]=\epsilon_{sh}[x,y]$.
%
Now we can use the chain rule to calculate the derivative of $\epsilon_{s'h'}[x',y']$ with respect to $X_k[x,y,\tau]$ for a specific $\tau$
%
\begin{align}
 \del{\epsilon_{s'h'}[x',y']}{X_k[x,y,\tau]} = \left( \prod_{j=\tau}^{N_\tau-1} \del{\kappa_{s_{j+1}h_{j+1}}[x_{j+1},y_{j+1},j+1]}{\kappa_{s_{j}h_{j}}[x_{j},y_{j},j]} \right) \del{\kappa_{s_{\tau}h_{\tau}}[x_{\tau},y_{\tau},\tau]}{X_k[x,y,\tau]}
\end{align}
with the boundary conditions for the auxiliary indices $x_{N_\tau}\equiv x', y_{N_\tau} \equiv y', h_{N_\tau} \equiv h', s_{N_\tau} \equiv s'$.
%
This can be inserted into the formula for $\omega_{rk}[x,y,\tau]$ yielding
\begin{tcolorbox}[ams align]
\omega_{rk}[x,y,\tau] = \del{\phi_r}{\tilde{H}_{h''}[x'',y'']} \del{\tilde{H}_{h''}[x'',y'']}{\epsilon_{s_{N_\tau} h_{N_\tau}}[x_{N_\tau},y_{N_\tau}]} \left( \prod_{j=\tau}^{N_\tau-1} \del{\kappa_{s_{j+1}h_{j+1}}[x_{j+1},y_{j+1},j+1]}{\kappa_{s_{j}h_{j}}[x_{j},y_{j},j]} \right) \del{\kappa_{s_{\tau}h_{\tau}}[x_{\tau},y_{\tau},\tau]}{X_k[x,y,\tau]}
\end{tcolorbox}

At this point we have to evaluate the appearing derivatives explicitly.
%
We start with the dependence of the $\kappa$ at sequence time $j$ on the previous $\kappa$ at $j-1$.
%
\begin{align}
\del{\kappa_{1h'}[x',y',j]}{\kappa_{1h}[x,y,j-1]}  =\del{C_{h'}[x',y',j]}{C_h[x,y,j-1]} & = F_{h'}[x',y',j]\delta_{hh'}\delta_{xx'}\delta_{yy'} \\
\del{\kappa_{2h'}[x',y',j]}{\kappa_{1h}[x,y,j-1]}  = \del{H_{h'}[x',y',j]}{C_h[x,y,j-1]} & = \frac{O_{h'}[x',y',j] F_{h'}[x',y',j]}{\cosh^2(C_{h'}[x',y',j])}  \delta_{hh'}\delta_{xx'}\delta_{yy'} \\
\del{\kappa_{1h'}[x',y',j]}{\kappa_{2h}[x,y,j-1]} = \del{C_{h'}[x',y',j]}{H_h[x,y,j-1]} & = 
\dot{\sigma}(f_{h'}[x',y',j])\del{f_{h'}[x',y',j]}{H_h[x,y,j-1]}C_{h'}[x',y',j-1]\\
& + \dot{\sigma}(i_{h'}[x',y',j])\del{i_{h'}[x',y',j]}{H_h[x,y,j-1]} \tanh(c_{h'}[x',y',j])\\
& +  \frac{\sigma(i_{h'}[x',y',j])}{\cosh^2(c_{h'}[x',y',j])} \del{c_{h'}[x',y',j]}{H_h[x,y,j-1]} \\
\del{\kappa_{2h'}[x',y',j]}{\kappa_{2h}[x,y,j-1]} = \del{H_{h'}[x',y',j]}{H_h[x,y,j-1]} & = 
\dot{\sigma}(o_{h'}[x',y',j])\del{{o_{h'}[x',y',j]}}{H_h[x,y,j-1]}\tanh(C_{h'}[x',y',j]) \\
& + \frac{\sigma(o_{h'}[x',y',j])}{\cosh^2(C_{h'}[x',y',j])} \del{C_{h'}[x',y',j]}{H_h[x,y,j-1]}
\end{align}
%
Note that $ \partial \sigma(x)/ \partial {x} = \sigma(x)(1-\sigma(x)) =: \dot{\sigma}(x)$ and $\partial \tanh(x)/ \partial {x} = 1/\cosh^2(x)$.
%
The remaining task is to calculate the derivative of the gate functions $g_h[x,y,j]$ with respect to the previous hidden state $H_h[x,y,j-1]$, i.e. 
%
\begin{align}
\del{{g_{h'}[x',y',j]}}{H_h[x,y,j-1]} = \mathcal{N}^{g}_{h'h}[x'-x,y'-y]
\end{align}
%
Then we finally have
\begin{tcolorbox}[ams align]
\del{\kappa_{1h'}[x',y',j]}{\kappa_{1h}[x,y,j-1]} & = F_{h'}[x',y',j]\delta_{hh'}\delta_{xx'}\delta_{yy'} \\
\del{\kappa_{2h'}[x',y',j]}{\kappa_{1h}[x,y,j-1]} & = \frac{O_{h'}[x',y',j] F_{h'}[x',y',j]}{\cosh^2(C_{h'}[x',y',j])}  \delta_{hh'}\delta_{xx'}\delta_{yy'} \\
\del{\kappa_{1h'}[x',y',j]}{\kappa_{2h}[x,y,j-1]} & = 
\dot{\sigma}(f_{h'}[x',y',j]) C_{h'}[x',y',j-1] \mathcal{N}^{f}_{h'h}[x'-x,y'-y]\\
& + \dot{\sigma}(i_{h'}[x',y',j]) \tanh(c_{h'}[x',y',j])\mathcal{N}^{i}_{h'h}[x'-x,y'-y]\\
& +  \frac{\sigma(i_{h'}[x',y',j])}{\cosh^2(c_{h'}[x',y',j])} \mathcal{N}^{c}_{h'h}[x'-x,y'-y] \\
\del{\kappa_{2h'}[x',y',j]}{\kappa_{2h}[x,y,j-1]} & = 
\dot{\sigma}(o_{h'}[x',y',j])\tanh(C_{h'}[x',y',j]) \mathcal{N}^{o}_{h'h}[x'-x,y'-y] \\
& + \frac{\sigma(o_{h'}[x',y',j])}{\cosh^2(C_{h'}[x',y',j])} \del{\kappa_{1h'}[x',y',j]}{\kappa_{2h}[x,y,j-1]}
\end{tcolorbox}
%

The next task is to calculate the dependence of the $\kappa$ on the input quantity $X$.
\begin{align}
\del{\kappa_{1h'}[x',y',\tau]}{X_k[x,y,\tau]} = \del{C_{h'}[x',y',\tau]}{X_k[x,y,\tau]} & = 
\dot{\sigma}(f_{h'}[x',y',\tau]) \del{f_{h'}[x',y',\tau]}{X_k[x,y,\tau]} C_{h'}[x',y',\tau-1] \\
& + \dot{\sigma}(i_{h'}[x',y',\tau]) \del{i_{h'}[x',y',\tau]}{X_k[x,y,\tau]} \tanh(c_{h'}[x',y',\tau])\\
& +  \frac{\sigma(i_{h'}[x',y',\tau])}{\cosh^2(c_{h'}[x',y',\tau])} \del{c_{h'}[x',y',\tau]}{X_k[x,y,\tau]} \\
\del{\kappa_{2h'}[x',y',\tau]}{X_k[x,y,\tau]}  = \del{H_{h'}[x',y',\tau]}{X_k[x,y,\tau]} &=
\dot{\sigma}(o_{h'}[x',y',\tau])\del{{o_{h'}[x',y',\tau]}}{X_k[x,y,\tau]}\tanh(C_{h'}[x',y',\tau]) \\
& + \frac{\sigma(o_{h'}[x',y',\tau])}{\cosh^2(C_{h'}[x',y',\tau])} \del{C_{h'}[x',y',\tau]}{X_k[x,y,\tau]}
\end{align} 
%
The remaining task is to calculate the derivative of the gate functions $g_h[x,y,\tau]$ with respect to the current input $X_k[x,y,\tau]$, i.e. 
%
\begin{align}
\del{{g_{h'}[x',y',\tau]}}{X_k[x,y,\tau]} = \mathcal{M}^{g}_{h'k}[x'-x,y'-y]
\end{align}
%
Inserted into the equations above yields
%
\begin{tcolorbox}[ams align]
\del{\kappa_{1h'}[x',y',\tau]}{X_k[x,y,\tau]}  & = 
\dot{\sigma}(f_{h'}[x',y',\tau])  C_{h'}[x',y',\tau-1] \mathcal{M}^{f}_{h'k}[x'-x,y'-y] \\
& + \dot{\sigma}(i_{h'}[x',y',\tau])  \tanh(c_{h'}[x',y',\tau])\mathcal{M}^{i}_{h'k}[x'-x,y'-y]\\
& +  \frac{\sigma(i_{h'}[x',y',\tau])}{\cosh^2(c_{h'}[x',y',\tau])} \mathcal{M}^{c}_{h'k}[x'-x,y'-y] \\
\del{\kappa_{2h'}[x',y',\tau]}{X_k[x,y,\tau]}  &=
\dot{\sigma}(o_{h'}[x',y',\tau])\tanh(C_{h'}[x',y',\tau])\mathcal{M}^{o}_{h'k}[x'-x,y'-y] \\
& + \frac{\sigma(o_{h'}[x',y',\tau])}{\cosh^2(C_{h'}[x',y',\tau])} \del{\kappa_{1h'}[x',y',\tau]}{X_k[x,y,\tau]} 
\end{tcolorbox} 
%

Now that we have expressions for the derivatives we can give them names
\begin{tcolorbox}[ams align]
\Lambda_{s's,h'h}[x', x, y', y, j] = \del{\kappa_{s'h'}[x',y',j]}{\kappa_{sh}[x,y,j-1]}  \\
\lambda_{s'h',k}[x', x, y', y, \tau] = \del{\kappa_{s'h'}[x',y',\tau]}{X_k[x,y,\tau]} 
\end{tcolorbox}
and insert it into the full expression
\begin{align}
\omega_{rk}[x,y,\tau] = \del{\phi_r}{\tilde{H}_{h''}[x'',y'']} \del{\tilde{H}_{h''}[x'',y'']}{\epsilon_{s_{N_\tau} h_{N_\tau}}[x_{N_\tau},y_{N_\tau}]} \left( \prod_{j=\tau}^{N_\tau-1} \Lambda_{s_{j+1}s_j,h_{j+1}h_j}[x_{j+1}, x_j, y_{j+1}, y_j, j+1] \right)
\lambda_{s_\tau h_\tau,k}[x_\tau, x, y_\tau, y, \tau]
\end{align}
%
To further shorten the expression we can formally carry out the product as 
\begin{align}
\chi_{s_{n} h_{n},k}[x_{n}, x, y_{n}, y, n, \tau] = \left( \prod_{j=\tau}^{n-1} \Lambda_{s_{j+1}s_j,h_{j+1}h_j}[x_{j+1}, x_j, y_{j+1}, y_j, j+1] \right)
\lambda_{s_\tau h_\tau,k}[x_\tau, x, y_\tau, y, \tau]
\end{align}
with $\tau \leq n \leq N_\tau$ then
%
\begin{tcolorbox}[ams align]
\chi_{s_{n+1} h_{n+1},k}[x_{n+1}, x, y_{n+1}, y, n+1, \tau] =  \Lambda_{s_{n+1}s_n,h_{n+1}h_n}[x_{n+1}, x_n, y_{n+1}, y_n, n+1] 
\chi_{s_{n} h_{n},k}[x_{n}, x, y_{n}, y, n, \tau]
\end{tcolorbox}
%
for $n=\tau,\ldots N_\tau-1$ and $\chi_{s_{\tau} h_{\tau},k}[x_{\tau}, x, y_{\tau}, y, \tau, \tau] = \lambda_{s_\tau h_\tau,k}[x_\tau, x, y_\tau, y, \tau]$.

The next task is to evaluate the derivative of the decoder with respect to the encoders output.
%
Since the decoder is itself a ConvLSTM cell with input from encoders last hidden state we can write
\begin{align}
\{\tilde{H}_{h}[x,y]\} = \tilde{K}_2 \left( \{ \epsilon_{2 h}[x,y]\}, \{ \epsilon_{s h}[x,y]\} \right)
\end{align}
%
Following the same line of reasoning as above we obtain for the derivatives
%
\begin{tcolorbox}[ams align]
\del{\tilde{H}_{h'}[x',y']}{\epsilon_{1h}[x,y]} = \del{\tilde{H}_{h'}[x',y']}{C_h[x,y,N_\tau]}  & = \frac{\tilde{O}_{h'}[x',y'] \tilde{F}_{h'}[x',y']}{\cosh^2(\tilde{C}_{h'}[x',y'])}  \delta_{hh'}\delta_{xx'}\delta_{yy'} \\
\del{\tilde{H}_{h'}[x',y']}{\epsilon_{2h}[x,y]} = \del{\tilde{H}_{h'}[x',y']}{H_h[x,y,N_\tau]}     & = \dot{\sigma}(\tilde{o}_{h'}[x',y'])\tanh(\tilde{C}_{h'}[x',y']) \tilde{\mathcal{Q}}^{o}_{h'h}[x'-x,y'-y] \\
& + \frac{\sigma(\tilde{o}_{h'}[x',y'])}{\cosh^2(\tilde{C}_{h'}[x',y'])} \tilde{\Lambda}_{1,2,h'h}[x', x, y', y]
\end{tcolorbox}
%
with
\begin{align}
\tilde{\Lambda}_{1,2,h'h}[x', x, y', y] & = \dot{\sigma}(\tilde{f}_{h'}[x',y']) C_{h'}[x',y',N_\tau] \tilde{\mathcal{Q}}^{f}_{h'h}[x'-x,y'-y]\\
& + \dot{\sigma}(\tilde{i}_{h'}[x',y']) \tanh(\tilde{c}_{h'}[x',y'])\tilde{\mathcal{Q}}^{i}_{h'h}[x'-x,y'-y]\\
& +  \frac{\sigma(\tilde{i}_{h'}[x',y'])}{\cosh^2(\tilde{c}_{h'}[x',y'])} \tilde{\mathcal{Q}}^{c}_{h'h}[x'-x,y'-y]
\end{align}
%
Now we can also rename these derivatives as
\begin{tcolorbox}[ams align]
\tilde{\chi}_{s_{N_\tau}h_{N_\tau}, h''}[x'', x_{N_\tau}, y'', y_{N_\tau}] = \del{\tilde{H}_{h''}[x'',y'']}{\epsilon_{s_{N_\tau} h_{N_\tau}}[x_{N_\tau},y_{N_\tau}]}
\end{tcolorbox}
%
and insert into the full expression
%
\begin{align}
\omega_{rk}[x,y,\tau] = \del{\phi_r}{\tilde{H}_{h''}[x'',y'']} \tilde{\chi}_{s_{N_\tau}h_{N_\tau}, h''}[x'', x_{N_\tau}, y'', y_{N_\tau}] \chi_{s_{N_\tau} h_{N_\tau},k}[x_{N_\tau}, x, y_{N_\tau}, y, N_\tau, \tau]
\end{align}
%
If we define the product 
%
\begin{tcolorbox}[ams align]
\Xi_{h''k}[x'', x, y'', y, \tau] = \tilde{\chi}_{s_{N_\tau}h_{N_\tau}, h''}[x'', x_{N_\tau}, y'', y_{N_\tau}] \chi_{s_{N_\tau} h_{N_\tau},k}[x_{N_\tau}, x, y_{N_\tau}, y, N_\tau, \tau]
\end{tcolorbox}
%
we can write
\begin{tcolorbox}[ams align]
\omega_{rk}[x,y,\tau] = \del{\phi_r}{\tilde{H}_{h''}[x'',y'']} \Xi_{h''k}[x'', x, y'', y, \tau]
\end{tcolorbox}

The remaining task is to find the derivative of the fully connected layser with respect to the decoder's output.
%
For this purpose we define gain the output variables of the nested ReLU functions as individual variables, i.e.
\begin{tcolorbox}[ams align]
\rho^1_a & = \mathcal{W}^{1}_{ah} [x,y] \tilde{H}_h[x,y] \\
\rho^2_b & = \mathcal{W}^{2}_{ba} \mathrm{ReLU}\left(\rho^1_a\right) \\
\phi_r & = \mathcal{W}^{3}_{rb} \mathrm{ReLU}\left(\rho^2_b \right)
\end{tcolorbox}
%
With these definitions we can write after using the chain rule
\begin{align}
\del{\phi_r}{\tilde{H}_{h''}[x'',y'']} = \del{\phi_r}{\rho^2_{b'}} \del{\rho^2_{b'}}{\rho^1_{a'}} \del{\rho^1_{a'}}{\tilde{H}_{h''}[x'',y'']}
\end{align}
The individual derivatives can be easily carried out by noting $\partial \mathrm{ReLU}(x) / \partial x = \theta(x)$ which is the Heaviside step function
\begin{align}
\del{\phi_r}{\rho^2_{b'}} & = \mathcal{W}^{3}_{rb'} \theta \left( \rho^2_{b'} \right)  \\
 \del{\rho^2_{b'}}{\rho^1_{a'}} & =  \mathcal{W}^{2}_{b'a'} \theta \left( \rho^1_{a'} \right) \\
 \del{\rho^1_{a'}}{\tilde{H}_{h''}[x'',y'']} & = \mathcal{W}^{1}_{a'h''} [x'',y'']
\end{align}
%
\begin{tcolorbox}[ams align]
\del{\phi_r}{\tilde{H}_{h''}[x'',y'']} = \mathcal{W}^{3}_{rb'} \theta \left( \rho^2_{b'} \right) \cdot  \mathcal{W}^{2}_{b'a'} \theta \left( \rho^1_{a'} \right) \cdot \mathcal{W}^{1}_{a'h''} [x'',y''] = \Omega_{r h''}[x'',y'']
\end{tcolorbox}

As the final result we can then write 
%
\begin{tcolorbox}[ams align]
\omega_{rk}[x,y,\tau] = \Omega_{r h''}[x'',y''] \cdot \Xi_{h''k}[x'', x, y'', y, \tau]
\end{tcolorbox}

\subsection{Recipe}

Evaluating the result from the right to the left suggests the following recipe:

\begin{itemize}
\item take a specific reference sequence $\{X_k[x,y,\tau]\}$ and go over $\tau \in [1,N_\tau]$ in ascending order
\begin{itemize}
\item calculate for each $\tau$
$$ \chi_{s_\tau h_\tau,k}[x_\tau, x, y_\tau, y, \tau, \tau] = \lambda_{s_\tau h_\tau,k}[x_\tau, x, y_\tau, y, \tau] $$
where $\lambda$ is a function of the current gates $g_{h_\tau}[x_\tau,y_\tau,\tau]$, the last cell state $C_{h_\tau}[x_\tau,y_\tau,\tau-1]$ and the parameters $\mathcal{M}^{g}_{h_\tau k}[x_\tau-x,y_\tau-y]$, \begin{tcolorbox} expense: $N_k (2 N_h N_x^2 N_y^2)$ \end{tcolorbox}
\item go then over $n \in [\tau, N_\tau-1]$ in ascending order
\begin{itemize}
\item for each $n$ calculate the tensor $\Lambda_{s_{n+1}s_n,h_{n+1}h_n}[x_{n+1}, x_n, y_{n+1}, y_j, n+1]$ as a function of the input gates and hidden states of the $n+1$-th and $n$-th ConvLSTM cells and the parameters $\mathcal{M}^{g}_{h_\tau k}[x_\tau-x,y_\tau-y]$ and $\mathcal{N}^{g}_{h_\tau k}[x_\tau-x,y_\tau-y]$
and perform the summation over $s_n, h_n, x_n, y_n$ for each $k, s_{n+1}, h_{n+1}, x, x_{n+1}, y, y_{n+1}$
$$ \chi_{s_{n+1} h_{n+1},k}[x_{n+1}, x, y_{n+1}, y, n+1, \tau] =  \Lambda_{s_{n+1}s_n,h_{n+1}h_n}[x_{n+1}, x_n, y_{n+1}, y_n, n+1] 
\chi_{s_{n} h_{n},k}[x_{n}, x, y_{n}, y, n, \tau]$$
\begin{tcolorbox} expense: $N_k (2 N_h N_x^2 N_y^2) (2 N_h N_x N_y)$ \end{tcolorbox}
\end{itemize}
\item the final result is $\chi_{s_{N_\tau} h_{N_\tau},k}[x_{N_\tau}, x, y_{N_\tau}, y, N_\tau, \tau]$
\begin{tcolorbox} accumulated expense: $N_k (2 N_h N_x^2 N_y^2) +(N_\tau - \tau) N_k (2 N_h N_x^2 N_y^2) (2 N_h N_x N_y)$ \end{tcolorbox}
\item calculate $\tilde{\chi}_{s_{N_\tau}h_{N_\tau}, h''}[x'', x_{N_\tau}, y'', y_{N_\tau}]$ as function of the decoder gates $\tilde{g}_{h''}[x'',y'']$, the encoder's last cell state $C_{h_{N_\tau}}[x_{N_\tau},y_{N_\tau},{N_\tau}]$ and the parameters $\tilde{\mathcal{Q}}^{g}_{h'' k}[x''-x_{N_\tau},y''-y_{N_\tau}]$ $\to$ expense: $2 N_h^2 N_x^2 N_y^2$
\item accumulated expense: $N_k (2 N_h N_x^2 N_y^2) +(N_\tau - \tau) N_k (2 N_h N_x^2 N_y^2) (2 N_h N_x N_y)+2 N_h^2 N_x^2 N_y^2$
\item then perform the summation over $s_{N_\tau}, h_{N_\tau}, x_{N_\tau}, y_{N_\tau}$ for each $h'', k, x'', x, y'', y$
$$\Xi_{h''k}[x'', x, y'', y, \tau] = \tilde{\chi}_{s_{N_\tau}h_{N_\tau}, h''}[x'', x_{N_\tau}, y'', y_{N_\tau}] \chi_{s_{N_\tau} h_{N_\tau},k}[x_{N_\tau}, x, y_{N_\tau}, y, N_\tau, \tau] $$
\begin{tcolorbox} expense: $N_k (N_h N_x^2 N_y^2) (2 N_h N_x N_y)$ \end{tcolorbox}
\item next calculate $\Omega_{r h''}[x'',y'']$ as a function of the decoder's output $\tilde{H}_{h''}[x'',y'']$ and the parameters $\mathcal{W}^{1}_{a'h''}[x'', y''], \mathcal{W}^{2}_{b'a'}$ and $ \mathcal{W}^{3}_{rb'}$
\begin{tcolorbox} expense: $N_h N_x N_y$ \end{tcolorbox}
\item finally perform the summation over $h'', x'', y''$ for each $k,x,y$
$$ \omega_{rk}[x,y,\tau] = \Omega_{r h''}[x'',y''] \cdot \Xi_{h''k}[x'', x, y'', y, \tau] $$
\begin{tcolorbox} expense: $N_k (N_h N^2_x N^2_y)$ \end{tcolorbox}

\end{itemize}

\begin{tcolorbox}
accumulated expense: 
\begin{align}
 N_\tau & \left[ N_k (2 N_h N_x^2 N_y^2) +(N_\tau - \tau) N_k (2 N_h N_x^2 N_y^2) (2 N_h N_x N_y) \right. \nonumber \\ 
 & \left.  + 2 N_h^2 N_x^2 N_y^2 + N_k (N_h N_x^2 N_y^2) (2 N_h N_x N_y) + N_h N_x N_y + N_k (N_h N^2_x N^2_y) \right]
\end{align}
\end{tcolorbox}
\end{itemize}

\subsection{Choice of the reference sequence $\{X_k[x,y,\tau]\}$}

\subsubsection{Average over all sequences in the data set}

Suppose the number of all sequences with length $N_\tau$ in the data set is $N_s$
\begin{align}
\bar{A}_r[t] & = \frac{1}{N_s}\sum_{\{X_k[x,y,\tau]\} \in \{X^t_k[x,y,\tau]\}} \left( M_r(\{X_k[x,y,\tau]\}) + \del{ M_r(\{X_k[x,y,\tau]\})}{X_k[x,y,\tau]}(X^t_k[x,y,\tau]-X_k[x,y,\tau]) \right) \\
& = \left(  \frac{1}{N_s}\sum_{\{X_k[x,y,\tau]\} \in \{X^t_k[x,y,\tau]\}} M_r(\{X_k[x,y,\tau]\})) - \del{ M_r(\{X_k[x,y,\tau]\})}{X_k[x,y,\tau]}X_k[x,y,\tau] \right) \\
& + \left( \frac{1}{N_s} \sum_{\{X_k[x,y,\tau]\} \in \{X^t_k[x,y,\tau]\}} \del{ M_r(\{X_k[x,y,\tau]\})}{X_k[x,y,\tau]} \right) X^t_k[x,y,\tau] \\
& = \bar{a}_r + \bar{\omega}_{rk}[x,y,\tau] X^t_k[x,y,\tau] 
\end{align}

\subsubsection{Optimal sequence}

Find a specific sequence $\{X^*_k[x,y,\tau]\}$ such that
\begin{align}
{A}^*_r[t] = M_r(\{X^*_k[x,y,\tau]\}) + \del{ M_r(\{X^*_k[x,y,\tau]\})}{X^*_k[x,y,\tau]}(X^t_k[x,y,\tau]-X^*_k[x,y,\tau])
\end{align}
minimizes the integrated squared error
\begin{align}
\sum_t (R_r[t]-{A}^*_r[t])^2
\end{align}

\end{document}

### 2024-04-23 paper/paper.tex

--- 2024-03-18 paper/paper.tex
@@
 \usepackage{amsmath,amssymb}
-\usepackage{lmodern}
 \usepackage{iftex}
@@
   \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+\usepackage{lmodern}
+\ifPDFTeX\else  
+    % xetex/luatex font selection
 \fi
@@
 \makeatother
+% definitions for citeproc citations
+\NewDocumentCommand\citeproctext{}{}
+\NewDocumentCommand\citeproc{mm}{%
+  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
+\makeatletter
+ % allow citations to break across lines
+ \let\@cite@ofmt\@firstofone
+ % avoid brackets around text for \cite:
+ \def\@biblabel#1{}
+ \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
+\makeatother
 \newlength{\cslhangindent}
@@
 \setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
+\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
+ {\begin{list}{}{%
+  \setlength{\itemindent}{0pt}
+  \setlength{\leftmargin}{0pt}
+  \setlength{\parsep}{0pt}
   % turn on hanging indent if param 1 is 1
   \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
+   \setlength{\leftmargin}{\cslhangindent}
+   \setlength{\itemindent}{-1\cslhangindent}
   \fi
   % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
+  \setlength{\itemsep}{#2\baselineskip}}}
+ {\end{list}}
 \usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
@@
 \linenumbers
-\makeatletter
-\makeatother
-\makeatletter
-\makeatother
 \makeatletter
@@
 \makeatletter
+\makeatother
+\makeatletter
 \@ifpackageloaded{caption}{}{\usepackage{caption}}
 \@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
-\makeatother
-\makeatletter
-\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
-\makeatother
-\makeatletter
-\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
-\makeatother
-\makeatletter
 \makeatother
@@
 \fi
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\usepackage{bookmark}
+
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
@@
   pdftitle={From Precipitation to Prediction: Integrating ConvLSTM Models for Comprehensive River Runoff Forecasting},
-  pdfauthor={Florian Börgel; Sven Karsten},
+  pdfauthor={Florian Börgel; Sven Karsten; Karoline Rummel},
   colorlinks=true,
@@
 
-\authors{Florian Börgel\affil{1}, Sven Karsten\affil{1}}
+\authors{Florian Börgel\affil{1}, Sven Karsten\affil{1}, Karoline
+Rummel\affil{1}}
 \affiliation{1}{Leibniz-Institute for Baltic Sea Research Warnemünde, }
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced, interior hidden, sharp corners, frame hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
-
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+\section{Introduction}\label{introduction}
 
@@
 
-\hypertarget{methods}{%
-\section{Methods}\label{methods}}
+\section{Methods}\label{methods}
 
-\hypertarget{runoff-data-used-for-training}{%
 \subsection{Runoff data used for
-training}\label{runoff-data-used-for-training}}
+training}\label{runoff-data-used-for-training}
 
@@
 
-\hypertarget{atmospheric-forcing}{%
-\subsection{Atmospheric Forcing}\label{atmospheric-forcing}}
+\subsection{Atmospheric Forcing}\label{atmospheric-forcing}
 
@@
 
-\hypertarget{ocean-model}{%
-\subsection{Ocean Model}\label{ocean-model}}
+\subsection{Ocean Model}\label{ocean-model}
 
@@
 
-\hypertarget{lstm-network}{%
-\subsection{LSTM network}\label{lstm-network}}
+\subsection{LSTM network}\label{lstm-network}
+
+Before turning to the convolutional LSTM, the simpler architecture of
+the plain LSTM is examined and should serve as a role model for the
+later consideration of the full ConvLSTM.
 
@@
 Neural Networks (RNNs), is specifically tailored for modeling temporal
-sequences. Its unique design allows it to adeptly handle long-range
-dependencies, setting it apart from traditional RNNs in terms of
-accuracy (see Figure~\ref{fig-lstm}).
+sequences
+\(\vec{X}^t[1], \ldots \vec{X}^t[\tau],\ldots \vec{X}^t[N_\tau]\) of
+\(N_\tau\) input quantities
+\(\vec{X}^t[\tau] = (X^t_k[\tau]) \in \mathbb{R}^{N_k}\). The sequence
+is taken from a dataset given in form of a time series \(\{\vec{X}^t\}\)
+and the endpoint should coincide with the specific element in the time
+series connected to time \(t\),
+i.e.~\(\vec{X}^t[N_\tau] \equiv \vec{X}^t\). The \(N_k\) is the number
+of used input ``channels'' and can, for example, represent different
+measurable quantities. Its unique design allows it to adeptly handle
+long-range dependencies, setting it apart from traditional RNNs in terms
+of accuracy (see Figure~\ref{fig-lstm}).
 
@@
 
-{\centering \includegraphics{draw_lstm.png}
+\centering{
+
+\includegraphics{draw_lstm.png}
 
@@
 
-\end{figure}
+\end{figure}%
 
@@
 in various studies. The key component of LSTM's innovation is its memory
-cell, \(c_t\)t, which stores state information, also refered to as
-long-term memory. This cell is accessed, modified, and reset through
-several self-parameterized gates. For the input of the sequence \(x_t\)
-input, the forget gate \(f_t\) defines the percentage of the previous
-long-term memory status \(c_{t-1}\) that should be retained
-stored\hspace{0pt}. Next the input gate \(i_t\) decides how much of the
-input is added to the the long-term memory, forming the updated cell
-state \(c_{t}\). The decision to propagate the latest cell output,
-\(c_t\), to the final state, \(h_t\), is governed by the output gate,
-\(o_t\), representing the updated short-term memory of the hidden state
-\(h_t\). A significant advantage of this architecture is the memory
-cell's ability to retain gradients. This mechanism addresses the
-vanishing gradient problem, where, as input sequences elongate, the
-influence of initial stages becomes harder to capture, causing gradients
-of early input points to approach zero. The LSTM's activation function,
+cell, \(\vec{C}^t[\tau] = (C^t_h[\tau]) \in \mathbb{R}^{N_h}\), which
+stores state information, also referred to as long-term memory. The
+index \(\tau \in [1, N_\tau]\) corresponds to the \(\tau\)-th element in
+the input sequence. The cell state is a vector consists of elements,
+each associated with one of the \(N_h\) hidden layers, labeled by \(h\).
+This vector is determined through several self-parameterized gates, all
+in the same vector space as \(\vec{C}^t[\tau]\). The forget gate
+\(\vec{F}^t[\tau]\) defines the percentage of the previous long-term
+memory status \(\vec{C}^t[\tau-1]\) that should be retained
+stored\hspace{0pt}. The input gate \(\vec{I}^t[\tau]\) decides how much
+of the input is added to the the long-term memory, forming the updated
+cell state. The so-called output gate \(\vec{O}^t[\tau]\) then
+determines how much of the latest cell output, \(\vec{C}^t[\tau]\), is
+propagated to the final state, \(\vec{H}^t[\tau]\) representing the
+updated short-term memory of the hidden state.
+
+For a fixed point \(\tau\) in the sequence, the action of a LSTM cell,
+i.e.~the connection between the input \(\vec{X}^t[\tau]\), the various
+gates and the state vectors, is given as follows. First, the elements of
+the input sequence together with the hidden state are mapped onto
+auxiliary gate vectors, collectively denoted by
+\(\vec{g}^t[\tau] = (g^t_h[\tau]) \in \mathbb{R}^{N_h}\), via \[
+\begin{aligned}
+g_h^t[\tau] =  \mathcal{M}^{g}_{hk} \, X^t_k[\tau] +  \mathcal{N}^{g}_{hh'} \, H^t_{h'}[\tau-1] + \mathcal{B}^g_{h} \ ,
+\end{aligned}
+\]
+
+where \(g=i,f,o,c\) stands for the input, forget, output and cell-state
+gate, respectively and Eistein's summation convention is employed,
+i.e.~indices that appear twice are summed over. The calligraphic symbols
+\(\mathcal{M}^{g}_{hk}, \mathcal{N}^{g}_{hh'}\) and
+\(\mathcal{B}^g_{h}\) are the free parameters of the network that are
+optimized for the given problem during the training, which is at the
+heart of the any machine learning approach. The matrix
+\(\pmb{\mathcal{M}}^{g} = (\mathcal{M}^{g}_{hk}) \in \mathbb{R}^{N_h \times N_k}\)
+can be interpreted as a Markovian-like contribution of the current input
+\(\vec{X}^t[\tau]\) to the gates, whereas the
+\(\pmb{\mathcal{N}}^{g} = (\mathcal{N}^{g}_{hh'}) \in \mathbb{R}^{N_h \times N_h}\)
+scales a non-Markovian part determined by the hidden state of the last
+sequence point \(\tau-1\). The vector
+\(\vec{\mathcal{B}}^g = (\mathcal{B}^g_{h}) \in \mathbb{R}^{N_h}\) is a
+learnable bias. It should be stressed that these parameters do neither
+depend on \(t\) nor on \(\tau\) and are thus optimized once for the
+complete dataset \(\{\vec{X}^t\}\).
+
+Note that this mapping is sometimes extended by a contribution to the
+\(g_h^t[\tau]\) from the past cell state \(\vec{C}^t[\tau-1]\)
+(\textbf{XXX?}). Nevertheless, this mechanism called ``peeping'' is not
+further considered in this work.
+
+For the sake of brevity, we write the mapping more compactly in
+matrix-vector form as
+\begin{equation}\phantomsection\label{eq-LSTM-gates}{
+\begin{aligned}
+\vec{g}^t[\tau] = \pmb{\mathcal{M}}^{g}\vec{X}^t[\tau] + \pmb{\mathcal{N}}^{g}  \vec{H}^t[ \tau-1]+ \vec{\mathcal{B}}^g \,
+\end{aligned}
+}\end{equation}
+
+Second, the actual gate vectors are computed by the core equations of
+the LSTM as proposed by (\textbf{XXX?}):
+
+\begin{equation}\phantomsection\label{eq-LSTM}{
+\begin{aligned}
+\vec{I}^t[\tau] &= \sigma(\vec{i}^t[\tau]) \\
+\vec{F}^t[\tau] &= \sigma(\vec{f}^t[\tau]) \\
+\vec{O}^t[\tau] &= \sigma(\vec{o}^t[\tau]) \\
+\vec{C}^t[\tau] &= \vec{F}^t[\tau] \circ \vec{C}^t[\tau-1]  + \vec{I}^t[\tau] \circ \tanh(\vec{c}^t[\tau]) \\
+\vec{H}^t[\tau]  &= \vec{O}^t[\tau] \circ \tanh(\vec{C}^t[\tau]) \ ,
+\end{aligned}
+}\end{equation}
+
+where \(\sigma\) denotes the logisitc sigmoid function, \(\tanh\) is the
+hyperbolic tangent and the \(\circ\) stands for the Hadamard product
+(all applied in an element-wise fashion to the vectors). In the last two
+equations the role of the input, forget and output gates as described
+above become apparent.
+
+The third step in a single layer LSTM (as employed for the work
+presented here) is then to provide the output of the current LSTM cell,
+i.e.~\(\vec{H}^t[\tau]\) and \(\vec{C}^t[\tau]\), to the subsequent LSTM
+cell that processes the next element \(\vec{X}^t[\tau+1]\) of the input
+sequence.
+
+The full action of the LSTM network up to the end of the sequence can
+therefore be written as a nested function call
+
+\[
+\left(\vec{H}^t[N_\tau], \vec{C}^t[N_\tau] \right) = L \left( \vec{X}^t[N_\tau], L \left( \vec{X}^t[N_\tau-1], \ldots L \left( \vec{X}^t[1], (\vec{H}^t[0], \vec{C}^t[0] \right) \ldots \right) \right) \ ,
+\]
+
+where \(L\) represents Eqs. (Equation~\ref{eq-LSTM-gates},
+Equation~\ref{eq-LSTM}) and the initial conditions are usually chosen as
+\(\vec{H}^t[0]=\vec{C}^t[0]=0\).
+
+The final output \(\vec{H}^t[N_\tau]\) and \(\vec{C}^t[N_\tau]\) encode
+information on the full input sequence ending at time \(t\). This
+information has to be decoded to obtain usueful information via an
+appropriate subsequent layer, as it is described in the Sec.
+(\textbf{XXX?}).
+
+A significant advantage of this architecture is the memory cell's
+ability to retain gradients. This mechanism addresses the vanishing
+gradient problem, where, as input sequences elongate, the influence of
+initial stages becomes harder to capture, causing gradients of early
+input points to approach zero. The LSTM's activation function,
 inherently recurrent, mirrors the identity function with a consistent
@@
 
-One LSTM cell hence maybe expressed as:
+\subsection{ConvLSTM network}\label{convlstm-network}
 
-\[
+Although the plain LSTM has high performance in handling temporal
+sequences of point-like quantities it is not designed to recognize
+spatial features in seuqences of two-dimensional maps. To adress this
+limitation we use a convLSTM architecture.
+
+In this kind of network the elements of the input sequence are given as
+(kind of) vector fields
+\(\vec{X}^t[\tau] = (X^t_k[x,y,\tau]) \in \mathbb{R}^{N_k \times (N_x \times N_y)}\),
+where \(x\in[1, N_x]\) and \(y \in [1, N_y]\) run over the horizontal
+and vertical dimensions of the map, respectively. In order to enable the
+``learning'' of spatial patterns, the free parameters of the network are
+replaced by two-dimenmsional convolutional kernels
+\(\pmb{\mathcal{M}}^{g} = (\mathcal{M}^{g}_{hk}[\xi, \eta]) \in \mathbb{R}^{(N_h \times N_k)\times (N_\xi \times N_\eta)}\)
+and
+\(\pmb{\mathcal{N}}^{g} = (\mathcal{N}^{g}_{hh'}[\xi, \eta]) \in \mathbb{R}^{(N_h \times N_h)\times (N_\xi \times N_\eta)}\).
+The width and the height of the kernels are given by \(N_\xi\) and
+\(N_\eta\), respectively and
+\(\xi\in [-(N_\xi-1)/2,(N_\xi-1)/2], \eta\in [-(N_\eta-1)/2,(N_\eta-1)/2]\),
+where we assume odd numbers for the kernel sizes.
+
+The mapping from the input quantities to the gates is then given by a
+convolution with these kernels \[
 \begin{aligned}
-i_t &= \sigma(W_{xi} x_t + W_{hi} h_{t-1} + W_{ci} \circ c_{t-1} + b_i) \\
-f_t &= \sigma(W_{xf} x_t + W_{hf} h_{t-1} + W_{cf} \circ c_{t-1} + b_f) \\
-c_t &= f_t \circ c_{t-1} + i_t \circ \tanh(W_{xc} x_t + W_{hc} h_{t-1} + b_c) \\
-o_t &= \sigma(W_{xo} x_t + W_{ho} h_{t-1} + W_{co} \circ c_t + b_o) \\
-h_t &= o_t \circ \tanh(c_t)
+g^t_h[x,y,\tau] & =  \mathcal{M}^{g}_{hk} [\xi,\eta]\, X^t_k[x-\xi, y-\eta, \tau]  +  \mathcal{N}^{g}_{hh'}[\xi,\eta] \, H^t_{h'}[x-\xi, y-\eta, \tau-1] + \mathcal{B}^g_{h}[x,y] \ .
 \end{aligned}
@@
 
-with
-
-\begin{itemize}
-\tightlist
-\item
-  \(x_t\): Input vector at time step \(t\).
-\item
-  \(h_{t-1}\): Hidden state from the previous time step.
-\item
-  \(C_{t-1}\): Cell state from the previous time step.
-\item
-  \(W\) and \(b\): Weight matrices and bias vectors, respectively,
-  associated with the gates of the LSTM. The subscripts denote the
-  specific gate or operation they are associated with (e.g., \(W_f\) and
-  \(b_f\) are the weight matrix and bias for the forget gate).
-\item
-  \(\sigma\): Sigmoid activation function (\(0,\ldots 1\))
-\item
-  \(\tanh\): Hyperbolic tangent activation function (\$ -1 \ldots 1\$)
-\end{itemize}
-
-\hypertarget{convlstm-network}{%
-\subsection{ConvLSTM network}\label{convlstm-network}}
-
-The FC-LSTM fails to handle information when handling spatiotemporal
-data due to its reliance on full connections in both input-to-state and
-state-to-state transitions. To adress this limitation we use a convLSTM
-architecture. convLSTM replaces the fully connected operations in the
-LSTM with convolutional operations. Hence, all inputs
-\(X_1, \ldots, X_t\), cell outputs \(C_1, \ldots, C_t\), hidden states
-\(H_1, \ldots, H_t\), and gates \(i_t, f_t, o_t\) of the ConvLSTM are 3D
-tensors. The last two dimensions of these tensors represent spatial
-dimensions, specifically rows and columns. Conceptually, these inputs
-and states can be visualized as vectors positioned on a spatial grid.
-
-In the ConvLSTM, the future state of a specific cell on this grid is
-determined by the inputs and past states of its neighboring cells. This
-spatial consideration is integrated by employing a convolution operator
-in both state-to-state and input-to-state transitions, as illustrated in
-Fig. 2. The foundational equations for ConvLSTM are:
-
+It becomes immediatly apparent that in case of the convLSTM, the bias,
+gate and state vectors must become vector fields
+(\(\in \mathbb{R}^{N_h \times (N_x \times N_y)}\)) as well. We can write
+this mapping in the same fashion as Eq. (Equation~\ref{eq-LSTM-gates})
+but with replacing the normal matrix-vector multiplication by a
+convolution (denoted with \(\ast\)), i.e.\\
 \[
 \begin{aligned}
-i_t &= \sigma(W_{xi} \ast X_t + W_{hi} \ast H_{t-1} + W_{ci} \circ C_{t-1} + b_i) \\
-f_t &= \sigma(W_{xf} \ast X_t + W_{hf} \ast H_{t-1} + W_{cf} \circ C_{t-1} + b_f) \\
-C_t &= f_t \circ C_{t-1} + i_t \circ \tanh(W_{xc} \ast X_t + W_{hc} \ast H_{t-1} + b_c) \\
-o_t &= \sigma(W_{xo} \ast X_t + W_{ho} \ast H_{t-1} + W_{co} \circ C_t + b_o) \\
-H_t &= o_t \circ \tanh(C_t)
+\vec{g}^t[\tau] = \pmb{\mathcal{M}}^{g} \ast \vec{X}^t[\tau] + \pmb{\mathcal{N}}^{g} \ast  \vec{H}^t[ \tau-1]+ \vec{\mathcal{B}}^g \,
 \end{aligned}
 \]
+
+The susebsequent processing of the \(\vec{g}^t[\tau]\) remains
+symbolically the same as presented in Eq. (Equation~\ref{eq-LSTM}) but
+with all appearing quantities meaning standing now for vector fields
+instead of simple vectors.
 
@@
 
-\hypertarget{implemented-model-architecture}{%
 \subsection{Implemented model
-architecture}\label{implemented-model-architecture}}
+architecture}\label{implemented-model-architecture}
 
@@
 
-{\centering \includegraphics{model_structure.png}
+\centering{
+
+\includegraphics{model_structure.png}
 
@@
 
-\hypertarget{tbl-letters}{}
 \begin{longtable}[]{@{}ll@{}}
-\caption{\label{tbl-letters}Hyperparameters}\tabularnewline
-\toprule()
+\caption{Hyperparameters}\label{tbl-letters}\tabularnewline
+\toprule\noalign{}
 Parameter name & Parameter size \\
-\midrule()
+\midrule\noalign{}
 \endfirsthead
-\toprule()
+\toprule\noalign{}
 Parameter name & Parameter size \\
-\midrule()
+\midrule\noalign{}
 \endhead
+\bottomrule\noalign{}
+\endlastfoot
 Channel size & 4 \\
@@
 ~Learning Rate & 1e-3 with CosineAnnealing \\
-\bottomrule()
 \end{longtable}
@@
 
-\hypertarget{results}{%
-\section{Results}\label{results}}
+\section{Results}\label{results}
 
@@
 
-Figure~\ref{fig-PerformanceNeuralNetworkRunoff} shows the perfomance of
+The accuracy of the model is further displayed in
+\textbf{?@fig-statistical-evaluationNN} . The correlation is close to 1
+and the residuals show the error of the model is below 1\% of the
+original data.
+
+\includegraphics{images/paste-13.png}
+
+\textbf{?@fig-PerformanceNeuralNetworkRunoff} shows the performance of
 the model using the test dataset. The predicted total river runoff for
@@
 
-\begin{figure}
-
-{\centering \includegraphics{images/paste-1.png}
-
-}
-
-\caption{\label{fig-PerformanceNeuralNetworkRunoff}Figure ..}
-
-\end{figure}
-
-The accuracy of the model is further displayed in
-Figure~\ref{fig-statistical-evaluationNN} . The correlation is close to
-1 and the residuals show the error of the model is below 1\% of the
-original data.
-
-\begin{figure}
-
-{\centering \includegraphics{images/paste-4.png}
-
-}
-
-\caption{\label{fig-statistical-evaluationNN}Figure \ldots{}}
-
-\end{figure}
+\includegraphics{images/paper_total_river_runoff.png}
 
@@
 
-\includegraphics{images/paste-5.png}
+\includegraphics{images/paste-9.png}
 
-\includegraphics{images/paste-6.png}
+\includegraphics{images/paste-12.png}
 
-\hypertarget{acknowledgments}{%
-\section{Acknowledgments}\label{acknowledgments}}
+\section{Acknowledgments}\label{acknowledgments}
 
@@
 
-\hypertarget{open-research}{%
-\section{Open research}\label{open-research}}
+\section{Open research}\label{open-research}
 
@@
 
-\hypertarget{references}{%
-\section*{References}\label{references}}
+\section*{References}\label{references}
 \addcontentsline{toc}{section}{References}
 
-\hypertarget{refs}{}
+\phantomsection\label{refs}
 \begin{CSLReferences}{1}{0}
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-dee2011}{}}%
+\bibitem[\citeproctext]{ref-dee2011}
 Dee, D. P., Uppala, S. M., Simmons, A. J., Berrisford, P., Poli, P.,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-fangExaminingApplicabilityDifferent2019}{}}%
+\bibitem[\citeproctext]{ref-fangExaminingApplicabilityDifferent2019}
 Fang, W., Huang, S., Ren, K., Huang, Q., Huang, G., Cheng, G., \& Li, K.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-gruxf6ger2022}{}}%
+\bibitem[\citeproctext]{ref-gruxf6ger2022}
 Gröger, M., Placke, M., Meier, M., Börgel, F., Brunnabend, S.-E.,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-ha2021}{}}%
+\bibitem[\citeproctext]{ref-ha2021}
 Ha, S., Liu, D., \& Mu, L. (2021). Prediction of Yangtze River
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-hagemannHighResolutionDischarge2020}{}}%
+\bibitem[\citeproctext]{ref-hagemannHighResolutionDischarge2020}
 Hagemann, S., Stacke, T., \& Ho-Hagemann, H. T. M. (2020). High
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-hordoir2019}{}}%
+\bibitem[\citeproctext]{ref-hordoir2019}
 Hordoir, R., Axell, L., Höglund, A., Dieterich, C., Fransner, F.,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-shi2015}{}}%
+\bibitem[\citeproctext]{ref-shi2015}
 SHI, X., Chen, Z., Wang, H., Yeung, D.-Y., Wong, W., \& WOO, W. (2015).
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-tanAdaptiveMiddleLongterm2018}{}}%
+\bibitem[\citeproctext]{ref-tanAdaptiveMiddleLongterm2018}
 Tan, Q.-F., Lei, X.-H., Wang, X., Wang, H., Wen, X., Ji, Y., \& Kang,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-theross}{}}%
+\bibitem[\citeproctext]{ref-theross}
 The rossby centre regional climate model RCA3: Model description and
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-yangDerivingOperatingRules2018}{}}%
+\bibitem[\citeproctext]{ref-yangDerivingOperatingRules2018}
 Yang, Z., Liu, P., Cheng, L., Wang, H., Ming, B., \& Gong, W. (2018).

### 2024-04-23 paper/paper.qmd
Update results

@@
 
-![BaltConvLSTM](model_structure.png){#fig-baltNet}
+![Schematic structure of the convLSTM implementation for river runoff forecasting.](convLSTM.png){#fig-baltNet}
 
@@
 | Channel size         | 4                         |
-| Num. hidden layer    | 10                        |
+| Num. hidden layer    | 9                         |
 | Num. timesteps       | 30                        |
-| Conv. Kernelsize     | (9,9)                     |
+| Conv. Kernelsize     | (7,7)                     |
 | Num. ConvLSTM layers | 1                         |
-| Batch size           | 64                        |
+| Batch size           | 50                        |
 | Learning Rate        | 1e-3 with CosineAnnealing |
@@
 
-The model was trained with daily data for the period 1979 to 2011, as this period represents the mapping between the original atmospheric forcing and E-HYPE without further bias correction to match observations.
-The data was then divided into randomly chosen splits of 80% training data, 10% validation data to evaluate the performance of the model during training, and 10% training data that is finally used to evaluate the performance of the model after training.
+The model was trained with daily data for the period 1979 to 2011, as this period represents the only period of E-HYPE without further bias correction applied to the runoff to match observations.
+The data was divided into randomly chosen splits of 80% training data, 10% validation data to evaluate the performance of the model during training, and 10% training data which is finally used to evaluate the performance of the model after training.
 The model was trained for 400 epochs and the model weights with the lowest mean squared error during training have been stored.
@@
 The accuracy of the model is displayed in Figure @fig-statistical-evaluationNN.
-For this evaluation, the test dataset was utilized.
-The left panel presents a scatter plot comparing the original E-HYPE data with the predicted river runoff, revealing a high correlation of 0.997.
-The middle panel illustrates the relative prediction error in relation to the original E-HYPE data, indicating that, on daily timescales, the model can predict river runoff with an accuracy of $\pm$ 5%.
-Partly, this may be attributed to the use of a different atmospheric dataset than that what was used to force the original E-HYPE model.
-However, by applying a rolling-mean with a 5-day window, the prediction error is reduced to less than 1%.
+As mention above for evaluation, the test dataset was utilized.
+The left panel panel (Figure @fig-statistical-evaluationNN a) ) illustrates the relative prediction error in relation to the original E-HYPE data, indicating that, on daily timescales, the model can predict river runoff with an accuracy of $\pm$ 5%.
+The overall correlation is 0.997 with the resulting error metrics yielding a RMSE of 323.99 $m^3$/s and MAE of 249.51 $m^3$/s.
+While the model's performance is satisfactory, the discrepancies between the actual values and the predictions could partly be attributed to the use of a different atmospheric dataset than the one originally used to drive the E-HYPE model.
+However, by applying a rolling mean with a 5-day window, the prediction error is reduced to less than 1%, which is acceptable for the purposes of climate modeling.
 Lastly, the right panel demonstrates that the distribution of residuals follows a Gaussian shape, suggesting that our model does not exhibit bias by systematically over or underestimating river runoff values.
@@
 
-@fig-PerformanceNeuralNetworkRunoff shows the predicted and the original river runoff using the test dataset.
-The predicted total river runoff for the Baltic Sea is closely matching the original data.
-Zooming in on the largest individual rivers (lower panels) it can be seen that that also the prediction of the individual rivers is close to the original data.
+In the following we will now address the overall performance of the total river runoff while also zooming in on the four largest rivers entering the Baltic Sea.
+\@fig-PerformanceNeuralNetworkRunoff shows the predicted and the original river runoff using the test dataset.
+The predicted total river runoff for the Baltic Sea is closely matching the original data (\@fig-PerformanceNeuralNetworkRunoff a).
+Zooming in on the largest individual rivers (\@fig-PerformanceNeuralNetworkRunoff b-e) it can be seen that that also the prediction of the individual rivers is close to the original data.
 
-![](images/paper_total_river_runoff.png)
+![](../src/figures/paper_total_river_runoff.png)
 
@@
 It can be seen that the model simulation using the predicted river runoff by the ConvLSTM is closely mirroring the control simulation.
-The left panel shows the surface salinity, representing the high-frequency variations in the salinity, which is heavily affected by river runoff.
-The right panel shows the bottom salinity which can be viewed as a low-pass filter in the Baltic Sea, which is also closely mirrored by the ConvLSTM predictions.
+The upper panel (@fig-by15 a) shows the surface salinity, representing the high-frequency variations in the salinity, which is heavily affected by river runoff.
+The lower panel (@fig-by15 b) shows the bottom salinity which can be viewed as a low-pass filter in the Baltic Sea, which is also closely mirrored by the ConvLSTM predictions.
 
-![](images/paste-9.png)
+![](../src/figures/by15_model.png)
 
-![](images/paste-12.png)
+The final evaluation of the ConvLSTM model concentrates on the spatial accuracy of river runoff predictions.
+Figure xxx a) shows the vertically averaged salinity for the period 1981 to 2011 in the reference simulation.
+It highlights the strong horizontal gradients and complex topographic features in the Baltic Sea, as evidenced by salinity variations in deeper waters, captured by the vertical integration.
+Figure 4b compares these results by showing the percentage difference in vertically averaged salinity between the ConvLSTM simulation and the reference simulation.
+Overall, the differences remain below 1%, except near a river mouth in the Gulf of Riga (22-24°E, 56.5-58.5°N for orientation), where the difference is approximately 1%.
+
+![](../src/figures/evaluation_MOM.png){#spatialBalticSea}
 

### 2024-04-23 paper/paper.tex
Update results

@@
 \usepackage{amsmath,amssymb}
+\usepackage{lmodern}
 \usepackage{iftex}
@@
   \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-\usepackage{lmodern}
-\ifPDFTeX\else  
-    % xetex/luatex font selection
 \fi
@@
 \makeatother
-% definitions for citeproc citations
-\NewDocumentCommand\citeproctext{}{}
-\NewDocumentCommand\citeproc{mm}{%
-  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
-\makeatletter
- % allow citations to break across lines
- \let\@cite@ofmt\@firstofone
- % avoid brackets around text for \cite:
- \def\@biblabel#1{}
- \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
-\makeatother
 \newlength{\cslhangindent}
@@
 \setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
- {\begin{list}{}{%
-  \setlength{\itemindent}{0pt}
-  \setlength{\leftmargin}{0pt}
-  \setlength{\parsep}{0pt}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
   % turn on hanging indent if param 1 is 1
   \ifodd #1
-   \setlength{\leftmargin}{\cslhangindent}
-   \setlength{\itemindent}{-1\cslhangindent}
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
   \fi
   % set entry spacing
-  \setlength{\itemsep}{#2\baselineskip}}}
- {\end{list}}
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
 \usepackage{calc}
-\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
@@
 \linenumbers
+\makeatletter
+\makeatother
+\makeatletter
+\makeatother
 \makeatletter
@@
 \makeatletter
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
 \makeatother
 \makeatletter
-\@ifpackageloaded{caption}{}{\usepackage{caption}}
-\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
+\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
+\makeatother
+\makeatletter
+\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
+\makeatother
+\makeatletter
 \makeatother
@@
 \fi
-\usepackage{bookmark}
-
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
@@
 
-\section{Introduction}\label{introduction}
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, boxrule=0pt, frame hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners]}{\end{tcolorbox}}\fi
+
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
@@
 area where Convolutional Neural Networks (CNNs) excel. Reconizing this
-limitation (SHI et al., 2015) proposed a convolutional LSTM (ConvLSTM)
+limitation (\textbf{shi2015?}) proposed a convolutional LSTM (ConvLSTM)
 architectures, which combines the strengths of both LSTM and CNN. The
 ConvLSTM network has been proven useful for precipitation nowcasting
-(SHI et al., 2015) or for river runoff forecasting {[}Ha et al. (2021);
-@niu2020{]}.
+(\textbf{shi2015?}) or for river runoff forecasting
+{[}(\textbf{ha2021?}); @niu2020{]}.
 
@@
 
-\section{Methods}\label{methods}
+\hypertarget{methods}{%
+\section{Methods}\label{methods}}
 
+\hypertarget{runoff-data-used-for-training}{%
 \subsection{Runoff data used for
-training}\label{runoff-data-used-for-training}
+training}\label{runoff-data-used-for-training}}
 
@@
 hindcast simulation that was forced by a regional downscaling of
-ERA-Interim (Dee et al., 2011) with RCA3 ({``The rossby centre regional
-climate model RCA3,''} n.d.) and implemented into NEMO-Nordic (Hordoir
-et al., 2019) as a mass flux. For the periods before (1961 to 1978) and
-after (2012 to 2018) additional spatial temporal corrections have been
-applied to the runoff data, and have therefore been ignored. For more
-information see Gröger et al. (2022) and references therein.
+ERA-Interim (\textbf{dee2011?}) with RCA3 (\textbf{theross?}) and
+implemented into NEMO-Nordic (\textbf{hordoir2019?}) as a mass flux. For
+the periods before (1961 to 1978) and after (2012 to 2018) additional
+spatial temporal corrections have been applied to the runoff data, and
+have therefore been ignored. For more information see
+(\textbf{gröger2022?}) and references therein.
 
-\subsection{Atmospheric Forcing}\label{atmospheric-forcing}
+\hypertarget{atmospheric-forcing}{%
+\subsection{Atmospheric Forcing}\label{atmospheric-forcing}}
 
@@
 
-\subsection{Ocean Model}\label{ocean-model}
+\hypertarget{ocean-model}{%
+\subsection{Ocean Model}\label{ocean-model}}
 
@@
 Baltic Sea to the North Sea. The maximum depth was set at 264 meters. A
-more detailed description of the setup can be found in (Gröger et al.,
-2022).
+more detailed description of the setup can be found in
+(\textbf{gröger2022?}).
 
-\subsection{LSTM network}\label{lstm-network}
+\hypertarget{lstm-network}{%
+\subsection{LSTM network}\label{lstm-network}}
 
@@
 
-\centering{
-
-\includegraphics{draw_lstm.png}
+{\centering \includegraphics{draw_lstm.png}
 
@@
 
-\end{figure}%
+\end{figure}
 
@@
 matrix-vector form as
-\begin{equation}\phantomsection\label{eq-LSTM-gates}{
+\begin{equation}\protect\hypertarget{eq-LSTM-gates}{}{
 \begin{aligned}
@@
 \end{aligned}
-}\end{equation}
+}\label{eq-LSTM-gates}\end{equation}
 
@@
 
-\begin{equation}\phantomsection\label{eq-LSTM}{
+\begin{equation}\protect\hypertarget{eq-LSTM}{}{
 \begin{aligned}
@@
 \end{aligned}
-}\end{equation}
+}\label{eq-LSTM}\end{equation}
 
@@
 
-\subsection{ConvLSTM network}\label{convlstm-network}
+\hypertarget{convlstm-network}{%
+\subsection{ConvLSTM network}\label{convlstm-network}}
 
@@
 
+\hypertarget{implemented-model-architecture}{%
 \subsection{Implemented model
-architecture}\label{implemented-model-architecture}
+architecture}\label{implemented-model-architecture}}
 
@@
 
-\centering{
-
-\includegraphics{model_structure.png}
+{\centering \includegraphics{convLSTM.png}
 
@@
 
-\caption{\label{fig-baltNet}BaltConvLSTM}
+\caption{\label{fig-baltNet}Schematic structure of the convLSTM
+implementation for river runoff forecasting.}
 
-\end{figure}%
+\end{figure}
 
@@
 
+\hypertarget{tbl-letters}{}
 \begin{longtable}[]{@{}ll@{}}
-\caption{Hyperparameters}\label{tbl-letters}\tabularnewline
-\toprule\noalign{}
+\caption{\label{tbl-letters}Hyperparameters}\tabularnewline
+\toprule()
 Parameter name & Parameter size \\
-\midrule\noalign{}
+\midrule()
 \endfirsthead
-\toprule\noalign{}
+\toprule()
 Parameter name & Parameter size \\
-\midrule\noalign{}
+\midrule()
 \endhead
-\bottomrule\noalign{}
-\endlastfoot
 Channel size & 4 \\
-Num. hidden layer & 10 \\
+Num. hidden layer & 9 \\
 Num. timesteps & 30 \\
-Conv. Kernelsize & (9,9) \\
+Conv. Kernelsize & (7,7) \\
 Num. ConvLSTM layers & 1 \\
-Batch size & 64 \\
-~Learning Rate & 1e-3 with CosineAnnealing \\
+Batch size & 50 \\
+Learning Rate & 1e-3 with CosineAnnealing \\
+\bottomrule()
 \end{longtable}
@@
 
-\section{Results}\label{results}
+\hypertarget{results}{%
+\section{Results}\label{results}}
 
-For the evaluation of the model performance we consider the period 1979
-to 2011. For this period no bias correction was applied to the orignal
-E-HYPE dataset. We chose a split of 80\% training data, 10\% validation
-data to evaluate the performance of the model during training, and 10\%
-training data that is finally used to evaluate the performance of the
-model after training.
+The model was trained with daily data for the period 1979 to 2011, as
+this period represents the only period of E-HYPE without further bias
+correction applied to the runoff to match observations. The data was
+divided into randomly chosen splits of 80\% training data, 10\%
+validation data to evaluate the performance of the model during
+training, and 10\% training data which is finally used to evaluate the
+performance of the model after training. The model was trained for 400
+epochs and the model weights with the lowest mean squared error during
+training have been stored.
 
-The accuracy of the model is further displayed in
-\textbf{?@fig-statistical-evaluationNN} . The correlation is close to 1
-and the residuals show the error of the model is below 1\% of the
-original data.
+The accuracy of the model is displayed in Figure
+\textbf{?@fig-statistical-evaluationNN}. As mention above for
+evaluation, the test dataset was utilized. The left panel panel (Figure
+\textbf{?@fig-statistical-evaluationNN} a) ) illustrates the relative
+prediction error in relation to the original E-HYPE data, indicating
+that, on daily timescales, the model can predict river runoff with an
+accuracy of \(\pm\) 5\%. The overall correlation is 0.997 with the
+resulting error metrics yielding a RMSE of 323.99 \(m^3\)/s and MAE of
+249.51 \(m^3\)/s. While the model's performance is satisfactory, the
+discrepancies between the actual values and the predictions could partly
+be attributed to the use of a different atmospheric dataset than the one
+originally used to drive the E-HYPE model. However, by applying a
+rolling mean with a 5-day window, the prediction error is reduced to
+less than 1\%, which is acceptable for the purposes of climate modeling.
+Lastly, the right panel demonstrates that the distribution of residuals
+follows a Gaussian shape, suggesting that our model does not exhibit
+bias by systematically over or underestimating river runoff values.
 
-\includegraphics{images/paste-13.png}
+\includegraphics{../src/figures/paper_error_metrics.png}
 
-\textbf{?@fig-PerformanceNeuralNetworkRunoff} shows the performance of
-the model using the test dataset. The predicted total river runoff for
-the Baltic Sea is closely matching the original data. Zooming in on the
-largest individual rivers (lower panels) it can be seen that that also
-the prediction of the inidivual rivers is close to the original data.
+In the following we will now address the overall performance of the
+total river runoff while also zooming in on the four largest rivers
+entering the Baltic Sea. @fig-PerformanceNeuralNetworkRunoff shows the
+predicted and the original river runoff using the test dataset. The
+predicted total river runoff for the Baltic Sea is closely matching the
+original data (@fig-PerformanceNeuralNetworkRunoff a). Zooming in on the
+largest individual rivers (@fig-PerformanceNeuralNetworkRunoff b-e) it
+can be seen that that also the prediction of the individual rivers is
+close to the original data.
 
-\includegraphics{images/paper_total_river_runoff.png}
+\includegraphics{../src/figures/paper_total_river_runoff.png}
 
 Lastly, we evaluated the performance of the runoff model by
-incorporating the predicted river runoff as forcing into the ocean model
-MOM5. This provides a robust validation of the runoff model against more
-complex real world conditions. This allows us to ensure that the
-predictions accurately reflect the impact of the river discharge on the
-ocean dynamics, validating the temporal and spatial variability of the
-the river discharge. \textbf{?@fig-by15} shows the salinity comparison
-between the original E-HYPE river runoff and the predicted river runoff
-at BY15 - a central stations in the Baltic Sea.
+incorporating the predicted river runoff as runoff forcing into the
+ocean model MOM5. This provides a robust validation of the runoff model
+against more complex real world conditions. This allows us to ensure
+that the predictions accurately reflect the impact of the river
+discharge on the ocean dynamics, validating the temporal and spatial
+variability of the the river discharge. \textbf{?@fig-by15} shows the
+salinity comparison between the original E-HYPE river runoff and the
+predicted river runoff at BY15 - a central stations in the Baltic Sea.
+It can be seen that the model simulation using the predicted river
+runoff by the ConvLSTM is closely mirroring the control simulation. The
+upper panel (\textbf{?@fig-by15} a) shows the surface salinity,
+representing the high-frequency variations in the salinity, which is
+heavily affected by river runoff. The lower panel (\textbf{?@fig-by15}
+b) shows the bottom salinity which can be viewed as a low-pass filter in
+the Baltic Sea, which is also closely mirrored by the ConvLSTM
+predictions.
 
-\includegraphics{images/paste-9.png}
+\includegraphics{../src/figures/by15_model.png}
 
-\includegraphics{images/paste-12.png}
+The final evaluation of the ConvLSTM model concentrates on the spatial
+accuracy of river runoff predictions. Figure xxx a) shows the vertically
+averaged salinity for the period 1981 to 2011 in the reference
+simulation. It highlights the strong horizontal gradients and complex
+topographic features in the Baltic Sea, as evidenced by salinity
+variations in deeper waters, captured by the vertical integration.
+Figure 4b compares these results by showing the percentage difference in
+vertically averaged salinity between the ConvLSTM simulation and the
+reference simulation. Overall, the differences remain below 1\%, except
+near a river mouth in the Gulf of Riga (22-24°E, 56.5-58.5°N for
+orientation), where the difference is approximately 1\%.
 
-\section{Acknowledgments}\label{acknowledgments}
+\includegraphics{../src/figures/evaluation_MOM.png}
+
+\hypertarget{acknowledgments}{%
+\section{Acknowledgments}\label{acknowledgments}}
 
@@
 
-\section{Open research}\label{open-research}
+\hypertarget{open-research}{%
+\section{Open research}\label{open-research}}
 
@@
 
-\section*{References}\label{references}
+\hypertarget{references}{%
+\section*{References}\label{references}}
 \addcontentsline{toc}{section}{References}
 
-\phantomsection\label{refs}
+\hypertarget{refs}{}
 \begin{CSLReferences}{1}{0}
@@
 
-\bibitem[\citeproctext]{ref-dee2011}
-Dee, D. P., Uppala, S. M., Simmons, A. J., Berrisford, P., Poli, P.,
-Kobayashi, S., et al. (2011). The ERA-Interim reanalysis: configuration
-and performance of the data assimilation system. \emph{Quarterly Journal
-of the Royal Meteorological Society}, \emph{137}(656), 553--597.
-\url{https://doi.org/10.1002/qj.828}
-
-\bibitem[\citeproctext]{ref-fangExaminingApplicabilityDifferent2019}
+\leavevmode\vadjust pre{\hypertarget{ref-fangExaminingApplicabilityDifferent2019}{}}%
 Fang, W., Huang, S., Ren, K., Huang, Q., Huang, G., Cheng, G., \& Li, K.
@@
 
-\bibitem[\citeproctext]{ref-gruxf6ger2022}
-Gröger, M., Placke, M., Meier, M., Börgel, F., Brunnabend, S.-E.,
-Dutheil, C., et al. (2022). \emph{The Baltic Sea model inter-comparison
-project BMIP {\textendash} a platform for model development, evaluation,
-and uncertainty assessment}. Retrieved from
-\url{https://gmd.copernicus.org/preprints/gmd-2022-160/}
-
-\bibitem[\citeproctext]{ref-ha2021}
-Ha, S., Liu, D., \& Mu, L. (2021). Prediction of Yangtze River
-streamflow based on deep learning neural network with El
-Niño{\textendash}Southern Oscillation. \emph{Scientific Reports},
-\emph{11}(1), 11738. \url{https://doi.org/10.1038/s41598-021-90964-3}
-
-\bibitem[\citeproctext]{ref-hagemannHighResolutionDischarge2020}
+\leavevmode\vadjust pre{\hypertarget{ref-hagemannHighResolutionDischarge2020}{}}%
 Hagemann, S., Stacke, T., \& Ho-Hagemann, H. T. M. (2020). High
@@
 
-\bibitem[\citeproctext]{ref-hordoir2019}
-Hordoir, R., Axell, L., Höglund, A., Dieterich, C., Fransner, F.,
-Gröger, M., et al. (2019). Nemo-nordic 1.0: A NEMO-based ocean model for
-the baltic and north seas {\textendash} research and operational
-applications. \emph{Geoscientific Model Development}, \emph{12}(1),
-363--386. \url{https://doi.org/10.5194/gmd-12-363-2019}
-
-\bibitem[\citeproctext]{ref-shi2015}
-SHI, X., Chen, Z., Wang, H., Yeung, D.-Y., Wong, W., \& WOO, W. (2015).
-Convolutional LSTM network: A machine learning approach for
-precipitation nowcasting. In (Vol. 28). Curran Associates, Inc.
-Retrieved from
-\url{https://proceedings.neurips.cc/paper/2015/hash/07563a3fe3bbe7e3ba84431ad9d055af-Abstract.html}
-
-\bibitem[\citeproctext]{ref-tanAdaptiveMiddleLongterm2018}
+\leavevmode\vadjust pre{\hypertarget{ref-tanAdaptiveMiddleLongterm2018}{}}%
 Tan, Q.-F., Lei, X.-H., Wang, X., Wang, H., Wen, X., Ji, Y., \& Kang,
@@
 
-\bibitem[\citeproctext]{ref-theross}
-The rossby centre regional climate model RCA3: Model description and
-performance - tellus a: Dynamic meteorology and oceanography. (n.d.).
-Retrieved from
-\url{https://a.tellusjournals.se/articles/10.1111/j.1600-0870.2010.00478.x}
-
-\bibitem[\citeproctext]{ref-yangDerivingOperatingRules2018}
+\leavevmode\vadjust pre{\hypertarget{ref-yangDerivingOperatingRules2018}{}}%
 Yang, Z., Liu, P., Cheng, L., Wang, H., Ming, B., \& Gong, W. (2018).

### 2024-04-23 paper/paper.qmd
First part of the discussion

@@
 
-It becomes immediatly apparent that in case of the convLSTM, the bias, gate and state vectors must become vector fields ($\in \mathbb{R}^{N_h \times (N_x \times N_y)}$) as well.
+It becomes immediately apparent that in case of the convLSTM, the bias, gate and state vectors must become vector fields ($\in \mathbb{R}^{N_h \times (N_x \times N_y)}$) as well.
 We can write this mapping in the same fashion as Eq.
@@
 
+## Discussion
+
+With the increasing demand of decision makers for regional climate projections, that allow to quantify regional climate change impacts, the availability of precise hydrological forecasting becomes invaluable.
+The quality of a projection for a coastal sea such as the Baltic Sea is too large parts based on the quality of the hyrdological conditions.
+In this work we analyze the implementation of Convolutional Long Short-Term Memory (ConvLSTM) networks for predicting river runoff, highlighting its potential to enhance river runoff forecasting across different coastal seas.
+Given the unique hydrological characteristics of the Baltic Sea, largely influenced by its limited connection to the open ocean and significant freshwater input from surrounding rivers, the region presents a critical case for the application of sophisticated forecasting models.
+
+The transition from traditional hydrological models to machine learning approaches, such as the ConvLSTM model, offers significant advantages.
+The ConvLSTM can be seamlessly integrated into regional climate models, allowing for the real-time computation of river runoff while performing climate projections.
+While the initial training of the model requires substantial computational resources, it remains less intensive than running comprehensive hydrological models.
+Furthermore, once trained, the ConvLSTM model is computationally efficient during inference, ensuring enhanced forecasting capabilities without significantly increasing computational demands.
+
+While the ConvLSTM represents an advancement for the climate community, given that running and tuning traditional hydrological models demands extensive expertise, models like E-HYPE maintain an essential role.
+They provide a comprehensive dataset that helps to train our ConvLSTM model effectively.
+This robust training enables the machine learning models to achieve highly accurate predictive weights.
+Thus, rather than rendering traditional methods obsolete, the integration of machine learning models builds upon and enhances the foundational data provided by them.
+
+The robust performance of the ConvLSTM model in simulating river runoff and its possible effective integration into regional climate models pave the way for a multitude of new storyline simulations.
+Hence, we can now explore various "what-if" scenarios more reliably, under the assumption that the model weights attained during training are robust.
+
+In summary, we showed that the ConvLSTM model demonstrated robust performance in forecasting river runoff across 97 rivers entering the Baltic Sea.
+Trained on data from 1979 to 2011, the model achieved an impressive daily prediction accuracy of ±5%.
+This capability to generate accurate and detailed simulations enables to examine the potential impacts of different climate scenarios.
+Such precision in forecasting and scenario testing is crucial for crafting effective water resource management strategies and adapting to the changing climate.
+
+Ultimately, the integration of ConvLSTM into regional climate models represents a significant step forward in our ability to understand and predict the complex dynamics of river systems and their impact on regional climate systems.
+
 ## Acknowledgments

### 2024-04-23 paper/paper.tex
First part of the discussion

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, boxrule=0pt, frame hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, boxrule=0pt, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable]}{\end{tcolorbox}}\fi
 
@@
 
-It becomes immediatly apparent that in case of the convLSTM, the bias,
+It becomes immediately apparent that in case of the convLSTM, the bias,
 gate and state vectors must become vector fields
@@
 
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+With the increasing demand of decision makers for regional climate
+projections, that allow to quantify regional climate change impacts, the
+availability of precise hydrological forecasting becomes invaluable. The
+quality of a projection for a coastal sea such as the Baltic Sea is too
+large parts based on the quality of the hyrdological conditions. In this
+work we analyze the implementation of Convolutional Long Short-Term
+Memory (ConvLSTM) networks for predicting river runoff, highlighting its
+potential to enhance river runoff forecasting across different coastal
+seas. Given the unique hydrological characteristics of the Baltic Sea,
+largely influenced by its limited connection to the open ocean and
+significant freshwater input from surrounding rivers, the region
+presents a critical case for the application of sophisticated
+forecasting models.
+
+The transition from traditional hydrological models to machine learning
+approaches, such as the ConvLSTM model, offers significant advantages.
+The ConvLSTM can be seamlessly integrated into regional climate models,
+allowing for the real-time computation of river runoff while performing
+climate projections. While the initial training of the model requires
+substantial computational resources, it remains less intensive than
+running comprehensive hydrological models. Furthermore, once trained,
+the ConvLSTM model is computationally efficient during inference,
+ensuring enhanced forecasting capabilities without significantly
+increasing computational demands.
+
+While the ConvLSTM represents an advancement for the climate community,
+given that running and tuning traditional hydrological models demands
+extensive expertise, models like E-HYPE maintain an essential role. They
+provide a comprehensive dataset that helps to train our ConvLSTM model
+effectively. This robust training enables the machine learning models to
+achieve highly accurate predictive weights. Thus, rather than rendering
+traditional methods obsolete, the integration of machine learning models
+builds upon and enhances the foundational data provided by them.
+
+The robust performance of the ConvLSTM model in simulating river runoff
+and its possible effective integration into regional climate models pave
+the way for a multitude of new storyline simulations. Hence, we can now
+explore various ``what-if'' scenarios more reliably, under the
+assumption that the model weights attained during training are robust.
+
+In summary, we showed that the ConvLSTM model demonstrated robust
+performance in forecasting river runoff across 97 rivers entering the
+Baltic Sea. Trained on data from 1979 to 2011, the model achieved an
+impressive daily prediction accuracy of ±5\%. This capability to
+generate accurate and detailed simulations enables to examine the
+potential impacts of different climate scenarios. Such precision in
+forecasting and scenario testing is crucial for crafting effective water
+resource management strategies and adapting to the changing climate.
+
+Ultimately, the integration of ConvLSTM into regional climate models
+represents a significant step forward in our ability to understand and
+predict the complex dynamics of river systems and their impact on
+regional climate systems.
+
 \hypertarget{acknowledgments}{%

### 2024-04-25 paper/paper.qmd
Modify introduction

@@
 In the context of climate change studies, river runoff is usually generated in two ways.
-First, river runoff as input for ocean models can be created using hyrdological models such as the Hydrological Discharge (HD) model [@hagemannHighResolutionDischarge2020].
+First, river runoff as input for ocean models can be created using hydrological models such as the Hydrological Discharge (HD) model [@hagemannHighResolutionDischarge2020].
 HD models calculates the water balance using hydrological processes (e.g.
@@
 It represents a complex forecasting tool that uses underlying physical processes.
-A different approach would use data-based models that intergrate statistical correction, using the land surface schemes of global or regional climate models.
+A different approach would use data-based models that integrate statistical correction, using the land surface schemes of global or regional climate models.
 
-The releatively recent rise of machine learning (ML) models has been mostly explored for river runoff forecasting, as accurate runoff forecasting, especially over extended periods, is pivotal for effective water resources management [@fangExaminingApplicabilityDifferent2019; @tanAdaptiveMiddleLongterm2018; @yangDerivingOperatingRules2018].
-Common approaches employ artificial neural networks, support vector machines, adaptive neuro-fuzzy inference systems, and notably, Long Short-Term Memory (LSTM) neural networks that have gained traction for long-term hydrological forecasting due to their excellent performance (Humphrey et al 2016, Huang et al 2014, Ashrafi et al 2017, Yuan et al 2018, Xu et al 2021).
+The relatively recent emergence of machine learning (ML) models has offered different approaches to river runoff forecasting.
+Accurate runoff forecasting is crucial for effective water resources management, particularly over extended periods [@fangExaminingApplicabilityDifferent2019; @tanAdaptiveMiddleLongterm2018].
+Common approaches employ feed-forward artificial neural networks, support vector machines, adaptive neuro-fuzzy inference systems, and notably, Long Short-Term Memory (LSTM) neural networks that have gained traction for long-term hydrological forecasting due to their excellent performance [@humphrey2016hybrid; @huang2014monthly; @ashrafi2017fully; @liu2020streamflow; @fang2022application; @kratzert2018rainfall].
 
-LSTM networks, an evolution of the classical Recurrent Neural Networks (RNNs), have shown stability and efficacy in sequence-to-sequence predictions, such as using climatic indices for rainfall estimation or long-term hydrological forecasting.
+LSTM networks, first introduced by @hochreiter1997long , are an evolution of the classical Recurrent Neural Networks (RNNs).
+Their structure enables them to learn long-term dependencies while avoiding the vanishing or exploding gradient problem.
+They have shown stability and efficacy in sequence-to-sequence predictions.
 However, a limitation of LSTMs is their inability to effectively capture two-dimensional structures, an area where Convolutional Neural Networks (CNNs) excel.
-Reconizing this limitation [@shi2015] proposed a convolutional LSTM (ConvLSTM) architectures, which combines the strengths of both LSTM and CNN.
-The ConvLSTM network has been proven useful for precipitation nowcasting [@shi2015] or for river runoff forecasting [@ha2021; \@niu2020].
+Recognizing this limitation @shiConvolutionalLSTMNetwork2015 proposed a Convolutional LSTM (ConvLSTM) architecture, which combines the strengths of both LSTM and CNN.
+The ConvLSTM network has been proven useful for spatio-temporal applications such as precipitation nowcasting [@shiConvolutionalLSTMNetwork2015], flood forecasting [@moishin2021designing], and river runoff forecasting [@ha2021; @zhu2023spatiotemporal].
 
-In the following we will show that in absence of a fully functioning hydrogolocial model, that also uses a rather complex parametrization, ConvLSTM also represent a robust way to predict multiple rivers at once for any given period using only atmospheric forcing.
-In this work we use the Baltic Sea catchment to illustrate our approach, while in principal the methodology we propose is universally applicable across various geographic regions.
-The Baltic Sea serves as a challenging example due to its unique hydrological characteristics, being nearly decoupled from the open ocean (see Figure).
+In this work, we demonstrate that ConvLSTM networks are a reliable method for predicting multiple rivers simultaneously, using only atmospheric forcing, even in the absence of a fully functional hydrological model with a complex parameterization.
+We use the Baltic Sea catchment as an example to illustrate our approach.
+Although the methodology we propose is universally applicable across various geographic regions, the Baltic Sea presents a challenge due to its unique hydrological characteristics, being nearly decoupled from the open ocean (see Figure).
 As a consequence, the salinity of the Baltic Sea is driven to a large part by freshwater supply from rivers.
-More generally, the freshwater input into the Baltic Sea comes either as river runoff or a positive net precipitation (precipitation minus evaporation) over the sea surface.
-The net precipitation accounts for 11 % and the river input for 89 % of the total freshwater input (Meier and Doescher, 2002).
-Modeling the Baltic Sea is therefore to a large part the result of the quality of the river input, that is used for the simulation.
+Freshwater enters the Baltic Sea through river runoff or positive net precipitation (precipitation minus evaporation) over the sea surface.
+The net precipitation accounts for 11 % and the river input for 89 % of the total freshwater input [@meier2002simulated].
+Modelling the Baltic Sea is therefore to a large part the result of the quality of the river input, that is used for the simulation.
 This makes the accurate modeling of river runoff especially critical for simulations pertaining to the Baltic Sea.
 
-In this work we will, we present a ConvLSTM architecture that is able to predict daily river runoff for 97 rivers across the Baltic Sea catchment.
+In this work we will, we present a ConvLSTM architecture that is able to predict daily river runoff for 97 rivers across the Baltic Sea catchment. ***mehr Fokus auf Neues?***
 

### 2024-05-13 paper/paper.qmd
Small fix

--- 2024-04-23 paper/paper.qmd
@@
 
-where $L$ represents Eqs. (@eq-LSTM-gates, @eq-LSTM) and the initial conditions are usually chosen as $\vec{H}^t[0]=\vec{C}^t[0]=0$.
+where $L\left(\vec{X}^t[\tau], \left(\vec{H}^t[\tau-1], \vec{C}^t[\tau-1] \right) \right)$ represents Eqs.
+(@eq-LSTM-gates, @eq-LSTM) and the initial conditions are usually chosen as $\vec{H}^t[0]=\vec{C}^t[0]=0$.
 
@@
 
-![Schematic structure of the convLSTM implementation for river runoff forecasting.](convLSTM.png){#fig-baltNet}
+![BaltConvLSTM](model_structure.png){#fig-baltNet}
 
@@
 | Channel size         | 4                         |
-| Num. hidden layer    | 9                         |
+| Num. hidden layer    | 10                        |
 | Num. timesteps       | 30                        |
-| Conv. Kernelsize     | (7,7)                     |
+| Conv. Kernelsize     | (9,9)                     |
 | Num. ConvLSTM layers | 1                         |
-| Batch size           | 50                        |
+| Batch size           | 64                        |
 |  Learning Rate       | 1e-3 with CosineAnnealing |
@@
 
-The model was trained with daily data for the period 1979 to 2011, as this period represents the only period of E-HYPE without further bias correction applied to the runoff to match observations.
-The data was divided into randomly chosen splits of 80% training data, 10% validation data to evaluate the performance of the model during training, and 10% training data which is finally used to evaluate the performance of the model after training.
+The model was trained with daily data for the period 1979 to 2011, as this period represents the mapping between the original atmospheric forcing and E-HYPE without further bias correction to match observations.
+The data was then divided into randomly chosen splits of 80% training data, 10% validation data to evaluate the performance of the model during training, and 10% training data that is finally used to evaluate the performance of the model after training.
 The model was trained for 400 epochs and the model weights with the lowest mean squared error during training have been stored.
@@
 The accuracy of the model is displayed in Figure @fig-statistical-evaluationNN.
-As mention above for evaluation, the test dataset was utilized.
-The left panel panel (Figure @fig-statistical-evaluationNN a) ) illustrates the relative prediction error in relation to the original E-HYPE data, indicating that, on daily timescales, the model can predict river runoff with an accuracy of $\pm$ 5%.
-The overall correlation is 0.997 with the resulting error metrics yielding a RMSE of 323.99 $m^3$/s and MAE of 249.51 $m^3$/s.
-While the model's performance is satisfactory, the discrepancies between the actual values and the predictions could partly be attributed to the use of a different atmospheric dataset than the one originally used to drive the E-HYPE model.
-However, by applying a rolling mean with a 5-day window, the prediction error is reduced to less than 1%, which is acceptable for the purposes of climate modeling.
+For this evaluation, the test dataset was utilized.
+The left panel presents a scatter plot comparing the original E-HYPE data with the predicted river runoff, revealing a high correlation of 0.997.
+The middle panel illustrates the relative prediction error in relation to the original E-HYPE data, indicating that, on daily timescales, the model can predict river runoff with an accuracy of $\pm$ 5%.
+Partly, this may be attributed to the use of a different atmospheric dataset than that what was used to force the original E-HYPE model.
+However, by applying a rolling-mean with a 5-day window, the prediction error is reduced to less than 1%.
 Lastly, the right panel demonstrates that the distribution of residuals follows a Gaussian shape, suggesting that our model does not exhibit bias by systematically over or underestimating river runoff values.
@@
 
-In the following we will now address the overall performance of the total river runoff while also zooming in on the four largest rivers entering the Baltic Sea.
-\@fig-PerformanceNeuralNetworkRunoff shows the predicted and the original river runoff using the test dataset.
-The predicted total river runoff for the Baltic Sea is closely matching the original data (\@fig-PerformanceNeuralNetworkRunoff a).
-Zooming in on the largest individual rivers (\@fig-PerformanceNeuralNetworkRunoff b-e) it can be seen that that also the prediction of the individual rivers is close to the original data.
+@fig-PerformanceNeuralNetworkRunoff shows the predicted and the original river runoff using the test dataset.
+The predicted total river runoff for the Baltic Sea is closely matching the original data.
+Zooming in on the largest individual rivers (lower panels) it can be seen that that also the prediction of the individual rivers is close to the original data.
 
-![](../src/figures/paper_total_river_runoff.png)
+![](images/paper_total_river_runoff.png)
 
@@
 It can be seen that the model simulation using the predicted river runoff by the ConvLSTM is closely mirroring the control simulation.
-The upper panel (@fig-by15 a) shows the surface salinity, representing the high-frequency variations in the salinity, which is heavily affected by river runoff.
-The lower panel (@fig-by15 b) shows the bottom salinity which can be viewed as a low-pass filter in the Baltic Sea, which is also closely mirrored by the ConvLSTM predictions.
+The left panel shows the surface salinity, representing the high-frequency variations in the salinity, which is heavily affected by river runoff.
+The right panel shows the bottom salinity which can be viewed as a low-pass filter in the Baltic Sea, which is also closely mirrored by the ConvLSTM predictions.
 
-![](../src/figures/by15_model.png)
+![](images/paste-9.png)
 
-The final evaluation of the ConvLSTM model concentrates on the spatial accuracy of river runoff predictions.
-Figure xxx a) shows the vertically averaged salinity for the period 1981 to 2011 in the reference simulation.
-It highlights the strong horizontal gradients and complex topographic features in the Baltic Sea, as evidenced by salinity variations in deeper waters, captured by the vertical integration.
-Figure 4b compares these results by showing the percentage difference in vertically averaged salinity between the ConvLSTM simulation and the reference simulation.
-Overall, the differences remain below 1%, except near a river mouth in the Gulf of Riga (22-24°E, 56.5-58.5°N for orientation), where the difference is approximately 1%.
-
-![](../src/figures/evaluation_MOM.png){#spatialBalticSea}
+![](images/paste-12.png)
 

### 2024-05-13 paper/paper.tex

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, boxrule=0pt, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable]}{\end{tcolorbox}}\fi
-
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+\section{Introduction}\label{introduction}
 
@@
 usually generated in two ways. First, river runoff as input for ocean
-models can be created using hyrdological models such as the Hydrological
+models can be created using hydrological models such as the Hydrological
 Discharge (HD) model (Hagemann et al., 2020). HD models calculates the
@@
 tool that uses underlying physical processes. A different approach would
-use data-based models that intergrate statistical correction, using the
+use data-based models that integrate statistical correction, using the
 land surface schemes of global or regional climate models.
 
-The releatively recent rise of machine learning (ML) models has been
-mostly explored for river runoff forecasting, as accurate runoff
-forecasting, especially over extended periods, is pivotal for effective
-water resources management (Fang et al., 2019; Tan et al., 2018; Yang et
-al., 2018). Common approaches employ artificial neural networks, support
-vector machines, adaptive neuro-fuzzy inference systems, and notably,
-Long Short-Term Memory (LSTM) neural networks that have gained traction
-for long-term hydrological forecasting due to their excellent
-performance (Humphrey et al 2016, Huang et al 2014, Ashrafi et al 2017,
-Yuan et al 2018, Xu et al 2021).
+The relatively recent emergence of machine learning (ML) models has
+offered different approaches to river runoff forecasting. Accurate
+runoff forecasting is crucial for effective water resources management,
+particularly over extended periods (W. Fang et al., 2019; Tan et al.,
+2018). Common approaches employ feed-forward artificial neural networks,
+support vector machines, adaptive neuro-fuzzy inference systems, and
+notably, Long Short-Term Memory (LSTM) neural networks that have gained
+traction for long-term hydrological forecasting due to their excellent
+performance (Ashrafi et al., 2017; L. Fang \& Shao, 2022; Huang et al.,
+2014; Humphrey et al., 2016; Kratzert et al., 2018; Liu et al., 2020).
 
-LSTM networks, an evolution of the classical Recurrent Neural Networks
-(RNNs), have shown stability and efficacy in sequence-to-sequence
-predictions, such as using climatic indices for rainfall estimation or
-long-term hydrological forecasting. However, a limitation of LSTMs is
-their inability to effectively capture two-dimensional structures, an
-area where Convolutional Neural Networks (CNNs) excel. Reconizing this
-limitation (\textbf{shi2015?}) proposed a convolutional LSTM (ConvLSTM)
-architectures, which combines the strengths of both LSTM and CNN. The
-ConvLSTM network has been proven useful for precipitation nowcasting
-(\textbf{shi2015?}) or for river runoff forecasting
-{[}(\textbf{ha2021?}); @niu2020{]}.
+LSTM networks, first introduced by Hochreiter \& Schmidhuber (1997) ,
+are an evolution of the classical Recurrent Neural Networks (RNNs).
+Their structure enables them to learn long-term dependencies while
+avoiding the vanishing or exploding gradient problem. They have shown
+stability and efficacy in sequence-to-sequence predictions. However, a
+limitation of LSTMs is their inability to effectively capture
+two-dimensional structures, an area where Convolutional Neural Networks
+(CNNs) excel. Recognizing this limitation SHI et al. (2015) proposed a
+Convolutional LSTM (ConvLSTM) architecture, which combines the strengths
+of both LSTM and CNN. The ConvLSTM network has been proven useful for
+spatio-temporal applications such as precipitation nowcasting (SHI et
+al., 2015), flood forecasting (Moishin et al., 2021), and river runoff
+forecasting (Ha et al., 2021; Zhu et al., 2023).
 
-In the following we will show that in absence of a fully functioning
-hydrogolocial model, that also uses a rather complex parametrization,
-ConvLSTM also represent a robust way to predict multiple rivers at once
-for any given period using only atmospheric forcing. In this work we use
-the Baltic Sea catchment to illustrate our approach, while in principal
-the methodology we propose is universally applicable across various
-geographic regions. The Baltic Sea serves as a challenging example due
-to its unique hydrological characteristics, being nearly decoupled from
-the open ocean (see Figure). As a consequence, the salinity of the
-Baltic Sea is driven to a large part by freshwater supply from rivers.
-More generally, the freshwater input into the Baltic Sea comes either as
-river runoff or a positive net precipitation (precipitation minus
-evaporation) over the sea surface. The net precipitation accounts for 11
-\% and the river input for 89 \% of the total freshwater input (Meier
-and Doescher, 2002). Modeling the Baltic Sea is therefore to a large
-part the result of the quality of the river input, that is used for the
-simulation. This makes the accurate modeling of river runoff especially
-critical for simulations pertaining to the Baltic Sea.
+In this work, we demonstrate that ConvLSTM networks are a reliable
+method for predicting multiple rivers simultaneously, using only
+atmospheric forcing, even in the absence of a fully functional
+hydrological model with a complex parameterization. We use the Baltic
+Sea catchment as an example to illustrate our approach. Although the
+methodology we propose is universally applicable across various
+geographic regions, the Baltic Sea presents a challenge due to its
+unique hydrological characteristics, being nearly decoupled from the
+open ocean (see Figure). As a consequence, the salinity of the Baltic
+Sea is driven to a large part by freshwater supply from rivers.
+Freshwater enters the Baltic Sea through river runoff or positive net
+precipitation (precipitation minus evaporation) over the sea surface.
+The net precipitation accounts for 11 \% and the river input for 89 \%
+of the total freshwater input (Meier \& Döscher, 2002). Modelling the
+Baltic Sea is therefore to a large part the result of the quality of the
+river input, that is used for the simulation. This makes the accurate
+modeling of river runoff especially critical for simulations pertaining
+to the Baltic Sea.
 
@@
 predict daily river runoff for 97 rivers across the Baltic Sea
-catchment.
+catchment. \textbf{\emph{mehr Fokus auf Neues?}}
 
-\hypertarget{methods}{%
-\section{Methods}\label{methods}}
+\section{Methods}\label{methods}
 
-\hypertarget{runoff-data-used-for-training}{%
 \subsection{Runoff data used for
-training}\label{runoff-data-used-for-training}}
+training}\label{runoff-data-used-for-training}
 
@@
 hindcast simulation that was forced by a regional downscaling of
-ERA-Interim (\textbf{dee2011?}) with RCA3 (\textbf{theross?}) and
-implemented into NEMO-Nordic (\textbf{hordoir2019?}) as a mass flux. For
-the periods before (1961 to 1978) and after (2012 to 2018) additional
-spatial temporal corrections have been applied to the runoff data, and
-have therefore been ignored. For more information see
-(\textbf{gröger2022?}) and references therein.
+ERA-Interim (Dee et al., 2011) with RCA3 ({``The rossby centre regional
+climate model RCA3,''} n.d.) and implemented into NEMO-Nordic (Hordoir
+et al., 2019) as a mass flux. For the periods before (1961 to 1978) and
+after (2012 to 2018) additional spatial temporal corrections have been
+applied to the runoff data, and have therefore been ignored. For more
+information see Gröger et al. (2022) and references therein.
 
-\hypertarget{atmospheric-forcing}{%
-\subsection{Atmospheric Forcing}\label{atmospheric-forcing}}
+\subsection{Atmospheric Forcing}\label{atmospheric-forcing}
 
@@
 Baltic Sea to the North Sea. The maximum depth was set at 264 meters. A
-more detailed description of the setup can be found in
-(\textbf{gröger2022?}).
+more detailed description of the setup can be found in (Gröger et al.,
+2022).
 
-\hypertarget{lstm-network}{%
-\subsection{LSTM network}\label{lstm-network}}
+\subsection{LSTM network}\label{lstm-network}
 
@@
 matrix-vector form as
-\begin{equation}\protect\hypertarget{eq-LSTM-gates}{}{
+\begin{equation}\phantomsection\label{eq-LSTM-gates}{
 \begin{aligned}
@@
 \end{aligned}
-}\label{eq-LSTM-gates}\end{equation}
+}\end{equation}
 
@@
 
-\begin{equation}\protect\hypertarget{eq-LSTM}{}{
+\begin{equation}\phantomsection\label{eq-LSTM}{
 \begin{aligned}
@@
 \end{aligned}
-}\label{eq-LSTM}\end{equation}
+}\end{equation}
 
@@
 
-\hypertarget{convlstm-network}{%
-\subsection{ConvLSTM network}\label{convlstm-network}}
+\subsection{ConvLSTM network}\label{convlstm-network}
 
@@
 
-{\centering \includegraphics{convLSTM.png}
+\centering{
+
+\includegraphics{convLSTM.png}
 
@@
 Learning Rate & 1e-3 with CosineAnnealing \\
-\bottomrule()
 \end{longtable}
@@
 
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
+\section{Discussion}\label{discussion}
 
@@
 
-\hypertarget{acknowledgments}{%
-\section{Acknowledgments}\label{acknowledgments}}
+\section{Acknowledgments}\label{acknowledgments}
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-fangExaminingApplicabilityDifferent2019}{}}%
+\bibitem[\citeproctext]{ref-ashrafi2017fully}
+Ashrafi, M., Chua, L. H. C., Quek, C., \& Qin, X. (2017). A fully-online
+neuro-fuzzy model for flow forecasting in basins with limited data.
+\emph{Journal of Hydrology}, \emph{545}, 424--435.
+
+\bibitem[\citeproctext]{ref-dee2011}
+Dee, D. P., Uppala, S. M., Simmons, A. J., Berrisford, P., Poli, P.,
+Kobayashi, S., et al. (2011). The ERA-Interim reanalysis: configuration
+and performance of the data assimilation system. \emph{Quarterly Journal
+of the Royal Meteorological Society}, \emph{137}(656), 553--597.
+\url{https://doi.org/10.1002/qj.828}
+
+\bibitem[\citeproctext]{ref-fang2022application}
+Fang, L., \& Shao, D. (2022). Application of long short-term memory
+(LSTM) on the prediction of rainfall-runoff in karst area.
+\emph{Frontiers in Physics}, \emph{9}, 790687.
+
+\bibitem[\citeproctext]{ref-fangExaminingApplicabilityDifferent2019}
 Fang, W., Huang, S., Ren, K., Huang, Q., Huang, G., Cheng, G., \& Li, K.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-hagemannHighResolutionDischarge2020}{}}%
+\bibitem[\citeproctext]{ref-gruxf6ger2022}
+Gröger, M., Placke, M., Meier, M., Börgel, F., Brunnabend, S.-E.,
+Dutheil, C., et al. (2022). \emph{The Baltic Sea model inter-comparison
+project BMIP {\textendash} a platform for model development, evaluation,
+and uncertainty assessment}. Retrieved from
+\url{https://gmd.copernicus.org/preprints/gmd-2022-160/}
+
+\bibitem[\citeproctext]{ref-ha2021}
+Ha, S., Liu, D., \& Mu, L. (2021). Prediction of Yangtze River
+streamflow based on deep learning neural network with El
+Niño{\textendash}Southern Oscillation. \emph{Scientific Reports},
+\emph{11}(1), 11738. \url{https://doi.org/10.1038/s41598-021-90964-3}
+
+\bibitem[\citeproctext]{ref-hagemannHighResolutionDischarge2020}
 Hagemann, S., Stacke, T., \& Ho-Hagemann, H. T. M. (2020). High
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-tanAdaptiveMiddleLongterm2018}{}}%
+\bibitem[\citeproctext]{ref-hochreiter1997long}
+Hochreiter, S., \& Schmidhuber, J. (1997). Long short-term memory.
+\emph{Neural Computation}, \emph{9}(8), 1735--1780.
+
+\bibitem[\citeproctext]{ref-hordoir2019}
+Hordoir, R., Axell, L., Höglund, A., Dieterich, C., Fransner, F.,
+Gröger, M., et al. (2019). Nemo-nordic 1.0: A NEMO-based ocean model for
+the baltic and north seas {\textendash} research and operational
+applications. \emph{Geoscientific Model Development}, \emph{12}(1),
+363--386. \url{https://doi.org/10.5194/gmd-12-363-2019}
+
+\bibitem[\citeproctext]{ref-huang2014monthly}
+Huang, S., Chang, J., Huang, Q., \& Chen, Y. (2014). Monthly streamflow
+prediction using modified EMD-based support vector machine.
+\emph{Journal of Hydrology}, \emph{511}, 764--775.
+
+\bibitem[\citeproctext]{ref-humphrey2016hybrid}
+Humphrey, G. B., Gibbs, M. S., Dandy, G. C., \& Maier, H. R. (2016). A
+hybrid approach to monthly streamflow forecasting: Integrating
+hydrological model outputs into a bayesian artificial neural network.
+\emph{Journal of Hydrology}, \emph{540}, 623--640.
+
+\bibitem[\citeproctext]{ref-kratzert2018rainfall}
+Kratzert, F., Klotz, D., Brenner, C., Schulz, K., \& Herrnegger, M.
+(2018). Rainfall--runoff modelling using long short-term memory (LSTM)
+networks. \emph{Hydrology and Earth System Sciences}, \emph{22}(11),
+6005--6022.
+
+\bibitem[\citeproctext]{ref-liu2020streamflow}
+Liu, D., Jiang, W., Mu, L., \& Wang, S. (2020). Streamflow prediction
+using deep learning neural network: Case study of yangtze river.
+\emph{IEEE Access}, \emph{8}, 90069--90086.
+
+\bibitem[\citeproctext]{ref-meier2002simulated}
+Meier, H. M., \& Döscher, R. (2002). Simulated water and heat cycles of
+the baltic sea using a 3D coupled atmosphere--ice--ocean model.
+\emph{Boreal Environment Research}, \emph{7}(4), 327.
+
+\bibitem[\citeproctext]{ref-moishin2021designing}
+Moishin, M., Deo, R. C., Prasad, R., Raj, N., \& Abdulla, S. (2021).
+Designing deep-based learning flood forecast model with ConvLSTM hybrid
+algorithm. \emph{IEEE Access}, \emph{9}, 50982--50993.
+
+\bibitem[\citeproctext]{ref-shiConvolutionalLSTMNetwork2015}
+SHI, X., Chen, Z., Wang, H., Yeung, D.-Y., Wong, W., \& WOO, W. (2015).
+Convolutional {LSTM Network}: {A Machine Learning Approach} for
+{Precipitation Nowcasting}. In \emph{Advances in {Neural Information
+Processing Systems}} (Vol. 28). Curran Associates, Inc. Retrieved from
+\url{https://proceedings.neurips.cc/paper/2015/hash/07563a3fe3bbe7e3ba84431ad9d055af-Abstract.html}
+
+\bibitem[\citeproctext]{ref-tanAdaptiveMiddleLongterm2018}
 Tan, Q.-F., Lei, X.-H., Wang, X., Wang, H., Wen, X., Ji, Y., \& Kang,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-yangDerivingOperatingRules2018}{}}%
-Yang, Z., Liu, P., Cheng, L., Wang, H., Ming, B., \& Gong, W. (2018).
-Deriving operating rules for a large-scale hydro-photovoltaic power
-system using implicit stochastic optimization. \emph{Journal of Cleaner
-Production}, \emph{195}, 562--572.
-\url{https://doi.org/10.1016/j.jclepro.2018.05.154}
+\bibitem[\citeproctext]{ref-theross}
+The rossby centre regional climate model RCA3: Model description and
+performance - tellus a: Dynamic meteorology and oceanography. (n.d.).
+Retrieved from
+\url{https://a.tellusjournals.se/articles/10.1111/j.1600-0870.2010.00478.x}
+
+\bibitem[\citeproctext]{ref-zhu2023spatiotemporal}
+Zhu, S., Wei, J., Zhang, H., Xu, Y., \& Qin, H. (2023). Spatiotemporal
+deep learning rainfall-runoff forecasting combined with remote sensing
+precipitation products in large scale basins. \emph{Journal of
+Hydrology}, \emph{616}, 128727.
 

### 2024-05-14 equations/equations.tex
Updated equations.tex to new reduced architecture

@@
 
-\subsection{Encoder}
+\subsection{ConvLSTM}
 
@@
 \begin{align}
-\left(\vec{H}^t[\tau], \vec{C}^t[\tau] \right) = L \left( \left\{ \vec{g}^t[\tau] \right\}, \vec{C}^t[\tau-1] \right) 
+\left(\vec{H}^t, \vec{C}^t \right) = L \left( \vec{X}^t[N_\tau], L \left( \vec{X}^t[N_\tau-1], \ldots L \left( \vec{X}^t[1], (\vec{H}^t[0], \vec{C}^t[0] \right) \ldots \right) \right) \ ,
 \end{align}
-
-\subsection{Decoder}
-
-Take last hidden state of encoder $H^t_h[x,y,N_\tau]$ and use it as input for the decoder.
-%
-Perform a single step (no iteration anymore!) with initial conditions $H^t_h[x,y,N_\tau]$ and $C^t_h[x,y,N_\tau]$, thus
-%
-\begin{align}
-\tilde{g}^t_h[x,y] & =  \tilde{\mathcal{M}}^{g}_{hh'} [\xi,\eta]\, H^t_{h'}[x-\xi, y-\eta,N_\tau] \\
-& +  \tilde{\mathcal{N}}^{g}_{hh'}[\xi,\eta] \, H^t_{h'}[x-\xi, y-\eta, N_\tau] + \tilde{\mathcal{B}}^g_{h}[x,y] \\
-& = \tilde{\mathcal{Q}}^{g}_{hh'} [\xi,\eta]\, H^t_{h'}[x-\xi, y-\eta,N_\tau] + \tilde{\mathcal{B}}^g_{h}[x,y]
-\end{align}
-%
-\begin{align}
-\tilde{I}^t_h[x,y] & = \sigma (\tilde{i}^t_h[x,y]) \\
-\tilde{F}^t_h[x,y] & = \sigma (\tilde{f}^t_h[x,y]) \\
-\tilde{O}^t_h[x,y] & = \sigma (\tilde{o}^t_h[x,y]) \\
-\tilde{C}^t_h[x,y] & = \tilde{F}^t_h[x,y,\tau] \, C^t_h[x,y,N_\tau] + \tilde{I}^t_h[x,y] \, \tanh (\tilde{c}^t_h[x,y]) \\
-\tilde{H}^t_h[x,y] & = \tilde{O}^t_h[x,y]\, \tanh(\tilde{C}^t_h[x,y])
-\end{align}
-%
-The action of the decoder can be compactly written as
-\begin{align}
-\left(\tilde{\vec{H}}^t, \tilde{\vec{C}}^t \right) = L \left( \left\{ \tilde{\vec{g}}^t \right\}, \vec{C}^t[N_\tau] \right) 
-\end{align}
+where $L\left(\vec{X}^t[\tau], \left(\vec{H}^t[\tau-1], \vec{C}^t[\tau-1] \right) \right)$ represents LSTM network equations and the initial conditions are chosen as $\vec{H}^t[0]=\vec{C}^t[0]=0$ and with the abbreviation for final output $\vec{H}^t[N_\tau] \equiv \vec{H}^t$ and $\vec{C}^t[N_\tau] \equiv \vec{C}^t$
 
@@
 \begin{align}
-R_r[t] = \mathcal{W}^{3}_{rb}\mathrm{ReLU}\left(\mathcal{W}^{2}_{ba}\mathrm{ReLU}\left(\mathcal{W}^{1}_{ah} [x,y] \tilde{H}^t_h[x,y]\right)\right)
+R_r[t] = \mathcal{W}^{3}_{rb}\mathrm{ReLU}\left(\mathcal{W}^{2}_{ba}\mathrm{ReLU}\left(\mathcal{W}^{1}_{ah} [x,y] H^t_h[x,y] + \mathcal{B}^1_a\right) + \mathcal{B}^2_b \right) + \mathcal{B}^3_r
 \end{align}
@@
 
-We can write our final result $R_r[t]$ for the $r$-th river as a function $M_r$ (for model) which itself is a nested function call of the $r$-th component of the fully connected layer $F$, the decoder $D$ and the encoder $E$ on a given input time sequence  $\{X^t_k[x,y,\tau]\}$, i.e.
+We can write our final result $R_r[t]$ for the $r$-th river as a function $M_r$ (for model) which itself is a nested function call of the $r$-th component of the fully connected layer $F_r$ and the ConvLSTM $K$ on a given input time sequence  $\{X^t_k[x,y,\tau]\}$, i.e.
 \begin{tcolorbox}[ams align]
-R_r[t] = M_r(\{X^t_k[x,y,\tau]\}) = F_r(D(E(\{X^t_k[x,y,\tau]\})))
+R_r[t] = M_r(\{X^t_k[x,y,\tau]\}) = F_r(K(\{X^t_k[x,y,\tau]\}))
 \end{tcolorbox}
@@
 \begin{tcolorbox}[ams align]
-\{\epsilon_{sh}[x,y]\} = E(\{X_k[x,y,\tau]\}) \\
-\{\tilde{H}_h[x,y]\} = D(\{\epsilon^s_h[x,y]\}) \\
-\phi_r = F_r(\{\tilde{H}_h[x,y]\})
+\{H_h[x,y]\} = K(\{X_k[x,y,\tau]\}) \\
+\phi_r = F_r(\{H_h[x,y]\})
 \end{tcolorbox}
-where the additional index $s$ in $\epsilon_{sh}[x,y]$ goes over the internal states of the encoder's output, i.e. 
-\begin{tcolorbox}[ams align]
-\epsilon_{1h}[x,y] = C_h[x,y,N_\tau], \quad \epsilon_{2h}[x,y] = H_h[x,y,N_\tau].
-\end{tcolorbox}
+
 %
 Note that the superscript $t$ is now omitted in the internal states since we calculate these quantities with respect to the reference sequence $\{X_k[x,y,\tau]\}$ that should not be connected to any specific time.
-%
-Note further that the decoder's output state $\tilde{H}_h[x,y]$ has already been defined above.
 
@@
 \begin{align}
-\omega_{rk}[x,y,\tau] = \del{ M_r(\{X_k[x,y,\tau]\})}{X_k[x,y,\tau]} = \del{\phi_r}{\tilde{H}_{h''}[x'',y'']} \del{\tilde{H}_{h''}[x'',y'']}{\epsilon_{s'h'}[x',y']} \del{\epsilon_{s'h'}[x',y']}{X_k[x,y,\tau]}
+\omega_{rk}[x,y,\tau] = \del{ M_r(\{X_k[x,y,\tau]\})}{X_k[x,y,\tau]} = \del{\phi_r}{H_{h'}[x',y']} \del{H_{h'}[x',y']}{X_k[x,y,\tau]}
 \end{align}
 %
-The next task is to unfold the dependence of $\epsilon_{sh}[x,y]$ on $\{X_k[x,y,\tau]\}$.
+The next task is to unfold the dependence of $H_{h}[x,y]$ on $\{X_k[x,y,\tau]\}$.
 %
-For that purpose we define the action of one ConvLSTM cell as function $K$. 
+For that purpose we define the action of one ConvLSTM cell as function $L$. 
 Then $\epsilon_{sh}[x,y]$ is the $N_\tau$-fold nested function
 \begin{align}
-\epsilon_{sh}[x,y] = E(\{X_k[x,y,\tau]\}) = K(\{X_k[x,y,N_\tau]\},K(\{X_k[x,y,N_\tau-1]\},\ldots, \{X_k[x,y,2]\}, K(\{X_k[x,y,1]\},0)\ldots ))
+H_{h}[x,y] = K(\{X_k[x,y,\tau]\}) = L(\{X_k[x,y,N_\tau]\},L(\{X_k[x,y,N_\tau-1]\},\ldots, \{X_k[x,y,2]\}, L(\{X_k[x,y,1]\},0)\ldots ))
 \end{align}
 %
-We can now define the output of the $K$-calls recursively for a fixed $\tau$ as 
+We can now define the output of the $L$-calls recursively for a fixed $\tau$ as 
 %
 \begin{align}
-\{\kappa_{sh}[x,y,\tau]\} = K\left(\{X_k[x,y,\tau]\}, \{\kappa_{sh}[x,y,\tau-1]\} \right)
+\{\kappa_{sh}[x,y,\tau]\} = L\left(\{X_k[x,y,\tau]\}, \{\kappa_{sh}[x,y,\tau-1]\} \right)
 \end{align}
-with $\kappa_{sh}[x,y,0]=0$ and $\kappa_{sh}[x,y,N_\tau]=\epsilon_{sh}[x,y]$.
+where the additional index $s$ labels the two outputs of the $L$-function, i.e. $C_{h}[x,y,\tau]=\kappa_{1h}[x,y,\tau]$ and $H_{h}[x,y,\tau]=\kappa_{2h}[x,y,\tau]$.
+The initial condition is $\kappa_{sh}[x,y,0]=0$ and the endpoint is by construction $\kappa_{2h}[x,y,N_\tau]=H_{h}[x,y]$.
 %
-Now we can use the chain rule to calculate the derivative of $\epsilon_{s'h'}[x',y']$ with respect to $X_k[x,y,\tau]$ for a specific $\tau$
+Now we can use the chain rule to calculate for a specific $\tau$
 %
 \begin{align}
- \del{\epsilon_{s'h'}[x',y']}{X_k[x,y,\tau]} = \left( \prod_{j=\tau}^{N_\tau-1} \del{\kappa_{s_{j+1}h_{j+1}}[x_{j+1},y_{j+1},j+1]}{\kappa_{s_{j}h_{j}}[x_{j},y_{j},j]} \right) \del{\kappa_{s_{\tau}h_{\tau}}[x_{\tau},y_{\tau},\tau]}{X_k[x,y,\tau]}
+ \del{H_{h'}[x',y']}{X_k[x,y,\tau]} = \left.
+\del{\kappa_{s' h'}[x',y',N_\tau]}{X_k[x,y,\tau]} \right |_{s'=2} = \left.
+\left( \prod_{j=\tau}^{N_\tau-1} \del{\kappa_{s_{j+1}h_{j+1}}[x_{j+1},y_{j+1},j+1]}{\kappa_{s_{j}h_{j}}[x_{j},y_{j},j]} \right) \right |_{s_{N_\tau} = s' = 2} \del{\kappa_{s_{\tau}h_{\tau}}[x_{\tau},y_{\tau},\tau]}{X_k[x,y,\tau]}
 \end{align}
@@
 \begin{tcolorbox}[ams align]
-\omega_{rk}[x,y,\tau] = \del{\phi_r}{\tilde{H}_{h''}[x'',y'']} \del{\tilde{H}_{h''}[x'',y'']}{\epsilon_{s_{N_\tau} h_{N_\tau}}[x_{N_\tau},y_{N_\tau}]} \left( \prod_{j=\tau}^{N_\tau-1} \del{\kappa_{s_{j+1}h_{j+1}}[x_{j+1},y_{j+1},j+1]}{\kappa_{s_{j}h_{j}}[x_{j},y_{j},j]} \right) \del{\kappa_{s_{\tau}h_{\tau}}[x_{\tau},y_{\tau},\tau]}{X_k[x,y,\tau]}
+\omega_{rk}[x,y,\tau] = \del{\phi_r}{H_{h'}[x',y']} \left.\left( \prod_{j=\tau}^{N_\tau-1} \del{\kappa_{s_{j+1}h_{j+1}}[x_{j+1},y_{j+1},j+1]}{\kappa_{s_{j}h_{j}}[x_{j},y_{j},j]} \right) \right |_{s_{N_\tau} = s' = 2} \del{\kappa_{s_{\tau}h_{\tau}}[x_{\tau},y_{\tau},\tau]}{X_k[x,y,\tau]}
 \end{tcolorbox}
@@
 \begin{align}
-\omega_{rk}[x,y,\tau] = \del{\phi_r}{\tilde{H}_{h''}[x'',y'']} \del{\tilde{H}_{h''}[x'',y'']}{\epsilon_{s_{N_\tau} h_{N_\tau}}[x_{N_\tau},y_{N_\tau}]} \left( \prod_{j=\tau}^{N_\tau-1} \Lambda_{s_{j+1}s_j,h_{j+1}h_j}[x_{j+1}, x_j, y_{j+1}, y_j, j+1] \right)
+\omega_{rk}[x,y,\tau] = \del{\phi_r}{\tilde{H}_{h'}[x',y']} \left.
+\left( \prod_{j=\tau}^{N_\tau-1} \Lambda_{s_{j+1}s_j,h_{j+1}h_j}[x_{j+1}, x_j, y_{j+1}, y_j, j+1] \right) \right |_{s_{N_\tau} = s' = 2}
 \lambda_{s_\tau h_\tau,k}[x_\tau, x, y_\tau, y, \tau]
@@
 
-The next task is to evaluate the derivative of the decoder with respect to the encoders output.
-%
-Since the decoder is itself a ConvLSTM cell with input from encoders last hidden state we can write
-\begin{align}
-\{\tilde{H}_{h}[x,y]\} = \tilde{K}_2 \left( \{ \epsilon_{2 h}[x,y]\}, \{ \epsilon_{s h}[x,y]\} \right)
-\end{align}
-%
-Following the same line of reasoning as above we obtain for the derivatives
-%
-\begin{tcolorbox}[ams align]
-\del{\tilde{H}_{h'}[x',y']}{\epsilon_{1h}[x,y]} = \del{\tilde{H}_{h'}[x',y']}{C_h[x,y,N_\tau]}  & = \frac{\tilde{O}_{h'}[x',y'] \tilde{F}_{h'}[x',y']}{\cosh^2(\tilde{C}_{h'}[x',y'])}  \delta_{hh'}\delta_{xx'}\delta_{yy'} \\
-\del{\tilde{H}_{h'}[x',y']}{\epsilon_{2h}[x,y]} = \del{\tilde{H}_{h'}[x',y']}{H_h[x,y,N_\tau]}     & = \dot{\sigma}(\tilde{o}_{h'}[x',y'])\tanh(\tilde{C}_{h'}[x',y']) \tilde{\mathcal{Q}}^{o}_{h'h}[x'-x,y'-y] \\
-& + \frac{\sigma(\tilde{o}_{h'}[x',y'])}{\cosh^2(\tilde{C}_{h'}[x',y'])} \tilde{\Lambda}_{1,2,h'h}[x', x, y', y]
-\end{tcolorbox}
-%
-with
-\begin{align}
-\tilde{\Lambda}_{1,2,h'h}[x', x, y', y] & = \dot{\sigma}(\tilde{f}_{h'}[x',y']) C_{h'}[x',y',N_\tau] \tilde{\mathcal{Q}}^{f}_{h'h}[x'-x,y'-y]\\
-& + \dot{\sigma}(\tilde{i}_{h'}[x',y']) \tanh(\tilde{c}_{h'}[x',y'])\tilde{\mathcal{Q}}^{i}_{h'h}[x'-x,y'-y]\\
-& +  \frac{\sigma(\tilde{i}_{h'}[x',y'])}{\cosh^2(\tilde{c}_{h'}[x',y'])} \tilde{\mathcal{Q}}^{c}_{h'h}[x'-x,y'-y]
-\end{align}
-%
-Now we can also rename these derivatives as
-\begin{tcolorbox}[ams align]
-\tilde{\chi}_{s_{N_\tau}h_{N_\tau}, h''}[x'', x_{N_\tau}, y'', y_{N_\tau}] = \del{\tilde{H}_{h''}[x'',y'']}{\epsilon_{s_{N_\tau} h_{N_\tau}}[x_{N_\tau},y_{N_\tau}]}
-\end{tcolorbox}
 %
@@
 \begin{align}
-\omega_{rk}[x,y,\tau] = \del{\phi_r}{\tilde{H}_{h''}[x'',y'']} \tilde{\chi}_{s_{N_\tau}h_{N_\tau}, h''}[x'', x_{N_\tau}, y'', y_{N_\tau}] \chi_{s_{N_\tau} h_{N_\tau},k}[x_{N_\tau}, x, y_{N_\tau}, y, N_\tau, \tau]
+\omega_{rk}[x,y,\tau] = \del{\phi_r}{H_{h'}[x',y']} \chi_{2,h',k}[x', x, y', y, N_\tau, \tau]
 \end{align}
 %
-If we define the product 
+If we define the symbol 
 %
 \begin{tcolorbox}[ams align]
-\Xi_{h''k}[x'', x, y'', y, \tau] = \tilde{\chi}_{s_{N_\tau}h_{N_\tau}, h''}[x'', x_{N_\tau}, y'', y_{N_\tau}] \chi_{s_{N_\tau} h_{N_\tau},k}[x_{N_\tau}, x, y_{N_\tau}, y, N_\tau, \tau]
+\Xi_{h'k}[x', x, y', y, \tau] =  \chi_{2,h',k}[x', x, y', y, N_\tau, \tau]
 \end{tcolorbox}
@@
 \begin{tcolorbox}[ams align]
-\omega_{rk}[x,y,\tau] = \del{\phi_r}{\tilde{H}_{h''}[x'',y'']} \Xi_{h''k}[x'', x, y'', y, \tau]
+\omega_{rk}[x,y,\tau] = \del{\phi_r}{H_{h'}[x',y']} \Xi_{h'k}[x', x, y', y, \tau]
 \end{tcolorbox}
@@
 \begin{tcolorbox}[ams align]
-\rho^1_a & = \mathcal{W}^{1}_{ah} [x,y] \tilde{H}_h[x,y] \\
-\rho^2_b & = \mathcal{W}^{2}_{ba} \mathrm{ReLU}\left(\rho^1_a\right) \\
-\phi_r & = \mathcal{W}^{3}_{rb} \mathrm{ReLU}\left(\rho^2_b \right)
+\rho^1_a & = \mathcal{W}^{1}_{ah} [x,y] H_h[x,y] + \mathcal{B}^1_a \\
+\rho^2_b & = \mathcal{W}^{2}_{ba} \mathrm{ReLU}\left(\rho^1_a\right) + \mathcal{B}^2_b \\
+\phi_r & = \mathcal{W}^{3}_{rb} \mathrm{ReLU}\left(\rho^2_b \right) + \mathcal{B}^3_r
 \end{tcolorbox}
@@
 \begin{align}
-\del{\phi_r}{\tilde{H}_{h''}[x'',y'']} = \del{\phi_r}{\rho^2_{b'}} \del{\rho^2_{b'}}{\rho^1_{a'}} \del{\rho^1_{a'}}{\tilde{H}_{h''}[x'',y'']}
+\del{\phi_r}{H_{h'}[x',y']} = \del{\phi_r}{\rho^2_{b'}} \del{\rho^2_{b'}}{\rho^1_{a'}} \del{\rho^1_{a'}}{\tilde{H}_{h''}[x'',y'']}
 \end{align}
@@
  \del{\rho^2_{b'}}{\rho^1_{a'}} & =  \mathcal{W}^{2}_{b'a'} \theta \left( \rho^1_{a'} \right) \\
- \del{\rho^1_{a'}}{\tilde{H}_{h''}[x'',y'']} & = \mathcal{W}^{1}_{a'h''} [x'',y'']
+ \del{\rho^1_{a'}}{H_{h'}[x',y']} & = \mathcal{W}^{1}_{a'h'} [x',y']
 \end{align}
@@
 \begin{tcolorbox}[ams align]
-\del{\phi_r}{\tilde{H}_{h''}[x'',y'']} = \mathcal{W}^{3}_{rb'} \theta \left( \rho^2_{b'} \right) \cdot  \mathcal{W}^{2}_{b'a'} \theta \left( \rho^1_{a'} \right) \cdot \mathcal{W}^{1}_{a'h''} [x'',y''] = \Omega_{r h''}[x'',y'']
+\del{\phi_r}{H_{h'}[x',y']} = \mathcal{W}^{3}_{rb'} \theta \left( \rho^2_{b'} \right) \cdot  \mathcal{W}^{2}_{b'a'} \theta \left( \rho^1_{a'} \right) \cdot \mathcal{W}^{1}_{a'h'} [x',y'] = \Omega_{r h'}[x',y']
 \end{tcolorbox}
@@
 \begin{tcolorbox}[ams align]
-\omega_{rk}[x,y,\tau] = \Omega_{r h''}[x'',y''] \cdot \Xi_{h''k}[x'', x, y'', y, \tau]
+\omega_{rk}[x,y,\tau] = \Omega_{r h'}[x',y'] \cdot \Xi_{h'k}[x', x, y', y, \tau]
 \end{tcolorbox}

### 2024-05-14 equations/src/ndarray.c
Added small test code for linear Taylor expansion

@@ -1,0 +1,78 @@
+/*  Note:  For demonstration purposes only.  Depending on needs, other types
+    might be used for indices and sizes, and the array type might be wrapped
+    in an opaque struct rather than exposed as "int *".
+*/
+#include "ndarray.h"
+
+//  Create an array with N dimensions with sizes specified in D.
+double *CreateArray(int N, int D[])
+{
+    //  Calculate size needed.
+    long int s = sizeof(double);
+    for (int n = 0; n < N; ++n)
+        s *= D[n];
+
+    //  Allocate space.
+    return (double*) malloc(s);
+}
+
+
+
+//  Create an array with N dimensions with sizes specified in D.
+long int Len(int N, int D[])
+{
+    //  Calculate size needed.
+    long int s = 1;
+    for (int n = 0; n < N; ++n)
+        s *= D[n];
+
+    //  Allocate space.
+    return s;
+}
+
+void Fill(double value, double *A, int N, int D[])
+{
+    long int L = Len(N,D);
+
+    for (int l = 0; l < L; l++){
+        A[l] = value;
+    }
+}
+
+/*  Return a pointer to an element in an N-dimensional A array with sizes
+    specified in D and indices to the particular element specified in I.
+*/
+double *Element(double *A, int N, int D[], int I[])
+{
+    //  Handle degenerate case.
+    if (N == 0)
+        return A;
+
+    //  Map N-dimensional indices to one dimension.
+    long int index = I[0];
+    for (int n = 1; n < N; ++n)
+        index = index * D[n] + I[n];
+
+    //  Return address of element.
+    return &A[index];
+}
+
+
+extern void dgemm_(char *transa, char *transb, int *m, int *n, int *k, double
+        *alpha, double *a, int *lda, double *b, int *ldb, double *beta, double *c,
+        int *ldc );
+
+#define mm_success 0
+#define mm_failure 1
+
+// performs C = beta*C + alpha* A.B -> ac = br ! 
+int matrix_multiplication(double alpha, double* A, int ar, int ac, double* B, int br, int bc, double beta, double* C)
+{
+    if (ac != br){
+        printf("ERROR! %d != %d\n", ac, br);
+        return mm_failure;
+    } 
+    
+    dgemm_("n", "n", &bc, &ar, &br, &alpha, B, &bc, A, &br, &beta, C, &bc);
+    return mm_success;
+}

### 2024-05-14 equations/src/ndarray.h

@@ -1,0 +1,12 @@
+#include <stdio.h>
+#include <stdlib.h>
+
+double *CreateArray(int N, int D[]);
+
+long int Len(int N, int D[]);
+
+void Fill(double value, double *A, int N, int D[]);
+
+double *Element(double *A, int N, int D[], int I[]);
+
+int matrix_multiplication(double alpha, double* A, int ar, int ac, double* B, int br, int bc, double beta, double* C);

### 2024-05-14 paper/paper.qmd
Reworked theory part

--- 2024-04-25 paper/paper.qmd
@@
 
-## Methods
+## Implemented model architecture
 
-### Runoff data used for training
+### The main idea {#sec-main_idea}
 
-The runoff data covering the period 1979 to 2011 is based on an E-HYPE hindcast simulation that was forced by a regional downscaling of ERA-Interim [@dee2011] with RCA3 [@theross] and implemented into NEMO-Nordic [@hordoir2019] as a mass flux.
-For the periods before (1961 to 1978) and after (2012 to 2018) additional spatial temporal corrections have been applied to the runoff data, and have therefore been ignored.
-For more information see @gröger2022 and references therein.
+We assume that the runoff for a specific point in time of all $N_r$ considered rivers, collected in the vector $\vec{R}^t \in \mathbb{R}^{N_r}$, can be accurately approximated by a functional $\vec{M}(\{X^t_k[x,y,\tau]\})$ of $k=1,\ldots N_k$ atmospheric fields $X^t_k[x,y,\tau]$ that are known for the past $\tau=1,\ldots N_\tau$ time instances, i.e.
 
-### Atmospheric Forcing
+$$
+\begin{aligned}
+\vec{R}^t = \vec{M}(\{X^t_k[x,y,\tau]\}) \ .
+\end{aligned}
+$$
 
-The UERRA-HARMONIE regional reanalysis dataset was developed as part of the FP7 UERRA project (Uncertainties in Ensembles of Regional Re-Analyses, [http://www.uerra.eu/)](http://www.uerra.eu/),).
-The UERRA-HARMONIE system represents a comprehensive, high-resolution reanalysis covering a wide range of essential climate variables.
-This dataset encompasses data on air temperature, pressure, humidity, wind speed and direction, cloud cover, precipitation, albedo, surface heat fluxes, and radiation fluxes from January 1961 to July 2019.
-With a horizontal resolution of 11 km and analyses conducted at 00 UTC, 06 UTC, 12 UTC, and 18 UTC, it also provides hourly resolution forecast model data.
-UERRA-HARMONIE is accessible through the Copernicus Climate Data Store (CDS, <https://cds.climate.copernicus.eu/#!/home),> initially produced during the UERRA project and later transitioned to the Copernicus Climate Change Service (C3S, <https://climate.copernicus.eu/copernicus-regionalreanalysis-europe).>
+The fields are evaluated on a spatial domain $x=1,\ldots N_x$ and $y=1,\ldots N_y$ that is sufficiently large to capture all significant contributions to the modelled runoff.
 
-### Ocean Model
+We propose that this functional can be adequately represented by a combination of a ConvLSTM with a subsequent FC network as it is visualized in @fig-ConvLSTM_FC and described in detail in the following.
+In order to provide an overview, we will go through the main ingredients of this architecture one-by-one.
 
-To simulate the Baltic Sea, we use a coupled 3-dimensional ocean model Baltic Sea, called the Modular Ocean Model (MOM).
-This model uses a finite-difference method to solve the full set of primitive equations to calculate the motion of water and the transport of heat and salt.
-The K-profile parameterization (KPP) was used as turbulence closure scheme.
-The model's western boundary opens into the Skagerrak and connects the Baltic Sea to the North Sea.
-The maximum depth was set at 264 meters.
-A more detailed description of the setup can be found in [@gröger2022].
+![](./ConvLSTM_FC.png){#fig-ConvLSTM_FC fig-align="center"}
 
-### LSTM network
+### ConvLSTM network
+
+#### The LSTM approach {#sec-LSTM}
 
@@
 
-![Inner structure of a Long Short-Term Memory Cell](draw_lstm.png){#fig-lstm fig-align="center"}
+![Inner structure of a Long Short-Term Memory Cell](./LSTM.png){#fig-lstm fig-align="center"}
 
 This performance in modeling long-range dependencies has been validated in various studies.
-The key component of LSTM's innovation is its memory cell, $\vec{C}^t[\tau] = (C^t_h[\tau]) \in \mathbb{R}^{N_h}$, which stores state information, also referred to as long-term memory.
-The index $\tau \in [1, N_\tau]$ corresponds to the $\tau$-th element in the input sequence.
-The cell state is a vector consists of elements, each associated with one of the $N_h$ hidden layers, labeled by $h$.
-This vector is determined through several self-parameterized gates, all in the same vector space as $\vec{C}^t[\tau]$.
-The forget gate $\vec{F}^t[\tau]$ defines the percentage of the previous long-term memory status $\vec{C}^t[\tau-1]$ that should be retained stored​.
-The input gate $\vec{I}^t[\tau]$ decides how much of the input is added to the the long-term memory, forming the updated cell state.
-The so-called output gate $\vec{O}^t[\tau]$ then determines how much of the latest cell output, $\vec{C}^t[\tau]$, is propagated to the final state, $\vec{H}^t[\tau]$ representing the updated short-term memory of the hidden state.
+The key component of the LSTM's innovation is its memory cell, $\vec{C}^t[\tau] = (C^t_h[\tau]) \in \mathbb{R}^{N_h}$, which stores state information, also referred to as long-term memory.
+This state information complements the the so-called hidden state $\vec{H}^t[\tau] = (H^t_h[\tau]) \in \mathbb{R}^{N_h}$ vector, which is also known from simpler neural network architectures.
+In case of the LSTM, the hidden state vector plays the role of the short-term meory.
+The cell state and the hidden state are vectors, where each element is associated with one of the so-called $N_h$ hidden layers, labeled by $h$.
+These two state vectors are determined through several self-parameterized gates, all in the same vector space as $\vec{C}^t[\tau]$.
+
+In particular, the forget gate $\vec{F}^t[\tau]$ defines the portion of the previous (long-term memory) cell state $\vec{C}^t[\tau-1]$ that should be kept​.
+The input gate $\vec{I}^t[\tau]$ controls the contribution of the current input that is used to update the long-term memory,$\vec{C}^t[\tau]$.
+The output gate $\vec{O}^t[\tau]$ then determines how much of this updated long-term memory contributes to the new (short-term memory) hidden state, $\vec{H}^t[\tau]$.
 
 For a fixed point $\tau$ in the sequence, the action of a LSTM cell, i.e.
-the connection between the input $\vec{X}^t[\tau]$, the various gates and the state vectors, is given as follows.
+the connection between the input $\vec{X}^t[\tau]$, the various gates and the state vectors, is mathematically given as follows.
+
 First, the elements of the input sequence together with the hidden state are mapped onto auxiliary gate vectors, collectively denoted by $\vec{g}^t[\tau] = (g^t_h[\tau]) \in \mathbb{R}^{N_h}$, via $$
@@
 where $\sigma$ denotes the logisitc sigmoid function, $\tanh$ is the hyperbolic tangent and the $\circ$ stands for the Hadamard product (all applied in an element-wise fashion to the vectors).
-In the last two equations the role of the input, forget and output gates as described above become apparent.
+In the last two equations the role of the input, forget and output gates as described above becomes apparent.
 
@@
 $$
-\left(\vec{H}^t[N_\tau], \vec{C}^t[N_\tau] \right) = L \left( \vec{X}^t[N_\tau], L \left( \vec{X}^t[N_\tau-1], \ldots L \left( \vec{X}^t[1], (\vec{H}^t[0], \vec{C}^t[0] \right) \ldots \right) \right) \ ,
-$$
+\left(\vec{H}^t[N_\tau], \vec{C}^t[N_\tau] \right) = L \left( \vec{X}^t[N_\tau], L \left( \vec{X}^t[N_\tau-1], \ldots L \left( \vec{X}^t[1], (\vec{H}^t[0], \vec{C}^t[0]) \right) \ldots \right) \right) \ ,
+$$ {#eq-L_nested}
 
-where $L$ represents Eqs. (@eq-LSTM-gates, @eq-LSTM) and the initial conditions are usually chosen as $\vec{H}^t[0]=\vec{C}^t[0]=0$.
+where $L\left(\vec{X}^t[\tau], \left(\vec{H}^t[\tau-1], \vec{C}^t[\tau-1] \right) \right)$ represents @eq-LSTM-gates and @eq-LSTM and the initial conditions are chosen as $\vec{H}^t[0]=\vec{C}^t[0]=0$.
 
 The final output $\vec{H}^t[N_\tau]$ and $\vec{C}^t[N_\tau]$ encode information on the full input sequence ending at time $t$.
-This information has to be decoded to obtain usueful information via an appropriate subsequent layer, as it is described in the Sec.
-@XXX.
+This information has to be decoded to obtain useful information via an appropriate subsequent layer, as it is described in the @sec-FC_layer.
 
@@
 
-### ConvLSTM network
+#### Combining LSTM with spatial convolution
 
-Although the plain LSTM has high performance in handling temporal sequences of point-like quantities it is not designed to recognize spatial features in seuqences of two-dimensional maps.
-To adress this limitation we use a convLSTM architecture.
+Although the plain LSTM has high performance in handling temporal sequences of point-like quantities it is not designed to recognize spatial features in seuqences of, e.g., two-dimensional maps.
+To adress this limitation we employ a ConvLSTM architecture as described in the following.
 
-In this kind of network the elements of the input sequence are given as (kind of) vector fields $\vec{X}^t[\tau] = (X^t_k[x,y,\tau]) \in \mathbb{R}^{N_k \times (N_x \times N_y)}$, where $x\in[1, N_x]$ and $y \in [1, N_y]$ run over the horizontal and vertical dimensions of the map, respectively.
+In this kind of network the elements of the input sequence are given as spatially varying fields $\vec{X}^t[\tau] = (X^t_k[x,y,\tau]) \in \mathbb{R}^{N_k \times (N_x \times N_y)}$, where $x\in[1, N_x]$ and $y \in [1, N_y]$ run over the horizontal and vertical dimensions of the map, respectively.
 In order to enable the "learning" of spatial patterns, the free parameters of the network are replaced by two-dimenmsional convolutional kernels $\pmb{\mathcal{M}}^{g} = (\mathcal{M}^{g}_{hk}[\xi, \eta]) \in \mathbb{R}^{(N_h \times N_k)\times (N_\xi \times N_\eta)}$ and $\pmb{\mathcal{N}}^{g} = (\mathcal{N}^{g}_{hh'}[\xi, \eta]) \in \mathbb{R}^{(N_h \times N_h)\times (N_\xi \times N_\eta)}$.
@@
 \begin{aligned}
-g^t_h[x,y,\tau] & =  \mathcal{M}^{g}_{hk} [\xi,\eta]\, X^t_k[x-\xi, y-\eta, \tau]  +  \mathcal{N}^{g}_{hh'}[\xi,\eta] \, H^t_{h'}[x-\xi, y-\eta, \tau-1] + \mathcal{B}^g_{h}[x,y] \ .
+g^t_h[x,y,\tau] & =  \mathcal{M}^{g}_{hk} [\xi,\eta]\, X^t_k[x-\xi, y-\eta, \tau]  +  \mathcal{N}^{g}_{hh'}[\xi,\eta] \, H^t_{h'}[x-\xi, y-\eta, \tau-1] + \mathcal{B}^g_{h}\ .
 \end{aligned}
@@
 
-It becomes immediately apparent that in case of the convLSTM, the bias, gate and state vectors must become vector fields ($\in \mathbb{R}^{N_h \times (N_x \times N_y)}$) as well.
-We can write this mapping in the same fashion as Eq.
-(@eq-LSTM-gates) but with replacing the normal matrix-vector multiplication by a convolution (denoted with $\ast$), i.e.\
+again with Einstein's convention imposed.
+
+It becomes immediately apparent that in case of the ConvLSTM, gate and state vectors must become vector fields ($\in \mathbb{R}^{N_h \times (N_x \times N_y)}$) as well.
+We can write this mapping in the same fashion as @eq-LSTM-gates but with replacing the normal matrix-vector multiplication by a convolution (denoted with $\ast$), i.e.\
 $$
@@
 
-The susebsequent processing of the $\vec{g}^t[\tau]$ remains symbolically the same as presented in Eq.
-(@eq-LSTM) but with all appearing quantities meaning standing now for vector fields instead of simple vectors.
+The susebsequent processing of the $\vec{g}^t[\tau]$ remains symbolically the same as presented in @eq-LSTM but with all appearing quantities meaning standing now for vector fields instead of simple vectors.
 
@@
 
-### Implemented model architecture
+### Fully connected layer {#sec-FC_layer}
 
-The ConvLSTM architectures uses and encoder/decoder structure as discussed in TODO.
-To predict all 97 rivers entering the Baltic Sea, we flatten the output and use fully connected layers to map onto the individual rivers outputs.
+As stated in Sec.
+@sec-LSTM, the final output $\vec{H}^t[N_\tau]$ and $\vec{C}^t[N_\tau]$ of the ConvLSTM encode information on the full input sequence.
+In order to contract this information to the runoff vector $\vec{R}^t$ representing the $N_r$ rivers, we propose to subject the final short-term memory (i.e.
+the hidden state $\vec{H}^t[N_\tau]$) to an additional FC network.
 
-An overview of the model structure is given below
+In particular, the dimensionality of the vector field $\vec{H}^t[N_\tau]$ is sequentially reduced by three nested FC layers, each connected to the other by the Rectified Linear Unit (ReLU), see @fig-ConvLSTM_FC.
+Integrating over artificial degrees of freedom in a step-wise fashion has turned out to be benificial (@XXX).
 
-![Schematic structure of the convLSTM implementation for river runoff forecasting.](convLSTM.png){#fig-baltNet}
+Spelled out in mathematics the runoff of the $r$-th river is then obtained via (using Einstein's convention)
+
+$$
+\begin{aligned}
+R_r^t = \mathcal{W}^{3}_{rb}\mathrm{ReLU}\left(\mathcal{W}^{2}_{ba}\mathrm{ReLU}\left(\mathcal{W}^{1}_{ah} [x,y] H^t_h[x,y, N_\tau] + \mathcal{B}^1_a\right) + \mathcal{B}^2_b \right) + \mathcal{B}^3_r \ ,
+\end{aligned}
+$$
+
+where $a=1,\ldots N_a$ and $b=1,\ldots N_b$ and $N_h\cdot N_x\cdot N_y > N_a > N_b > N_r$ in order to achieve the aforementioned step-by-step reduction of dimensionality.
+The weights $\mathcal{W}$ and biases $\mathcal{B}$ stand for parameters that are optimized during the training of the network.
+
+In matrix-vector notation this can compactified to $$
+\begin{aligned}
+\vec{R}^t = \pmb{\mathcal{W}}^{3}\mathrm{ReLU}\left(\pmb{\mathcal{W}}^{2}\mathrm{ReLU}\left(\pmb{\mathcal{W}}^{1} \vec{H}^t[N_\tau] + \vec{\mathcal{B}}^1\right) + \vec{\mathcal{B}}^2 \right) + \vec{\mathcal{B}}^3 \ .
+\end{aligned}
+$$
+
+Combining this equation with @eq-L_nested provides finally an explicit formula for the initial assumption of modelling the runoff for time $t$ as a functional of a sequence of atmospheric fields, i.e.
+
+$$
+\begin{aligned}
+\vec{R}^t & = \vec{M}(\{X^t_k[x,y,\tau]\}) \\
+& = \pmb{\mathcal{W}}^{3}\mathrm{ReLU}\left(\pmb{\mathcal{W}}^{2}\mathrm{ReLU}\left(\pmb{\mathcal{W}}^{1} \vec{L}_H \left( \vec{X}^t[N_\tau], L \left( \vec{X}^t[N_\tau-1], \ldots L \left( \vec{X}^t[1], (0, 0) \right) \ldots \right) \right) + \vec{\mathcal{B}}^1\right) + \vec{\mathcal{B}}^2 \right) + \vec{\mathcal{B}}^3 \ ,
+\end{aligned}
+$$ where the $\vec{L}_H$ means that only the hidden state vector of the final ConvLSTM call is forwarded to the FC layer.
+In oder to make this model accurate the hyperparameters have to be chosen reasonable and the free parameters are to be adjusted accordingly.
+
+## Technical details
+
+### Runoff data used for training
+
+The runoff data covering the period 1979 to 2011 is based on an E-HYPE hindcast simulation that was forced by a regional downscaling of ERA-Interim [@dee2011] with RCA3 [@theross] and implemented into NEMO-Nordic [@hordoir2019] as a mass flux.
+For the periods before (1961 to 1978) and after (2012 to 2018) additional spatial temporal corrections have been applied to the runoff data, and have therefore been ignored.
+For more information see @gröger2022 and references therein.
+
+### Atmospheric Forcing
+
+The UERRA-HARMONIE regional reanalysis dataset was developed as part of the FP7 UERRA project (Uncertainties in Ensembles of Regional Re-Analyses, [http://www.uerra.eu/)](http://www.uerra.eu/),).
+The UERRA-HARMONIE system represents a comprehensive, high-resolution reanalysis covering a wide range of essential climate variables.
+This dataset encompasses data on air temperature, pressure, humidity, wind speed and direction, cloud cover, precipitation, albedo, surface heat fluxes, and radiation fluxes from January 1961 to July 2019.
+With a horizontal resolution of 11 km and analyses conducted at 00 UTC, 06 UTC, 12 UTC, and 18 UTC, it also provides hourly resolution forecast model data.
+UERRA-HARMONIE is accessible through the Copernicus Climate Data Store (CDS, <https://cds.climate.copernicus.eu/#!/home),> initially produced during the UERRA project and later transitioned to the Copernicus Climate Change Service (C3S, <https://climate.copernicus.eu/copernicus-regionalreanalysis-europe).>
+
+### Ocean Model
+
+To simulate the Baltic Sea, we use a coupled 3-dimensional ocean model Baltic Sea, called the Modular Ocean Model (MOM).
+This model uses a finite-difference method to solve the full set of primitive equations to calculate the motion of water and the transport of heat and salt.
+The K-profile parameterization (KPP) was used as turbulence closure scheme.
+The model's western boundary opens into the Skagerrak and connects the Baltic Sea to the North Sea.
+The maximum depth was set at 264 meters.
+A more detailed description of the setup can be found in [@gröger2022].
+
+### Neural network hyper parameters
+
+![Schematic structure of the ConvLSTM implementation for river runoff forecasting.](ConvLSTM.png){#fig-baltNet}
 
@@
 
-![](../src/figures/evaluation_MOM.png){#spatialBalticSea}
+![](../src/figures/evaluation_MOM.png)
 

### 2024-05-14 paper/paper.tex
Reworked theory part

@@
 
-\section{Methods}\label{methods}
+\section{Implemented model
+architecture}\label{implemented-model-architecture}
 
-\subsection{Runoff data used for
-training}\label{runoff-data-used-for-training}
+\subsection{The main idea}\label{sec-main_idea}
 
-The runoff data covering the period 1979 to 2011 is based on an E-HYPE
-hindcast simulation that was forced by a regional downscaling of
-ERA-Interim (Dee et al., 2011) with RCA3 ({``The rossby centre regional
-climate model RCA3,''} n.d.) and implemented into NEMO-Nordic (Hordoir
-et al., 2019) as a mass flux. For the periods before (1961 to 1978) and
-after (2012 to 2018) additional spatial temporal corrections have been
-applied to the runoff data, and have therefore been ignored. For more
-information see Gröger et al. (2022) and references therein.
+We assume that the runoff for a specific point in time of all \(N_r\)
+considered rivers, collected in the vector
+\(\vec{R}^t \in \mathbb{R}^{N_r}\), can be accurately approximated by a
+functional \(\vec{M}(\{X^t_k[x,y,\tau]\})\) of \(k=1,\ldots N_k\)
+atmospheric fields \(X^t_k[x,y,\tau]\) that are known for the past
+\(\tau=1,\ldots N_\tau\) time instances, i.e.
 
-\subsection{Atmospheric Forcing}\label{atmospheric-forcing}
+\[
+\begin{aligned}
+\vec{R}^t = \vec{M}(\{X^t_k[x,y,\tau]\}) \ .
+\end{aligned}
+\]
 
-The UERRA-HARMONIE regional reanalysis dataset was developed as part of
-the FP7 UERRA project (Uncertainties in Ensembles of Regional
-Re-Analyses, \href{http://www.uerra.eu/}{http://www.uerra.eu/)},). The
-UERRA-HARMONIE system represents a comprehensive, high-resolution
-reanalysis covering a wide range of essential climate variables. This
-dataset encompasses data on air temperature, pressure, humidity, wind
-speed and direction, cloud cover, precipitation, albedo, surface heat
-fluxes, and radiation fluxes from January 1961 to July 2019. With a
-horizontal resolution of 11 km and analyses conducted at 00 UTC, 06 UTC,
-12 UTC, and 18 UTC, it also provides hourly resolution forecast model
-data. UERRA-HARMONIE is accessible through the Copernicus Climate Data
-Store (CDS, \url{https://cds.climate.copernicus.eu/\#!/home),} initially
-produced during the UERRA project and later transitioned to the
-Copernicus Climate Change Service (C3S,
-\url{https://climate.copernicus.eu/copernicus-regionalreanalysis-europe).}
+The fields are evaluated on a spatial domain \(x=1,\ldots N_x\) and
+\(y=1,\ldots N_y\) that is sufficiently large to capture all significant
+contributions to the modelled runoff.
 
-\subsection{Ocean Model}\label{ocean-model}
+We propose that this functional can be adequately represented by a
+combination of a ConvLSTM with a subsequent FC network as it is
+visualized in Figure~\ref{fig-ConvLSTM_FC} and described in detail in
+the following. In order to provide an overview, we will go through the
+main ingredients of this architecture one-by-one.
 
-To simulate the Baltic Sea, we use a coupled 3-dimensional ocean model
-Baltic Sea, called the Modular Ocean Model (MOM). This model uses a
-finite-difference method to solve the full set of primitive equations to
-calculate the motion of water and the transport of heat and salt. The
-K-profile parameterization (KPP) was used as turbulence closure scheme.
-The model's western boundary opens into the Skagerrak and connects the
-Baltic Sea to the North Sea. The maximum depth was set at 264 meters. A
-more detailed description of the setup can be found in (Gröger et al.,
-2022).
+\begin{figure}
 
-\subsection{LSTM network}\label{lstm-network}
+\centering{
+
+\captionsetup{labelsep=none}\includegraphics{./ConvLSTM_FC.png}
+
+}
+
+\caption{\label{fig-ConvLSTM_FC}}
+
+\end{figure}%
+
+\subsection{ConvLSTM network}\label{convlstm-network}
+
+\subsubsection{The LSTM approach}\label{sec-LSTM}
 
@@
 
-\includegraphics{draw_lstm.png}
+\includegraphics{./LSTM.png}
 
@@
 This performance in modeling long-range dependencies has been validated
-in various studies. The key component of LSTM's innovation is its memory
-cell, \(\vec{C}^t[\tau] = (C^t_h[\tau]) \in \mathbb{R}^{N_h}\), which
-stores state information, also referred to as long-term memory. The
-index \(\tau \in [1, N_\tau]\) corresponds to the \(\tau\)-th element in
-the input sequence. The cell state is a vector consists of elements,
-each associated with one of the \(N_h\) hidden layers, labeled by \(h\).
-This vector is determined through several self-parameterized gates, all
-in the same vector space as \(\vec{C}^t[\tau]\). The forget gate
-\(\vec{F}^t[\tau]\) defines the percentage of the previous long-term
-memory status \(\vec{C}^t[\tau-1]\) that should be retained
-stored\hspace{0pt}. The input gate \(\vec{I}^t[\tau]\) decides how much
-of the input is added to the the long-term memory, forming the updated
-cell state. The so-called output gate \(\vec{O}^t[\tau]\) then
-determines how much of the latest cell output, \(\vec{C}^t[\tau]\), is
-propagated to the final state, \(\vec{H}^t[\tau]\) representing the
-updated short-term memory of the hidden state.
+in various studies. The key component of the LSTM's innovation is its
+memory cell, \(\vec{C}^t[\tau] = (C^t_h[\tau]) \in \mathbb{R}^{N_h}\),
+which stores state information, also referred to as long-term memory.
+This state information complements the the so-called hidden state
+\(\vec{H}^t[\tau] = (H^t_h[\tau]) \in \mathbb{R}^{N_h}\) vector, which
+is also known from simpler neural network architectures. In case of the
+LSTM, the hidden state vector plays the role of the short-term meory.
+The cell state and the hidden state are vectors, where each element is
+associated with one of the so-called \(N_h\) hidden layers, labeled by
+\(h\). These two state vectors are determined through several
+self-parameterized gates, all in the same vector space as
+\(\vec{C}^t[\tau]\).
+
+In particular, the forget gate \(\vec{F}^t[\tau]\) defines the portion
+of the previous (long-term memory) cell state \(\vec{C}^t[\tau-1]\) that
+should be kept\hspace{0pt}. The input gate \(\vec{I}^t[\tau]\) controls
+the contribution of the current input that is used to update the
+long-term memory,\(\vec{C}^t[\tau]\). The output gate
+\(\vec{O}^t[\tau]\) then determines how much of this updated long-term
+memory contributes to the new (short-term memory) hidden state,
+\(\vec{H}^t[\tau]\).
 
@@
 i.e.~the connection between the input \(\vec{X}^t[\tau]\), the various
-gates and the state vectors, is given as follows. First, the elements of
-the input sequence together with the hidden state are mapped onto
-auxiliary gate vectors, collectively denoted by
+gates and the state vectors, is mathematically given as follows.
+
+First, the elements of the input sequence together with the hidden state
+are mapped onto auxiliary gate vectors, collectively denoted by
 \(\vec{g}^t[\tau] = (g^t_h[\tau]) \in \mathbb{R}^{N_h}\), via \[
@@
 equations the role of the input, forget and output gates as described
-above become apparent.
+above becomes apparent.
 
@@
 
-\[
-\left(\vec{H}^t[N_\tau], \vec{C}^t[N_\tau] \right) = L \left( \vec{X}^t[N_\tau], L \left( \vec{X}^t[N_\tau-1], \ldots L \left( \vec{X}^t[1], (\vec{H}^t[0], \vec{C}^t[0] \right) \ldots \right) \right) \ ,
-\]
+\begin{equation}\phantomsection\label{eq-L_nested}{
+\left(\vec{H}^t[N_\tau], \vec{C}^t[N_\tau] \right) = L \left( \vec{X}^t[N_\tau], L \left( \vec{X}^t[N_\tau-1], \ldots L \left( \vec{X}^t[1], (\vec{H}^t[0], \vec{C}^t[0]) \right) \ldots \right) \right) \ ,
+}\end{equation}
 
-where \(L\) represents Eqs. (Equation~\ref{eq-LSTM-gates},
-Equation~\ref{eq-LSTM}) and the initial conditions are usually chosen as
-\(\vec{H}^t[0]=\vec{C}^t[0]=0\).
+where
+\(L\left(\vec{X}^t[\tau], \left(\vec{H}^t[\tau-1], \vec{C}^t[\tau-1] \right) \right)\)
+represents Equation~\ref{eq-LSTM-gates} and Equation~\ref{eq-LSTM} and
+the initial conditions are chosen as \(\vec{H}^t[0]=\vec{C}^t[0]=0\).
 
@@
 information on the full input sequence ending at time \(t\). This
-information has to be decoded to obtain usueful information via an
-appropriate subsequent layer, as it is described in the Sec.
-(\textbf{XXX?}).
+information has to be decoded to obtain useful information via an
+appropriate subsequent layer, as it is described in the
+Section~\ref{sec-FC_layer}.
 
@@
 
-\subsection{ConvLSTM network}\label{convlstm-network}
+\subsubsection{Combining LSTM with spatial
+convolution}\label{combining-lstm-with-spatial-convolution}
 
@@
 sequences of point-like quantities it is not designed to recognize
-spatial features in seuqences of two-dimensional maps. To adress this
-limitation we use a convLSTM architecture.
+spatial features in seuqences of, e.g., two-dimensional maps. To adress
+this limitation we employ a ConvLSTM architecture as described in the
+following.
 
 In this kind of network the elements of the input sequence are given as
-(kind of) vector fields
+spatially varying fields
 \(\vec{X}^t[\tau] = (X^t_k[x,y,\tau]) \in \mathbb{R}^{N_k \times (N_x \times N_y)}\),
@@
 
-It becomes immediately apparent that in case of the convLSTM, the bias,
-gate and state vectors must become vector fields
+again with Einstein's convention imposed.
+
+It becomes immediately apparent that in case of the ConvLSTM, gate and
+state vectors must become vector fields
 (\(\in \mathbb{R}^{N_h \times (N_x \times N_y)}\)) as well. We can write
-this mapping in the same fashion as Eq. (Equation~\ref{eq-LSTM-gates})
-but with replacing the normal matrix-vector multiplication by a
-convolution (denoted with \(\ast\)), i.e.\\
+this mapping in the same fashion as Equation~\ref{eq-LSTM-gates} but
+with replacing the normal matrix-vector multiplication by a convolution
+(denoted with \(\ast\)), i.e.\\
 \[
@@
 The susebsequent processing of the \(\vec{g}^t[\tau]\) remains
-symbolically the same as presented in Eq. (Equation~\ref{eq-LSTM}) but
-with all appearing quantities meaning standing now for vector fields
-instead of simple vectors.
+symbolically the same as presented in Equation~\ref{eq-LSTM} but with
+all appearing quantities meaning standing now for vector fields instead
+of simple vectors.
 
@@
 
-\subsection{Implemented model
-architecture}\label{implemented-model-architecture}
+\subsection{Fully connected layer}\label{sec-FC_layer}
 
-The ConvLSTM architectures uses and encoder/decoder structure as
-discussed in TODO. To predict all 97 rivers entering the Baltic Sea, we
-flatten the output and use fully connected layers to map onto the
-individual rivers outputs.
+As stated in Sec. Section~\ref{sec-LSTM}, the final output
+\(\vec{H}^t[N_\tau]\) and \(\vec{C}^t[N_\tau]\) of the ConvLSTM encode
+information on the full input sequence. In order to contract this
+information to the runoff vector \(\vec{R}^t\) representing the \(N_r\)
+rivers, we propose to subject the final short-term memory (i.e.~the
+hidden state \(\vec{H}^t[N_\tau]\)) to an additional FC network.
 
-An overview of the model structure is given below
+In particular, the dimensionality of the vector field
+\(\vec{H}^t[N_\tau]\) is sequentially reduced by three nested FC layers,
+each connected to the other by the Rectified Linear Unit (ReLU), see
+Figure~\ref{fig-ConvLSTM_FC}. Integrating over artificial degrees of
+freedom in a step-wise fashion has turned out to be benificial
+((\textbf{XXX?})).
+
+Spelled out in mathematics the runoff of the \(r\)-th river is then
+obtained via (using Einstein's convention)
+
+\[
+\begin{aligned}
+R_r^t = \mathcal{W}^{3}_{rb}\mathrm{ReLU}\left(\mathcal{W}^{2}_{ba}\mathrm{ReLU}\left(\mathcal{W}^{1}_{ah} [x,y] H^t_h[x,y, N_\tau] + \mathcal{B}^1_a\right) + \mathcal{B}^2_b \right) + \mathcal{B}^3_r \ ,
+\end{aligned}
+\]
+
+where \(a=1,\ldots N_a\) and \(b=1,\ldots N_b\) and
+\(N_h\cdot N_x\cdot N_y > N_a > N_b > N_r\) in order to achieve the
+aforementioned step-by-step reduction of dimensionality. The weights
+\(\mathcal{W}\) and biases \(\mathcal{B}\) stand for parameters that are
+optimized during the training of the network.
+
+In matrix-vector notation this can compactified to \[
+\begin{aligned}
+\vec{R}^t = \pmb{\mathcal{W}}^{3}\mathrm{ReLU}\left(\pmb{\mathcal{W}}^{2}\mathrm{ReLU}\left(\pmb{\mathcal{W}}^{1} \vec{H}^t[N_\tau] + \vec{\mathcal{B}}^1\right) + \vec{\mathcal{B}}^2 \right) + \vec{\mathcal{B}}^3 \ .
+\end{aligned}
+\]
+
+Combining this equation with Equation~\ref{eq-L_nested} provides finally
+an explicit formula for the initial assumption of modelling the runoff
+for time \(t\) as a functional of a sequence of atmospheric fields, i.e.
+
+\[
+\begin{aligned}
+\vec{R}^t & = \vec{M}(\{X^t_k[x,y,\tau]\}) \\
+& = \pmb{\mathcal{W}}^{3}\mathrm{ReLU}\left(\pmb{\mathcal{W}}^{2}\mathrm{ReLU}\left(\pmb{\mathcal{W}}^{1} \vec{L}_H \left( \vec{X}^t[N_\tau], L \left( \vec{X}^t[N_\tau-1], \ldots L \left( \vec{X}^t[1], (0, 0) \right) \ldots \right) \right) + \vec{\mathcal{B}}^1\right) + \vec{\mathcal{B}}^2 \right) + \vec{\mathcal{B}}^3 \ ,
+\end{aligned}
+\] where the \(\vec{L}_H\) means that only the hidden state vector of
+the final ConvLSTM call is forwarded to the FC layer. In oder to make
+this model accurate the hyperparameters have to be chosen reasonable and
+the free parameters are to be adjusted accordingly.
+
+\section{Technical details}\label{technical-details}
+
+\subsection{Runoff data used for
+training}\label{runoff-data-used-for-training}
+
+The runoff data covering the period 1979 to 2011 is based on an E-HYPE
+hindcast simulation that was forced by a regional downscaling of
+ERA-Interim (Dee et al., 2011) with RCA3 ({``The rossby centre regional
+climate model RCA3,''} n.d.) and implemented into NEMO-Nordic (Hordoir
+et al., 2019) as a mass flux. For the periods before (1961 to 1978) and
+after (2012 to 2018) additional spatial temporal corrections have been
+applied to the runoff data, and have therefore been ignored. For more
+information see Gröger et al. (2022) and references therein.
+
+\subsection{Atmospheric Forcing}\label{atmospheric-forcing}
+
+The UERRA-HARMONIE regional reanalysis dataset was developed as part of
+the FP7 UERRA project (Uncertainties in Ensembles of Regional
+Re-Analyses, \href{http://www.uerra.eu/}{http://www.uerra.eu/)},). The
+UERRA-HARMONIE system represents a comprehensive, high-resolution
+reanalysis covering a wide range of essential climate variables. This
+dataset encompasses data on air temperature, pressure, humidity, wind
+speed and direction, cloud cover, precipitation, albedo, surface heat
+fluxes, and radiation fluxes from January 1961 to July 2019. With a
+horizontal resolution of 11 km and analyses conducted at 00 UTC, 06 UTC,
+12 UTC, and 18 UTC, it also provides hourly resolution forecast model
+data. UERRA-HARMONIE is accessible through the Copernicus Climate Data
+Store (CDS, \url{https://cds.climate.copernicus.eu/\#!/home),} initially
+produced during the UERRA project and later transitioned to the
+Copernicus Climate Change Service (C3S,
+\url{https://climate.copernicus.eu/copernicus-regionalreanalysis-europe).}
+
+\subsection{Ocean Model}\label{ocean-model}
+
+To simulate the Baltic Sea, we use a coupled 3-dimensional ocean model
+Baltic Sea, called the Modular Ocean Model (MOM). This model uses a
+finite-difference method to solve the full set of primitive equations to
+calculate the motion of water and the transport of heat and salt. The
+K-profile parameterization (KPP) was used as turbulence closure scheme.
+The model's western boundary opens into the Skagerrak and connects the
+Baltic Sea to the North Sea. The maximum depth was set at 264 meters. A
+more detailed description of the setup can be found in (Gröger et al.,
+2022).
+
+\subsection{Neural network hyper
+parameters}\label{neural-network-hyper-parameters}
 
@@
 
-\includegraphics{convLSTM.png}
+\includegraphics{ConvLSTM.png}
 
@@
 
-\caption{\label{fig-baltNet}Schematic structure of the convLSTM
+\caption{\label{fig-baltNet}Schematic structure of the ConvLSTM
 implementation for river runoff forecasting.}

### 2024-05-15 paper/paper.qmd
Polished theory part

@@
 
-We assume that the runoff for a specific point in time of all $N_r$ considered rivers, collected in the vector $\vec{R}^t \in \mathbb{R}^{N_r}$, can be accurately approximated by a functional $\vec{M}(\{X^t_k[x,y,\tau]\})$ of $k=1,\ldots N_k$ atmospheric fields $X^t_k[x,y,\tau]$ that are known for the past $\tau=1,\ldots N_\tau$ time instances, i.e.
+We assume that the runoff for a specific point in time for all $N_r$ considered rivers, collected in the vector $\vec{R}^t \in \mathbb{R}^{N_r}$, can be accurately approximated by a functional $\vec{M}(\{X^t_k[x,y,\tau]\})$ of $k=1,\ldots N_k$ atmospheric fields $X^t_k[x,y,\tau]$ that are known for the past $\tau=1,\ldots N_\tau$ time instances, i.e.
 
@@
 
-The fields are evaluated on a spatial domain $x=1,\ldots N_x$ and $y=1,\ldots N_y$ that is sufficiently large to capture all significant contributions to the modelled runoff.
+The fields are evaluated on a spatial domain $x=1,\ldots N_x$ and $y=1,\ldots N_y$ that is sufficiently large to capture all significant local and non-local contributions of atmospheric fields to the river runoff.
 
-We propose that this functional can be adequately represented by a combination of a ConvLSTM with a subsequent FC network as it is visualized in @fig-ConvLSTM_FC and described in detail in the following.
+Usually such a mapping is realized in a suitable hydrological model that simulates all relevant physical processes that transform variables like, e.g., precipitation and evaproration to the river runoff, relying on expertise for tuning all appearing parameters to reasonable values.
+
+As an alternative, we propose that this functional can be adequately represented by a combination of a ConvLSTM with a subsequent FC network.
+That way, there is no need to have knowledge of involved processes and how they can be modeled, instead, these features can be "lerned" by the network in an automated manner.
+Our poroposed network architecture is visualized in @fig-ConvLSTM_FC and described in detail in the following.
 In order to provide an overview, we will go through the main ingredients of this architecture one-by-one.
 
-![](./ConvLSTM_FC.png){#fig-ConvLSTM_FC fig-align="center"}
+![**Combined ConvLSTM and FC network architecture**.
+The starting point is the continuous timeseries of input data $\{\vec{X}^t\}$, upper white block.
+From this series, a contiguous sequence of $N_\tau$ elements (yellow block) is used to feed a chain of $N_\tau$ connected ConvLSTM cells (light blue block) building the ConvLSTM network (grey block).
+The input seuqence is mapped via weighting matrices $\pmb{\mathcal{M}}^g$ (green blocks) onto gate vectors $\vec{g}^t[\tau]$.
+The gate vectors are then used to update the cell state $\vec{C}^t[\tau-1]$ and the hidden state $\vec{H}^t[\tau-1]$ of the last ConvLSTM cell to the current values $\vec{C}^t[\tau]$ and $\vec{H}^t[\tau]$, respectively.
+The update is performed with LSTM core equations collectively described by the mapping L, see @eq-LSTM.
+The weighting matrices $\pmb{\mathcal{N}}^g$ (green blocks) control how much of the last hidden state enters the updated state.
+The final output of the ConvLSTM $\vec{H}^t[\tau]$ is then propagated to a FC network, which itself is a chain of three FC layers consisiting of weigthing matrices $\pmb{\mathcal{W}}$ and connected via ReLU functions, see @sec-FC_layer.
+The final result is then the river runoff $\vec{R}^t$ for all rivers considered at the current time instance $t$ (white block on the lower left).
+Note that all bias vectors are ommitted for the sake of clarity.
+See text for more information.](./ConvLSTM_FC.png){#fig-ConvLSTM_FC fig-align="center"}
 
@@
 
-Before turning to the convolutional LSTM, the simpler architecture of the plain LSTM is examined and should serve as a role model for the later consideration of the full ConvLSTM.
+Before turning directly to the convolutional LSTM, the simpler architecture of the plain LSTM is examined and should serve as a role model for the later consideration of the full ConvLSTM.
 
@@
 The sequence is taken from a dataset given in form of a time series $\{\vec{X}^t\}$ and the endpoint should coincide with the specific element in the time series connected to time $t$, i.e.
-$\vec{X}^t[N_\tau] \equiv \vec{X}^t$.
+$\vec{X}^t[N_\tau] \equiv \vec{X}^t$, see @fig-ConvLSTM_FC.
 The $N_k$ is the number of used input "channels" and can, for example, represent different measurable quantities.
@@
 
-![Inner structure of a Long Short-Term Memory Cell](./LSTM.png){#fig-lstm fig-align="center"}
+![**Inner structure of a Long Short-Term Memory Cell**. See @fig-ConvLSTM_FC and text for information.](./LSTM.png){#fig-lstm fig-align="center"}
 
-This performance in modeling long-range dependencies has been validated in various studies.
-The key component of the LSTM's innovation is its memory cell, $\vec{C}^t[\tau] = (C^t_h[\tau]) \in \mathbb{R}^{N_h}$, which stores state information, also referred to as long-term memory.
-This state information complements the the so-called hidden state $\vec{H}^t[\tau] = (H^t_h[\tau]) \in \mathbb{R}^{N_h}$ vector, which is also known from simpler neural network architectures.
-In case of the LSTM, the hidden state vector plays the role of the short-term meory.
-The cell state and the hidden state are vectors, where each element is associated with one of the so-called $N_h$ hidden layers, labeled by $h$.
-These two state vectors are determined through several self-parameterized gates, all in the same vector space as $\vec{C}^t[\tau]$.
+This performance in modeling long-range dependencies has been validated in various studies [@XXX].
+The key component of the LSTM's innovation is its cell state, $\vec{C}^t[\tau] = (C^t_h[\tau]) \in \mathbb{R}^{N_h}$, which stores state information, also referred to as *long-term memory*.
+This state information complements the so-called hidden state $\vec{H}^t[\tau] = (H^t_h[\tau]) \in \mathbb{R}^{N_h}$ vector, which is also known from simpler neural network architectures.
+In case of the LSTM, the hidden state vector plays the role of the *short-term meory*.
 
-In particular, the forget gate $\vec{F}^t[\tau]$ defines the portion of the previous (long-term memory) cell state $\vec{C}^t[\tau-1]$ that should be kept​.
-The input gate $\vec{I}^t[\tau]$ controls the contribution of the current input that is used to update the long-term memory,$\vec{C}^t[\tau]$.
-The output gate $\vec{O}^t[\tau]$ then determines how much of this updated long-term memory contributes to the new (short-term memory) hidden state, $\vec{H}^t[\tau]$.
+A significant advantage of this architecture is the memory cell's ability to retain gradients.
+This mechanism addresses the vanishing gradient problem, where, as input sequences elongate, the influence of initial stages becomes harder to capture, causing gradients of early input points to approach zero.
+The LSTM's activation function, inherently recurrent, mirrors the identity function with a consistent derivative of 1.0, ensuring the gradient remains stable throughout backpropagation.
+
+The cell state and the hidden state are vectors, where each element is associated with one of the $N_h$ hidden layers, labeled by $h$, which are internal, *artificial* degrees of freedom that enable the high adaptibility of neural networks.
+These two state vectors are determined through several self-parameterized gates, all in the same vector space as $\vec{C}^t[\tau]$, see @fig-lstm for a visualization.
+
+In particular, the forget gate $\vec{F}^t[\tau]$ defines the portion of the previous (long-term memory) cell state $\vec{C}^t[\tau-1]$ that should be kept, see dashed cyan box therin​.
+The input gate $\vec{I}^t[\tau]$ controls the contribution of the current input used to update the long-term memory, $\vec{C}^t[\tau]$ (magenta and yellow boxes).
+The output gate, $\vec{O}^t[\tau]$, then determines how much of this updated long-term memory contributes to the new (short-term memory) hidden state, $\vec{H}^t[\tau]$ (black dashed box).
 
@@
 
-For the sake of brevity, we write the mapping more compactly in matrix-vector form as $$
+For the sake of brevity, we can write the mapping more compactly in matrix-vector form as $$
 \begin{aligned}
@@
 
-Second, the actual gate vectors are computed by the core equations of the LSTM as proposed by [@XXX]:
+Second, the actual gate vectors are computed by the core equations of the LSTM as proposed by @hochreiter1997long:
 
@@
 
-The full action of the LSTM network up to the end of the sequence can therefore be written as a nested function call
+The full action of the LSTM network up to the end of the sequence can be written as a nested function call
 
@@
 
-where $L\left(\vec{X}^t[\tau], \left(\vec{H}^t[\tau-1], \vec{C}^t[\tau-1] \right) \right)$ represents @eq-LSTM-gates and @eq-LSTM and the initial conditions are chosen as $\vec{H}^t[0]=\vec{C}^t[0]=0$.
+where $L\left(\vec{X}^t[\tau], \left(\vec{H}^t[\tau-1], \vec{C}^t[\tau-1] \right) \right)$ represents @eq-LSTM-gates and @eq-LSTM.
+For the present work, the initial conditions are chosen as $\vec{H}^t[0]=\vec{C}^t[0]=0$, which simply means that there is no memory longer than $N_\tau$ time steps.
 
-The final output $\vec{H}^t[N_\tau]$ and $\vec{C}^t[N_\tau]$ encode information on the full input sequence ending at time $t$.
-This information has to be decoded to obtain useful information via an appropriate subsequent layer, as it is described in the @sec-FC_layer.
-
-A significant advantage of this architecture is the memory cell's ability to retain gradients.
-This mechanism addresses the vanishing gradient problem, where, as input sequences elongate, the influence of initial stages becomes harder to capture, causing gradients of early input points to approach zero.
-The LSTM's activation function, inherently recurrent, mirrors the identity function with a consistent derivative of 1.0, ensuring the gradient remains stable throughout backpropagation.
+The final output of the ConvLSTM chain, $\vec{H}^t[N_\tau]$ and $\vec{C}^t[N_\tau]$, encode information on the full input sequence ending at time $t$.
+This information has to be decoded to obtain useful information via an appropriate subsequent network, as it is described in the @sec-FC_layer.
 
@@
 
-Although the plain LSTM has high performance in handling temporal sequences of point-like quantities it is not designed to recognize spatial features in seuqences of, e.g., two-dimensional maps.
+Although the plain LSTM has high performance in handling temporal sequences of point-like quantities it is not designed to recognize spatial features in seuqences of, e.g., two-dimensional maps as atmospheric-ocean interface fields.
 To adress this limitation we employ a ConvLSTM architecture as described in the following.
@@
 In order to enable the "learning" of spatial patterns, the free parameters of the network are replaced by two-dimenmsional convolutional kernels $\pmb{\mathcal{M}}^{g} = (\mathcal{M}^{g}_{hk}[\xi, \eta]) \in \mathbb{R}^{(N_h \times N_k)\times (N_\xi \times N_\eta)}$ and $\pmb{\mathcal{N}}^{g} = (\mathcal{N}^{g}_{hh'}[\xi, \eta]) \in \mathbb{R}^{(N_h \times N_h)\times (N_\xi \times N_\eta)}$.
-The width and the height of the kernels are given by $N_\xi$ and $N_\eta$, respectively and $\xi\in [-(N_\xi-1)/2,(N_\xi-1)/2], \eta\in [-(N_\eta-1)/2,(N_\eta-1)/2]$, where we assume odd numbers for the kernel sizes.
+The width and the height of the kernels are given by $N_\xi$ and $N_\eta$, respectively and $\xi\in [-(N_\xi-1)/2,(N_\xi-1)/2], \eta\in [-(N_\eta-1)/2,(N_\eta-1)/2]$, where, without loss of generality, we assume odd numbers for the kernel sizes.
 
@@
 
-It becomes immediately apparent that in case of the ConvLSTM, gate and state vectors must become vector fields ($\in \mathbb{R}^{N_h \times (N_x \times N_y)}$) as well.
+It becomes immediately apparent that in case of the ConvLSTM, the gate and state vectors must become vector fields ($\in \mathbb{R}^{N_h \times (N_x \times N_y)}$) as well.
 We can write this mapping in the same fashion as @eq-LSTM-gates but with replacing the normal matrix-vector multiplication by a convolution (denoted with $\ast$), i.e.\
@@
 
-The susebsequent processing of the $\vec{g}^t[\tau]$ remains symbolically the same as presented in @eq-LSTM but with all appearing quantities meaning standing now for vector fields instead of simple vectors.
+The susebsequent processing of the $\vec{g}^t[\tau]$ remains symbolically the same as presented in @eq-LSTM but with all appearing quantities now meaning vector fields instead of simple vectors.
 
@@
 @sec-LSTM, the final output $\vec{H}^t[N_\tau]$ and $\vec{C}^t[N_\tau]$ of the ConvLSTM encode information on the full input sequence.
-In order to contract this information to the runoff vector $\vec{R}^t$ representing the $N_r$ rivers, we propose to subject the final short-term memory (i.e.
+In order to contract this information to obtain the runoff vector $\vec{R}^t$ representing the $N_r$ rivers, we propose to subject the final short-term memory (i.e.
 the hidden state $\vec{H}^t[N_\tau]$) to an additional FC network.
@@
 In particular, the dimensionality of the vector field $\vec{H}^t[N_\tau]$ is sequentially reduced by three nested FC layers, each connected to the other by the Rectified Linear Unit (ReLU), see @fig-ConvLSTM_FC.
-Integrating over artificial degrees of freedom in a step-wise fashion has turned out to be benificial (@XXX).
+Integrating over artificial degrees of freedom in a step-wise fashion has turned out to be benificial \[@XXX\].
 
-Spelled out in mathematics the runoff of the $r$-th river is then obtained via (using Einstein's convention)
+Spelled out in mathematics, the runoff of the $r$-th river is then obtained via (using Einstein's convention)
 
@@
 
-where $a=1,\ldots N_a$ and $b=1,\ldots N_b$ and $N_h\cdot N_x\cdot N_y > N_a > N_b > N_r$ in order to achieve the aforementioned step-by-step reduction of dimensionality.
+where $a=1,\ldots N_a$, $b=1,\ldots N_b$ and the hyper parameters $N_a$ and $N_b$ are chosen such that $N_h\cdot N_x\cdot N_y > N_a > N_b > N_r$ in order to achieve the aforementioned step-by-step reduction of dimensionality.
 The weights $\mathcal{W}$ and biases $\mathcal{B}$ stand for parameters that are optimized during the training of the network.
 
-In matrix-vector notation this can compactified to $$
+In matrix-vector notation this can be compactified to $$
 \begin{aligned}
@@
 
-Combining this equation with @eq-L_nested provides finally an explicit formula for the initial assumption of modelling the runoff for time $t$ as a functional of a sequence of atmospheric fields, i.e.
+Combining the last equation with @eq-L_nested provides finally an explicit formula for the initial assumption of modelling the runoff for time $t$ as a functional of a sequence of atmospheric fields, i.e.
 
@@
 $$ where the $\vec{L}_H$ means that only the hidden state vector of the final ConvLSTM call is forwarded to the FC layer.
-In oder to make this model accurate the hyperparameters have to be chosen reasonable and the free parameters are to be adjusted accordingly.
 
-## Technical details
+In @sec-technical_details, we present the employed model and data setup as well as the choice for all mentioned hyper parameters that lead to an adequate model performance after training.
+
+## Technical details {#sec-technical_details}
 

### 2024-05-15 paper/paper.tex

@@
 
-We assume that the runoff for a specific point in time of all \(N_r\)
+We assume that the runoff for a specific point in time for all \(N_r\)
 considered rivers, collected in the vector
@@
 \(y=1,\ldots N_y\) that is sufficiently large to capture all significant
-contributions to the modelled runoff.
+local and non-local contributions of atmospheric fields to the river
+runoff.
 
-We propose that this functional can be adequately represented by a
-combination of a ConvLSTM with a subsequent FC network as it is
-visualized in Figure~\ref{fig-ConvLSTM_FC} and described in detail in
+Usually such a mapping is realized in a suitable hydrological model that
+simulates all relevant physical processes that transform variables like,
+e.g., precipitation and evaproration to the river runoff, relying on
+expertise for tuning all appearing parameters to reasonable values.
+
+As an alternative, we propose that this functional can be adequately
+represented by a combination of a ConvLSTM with a subsequent FC network.
+That way, there is no need to have knowledge of involved processes and
+how they can be modeled, instead, these features can be ``lerned'' by
+the network in an automated manner. Our poroposed network architecture
+is visualized in Figure~\ref{fig-ConvLSTM_FC} and described in detail in
 the following. In order to provide an overview, we will go through the
@@
 
-\captionsetup{labelsep=none}\includegraphics{./ConvLSTM_FC.png}
+\includegraphics{./ConvLSTM_FC.png}
 
@@
 
-\caption{\label{fig-ConvLSTM_FC}}
+\caption{\label{fig-ConvLSTM_FC}\textbf{Combined ConvLSTM and FC network
+architecture}. The starting point is the continuous timeseries of input
+data \(\{\vec{X}^t\}\), upper white block. From this series, a
+contiguous sequence of \(N_\tau\) elements (yellow block) is used to
+feed a chain of \(N_\tau\) connected ConvLSTM cells (light blue block)
+building the ConvLSTM network (grey block). The input seuqence is mapped
+via weighting matrices \(\pmb{\mathcal{M}}^g\) (green blocks) onto gate
+vectors \(\vec{g}^t[\tau]\). The gate vectors are then used to update
+the cell state \(\vec{C}^t[\tau-1]\) and the hidden state
+\(\vec{H}^t[\tau-1]\) of the last ConvLSTM cell to the current values
+\(\vec{C}^t[\tau]\) and \(\vec{H}^t[\tau]\), respectively. The update is
+performed with LSTM core equations collectively described by the mapping
+L, see Equation~\ref{eq-LSTM}. The weighting matrices
+\(\pmb{\mathcal{N}}^g\) (green blocks) control how much of the last
+hidden state enters the updated state. The final output of the ConvLSTM
+\(\vec{H}^t[\tau]\) is then propagated to a FC network, which itself is
+a chain of three FC layers consisiting of weigthing matrices
+\(\pmb{\mathcal{W}}\) and connected via ReLU functions, see
+Section~\ref{sec-FC_layer}. The final result is then the river runoff
+\(\vec{R}^t\) for all rivers considered at the current time instance
+\(t\) (white block on the lower left). Note that all bias vectors are
+ommitted for the sake of clarity. See text for more information.}
 
@@
 
-Before turning to the convolutional LSTM, the simpler architecture of
-the plain LSTM is examined and should serve as a role model for the
-later consideration of the full ConvLSTM.
+Before turning directly to the convolutional LSTM, the simpler
+architecture of the plain LSTM is examined and should serve as a role
+model for the later consideration of the full ConvLSTM.
 
@@
 series connected to time \(t\),
-i.e.~\(\vec{X}^t[N_\tau] \equiv \vec{X}^t\). The \(N_k\) is the number
-of used input ``channels'' and can, for example, represent different
-measurable quantities. Its unique design allows it to adeptly handle
-long-range dependencies, setting it apart from traditional RNNs in terms
-of accuracy (see Figure~\ref{fig-lstm}).
+i.e.~\(\vec{X}^t[N_\tau] \equiv \vec{X}^t\), see
+Figure~\ref{fig-ConvLSTM_FC}. The \(N_k\) is the number of used input
+``channels'' and can, for example, represent different measurable
+quantities. Its unique design allows it to adeptly handle long-range
+dependencies, setting it apart from traditional RNNs in terms of
+accuracy (see Figure~\ref{fig-lstm}).
 
@@
 
-\caption{\label{fig-lstm}Inner structure of a Long Short-Term Memory
-Cell}
+\caption{\label{fig-lstm}\textbf{Inner structure of a Long Short-Term
+Memory Cell}. See Figure~\ref{fig-ConvLSTM_FC} and text for
+information.}
 
@@
 This performance in modeling long-range dependencies has been validated
-in various studies. The key component of the LSTM's innovation is its
-memory cell, \(\vec{C}^t[\tau] = (C^t_h[\tau]) \in \mathbb{R}^{N_h}\),
-which stores state information, also referred to as long-term memory.
-This state information complements the the so-called hidden state
+in various studies (\textbf{XXX?}). The key component of the LSTM's
+innovation is its cell state,
+\(\vec{C}^t[\tau] = (C^t_h[\tau]) \in \mathbb{R}^{N_h}\), which stores
+state information, also referred to as \emph{long-term memory}. This
+state information complements the so-called hidden state
 \(\vec{H}^t[\tau] = (H^t_h[\tau]) \in \mathbb{R}^{N_h}\) vector, which
 is also known from simpler neural network architectures. In case of the
-LSTM, the hidden state vector plays the role of the short-term meory.
+LSTM, the hidden state vector plays the role of the \emph{short-term
+meory}.
+
+A significant advantage of this architecture is the memory cell's
+ability to retain gradients. This mechanism addresses the vanishing
+gradient problem, where, as input sequences elongate, the influence of
+initial stages becomes harder to capture, causing gradients of early
+input points to approach zero. The LSTM's activation function,
+inherently recurrent, mirrors the identity function with a consistent
+derivative of 1.0, ensuring the gradient remains stable throughout
+backpropagation.
+
 The cell state and the hidden state are vectors, where each element is
-associated with one of the so-called \(N_h\) hidden layers, labeled by
-\(h\). These two state vectors are determined through several
-self-parameterized gates, all in the same vector space as
-\(\vec{C}^t[\tau]\).
+associated with one of the \(N_h\) hidden layers, labeled by \(h\),
+which are internal, \emph{artificial} degrees of freedom that enable the
+high adaptibility of neural networks. These two state vectors are
+determined through several self-parameterized gates, all in the same
+vector space as \(\vec{C}^t[\tau]\), see Figure~\ref{fig-lstm} for a
+visualization.
 
@@
 of the previous (long-term memory) cell state \(\vec{C}^t[\tau-1]\) that
-should be kept\hspace{0pt}. The input gate \(\vec{I}^t[\tau]\) controls
-the contribution of the current input that is used to update the
-long-term memory,\(\vec{C}^t[\tau]\). The output gate
-\(\vec{O}^t[\tau]\) then determines how much of this updated long-term
-memory contributes to the new (short-term memory) hidden state,
-\(\vec{H}^t[\tau]\).
+should be kept, see dashed cyan box therin\hspace{0pt}. The input gate
+\(\vec{I}^t[\tau]\) controls the contribution of the current input used
+to update the long-term memory, \(\vec{C}^t[\tau]\) (magenta and yellow
+boxes). The output gate, \(\vec{O}^t[\tau]\), then determines how much
+of this updated long-term memory contributes to the new (short-term
+memory) hidden state, \(\vec{H}^t[\tau]\) (black dashed box).
 
@@
 
-For the sake of brevity, we write the mapping more compactly in
+For the sake of brevity, we can write the mapping more compactly in
 matrix-vector form as
@@
 Second, the actual gate vectors are computed by the core equations of
-the LSTM as proposed by (\textbf{XXX?}):
+the LSTM as proposed by Hochreiter \& Schmidhuber (1997):
 
@@
 
-The full action of the LSTM network up to the end of the sequence can
-therefore be written as a nested function call
+The full action of the LSTM network up to the end of the sequence can be
+written as a nested function call
 
@@
 \(L\left(\vec{X}^t[\tau], \left(\vec{H}^t[\tau-1], \vec{C}^t[\tau-1] \right) \right)\)
-represents Equation~\ref{eq-LSTM-gates} and Equation~\ref{eq-LSTM} and
-the initial conditions are chosen as \(\vec{H}^t[0]=\vec{C}^t[0]=0\).
+represents Equation~\ref{eq-LSTM-gates} and Equation~\ref{eq-LSTM}. For
+the present work, the initial conditions are chosen as
+\(\vec{H}^t[0]=\vec{C}^t[0]=0\), which simply means that there is no
+memory longer than \(N_\tau\) time steps.
 
-The final output \(\vec{H}^t[N_\tau]\) and \(\vec{C}^t[N_\tau]\) encode
-information on the full input sequence ending at time \(t\). This
-information has to be decoded to obtain useful information via an
-appropriate subsequent layer, as it is described in the
-Section~\ref{sec-FC_layer}.
-
-A significant advantage of this architecture is the memory cell's
-ability to retain gradients. This mechanism addresses the vanishing
-gradient problem, where, as input sequences elongate, the influence of
-initial stages becomes harder to capture, causing gradients of early
-input points to approach zero. The LSTM's activation function,
-inherently recurrent, mirrors the identity function with a consistent
-derivative of 1.0, ensuring the gradient remains stable throughout
-backpropagation.
+The final output of the ConvLSTM chain, \(\vec{H}^t[N_\tau]\) and
+\(\vec{C}^t[N_\tau]\), encode information on the full input sequence
+ending at time \(t\). This information has to be decoded to obtain
+useful information via an appropriate subsequent network, as it is
+described in the Section~\ref{sec-FC_layer}.
 
@@
 sequences of point-like quantities it is not designed to recognize
-spatial features in seuqences of, e.g., two-dimensional maps. To adress
-this limitation we employ a ConvLSTM architecture as described in the
-following.
+spatial features in seuqences of, e.g., two-dimensional maps as
+atmospheric-ocean interface fields. To adress this limitation we employ
+a ConvLSTM architecture as described in the following.
 
@@
 \(\xi\in [-(N_\xi-1)/2,(N_\xi-1)/2], \eta\in [-(N_\eta-1)/2,(N_\eta-1)/2]\),
-where we assume odd numbers for the kernel sizes.
+where, without loss of generality, we assume odd numbers for the kernel
+sizes.
 
@@
 
-It becomes immediately apparent that in case of the ConvLSTM, gate and
-state vectors must become vector fields
+It becomes immediately apparent that in case of the ConvLSTM, the gate
+and state vectors must become vector fields
 (\(\in \mathbb{R}^{N_h \times (N_x \times N_y)}\)) as well. We can write
@@
 symbolically the same as presented in Equation~\ref{eq-LSTM} but with
-all appearing quantities meaning standing now for vector fields instead
-of simple vectors.
+all appearing quantities now meaning vector fields instead of simple
+vectors.
 
@@
 information on the full input sequence. In order to contract this
-information to the runoff vector \(\vec{R}^t\) representing the \(N_r\)
-rivers, we propose to subject the final short-term memory (i.e.~the
-hidden state \(\vec{H}^t[N_\tau]\)) to an additional FC network.
+information to obtain the runoff vector \(\vec{R}^t\) representing the
+\(N_r\) rivers, we propose to subject the final short-term memory
+(i.e.~the hidden state \(\vec{H}^t[N_\tau]\)) to an additional FC
+network.
 
@@
 freedom in a step-wise fashion has turned out to be benificial
-((\textbf{XXX?})).
+{[}(\textbf{XXX?}){]}.
 
-Spelled out in mathematics the runoff of the \(r\)-th river is then
+Spelled out in mathematics, the runoff of the \(r\)-th river is then
 obtained via (using Einstein's convention)
@@
 
-where \(a=1,\ldots N_a\) and \(b=1,\ldots N_b\) and
+where \(a=1,\ldots N_a\), \(b=1,\ldots N_b\) and the hyper parameters
+\(N_a\) and \(N_b\) are chosen such that
 \(N_h\cdot N_x\cdot N_y > N_a > N_b > N_r\) in order to achieve the
@@
 
-In matrix-vector notation this can compactified to \[
+In matrix-vector notation this can be compactified to \[
 \begin{aligned}
@@
 
-Combining this equation with Equation~\ref{eq-L_nested} provides finally
-an explicit formula for the initial assumption of modelling the runoff
-for time \(t\) as a functional of a sequence of atmospheric fields, i.e.
+Combining the last equation with Equation~\ref{eq-L_nested} provides
+finally an explicit formula for the initial assumption of modelling the
+runoff for time \(t\) as a functional of a sequence of atmospheric
+fields, i.e.
 
@@
 \] where the \(\vec{L}_H\) means that only the hidden state vector of
-the final ConvLSTM call is forwarded to the FC layer. In oder to make
-this model accurate the hyperparameters have to be chosen reasonable and
-the free parameters are to be adjusted accordingly.
+the final ConvLSTM call is forwarded to the FC layer.
 
-\section{Technical details}\label{technical-details}
+In Section~\ref{sec-technical_details}, we present the employed model
+and data setup as well as the choice for all mentioned hyper parameters
+that lead to an adequate model performance after training.
+
+\section{Technical details}\label{sec-technical_details}
 

### 2024-05-27 paper/paper.qmd
went through text

@@
 Freshwater enters the Baltic Sea through river runoff or positive net precipitation (precipitation minus evaporation) over the sea surface.
-The net precipitation accounts for 11 % and the river input for 89 % of the total freshwater input [@meier2002simulated].
+The net precipitation accounts for 11 $\%$ and the river input for 89 $\%$ of the total freshwater input [@meier2002simulated].
 Modelling the Baltic Sea is therefore to a large part the result of the quality of the river input, that is used for the simulation.
@@
 
-We assume that the runoff for a specific point in time for all $N_r$ considered rivers, collected in the vector $\vec{R}^t \in \mathbb{R}^{N_r}$, can be accurately approximated by a functional $\vec{M}(\{X^t_k[x,y,\tau]\})$ of $k=1,\ldots N_k$ atmospheric fields $X^t_k[x,y,\tau]$ that are known for the past $\tau=1,\ldots N_\tau$ time instances, i.e.
+We assume that the runoff at a specific point in time for all $N_r$ considered rivers, collected in the vector $\vec{R}^t \in \mathbb{R}^{N_r}$, can be accurately approximated by a functional $\vec{M}(\{X^t_k[x,y,\tau]\})$ of $k=1,\ldots N_k$ atmospheric fields $X^t_k[x,y,\tau]$ which are known for the past $\tau=1,\ldots N_\tau$ time instances.
+This relationship is expressed as:
 
@@
 
-The fields are evaluated on a spatial domain $x=1,\ldots N_x$ and $y=1,\ldots N_y$ that is sufficiently large to capture all significant local and non-local contributions of atmospheric fields to the river runoff.
+The atmospheric fields are evaluated over a spatial domain $x=1,\ldots N_x$ and $y=1,\ldots N_y$ which is sufficiently large to capture all significant local and non-local contributions of the atmospheric fields to the river runoff.
 
-Usually such a mapping is realized in a suitable hydrological model that simulates all relevant physical processes that transform variables like, e.g., precipitation and evaproration to the river runoff, relying on expertise for tuning all appearing parameters to reasonable values.
+Typically, such a mapping is realized using a hydrological model that simulates all relevant physical processes, transforming variables like precipitation and evaporation into river runoff.
+This process relies heavily on domain knowledge to tune all parameters to reasonable values.
 
 As an alternative, we propose that this functional can be adequately represented by a combination of a ConvLSTM with a subsequent FC network.
-That way, there is no need to have knowledge of involved processes and how they can be modeled, instead, these features can be "lerned" by the network in an automated manner.
-Our poroposed network architecture is visualized in @fig-ConvLSTM_FC and described in detail in the following.
-In order to provide an overview, we will go through the main ingredients of this architecture one-by-one.
+This approach eliminates the need for detailed knowledge of the involved processes and their modeling.
+Instead, these features can be "learned" by the network in an automated manner.
+Our proposed network architecture is visualized in @fig-ConvLSTM_FC and described in detail in the following sections.
+To provide an overview, we will discuss the main components of this architecture one-by-one.
 
@@
 
-Before turning directly to the convolutional LSTM, the simpler architecture of the plain LSTM is examined and should serve as a role model for the later consideration of the full ConvLSTM.
+Before turning directly to the convolutional LSTM, the simpler architecture of the plain LSTM is examined which serves as a foundation for understanding the more complex ConvLSTM.
 
-The Long Short-Term Memory (LSTM), a specialized form of Recurrent Neural Networks (RNNs), is specifically tailored for modeling temporal sequences $\vec{X}^t[1], \ldots \vec{X}^t[\tau],\ldots \vec{X}^t[N_\tau]$ of $N_\tau$ input quantities $\vec{X}^t[\tau] = (X^t_k[\tau]) \in \mathbb{R}^{N_k}$.
-The sequence is taken from a dataset given in form of a time series $\{\vec{X}^t\}$ and the endpoint should coincide with the specific element in the time series connected to time $t$, i.e.
+The Long Short-Term Memory (LSTM), a specialized form of Recurrent Neural Networks (RNNs), is specifically designed to model temporal sequences $\vec{X}^t[1], \ldots \vec{X}^t[\tau],\ldots \vec{X}^t[N_\tau]$ of $N_\tau$ input quantities $\vec{X}^t[\tau] = (X^t_k[\tau]) \in \mathbb{R}^{N_k}$.
+This sequence is taken from a dataset given in form of a time series $\{\vec{X}^t\}$ with the endpoint coinciding with the specific element in the time series connected to time $t$, i.e.
 $\vec{X}^t[N_\tau] \equiv \vec{X}^t$, see @fig-ConvLSTM_FC.
-The $N_k$ is the number of used input "channels" and can, for example, represent different measurable quantities.
-Its unique design allows it to adeptly handle long-range dependencies, setting it apart from traditional RNNs in terms of accuracy (see @fig-lstm).
+Here, $N_k$ represents the number of input "channels," which can correspond to different measurable quantities.
+The LSTM's unique design allows it to adeptly handle long-range dependencies, setting it apart from traditional RNNs in terms of accuracy (see @fig-lstm).
 
@@
 
-In particular, the forget gate $\vec{F}^t[\tau]$ defines the portion of the previous (long-term memory) cell state $\vec{C}^t[\tau-1]$ that should be kept, see dashed cyan box therin​.
+In particular, the forget gate $\vec{F}^t[\tau]$ defines the portion of the previous (long-term memory) cell state $\vec{C}^t[\tau-1]$ that should be kept, see dashed cyan box therein​.
 The input gate $\vec{I}^t[\tau]$ controls the contribution of the current input used to update the long-term memory, $\vec{C}^t[\tau]$ (magenta and yellow boxes).
@@
 
-where $g=i,f,o,c$ stands for the input, forget, output and cell-state gate, respectively and Eistein's summation convention is employed, i.e.
+where $g=i,f,o,c$ stands for the input, forget, output and cell-state gate, respectively and Einstein's summation convention is employed, i.e.
 indices that appear twice are summed over.
-The calligraphic symbols $\mathcal{M}^{g}_{hk}, \mathcal{N}^{g}_{hh'}$ and $\mathcal{B}^g_{h}$ are the free parameters of the network that are optimized for the given problem during the training, which is at the heart of the any machine learning approach.
+The calligraphic symbols $\mathcal{M}^{g}_{hk}, \mathcal{N}^{g}_{hh'}$ and $\mathcal{B}^g_{h}$ are the free parameters of the network that are optimized for the given problem during the training, which is at the heart of any machine learning approach.
 The matrix $\pmb{\mathcal{M}}^{g} = (\mathcal{M}^{g}_{hk}) \in \mathbb{R}^{N_h \times N_k}$ can be interpreted as a Markovian-like contribution of the current input $\vec{X}^t[\tau]$ to the gates, whereas the $\pmb{\mathcal{N}}^{g} = (\mathcal{N}^{g}_{hh'}) \in \mathbb{R}^{N_h \times N_h}$ scales a non-Markovian part determined by the hidden state of the last sequence point $\tau-1$.
@@
 
-where $\sigma$ denotes the logisitc sigmoid function, $\tanh$ is the hyperbolic tangent and the $\circ$ stands for the Hadamard product (all applied in an element-wise fashion to the vectors).
+where $\sigma$ denotes the logistic sigmoid function, $\tanh$ is the hyperbolic tangent and the $\circ$ stands for the Hadamard product (all applied in an element-wise fashion to the vectors).
 In the last two equations the role of the input, forget and output gates as described above becomes apparent.
@@
 
-Although the plain LSTM has high performance in handling temporal sequences of point-like quantities it is not designed to recognize spatial features in seuqences of, e.g., two-dimensional maps as atmospheric-ocean interface fields.
-To adress this limitation we employ a ConvLSTM architecture as described in the following.
+Although the plain LSTM has high performance in handling temporal sequences of point-like quantities it is not designed to recognize spatial features in sequences of, e.g., two-dimensional maps as atmospheric-ocean interface fields.
+To address this limitation we employ a ConvLSTM architecture as described in the following.
 
 In this kind of network the elements of the input sequence are given as spatially varying fields $\vec{X}^t[\tau] = (X^t_k[x,y,\tau]) \in \mathbb{R}^{N_k \times (N_x \times N_y)}$, where $x\in[1, N_x]$ and $y \in [1, N_y]$ run over the horizontal and vertical dimensions of the map, respectively.
-In order to enable the "learning" of spatial patterns, the free parameters of the network are replaced by two-dimenmsional convolutional kernels $\pmb{\mathcal{M}}^{g} = (\mathcal{M}^{g}_{hk}[\xi, \eta]) \in \mathbb{R}^{(N_h \times N_k)\times (N_\xi \times N_\eta)}$ and $\pmb{\mathcal{N}}^{g} = (\mathcal{N}^{g}_{hh'}[\xi, \eta]) \in \mathbb{R}^{(N_h \times N_h)\times (N_\xi \times N_\eta)}$.
+In order to enable the "learning" of spatial patterns, the free parameters of the network are replaced by two-dimensional convolution kernels $\pmb{\mathcal{M}}^{g} = (\mathcal{M}^{g}_{hk}[\xi, \eta]) \in \mathbb{R}^{(N_h \times N_k)\times (N_\xi \times N_\eta)}$ and $\pmb{\mathcal{N}}^{g} = (\mathcal{N}^{g}_{hh'}[\xi, \eta]) \in \mathbb{R}^{(N_h \times N_h)\times (N_\xi \times N_\eta)}$.
 The width and the height of the kernels are given by $N_\xi$ and $N_\eta$, respectively and $\xi\in [-(N_\xi-1)/2,(N_\xi-1)/2], \eta\in [-(N_\eta-1)/2,(N_\eta-1)/2]$, where, without loss of generality, we assume odd numbers for the kernel sizes.
@@
 It becomes immediately apparent that in case of the ConvLSTM, the gate and state vectors must become vector fields ($\in \mathbb{R}^{N_h \times (N_x \times N_y)}$) as well.
-We can write this mapping in the same fashion as @eq-LSTM-gates but with replacing the normal matrix-vector multiplication by a convolution (denoted with $\ast$), i.e.\
+We can write this mapping in the same way as @eq-LSTM-gates but with replacing the normal matrix-vector multiplication by a convolution (denoted with $\ast$), i.e.\
 $$
@@
 
-The susebsequent processing of the $\vec{g}^t[\tau]$ remains symbolically the same as presented in @eq-LSTM but with all appearing quantities now meaning vector fields instead of simple vectors.
+The subsequent processing of the $\vec{g}^t[\tau]$ remains symbolically the same as presented in @eq-LSTM but with all appearing quantities now meaning vector fields instead of simple vectors.
 
@@
 For the periods before (1961 to 1978) and after (2012 to 2018) additional spatial temporal corrections have been applied to the runoff data, and have therefore been ignored.
+The quality of the runoff was extensively evaluated.
 For more information see @gröger2022 and references therein.
@@
 
-The choice of atmospheric fields was based on the implemented river runoff calculation in the atmospheric model COSMO-CLM which uses these four fields to calculate an river runoff estimate.
+The choice of atmospheric fields was based on the implemented river runoff calculation in the atmospheric model COSMO-CLM which uses these atmospheric fields to calculate an river runoff estimate.
 
@@
 The model was trained with daily data for the period 1979 to 2011, as this period represents the only period of E-HYPE without further bias correction applied to the runoff to match observations.
-The data was divided into randomly chosen splits of 80% training data, 10% validation data to evaluate the performance of the model during training, and 10% training data which is finally used to evaluate the performance of the model after training.
+The data was divided into randomly chosen splits of 80$\%$ training data, 10$\%$ validation data to evaluate the performance of the model during training, and 10$\%$ training data which is finally used to evaluate the performance of the model after training.
 The model was trained for 400 epochs and the model weights with the lowest mean squared error during training have been stored.
@@
 As mention above for evaluation, the test dataset was utilized.
-The left panel panel (Figure @fig-statistical-evaluationNN a) ) illustrates the relative prediction error in relation to the original E-HYPE data, indicating that, on daily timescales, the model can predict river runoff with an accuracy of $\pm$ 5%.
+The left panel panel (Figure @fig-statistical-evaluationNN a) ) illustrates the relative prediction error in relation to the original E-HYPE data, indicating that, on daily timescales, the model can predict river runoff with an accuracy of $\pm$ 5$\%$.
 The overall correlation is 0.997 with the resulting error metrics yielding a RMSE of 323.99 $m^3$/s and MAE of 249.51 $m^3$/s.
 While the model's performance is satisfactory, the discrepancies between the actual values and the predictions could partly be attributed to the use of a different atmospheric dataset than the one originally used to drive the E-HYPE model.
-However, by applying a rolling mean with a 5-day window, the prediction error is reduced to less than 1%, which is acceptable for the purposes of climate modeling.
+However, by applying a rolling mean with a 5-day window, the prediction error is reduced to less than 1$\%$, which is acceptable for the purposes of climate modeling.
 Lastly, the right panel demonstrates that the distribution of residuals follows a Gaussian shape, suggesting that our model does not exhibit bias by systematically over or underestimating river runoff values.
@@
 Figure 4b compares these results by showing the percentage difference in vertically averaged salinity between the ConvLSTM simulation and the reference simulation.
-Overall, the differences remain below 1%, except near a river mouth in the Gulf of Riga (22-24°E, 56.5-58.5°N for orientation), where the difference is approximately 1%.
+Overall, the differences remain below 1$\%$, except near a river mouth in the Gulf of Riga (22-24°E, 56.5-58.5°N for orientation), where the difference is approximately 1$\%$.
 
@@
 In summary, we showed that the ConvLSTM model demonstrated robust performance in forecasting river runoff across 97 rivers entering the Baltic Sea.
-Trained on data from 1979 to 2011, the model achieved an impressive daily prediction accuracy of ±5%.
+Trained on data from 1979 to 2011, the model achieved an impressive daily prediction accuracy of ±5$\%$.
 This capability to generate accurate and detailed simulations enables to examine the potential impacts of different climate scenarios.

### 2024-05-27 paper/paper.tex
went through text

@@
 
-\section{Introduction}\label{introduction}
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden, boxrule=0pt, sharp corners]}{\end{tcolorbox}}\fi
+
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
@@
 precipitation (precipitation minus evaporation) over the sea surface.
-The net precipitation accounts for 11 \% and the river input for 89 \%
-of the total freshwater input (Meier \& Döscher, 2002). Modelling the
-Baltic Sea is therefore to a large part the result of the quality of the
-river input, that is used for the simulation. This makes the accurate
-modeling of river runoff especially critical for simulations pertaining
-to the Baltic Sea.
+The net precipitation accounts for 11 \(\%\) and the river input for 89
+\(\%\) of the total freshwater input (Meier \& Döscher, 2002). Modelling
+the Baltic Sea is therefore to a large part the result of the quality of
+the river input, that is used for the simulation. This makes the
+accurate modeling of river runoff especially critical for simulations
+pertaining to the Baltic Sea.
 
@@
 
+\hypertarget{implemented-model-architecture}{%
 \section{Implemented model
-architecture}\label{implemented-model-architecture}
+architecture}\label{implemented-model-architecture}}
 
-\subsection{The main idea}\label{sec-main_idea}
+\hypertarget{sec-main_idea}{%
+\subsection{The main idea}\label{sec-main_idea}}
 
-We assume that the runoff for a specific point in time for all \(N_r\)
+We assume that the runoff at a specific point in time for all \(N_r\)
 considered rivers, collected in the vector
@@
 functional \(\vec{M}(\{X^t_k[x,y,\tau]\})\) of \(k=1,\ldots N_k\)
-atmospheric fields \(X^t_k[x,y,\tau]\) that are known for the past
-\(\tau=1,\ldots N_\tau\) time instances, i.e.
+atmospheric fields \(X^t_k[x,y,\tau]\) which are known for the past
+\(\tau=1,\ldots N_\tau\) time instances. This relationship is expressed
+as:
 
@@
 
-The fields are evaluated on a spatial domain \(x=1,\ldots N_x\) and
-\(y=1,\ldots N_y\) that is sufficiently large to capture all significant
-local and non-local contributions of atmospheric fields to the river
-runoff.
+The atmospheric fields are evaluated over a spatial domain
+\(x=1,\ldots N_x\) and \(y=1,\ldots N_y\) which is sufficiently large to
+capture all significant local and non-local contributions of the
+atmospheric fields to the river runoff.
 
-Usually such a mapping is realized in a suitable hydrological model that
-simulates all relevant physical processes that transform variables like,
-e.g., precipitation and evaproration to the river runoff, relying on
-expertise for tuning all appearing parameters to reasonable values.
+Typically, such a mapping is realized using a hydrological model that
+simulates all relevant physical processes, transforming variables like
+precipitation and evaporation into river runoff. This process relies
+heavily on domain knowledge to tune all parameters to reasonable values.
 
@@
 represented by a combination of a ConvLSTM with a subsequent FC network.
-That way, there is no need to have knowledge of involved processes and
-how they can be modeled, instead, these features can be ``lerned'' by
-the network in an automated manner. Our poroposed network architecture
+This approach eliminates the need for detailed knowledge of the involved
+processes and their modeling. Instead, these features can be ``learned''
+by the network in an automated manner. Our proposed network architecture
 is visualized in Figure~\ref{fig-ConvLSTM_FC} and described in detail in
-the following. In order to provide an overview, we will go through the
-main ingredients of this architecture one-by-one.
+the following sections. To provide an overview, we will discuss the main
+components of this architecture one-by-one.
 
@@
 
-\centering{
-
-\includegraphics{./ConvLSTM_FC.png}
+{\centering \includegraphics{./ConvLSTM_FC.png}
 
@@
 
-\end{figure}%
+\end{figure}
 
-\subsection{ConvLSTM network}\label{convlstm-network}
+\hypertarget{convlstm-network}{%
+\subsection{ConvLSTM network}\label{convlstm-network}}
 
-\subsubsection{The LSTM approach}\label{sec-LSTM}
+\hypertarget{sec-LSTM}{%
+\subsubsection{The LSTM approach}\label{sec-LSTM}}
 
 Before turning directly to the convolutional LSTM, the simpler
-architecture of the plain LSTM is examined and should serve as a role
-model for the later consideration of the full ConvLSTM.
+architecture of the plain LSTM is examined which serves as a foundation
+for understanding the more complex ConvLSTM.
 
 The Long Short-Term Memory (LSTM), a specialized form of Recurrent
-Neural Networks (RNNs), is specifically tailored for modeling temporal
+Neural Networks (RNNs), is specifically designed to model temporal
 sequences
@@
 \(N_\tau\) input quantities
-\(\vec{X}^t[\tau] = (X^t_k[\tau]) \in \mathbb{R}^{N_k}\). The sequence
+\(\vec{X}^t[\tau] = (X^t_k[\tau]) \in \mathbb{R}^{N_k}\). This sequence
 is taken from a dataset given in form of a time series \(\{\vec{X}^t\}\)
-and the endpoint should coincide with the specific element in the time
+with the endpoint coinciding with the specific element in the time
 series connected to time \(t\),
 i.e.~\(\vec{X}^t[N_\tau] \equiv \vec{X}^t\), see
-Figure~\ref{fig-ConvLSTM_FC}. The \(N_k\) is the number of used input
-``channels'' and can, for example, represent different measurable
-quantities. Its unique design allows it to adeptly handle long-range
-dependencies, setting it apart from traditional RNNs in terms of
-accuracy (see Figure~\ref{fig-lstm}).
+Figure~\ref{fig-ConvLSTM_FC}. Here, \(N_k\) represents the number of
+input ``channels,'' which can correspond to different measurable
+quantities. The LSTM's unique design allows it to adeptly handle
+long-range dependencies, setting it apart from traditional RNNs in terms
+of accuracy (see Figure~\ref{fig-lstm}).
 
@@
 
-\centering{
-
-\includegraphics{./LSTM.png}
+{\centering \includegraphics{./LSTM.png}
 
@@
 of the previous (long-term memory) cell state \(\vec{C}^t[\tau-1]\) that
-should be kept, see dashed cyan box therin\hspace{0pt}. The input gate
+should be kept, see dashed cyan box therein\hspace{0pt}. The input gate
 \(\vec{I}^t[\tau]\) controls the contribution of the current input used
@@
 where \(g=i,f,o,c\) stands for the input, forget, output and cell-state
-gate, respectively and Eistein's summation convention is employed,
+gate, respectively and Einstein's summation convention is employed,
 i.e.~indices that appear twice are summed over. The calligraphic symbols
@@
 optimized for the given problem during the training, which is at the
-heart of the any machine learning approach. The matrix
+heart of any machine learning approach. The matrix
 \(\pmb{\mathcal{M}}^{g} = (\mathcal{M}^{g}_{hk}) \in \mathbb{R}^{N_h \times N_k}\)
@@
 \end{aligned}
-}\end{equation}
+}\label{eq-LSTM}\end{equation}
 
-where \(\sigma\) denotes the logisitc sigmoid function, \(\tanh\) is the
+where \(\sigma\) denotes the logistic sigmoid function, \(\tanh\) is the
 hyperbolic tangent and the \(\circ\) stands for the Hadamard product
@@
 
-\begin{equation}\phantomsection\label{eq-L_nested}{
+\begin{equation}\protect\hypertarget{eq-L_nested}{}{
 \left(\vec{H}^t[N_\tau], \vec{C}^t[N_\tau] \right) = L \left( \vec{X}^t[N_\tau], L \left( \vec{X}^t[N_\tau-1], \ldots L \left( \vec{X}^t[1], (\vec{H}^t[0], \vec{C}^t[0]) \right) \ldots \right) \right) \ ,
-}\end{equation}
+}\label{eq-L_nested}\end{equation}
 
@@
 
+\hypertarget{combining-lstm-with-spatial-convolution}{%
 \subsubsection{Combining LSTM with spatial
-convolution}\label{combining-lstm-with-spatial-convolution}
+convolution}\label{combining-lstm-with-spatial-convolution}}
 
@@
 sequences of point-like quantities it is not designed to recognize
-spatial features in seuqences of, e.g., two-dimensional maps as
-atmospheric-ocean interface fields. To adress this limitation we employ
+spatial features in sequences of, e.g., two-dimensional maps as
+atmospheric-ocean interface fields. To address this limitation we employ
 a ConvLSTM architecture as described in the following.
@@
 ``learning'' of spatial patterns, the free parameters of the network are
-replaced by two-dimenmsional convolutional kernels
+replaced by two-dimensional convolution kernels
 \(\pmb{\mathcal{M}}^{g} = (\mathcal{M}^{g}_{hk}[\xi, \eta]) \in \mathbb{R}^{(N_h \times N_k)\times (N_\xi \times N_\eta)}\)
@@
 (\(\in \mathbb{R}^{N_h \times (N_x \times N_y)}\)) as well. We can write
-this mapping in the same fashion as Equation~\ref{eq-LSTM-gates} but
-with replacing the normal matrix-vector multiplication by a convolution
+this mapping in the same way as Equation~\ref{eq-LSTM-gates} but with
+replacing the normal matrix-vector multiplication by a convolution
 (denoted with \(\ast\)), i.e.\\
@@
 
-The susebsequent processing of the \(\vec{g}^t[\tau]\) remains
+The subsequent processing of the \(\vec{g}^t[\tau]\) remains
 symbolically the same as presented in Equation~\ref{eq-LSTM} but with
@@
 
-\subsection{Fully connected layer}\label{sec-FC_layer}
+\hypertarget{sec-FC_layer}{%
+\subsection{Fully connected layer}\label{sec-FC_layer}}
 
@@
 
-\section{Technical details}\label{sec-technical_details}
+\hypertarget{sec-technical_details}{%
+\section{Technical details}\label{sec-technical_details}}
 
+\hypertarget{runoff-data-used-for-training}{%
 \subsection{Runoff data used for
-training}\label{runoff-data-used-for-training}
+training}\label{runoff-data-used-for-training}}
 
@@
 after (2012 to 2018) additional spatial temporal corrections have been
-applied to the runoff data, and have therefore been ignored. For more
-information see Gröger et al. (2022) and references therein.
+applied to the runoff data, and have therefore been ignored. The quality
+of the runoff was extensively evaluated. For more information see Gröger
+et al. (2022) and references therein.
 
-\subsection{Atmospheric Forcing}\label{atmospheric-forcing}
+\hypertarget{atmospheric-forcing}{%
+\subsection{Atmospheric Forcing}\label{atmospheric-forcing}}
 
@@
 
+\hypertarget{neural-network-hyper-parameters}{%
 \subsection{Neural network hyper
-parameters}\label{neural-network-hyper-parameters}
+parameters}\label{neural-network-hyper-parameters}}
 
@@
 
-\centering{
-
-\includegraphics{ConvLSTM.png}
+{\centering \includegraphics{ConvLSTM.png}
 
@@
 
+\hypertarget{tbl-letters}{}
 \begin{longtable}[]{@{}ll@{}}
-\caption{Hyperparameters}\label{tbl-letters}\tabularnewline
-\toprule\noalign{}
+\caption{\label{tbl-letters}Hyperparameters}\tabularnewline
+\toprule()
 Parameter name & Parameter size \\
-\midrule\noalign{}
+\midrule()
 \endfirsthead
-\toprule\noalign{}
+\toprule()
 Parameter name & Parameter size \\
-\midrule\noalign{}
+\midrule()
 \endhead
-\bottomrule\noalign{}
-\endlastfoot
 Channel size & 4 \\
@@
 Learning Rate & 1e-3 with CosineAnnealing \\
+\bottomrule()
 \end{longtable}
@@
 runoff calculation in the atmospheric model COSMO-CLM which uses these
-four fields to calculate an river runoff estimate.
+atmospheric fields to calculate an river runoff estimate.
 
-\section{Results}\label{results}
+\hypertarget{results}{%
+\section{Results}\label{results}}
 
@@
 correction applied to the runoff to match observations. The data was
-divided into randomly chosen splits of 80\% training data, 10\%
+divided into randomly chosen splits of 80\(\%\) training data, 10\(\%\)
 validation data to evaluate the performance of the model during
-training, and 10\% training data which is finally used to evaluate the
-performance of the model after training. The model was trained for 400
-epochs and the model weights with the lowest mean squared error during
-training have been stored.
+training, and 10\(\%\) training data which is finally used to evaluate
+the performance of the model after training. The model was trained for
+400 epochs and the model weights with the lowest mean squared error
+during training have been stored.
 
@@
 that, on daily timescales, the model can predict river runoff with an
-accuracy of \(\pm\) 5\%. The overall correlation is 0.997 with the
+accuracy of \(\pm\) 5\(\%\). The overall correlation is 0.997 with the
 resulting error metrics yielding a RMSE of 323.99 \(m^3\)/s and MAE of
@@
 rolling mean with a 5-day window, the prediction error is reduced to
-less than 1\%, which is acceptable for the purposes of climate modeling.
-Lastly, the right panel demonstrates that the distribution of residuals
-follows a Gaussian shape, suggesting that our model does not exhibit
-bias by systematically over or underestimating river runoff values.
+less than 1\(\%\), which is acceptable for the purposes of climate
+modeling. Lastly, the right panel demonstrates that the distribution of
+residuals follows a Gaussian shape, suggesting that our model does not
+exhibit bias by systematically over or underestimating river runoff
+values.
 
@@
 vertically averaged salinity between the ConvLSTM simulation and the
-reference simulation. Overall, the differences remain below 1\%, except
-near a river mouth in the Gulf of Riga (22-24°E, 56.5-58.5°N for
-orientation), where the difference is approximately 1\%.
+reference simulation. Overall, the differences remain below 1\(\%\),
+except near a river mouth in the Gulf of Riga (22-24°E, 56.5-58.5°N for
+orientation), where the difference is approximately 1\(\%\).
 
@@
 
-\section{Discussion}\label{discussion}
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
 
@@
 Baltic Sea. Trained on data from 1979 to 2011, the model achieved an
-impressive daily prediction accuracy of ±5\%. This capability to
+impressive daily prediction accuracy of ±5\(\%\). This capability to
 generate accurate and detailed simulations enables to examine the
@@
 
-\section{Acknowledgments}\label{acknowledgments}
+\hypertarget{acknowledgments}{%
+\section{Acknowledgments}\label{acknowledgments}}
 
@@
 
-\bibitem[\citeproctext]{ref-ashrafi2017fully}
+\leavevmode\vadjust pre{\hypertarget{ref-ashrafi2017fully}{}}%
 Ashrafi, M., Chua, L. H. C., Quek, C., \& Qin, X. (2017). A fully-online
@@
 
-\bibitem[\citeproctext]{ref-dee2011}
+\leavevmode\vadjust pre{\hypertarget{ref-dee2011}{}}%
 Dee, D. P., Uppala, S. M., Simmons, A. J., Berrisford, P., Poli, P.,
@@
 
-\bibitem[\citeproctext]{ref-fang2022application}
+\leavevmode\vadjust pre{\hypertarget{ref-fang2022application}{}}%
 Fang, L., \& Shao, D. (2022). Application of long short-term memory
@@
 
-\bibitem[\citeproctext]{ref-fangExaminingApplicabilityDifferent2019}
+\leavevmode\vadjust pre{\hypertarget{ref-fangExaminingApplicabilityDifferent2019}{}}%
 Fang, W., Huang, S., Ren, K., Huang, Q., Huang, G., Cheng, G., \& Li, K.
@@
 
-\bibitem[\citeproctext]{ref-gruxf6ger2022}
+\leavevmode\vadjust pre{\hypertarget{ref-gruxf6ger2022}{}}%
 Gröger, M., Placke, M., Meier, M., Börgel, F., Brunnabend, S.-E.,
@@
 
-\bibitem[\citeproctext]{ref-ha2021}
+\leavevmode\vadjust pre{\hypertarget{ref-ha2021}{}}%
 Ha, S., Liu, D., \& Mu, L. (2021). Prediction of Yangtze River
@@
 
-\bibitem[\citeproctext]{ref-hagemannHighResolutionDischarge2020}
+\leavevmode\vadjust pre{\hypertarget{ref-hagemannHighResolutionDischarge2020}{}}%
 Hagemann, S., Stacke, T., \& Ho-Hagemann, H. T. M. (2020). High
@@
 
-\bibitem[\citeproctext]{ref-hochreiter1997long}
+\leavevmode\vadjust pre{\hypertarget{ref-hochreiter1997long}{}}%
 Hochreiter, S., \& Schmidhuber, J. (1997). Long short-term memory.
@@
 
-\bibitem[\citeproctext]{ref-hordoir2019}
+\leavevmode\vadjust pre{\hypertarget{ref-hordoir2019}{}}%
 Hordoir, R., Axell, L., Höglund, A., Dieterich, C., Fransner, F.,
@@
 
-\bibitem[\citeproctext]{ref-huang2014monthly}
+\leavevmode\vadjust pre{\hypertarget{ref-huang2014monthly}{}}%
 Huang, S., Chang, J., Huang, Q., \& Chen, Y. (2014). Monthly streamflow
@@
 
-\bibitem[\citeproctext]{ref-humphrey2016hybrid}
+\leavevmode\vadjust pre{\hypertarget{ref-humphrey2016hybrid}{}}%
 Humphrey, G. B., Gibbs, M. S., Dandy, G. C., \& Maier, H. R. (2016). A
@@
 
-\bibitem[\citeproctext]{ref-kratzert2018rainfall}
+\leavevmode\vadjust pre{\hypertarget{ref-kratzert2018rainfall}{}}%
 Kratzert, F., Klotz, D., Brenner, C., Schulz, K., \& Herrnegger, M.
@@
 
-\bibitem[\citeproctext]{ref-liu2020streamflow}
+\leavevmode\vadjust pre{\hypertarget{ref-liu2020streamflow}{}}%
 Liu, D., Jiang, W., Mu, L., \& Wang, S. (2020). Streamflow prediction
@@
 
-\bibitem[\citeproctext]{ref-meier2002simulated}
+\leavevmode\vadjust pre{\hypertarget{ref-meier2002simulated}{}}%
 Meier, H. M., \& Döscher, R. (2002). Simulated water and heat cycles of
@@
 
-\bibitem[\citeproctext]{ref-moishin2021designing}
+\leavevmode\vadjust pre{\hypertarget{ref-moishin2021designing}{}}%
 Moishin, M., Deo, R. C., Prasad, R., Raj, N., \& Abdulla, S. (2021).
@@
 
-\bibitem[\citeproctext]{ref-shiConvolutionalLSTMNetwork2015}
+\leavevmode\vadjust pre{\hypertarget{ref-shiConvolutionalLSTMNetwork2015}{}}%
 SHI, X., Chen, Z., Wang, H., Yeung, D.-Y., Wong, W., \& WOO, W. (2015).
@@
 
-\bibitem[\citeproctext]{ref-tanAdaptiveMiddleLongterm2018}
+\leavevmode\vadjust pre{\hypertarget{ref-tanAdaptiveMiddleLongterm2018}{}}%
 Tan, Q.-F., Lei, X.-H., Wang, X., Wang, H., Wen, X., Ji, Y., \& Kang,
@@
 
-\bibitem[\citeproctext]{ref-theross}
+\leavevmode\vadjust pre{\hypertarget{ref-theross}{}}%
 The rossby centre regional climate model RCA3: Model description and
@@
 
-\bibitem[\citeproctext]{ref-zhu2023spatiotemporal}
+\leavevmode\vadjust pre{\hypertarget{ref-zhu2023spatiotemporal}{}}%
 Zhu, S., Wei, J., Zhang, H., Xu, Y., \& Qin, H. (2023). Spatiotemporal

### 2024-06-07 paper/paper.qmd
Update

@@
 ---
-title: "From Precipitation to Prediction: Integrating ConvLSTM Models for Comprehensive River Runoff Forecasting"
+title: "From Precipitation to Prediction: Using ConvLSTM Models for Comprehensive River Runoff Forecasting"
 format:
@@
 River runoff is an important component of the global water cycle as it comprises about one third of the precipitation over land areas [@hagemannHighResolutionDischarge2020].
+Therefore, accurate runoff forecasting is crucial for effective water resources management, particularly over extended periods [@fangExaminingApplicabilityDifferent2019; @tanAdaptiveMiddleLongterm2018].
 In the context of climate change studies, river runoff is usually generated in two ways.
 First, river runoff as input for ocean models can be created using hydrological models such as the Hydrological Discharge (HD) model [@hagemannHighResolutionDischarge2020].
-HD models calculates the water balance using hydrological processes (e.g.
+HD models calculate the water balance using hydrological processes (e.g.
 snow, glaciers, soil moisture, groundwater contribution).
 It represents a complex forecasting tool that uses underlying physical processes.
-A different approach would use data-based models that integrate statistical correction, using the land surface schemes of global or regional climate models.
+The second approach uses data-driven models that integrate statistical correction, using land surface schemes of global or regional climate models \[QUELLE\].
 
-The relatively recent emergence of machine learning (ML) models has offered different approaches to river runoff forecasting.
-Accurate runoff forecasting is crucial for effective water resources management, particularly over extended periods [@fangExaminingApplicabilityDifferent2019; @tanAdaptiveMiddleLongterm2018].
+With the recent rise of machine learning in climate research various model architectures have also been tested for river runoff forecasting.
 Common approaches employ feed-forward artificial neural networks, support vector machines, adaptive neuro-fuzzy inference systems, and notably, Long Short-Term Memory (LSTM) neural networks that have gained traction for long-term hydrological forecasting due to their excellent performance [@humphrey2016hybrid; @huang2014monthly; @ashrafi2017fully; @liu2020streamflow; @fang2022application; @kratzert2018rainfall].
@@
 We use the Baltic Sea catchment as an example to illustrate our approach.
-Although the methodology we propose is universally applicable across various geographic regions, the Baltic Sea presents a challenge due to its unique hydrological characteristics, being nearly decoupled from the open ocean (see Figure).
+Although the methodology we propose is universally applicable across various geographic regions, the Baltic Sea represents a challenging region due to its unique hydrological characteristics, being nearly decoupled from the open ocean (see Figure).
 As a consequence, the salinity of the Baltic Sea is driven to a large part by freshwater supply from rivers.
@@
 
-As an alternative, we propose that this functional can be adequately represented by a combination of a ConvLSTM with a subsequent FC network.
+As an alternative, we propose that this functional can be adequately represented by a combination of a convolutional Long-Short Term Memory (ConvLSTM) model with a subsequent fully connected neural network.
 This approach eliminates the need for detailed knowledge of the involved processes and their modeling.
@@
 
-Before turning directly to the convolutional LSTM, the simpler architecture of the plain LSTM is examined which serves as a foundation for understanding the more complex ConvLSTM.
+Before turning directly to the ConvLSTM the simpler architecture of the plain Long-Short Term Memory (LSTM) model is examined which serves as a foundation for understanding the more complex ConvLSTM.
 
-The Long Short-Term Memory (LSTM), a specialized form of Recurrent Neural Networks (RNNs), is specifically designed to model temporal sequences $\vec{X}^t[1], \ldots \vec{X}^t[\tau],\ldots \vec{X}^t[N_\tau]$ of $N_\tau$ input quantities $\vec{X}^t[\tau] = (X^t_k[\tau]) \in \mathbb{R}^{N_k}$.
+The LSTM, a specialized form of Recurrent Neural Networks (RNNs), is specifically designed to model temporal sequences $\vec{X}^t[1], \ldots \vec{X}^t[\tau],\ldots \vec{X}^t[N_\tau]$ of $N_\tau$ input quantities $\vec{X}^t[\tau] = (X^t_k[\tau]) \in \mathbb{R}^{N_k}$.
 This sequence is taken from a dataset given in form of a time series $\{\vec{X}^t\}$ with the endpoint coinciding with the specific element in the time series connected to time $t$, i.e.
@@
 
-As input the model receives 30 days of atmospheric surface fields temperature $T$, precipitation $P$, specific humidity $Q$ and wind speed $W$, with a daily resolution to predict the river runoff $R$ at the time step $\Delta t+1$ , which can be summarized as
-
-$R_{\Delta t+1} = f\left(T_{t-30:t}, P_{t-30:t}, Q_{t-30:t}, W_{t-30:t}\right)$
-
-with f being a function maps the 30 days of daily atmospheric surface fields data to the predicted river runoff.
-
+As input the model receives $N_{\tau}$ = 30 days of atmospheric surface fields temperature $T$, precipitation $P$, specific humidity $Q$ and wind speed $W$, with a daily resolution to predict the river runoff $\vec{R}$.
 The choice of atmospheric fields was based on the implemented river runoff calculation in the atmospheric model COSMO-CLM which uses these atmospheric fields to calculate an river runoff estimate.
@@
 As mention above for evaluation, the test dataset was utilized.
-The left panel panel (Figure @fig-statistical-evaluationNN a) ) illustrates the relative prediction error in relation to the original E-HYPE data, indicating that, on daily timescales, the model can predict river runoff with an accuracy of $\pm$ 5$\%$.
+The left panel panel (Figure @fig-statistical-evaluationNN a) illustrates the relative prediction error in relation to the original E-HYPE data, indicating that, on daily timescales, the model can predict river runoff with an accuracy of $\pm$ 5$\%$.
 The overall correlation is 0.997 with the resulting error metrics yielding a RMSE of 323.99 $m^3$/s and MAE of 249.51 $m^3$/s.
@@
 The quality of a projection for a coastal sea such as the Baltic Sea is too large parts based on the quality of the hyrdological conditions.
-In this work we analyze the implementation of Convolutional Long Short-Term Memory (ConvLSTM) networks for predicting river runoff, highlighting its potential to enhance river runoff forecasting across different coastal seas.
+In this work we analyze the implementation of a ConvLSTM networks for predicting river runoff, highlighting its potential to enhance river runoff forecasting across different coastal seas.
 Given the unique hydrological characteristics of the Baltic Sea, largely influenced by its limited connection to the open ocean and significant freshwater input from surrounding rivers, the region presents a critical case for the application of sophisticated forecasting models.
@@
 
-Phasellus interdum tincidunt ex, a euismod massa pulvinar at.
-Ut fringilla ut nisi nec volutpat.
-Morbi imperdiet congue tincidunt.
-Vivamus eget rutrum purus.
-Etiam et pretium justo.
-Donec et egestas sem.
-Donec molestie ex sit amet viverra egestas.
-Nullam justo nulla, fringilla at iaculis in, posuere non mauris.
-Ut eget imperdiet elit.
-
-## Open research
-
-Phasellus interdum tincidunt ex, a euismod massa pulvinar at.
-Ut fringilla ut nisi nec volutpat.
-Morbi imperdiet congue tincidunt.
-Vivamus eget rutrum purus.
-Etiam et pretium justo.
-Donec et egestas sem.
-Donec molestie ex sit amet viverra egestas.
-Nullam justo nulla, fringilla at iaculis in, posuere non mauris.
-Ut eget imperdiet elit.
+The research presented in this study is part of the Baltic Earth program (Earth System Science for the Baltic Sea region, see [https://www.baltic.earth](https://www.baltic.earth/).
 

### 2024-06-07 paper/paper.tex
Update

@@
 \hypersetup{
-  pdftitle={From Precipitation to Prediction: Integrating ConvLSTM Models for Comprehensive River Runoff Forecasting},
+  pdftitle={From Precipitation to Prediction: Using ConvLSTM Models for Comprehensive River Runoff Forecasting},
   pdfauthor={Florian Börgel; Sven Karsten; Karoline Rummel},
@@
 \begin{document}
-\title{From Precipitation to Prediction: Integrating ConvLSTM Models for
+\title{From Precipitation to Prediction: Using ConvLSTM Models for
 Comprehensive River Runoff Forecasting}
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden, boxrule=0pt, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, enhanced, boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 comprises about one third of the precipitation over land areas (Hagemann
-et al., 2020). In the context of climate change studies, river runoff is
-usually generated in two ways. First, river runoff as input for ocean
-models can be created using hydrological models such as the Hydrological
-Discharge (HD) model (Hagemann et al., 2020). HD models calculates the
-water balance using hydrological processes (e.g.~snow, glaciers, soil
-moisture, groundwater contribution). It represents a complex forecasting
-tool that uses underlying physical processes. A different approach would
-use data-based models that integrate statistical correction, using the
-land surface schemes of global or regional climate models.
+et al., 2020). Therefore, accurate runoff forecasting is crucial for
+effective water resources management, particularly over extended periods
+(Fang et al., 2019; Tan et al., 2018). In the context of climate change
+studies, river runoff is usually generated in two ways. First, river
+runoff as input for ocean models can be created using hydrological
+models such as the Hydrological Discharge (HD) model (Hagemann et al.,
+2020). HD models calculate the water balance using hydrological
+processes (e.g.~snow, glaciers, soil moisture, groundwater
+contribution). It represents a complex forecasting tool that uses
+underlying physical processes. The second approach uses data-driven
+models that integrate statistical correction, using land surface schemes
+of global or regional climate models {[}QUELLE{]}.
 
-The relatively recent emergence of machine learning (ML) models has
-offered different approaches to river runoff forecasting. Accurate
-runoff forecasting is crucial for effective water resources management,
-particularly over extended periods (W. Fang et al., 2019; Tan et al.,
-2018). Common approaches employ feed-forward artificial neural networks,
+With the recent rise of machine learning in climate research various
+model architectures have also been tested for river runoff forecasting.
+Common approaches employ feed-forward artificial neural networks,
 support vector machines, adaptive neuro-fuzzy inference systems, and
@@
 traction for long-term hydrological forecasting due to their excellent
-performance (Ashrafi et al., 2017; L. Fang \& Shao, 2022; Huang et al.,
-2014; Humphrey et al., 2016; Kratzert et al., 2018; Liu et al., 2020).
+performance (\textbf{humphrey2016hybrid?}; \textbf{huang2014monthly?};
+\textbf{ashrafi2017fully?}; \textbf{liu2020streamflow?};
+\textbf{fang2022application?}; \textbf{kratzert2018rainfall?}).
 
-LSTM networks, first introduced by Hochreiter \& Schmidhuber (1997) ,
-are an evolution of the classical Recurrent Neural Networks (RNNs).
-Their structure enables them to learn long-term dependencies while
-avoiding the vanishing or exploding gradient problem. They have shown
-stability and efficacy in sequence-to-sequence predictions. However, a
-limitation of LSTMs is their inability to effectively capture
-two-dimensional structures, an area where Convolutional Neural Networks
-(CNNs) excel. Recognizing this limitation SHI et al. (2015) proposed a
-Convolutional LSTM (ConvLSTM) architecture, which combines the strengths
-of both LSTM and CNN. The ConvLSTM network has been proven useful for
-spatio-temporal applications such as precipitation nowcasting (SHI et
-al., 2015), flood forecasting (Moishin et al., 2021), and river runoff
-forecasting (Ha et al., 2021; Zhu et al., 2023).
+LSTM networks, first introduced by (\textbf{hochreiter1997long?}) , are
+an evolution of the classical Recurrent Neural Networks (RNNs). Their
+structure enables them to learn long-term dependencies while avoiding
+the vanishing or exploding gradient problem. They have shown stability
+and efficacy in sequence-to-sequence predictions. However, a limitation
+of LSTMs is their inability to effectively capture two-dimensional
+structures, an area where Convolutional Neural Networks (CNNs) excel.
+Recognizing this limitation SHI et al. (2015) proposed a Convolutional
+LSTM (ConvLSTM) architecture, which combines the strengths of both LSTM
+and CNN. The ConvLSTM network has been proven useful for spatio-temporal
+applications such as precipitation nowcasting (SHI et al., 2015), flood
+forecasting (\textbf{moishin2021designing?}), and river runoff
+forecasting (\textbf{ha2021?}; \textbf{zhu2023spatiotemporal?}).
 
@@
 methodology we propose is universally applicable across various
-geographic regions, the Baltic Sea presents a challenge due to its
-unique hydrological characteristics, being nearly decoupled from the
-open ocean (see Figure). As a consequence, the salinity of the Baltic
-Sea is driven to a large part by freshwater supply from rivers.
+geographic regions, the Baltic Sea represents a challenging region due
+to its unique hydrological characteristics, being nearly decoupled from
+the open ocean (see Figure). As a consequence, the salinity of the
+Baltic Sea is driven to a large part by freshwater supply from rivers.
 Freshwater enters the Baltic Sea through river runoff or positive net
@@
 The net precipitation accounts for 11 \(\%\) and the river input for 89
-\(\%\) of the total freshwater input (Meier \& Döscher, 2002). Modelling
-the Baltic Sea is therefore to a large part the result of the quality of
-the river input, that is used for the simulation. This makes the
-accurate modeling of river runoff especially critical for simulations
-pertaining to the Baltic Sea.
+\(\%\) of the total freshwater input (\textbf{meier2002simulated?}).
+Modelling the Baltic Sea is therefore to a large part the result of the
+quality of the river input, that is used for the simulation. This makes
+the accurate modeling of river runoff especially critical for
+simulations pertaining to the Baltic Sea.
 
@@
 As an alternative, we propose that this functional can be adequately
-represented by a combination of a ConvLSTM with a subsequent FC network.
-This approach eliminates the need for detailed knowledge of the involved
+represented by a combination of a convolutional Long-Short Term Memory
+(ConvLSTM) model with a subsequent fully connected neural network. This
+approach eliminates the need for detailed knowledge of the involved
 processes and their modeling. Instead, these features can be ``learned''
@@
 
-Before turning directly to the convolutional LSTM, the simpler
-architecture of the plain LSTM is examined which serves as a foundation
-for understanding the more complex ConvLSTM.
+Before turning directly to the ConvLSTM the simpler architecture of the
+plain Long-Short Term Memory (LSTM) model is examined which serves as a
+foundation for understanding the more complex ConvLSTM.
 
-The Long Short-Term Memory (LSTM), a specialized form of Recurrent
-Neural Networks (RNNs), is specifically designed to model temporal
-sequences
+The LSTM, a specialized form of Recurrent Neural Networks (RNNs), is
+specifically designed to model temporal sequences
 \(\vec{X}^t[1], \ldots \vec{X}^t[\tau],\ldots \vec{X}^t[N_\tau]\) of
@@
 Second, the actual gate vectors are computed by the core equations of
-the LSTM as proposed by Hochreiter \& Schmidhuber (1997):
+the LSTM as proposed by (\textbf{hochreiter1997long?}):
 
@@
 hindcast simulation that was forced by a regional downscaling of
-ERA-Interim (Dee et al., 2011) with RCA3 ({``The rossby centre regional
-climate model RCA3,''} n.d.) and implemented into NEMO-Nordic (Hordoir
-et al., 2019) as a mass flux. For the periods before (1961 to 1978) and
-after (2012 to 2018) additional spatial temporal corrections have been
-applied to the runoff data, and have therefore been ignored. The quality
-of the runoff was extensively evaluated. For more information see Gröger
-et al. (2022) and references therein.
+ERA-Interim (\textbf{dee2011?}) with RCA3 (\textbf{theross?}) and
+implemented into NEMO-Nordic (\textbf{hordoir2019?}) as a mass flux. For
+the periods before (1961 to 1978) and after (2012 to 2018) additional
+spatial temporal corrections have been applied to the runoff data, and
+have therefore been ignored. The quality of the runoff was extensively
+evaluated. For more information see (\textbf{gröger2022?}) and
+references therein.
 
@@
 Baltic Sea to the North Sea. The maximum depth was set at 264 meters. A
-more detailed description of the setup can be found in (Gröger et al.,
-2022).
+more detailed description of the setup can be found in
+(\textbf{gröger2022?}).
 
@@
 
-As input the model receives 30 days of atmospheric surface fields
-temperature \(T\), precipitation \(P\), specific humidity \(Q\) and wind
-speed \(W\), with a daily resolution to predict the river runoff \(R\)
-at the time step \(\Delta t+1\) , which can be summarized as
-
-\(R_{\Delta t+1} = f\left(T_{t-30:t}, P_{t-30:t}, Q_{t-30:t}, W_{t-30:t}\right)\)
-
-with f being a function maps the 30 days of daily atmospheric surface
-fields data to the predicted river runoff.
-
-The choice of atmospheric fields was based on the implemented river
-runoff calculation in the atmospheric model COSMO-CLM which uses these
-atmospheric fields to calculate an river runoff estimate.
+As input the model receives \(N_{\tau}\) = 30 days of atmospheric
+surface fields temperature \(T\), precipitation \(P\), specific humidity
+\(Q\) and wind speed \(W\), with a daily resolution to predict the river
+runoff \(\vec{R}\). The choice of atmospheric fields was based on the
+implemented river runoff calculation in the atmospheric model COSMO-CLM
+which uses these atmospheric fields to calculate an river runoff
+estimate.
 
@@
 evaluation, the test dataset was utilized. The left panel panel (Figure
-\textbf{?@fig-statistical-evaluationNN} a) ) illustrates the relative
+\textbf{?@fig-statistical-evaluationNN} a) illustrates the relative
 prediction error in relation to the original E-HYPE data, indicating
@@
 large parts based on the quality of the hyrdological conditions. In this
-work we analyze the implementation of Convolutional Long Short-Term
-Memory (ConvLSTM) networks for predicting river runoff, highlighting its
-potential to enhance river runoff forecasting across different coastal
-seas. Given the unique hydrological characteristics of the Baltic Sea,
-largely influenced by its limited connection to the open ocean and
-significant freshwater input from surrounding rivers, the region
-presents a critical case for the application of sophisticated
-forecasting models.
+work we analyze the implementation of a ConvLSTM networks for predicting
+river runoff, highlighting its potential to enhance river runoff
+forecasting across different coastal seas. Given the unique hydrological
+characteristics of the Baltic Sea, largely influenced by its limited
+connection to the open ocean and significant freshwater input from
+surrounding rivers, the region presents a critical case for the
+application of sophisticated forecasting models.
 
@@
 
-Phasellus interdum tincidunt ex, a euismod massa pulvinar at. Ut
-fringilla ut nisi nec volutpat. Morbi imperdiet congue tincidunt.
-Vivamus eget rutrum purus. Etiam et pretium justo. Donec et egestas sem.
-Donec molestie ex sit amet viverra egestas. Nullam justo nulla,
-fringilla at iaculis in, posuere non mauris. Ut eget imperdiet elit.
-
-\hypertarget{open-research}{%
-\section{Open research}\label{open-research}}
-
-Phasellus interdum tincidunt ex, a euismod massa pulvinar at. Ut
-fringilla ut nisi nec volutpat. Morbi imperdiet congue tincidunt.
-Vivamus eget rutrum purus. Etiam et pretium justo. Donec et egestas sem.
-Donec molestie ex sit amet viverra egestas. Nullam justo nulla,
-fringilla at iaculis in, posuere non mauris. Ut eget imperdiet elit.
+The research presented in this study is part of the Baltic Earth program
+(Earth System Science for the Baltic Sea region, see
+\href{https://www.baltic.earth/}{https://www.baltic.earth}.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-ashrafi2017fully}{}}%
-Ashrafi, M., Chua, L. H. C., Quek, C., \& Qin, X. (2017). A fully-online
-neuro-fuzzy model for flow forecasting in basins with limited data.
-\emph{Journal of Hydrology}, \emph{545}, 424--435.
-
-\leavevmode\vadjust pre{\hypertarget{ref-dee2011}{}}%
-Dee, D. P., Uppala, S. M., Simmons, A. J., Berrisford, P., Poli, P.,
-Kobayashi, S., et al. (2011). The ERA-Interim reanalysis: configuration
-and performance of the data assimilation system. \emph{Quarterly Journal
-of the Royal Meteorological Society}, \emph{137}(656), 553--597.
-\url{https://doi.org/10.1002/qj.828}
-
-\leavevmode\vadjust pre{\hypertarget{ref-fang2022application}{}}%
-Fang, L., \& Shao, D. (2022). Application of long short-term memory
-(LSTM) on the prediction of rainfall-runoff in karst area.
-\emph{Frontiers in Physics}, \emph{9}, 790687.
-
 \leavevmode\vadjust pre{\hypertarget{ref-fangExaminingApplicabilityDifferent2019}{}}%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-gruxf6ger2022}{}}%
-Gröger, M., Placke, M., Meier, M., Börgel, F., Brunnabend, S.-E.,
-Dutheil, C., et al. (2022). \emph{The Baltic Sea model inter-comparison
-project BMIP {\textendash} a platform for model development, evaluation,
-and uncertainty assessment}. Retrieved from
-\url{https://gmd.copernicus.org/preprints/gmd-2022-160/}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ha2021}{}}%
-Ha, S., Liu, D., \& Mu, L. (2021). Prediction of Yangtze River
-streamflow based on deep learning neural network with El
-Niño{\textendash}Southern Oscillation. \emph{Scientific Reports},
-\emph{11}(1), 11738. \url{https://doi.org/10.1038/s41598-021-90964-3}
-
 \leavevmode\vadjust pre{\hypertarget{ref-hagemannHighResolutionDischarge2020}{}}%
@@
 \url{https://doi.org/10.3389/feart.2020.00012}
-
-\leavevmode\vadjust pre{\hypertarget{ref-hochreiter1997long}{}}%
-Hochreiter, S., \& Schmidhuber, J. (1997). Long short-term memory.
-\emph{Neural Computation}, \emph{9}(8), 1735--1780.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hordoir2019}{}}%
-Hordoir, R., Axell, L., Höglund, A., Dieterich, C., Fransner, F.,
-Gröger, M., et al. (2019). Nemo-nordic 1.0: A NEMO-based ocean model for
-the baltic and north seas {\textendash} research and operational
-applications. \emph{Geoscientific Model Development}, \emph{12}(1),
-363--386. \url{https://doi.org/10.5194/gmd-12-363-2019}
-
-\leavevmode\vadjust pre{\hypertarget{ref-huang2014monthly}{}}%
-Huang, S., Chang, J., Huang, Q., \& Chen, Y. (2014). Monthly streamflow
-prediction using modified EMD-based support vector machine.
-\emph{Journal of Hydrology}, \emph{511}, 764--775.
-
-\leavevmode\vadjust pre{\hypertarget{ref-humphrey2016hybrid}{}}%
-Humphrey, G. B., Gibbs, M. S., Dandy, G. C., \& Maier, H. R. (2016). A
-hybrid approach to monthly streamflow forecasting: Integrating
-hydrological model outputs into a bayesian artificial neural network.
-\emph{Journal of Hydrology}, \emph{540}, 623--640.
-
-\leavevmode\vadjust pre{\hypertarget{ref-kratzert2018rainfall}{}}%
-Kratzert, F., Klotz, D., Brenner, C., Schulz, K., \& Herrnegger, M.
-(2018). Rainfall--runoff modelling using long short-term memory (LSTM)
-networks. \emph{Hydrology and Earth System Sciences}, \emph{22}(11),
-6005--6022.
-
-\leavevmode\vadjust pre{\hypertarget{ref-liu2020streamflow}{}}%
-Liu, D., Jiang, W., Mu, L., \& Wang, S. (2020). Streamflow prediction
-using deep learning neural network: Case study of yangtze river.
-\emph{IEEE Access}, \emph{8}, 90069--90086.
-
-\leavevmode\vadjust pre{\hypertarget{ref-meier2002simulated}{}}%
-Meier, H. M., \& Döscher, R. (2002). Simulated water and heat cycles of
-the baltic sea using a 3D coupled atmosphere--ice--ocean model.
-\emph{Boreal Environment Research}, \emph{7}(4), 327.
-
-\leavevmode\vadjust pre{\hypertarget{ref-moishin2021designing}{}}%
-Moishin, M., Deo, R. C., Prasad, R., Raj, N., \& Abdulla, S. (2021).
-Designing deep-based learning flood forecast model with ConvLSTM hybrid
-algorithm. \emph{IEEE Access}, \emph{9}, 50982--50993.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-theross}{}}%
-The rossby centre regional climate model RCA3: Model description and
-performance - tellus a: Dynamic meteorology and oceanography. (n.d.).
-Retrieved from
-\url{https://a.tellusjournals.se/articles/10.1111/j.1600-0870.2010.00478.x}
-
-\leavevmode\vadjust pre{\hypertarget{ref-zhu2023spatiotemporal}{}}%
-Zhu, S., Wei, J., Zhang, H., Xu, Y., \& Qin, H. (2023). Spatiotemporal
-deep learning rainfall-runoff forecasting combined with remote sensing
-precipitation products in large scale basins. \emph{Journal of
-Hydrology}, \emph{616}, 128727.
-
 \end{CSLReferences}

### 2024-06-11 paper/paper.qmd
Small corrections

--- 2024-05-15 paper/paper.qmd
@@
 LSTM networks, first introduced by @hochreiter1997long , are an evolution of the classical Recurrent Neural Networks (RNNs).
-Their structure enables them to learn long-term dependencies while avoiding the vanishing or exploding gradient problem.
+A significant advantage of this architecture is the memory cell's ability to retain gradients.
+This mechanism addresses the vanishing gradient problem, where, as input sequences elongate, the influence of initial stages becomes harder to capture, causing gradients of early input points to approach zero.
 They have shown stability and efficacy in sequence-to-sequence predictions.
+This performance in modeling long-range dependencies has been validated in various studies [@XXX].
+
 However, a limitation of LSTMs is their inability to effectively capture two-dimensional structures, an area where Convolutional Neural Networks (CNNs) excel.
@@
 ![**Combined ConvLSTM and FC network architecture**.
-The starting point is the continuous timeseries of input data $\{\vec{X}^t\}$, upper white block.
+The starting point is the continuous time series of input data $\{\vec{X}^t\}$, upper white block.
 From this series, a contiguous sequence of $N_\tau$ elements (yellow block) is used to feed a chain of $N_\tau$ connected ConvLSTM cells (light blue block) building the ConvLSTM network (grey block).
-The input seuqence is mapped via weighting matrices $\pmb{\mathcal{M}}^g$ (green blocks) onto gate vectors $\vec{g}^t[\tau]$.
+The input sequence is mapped via weighting matrices $\pmb{\mathcal{M}}^g$ (green blocks) onto gate vectors $\vec{g}^t[\tau]$.
 The gate vectors are then used to update the cell state $\vec{C}^t[\tau-1]$ and the hidden state $\vec{H}^t[\tau-1]$ of the last ConvLSTM cell to the current values $\vec{C}^t[\tau]$ and $\vec{H}^t[\tau]$, respectively.
@@
 The weighting matrices $\pmb{\mathcal{N}}^g$ (green blocks) control how much of the last hidden state enters the updated state.
-The final output of the ConvLSTM $\vec{H}^t[\tau]$ is then propagated to a FC network, which itself is a chain of three FC layers consisiting of weigthing matrices $\pmb{\mathcal{W}}$ and connected via ReLU functions, see @sec-FC_layer.
+The final output of the ConvLSTM $\vec{H}^t[\tau]\$\$ is then propagated to a FC network, which itself is a chain of three FC layers consisting of weighting matrices $\pmb{\mathcal{W}}$ and connected via ReLU functions, see @sec-FC_layer.
 The final result is then the river runoff $\vec{R}^t$ for all rivers considered at the current time instance $t$ (white block on the lower left).
-Note that all bias vectors are ommitted for the sake of clarity.
+Note that all bias vectors are omitted for the sake of clarity.
 See text for more information.](./ConvLSTM_FC.png){#fig-ConvLSTM_FC fig-align="center"}
@@
 
-![**Inner structure of a Long Short-Term Memory Cell**. See @fig-ConvLSTM_FC and text for information.](./LSTM.png){#fig-lstm fig-align="center"}
+![**Inner structure of a Long Short-Term Memory Cell**. See @fig-ConvLSTM_FC and text for information.](./LSTM.png){#fig-lstm fig-align="center" width="10.5cm"}
 
-This performance in modeling long-range dependencies has been validated in various studies [@XXX].
 The key component of the LSTM's innovation is its cell state, $\vec{C}^t[\tau] = (C^t_h[\tau]) \in \mathbb{R}^{N_h}$, which stores state information, also referred to as *long-term memory*.
 This state information complements the so-called hidden state $\vec{H}^t[\tau] = (H^t_h[\tau]) \in \mathbb{R}^{N_h}$ vector, which is also known from simpler neural network architectures.
-In case of the LSTM, the hidden state vector plays the role of the *short-term meory*.
-
-A significant advantage of this architecture is the memory cell's ability to retain gradients.
-This mechanism addresses the vanishing gradient problem, where, as input sequences elongate, the influence of initial stages becomes harder to capture, causing gradients of early input points to approach zero.
-The LSTM's activation function, inherently recurrent, mirrors the identity function with a consistent derivative of 1.0, ensuring the gradient remains stable throughout backpropagation.
-
+In case of the LSTM, the hidden state vector plays the role of the *short-term memory*.
 The cell state and the hidden state are vectors, where each element is associated with one of the $N_h$ hidden layers, labeled by $h$, which are internal, *artificial* degrees of freedom that enable the high adaptibility of neural networks.
@@
 
-where $g=i,f,o,c$ stands for the input, forget, output and cell-state gate, respectively and Eistein's summation convention is employed, i.e.
+where $g=i,f,o,c$ stands for the input, forget, output and cell-state gate, respectively and Einstein's summation convention is employed, i.e.
 indices that appear twice are summed over.

### 2024-06-11 paper/paper.qmd

--- 2024-06-07 paper/paper.qmd
@@
 
-where $g=i,f,o,c$ stands for the input, forget, output and cell-state gate, respectively and Einstein's summation convention is employed, i.e.
+where $g=i,f,o,c$ stands for the input, forget, output and cell-state gate, respectively and Eistein's summation convention is employed, i.e.
 indices that appear twice are summed over.
-The calligraphic symbols $\mathcal{M}^{g}_{hk}, \mathcal{N}^{g}_{hh'}$ and $\mathcal{B}^g_{h}$ are the free parameters of the network that are optimized for the given problem during the training, which is at the heart of any machine learning approach.
+The calligraphic symbols $\mathcal{M}^{g}_{hk}, \mathcal{N}^{g}_{hh'}$ and $\mathcal{B}^g_{h}$ are the free parameters of the network that are optimized for the given problem during the training, which is at the heart of the any machine learning approach.
 The matrix $\pmb{\mathcal{M}}^{g} = (\mathcal{M}^{g}_{hk}) \in \mathbb{R}^{N_h \times N_k}$ can be interpreted as a Markovian-like contribution of the current input $\vec{X}^t[\tau]$ to the gates, whereas the $\pmb{\mathcal{N}}^{g} = (\mathcal{N}^{g}_{hh'}) \in \mathbb{R}^{N_h \times N_h}$ scales a non-Markovian part determined by the hidden state of the last sequence point $\tau-1$.

### 2024-06-12 paper/paper.qmd
Worked on the overall text.
Open questions are embraced by `...`.
A lot of the citations are not yet in the bibliography.bib

@@
 It represents a complex forecasting tool that uses underlying physical processes.
-The second approach uses data-driven models that integrate statistical correction, using land surface schemes of global or regional climate models \[QUELLE\].
+The second approach uses data-driven models that integrate statistical correction, using land surface schemes of global or regional climate models \[`missing`\].
 
@@
 This mechanism addresses the vanishing gradient problem, where, as input sequences elongate, the influence of initial stages becomes harder to capture, causing gradients of early input points to approach zero.
-They have shown stability and efficacy in sequence-to-sequence predictions.
-This performance in modeling long-range dependencies has been validated in various studies [@XXX].
+LSTMs have shown stability and efficacy in sequence-to-sequence predictions.
+This performance in modeling long-range dependencies has been validated in various studies \[`missing`\].
 
 However, a limitation of LSTMs is their inability to effectively capture two-dimensional structures, an area where Convolutional Neural Networks (CNNs) excel.
-Recognizing this limitation @shiConvolutionalLSTMNetwork2015 proposed a Convolutional LSTM (ConvLSTM) architecture, which combines the strengths of both LSTM and CNN.
+Recognizing this limitation, @shiConvolutionalLSTMNetwork2015 proposed a Convolutional LSTM (ConvLSTM) architecture, which combines the strengths of both LSTM and CNN.
 The ConvLSTM network has been proven useful for spatio-temporal applications such as precipitation nowcasting [@shiConvolutionalLSTMNetwork2015], flood forecasting [@moishin2021designing], and river runoff forecasting [@ha2021; @zhu2023spatiotemporal].
@@
 In this work, we demonstrate that ConvLSTM networks are a reliable method for predicting multiple rivers simultaneously, using only atmospheric forcing, even in the absence of a fully functional hydrological model with a complex parameterization.
+Still, data of a suitable HD model is used as reference to train the network.
 We use the Baltic Sea catchment as an example to illustrate our approach.
@@
 
-In this work we will, we present a ConvLSTM architecture that is able to predict daily river runoff for 97 rivers across the Baltic Sea catchment. ***mehr Fokus auf Neues?***
+The main focus of this work is to present a ConvLSTM architecture that is able to predict daily river runoff for 97 rivers across the Baltic Sea catchment (@sec-methodology).
+To train the network, we use data from the E-HYPE HD model \[`missing`\] as reference output data and data from the UERRA project (Uncertainties in Ensembles of Regional Re-Analyses, [http://www.uerra.eu/)](http://www.uerra.eu/),) as atmospheric forcing (@sec-technical_details).
+The quality of the model is evaluated in @sec-sec-evaluation and subsequently the trained ConvLSTM is used in combination with the MOM5 ocean model for the Baltic Sea [@neumann2021] (@sec-MOM-application).
+The obtained results are further discussed in @sec-discussion and concluded in @sec-conclusions.
 
-## Implemented model architecture
+## Implemented model architecture {#sec-methodology}
 
@@
 
-We assume that the runoff at a specific point in time for all $N_r$ considered rivers, collected in the vector $\vec{R}^t \in \mathbb{R}^{N_r}$, can be accurately approximated by a functional $\vec{M}(\{X^t_k[x,y,\tau]\})$ of $k=1,\ldots N_k$ atmospheric fields $X^t_k[x,y,\tau]$ which are known for the past $\tau=1,\ldots N_\tau$ time instances.
+We assume that the runoff at a specific point in time $t$ for all $N_r$ considered rivers, collected in the vector $\vec{R}^t \in \mathbb{R}^{N_r}$, can be accurately approximated by a functional $\vec{M}(\{X^t_k[x,y,\tau]\})$ of $k=1,\ldots N_k$ atmospheric fields $X^t_k[x,y,\tau]$ which are known for the past $\tau=1,\ldots N_\tau$ time instances.
 This relationship is expressed as:
@@
 
-As an alternative, we propose that this functional can be adequately represented by a combination of a convolutional Long-Short Term Memory (ConvLSTM) model with a subsequent fully connected neural network.
+As an alternative, we propose that this functional can be adequately represented by a combination of a convolutional Long-Short Term Memory (ConvLSTM) model with a subsequent fully connected (FC) neural network.
 This approach eliminates the need for detailed knowledge of the involved processes and their modeling.
-Instead, these features can be "learned" by the network in an automated manner.
+Instead, these features can be "learned" by the network in an automated manner, i.e.
+all free parameters are optimized such that the network's output reproduces best the HD model data with given atmospheric input fields.
 Our proposed network architecture is visualized in @fig-ConvLSTM_FC and described in detail in the following sections.
@@
 The gate vectors are then used to update the cell state $\vec{C}^t[\tau-1]$ and the hidden state $\vec{H}^t[\tau-1]$ of the last ConvLSTM cell to the current values $\vec{C}^t[\tau]$ and $\vec{H}^t[\tau]$, respectively.
-The update is performed with LSTM core equations collectively described by the mapping L, see @eq-LSTM.
+The update is performed with the LSTM core equations collectively described by the mapping $L$, see @eq-LSTM.
 The weighting matrices $\pmb{\mathcal{N}}^g$ (green blocks) control how much of the last hidden state enters the updated state.
-The final output of the ConvLSTM $\vec{H}^t[\tau]\$\$ is then propagated to a FC network, which itself is a chain of three FC layers consisting of weighting matrices $\pmb{\mathcal{W}}$ and connected via ReLU functions, see @sec-FC_layer.
-The final result is then the river runoff $\vec{R}^t$ for all rivers considered at the current time instance $t$ (white block on the lower left).
+The final output of the ConvLSTM $\vec{H}^t[\tau]$ is then propagated to a FC network, which itself is a chain of three FC layers consisting of weighting matrices $\pmb{\mathcal{W}}$ and connected via ReLU functions, see @sec-FC_layer.
+The final result is the river runoff $\vec{R}^t$ for all rivers considered at the current time instance $t$ (white block on the lower left).
 Note that all bias vectors are omitted for the sake of clarity.
@@
 
-### ConvLSTM network
+### The ConvLSTM network
 
@@
 The key component of the LSTM's innovation is its cell state, $\vec{C}^t[\tau] = (C^t_h[\tau]) \in \mathbb{R}^{N_h}$, which stores state information, also referred to as *long-term memory*.
-This state information complements the so-called hidden state $\vec{H}^t[\tau] = (H^t_h[\tau]) \in \mathbb{R}^{N_h}$ vector, which is also known from simpler neural network architectures.
+This state information complements the so-called hidden state vector $\vec{H}^t[\tau] = (H^t_h[\tau]) \in \mathbb{R}^{N_h}$, which is also known from simpler neural network architectures.
 In case of the LSTM, the hidden state vector plays the role of the *short-term memory*.
-The cell state and the hidden state are vectors, where each element is associated with one of the $N_h$ hidden layers, labeled by $h$, which are internal, *artificial* degrees of freedom that enable the high adaptibility of neural networks.
+The cell state and the hidden state are vectors, where each element is associated with one of the $N_h$ hidden layers, labeled by $h$, which are internal, *artificial* degrees of freedom that enable the high adaptability of neural networks.
 These two state vectors are determined through several self-parameterized gates, all in the same vector space as $\vec{C}^t[\tau]$, see @fig-lstm for a visualization.
@@
 
-Note that this mapping is sometimes extended by a contribution to the $g_h^t[\tau]$ from the past cell state $\vec{C}^t[\tau-1]$ [@XXX].
+Note that this mapping is sometimes extended by a contribution to the $g_h^t[\tau]$ from the past cell state $\vec{C}^t[\tau-1]$ \[`missing`\].
 Nevertheless, this mechanism called "peeping" is not further considered in this work.
@@
 The final output of the ConvLSTM chain, $\vec{H}^t[N_\tau]$ and $\vec{C}^t[N_\tau]$, encode information on the full input sequence ending at time $t$.
-This information has to be decoded to obtain useful information via an appropriate subsequent network, as it is described in the @sec-FC_layer.
+This information has to be decoded via an appropriate subsequent network, as it is described in the @sec-FC_layer.
 
-#### Combining LSTM with spatial convolution
+#### Combining the LSTM with spatial convolution
 
-Although the plain LSTM has high performance in handling temporal sequences of point-like quantities it is not designed to recognize spatial features in sequences of, e.g., two-dimensional maps as atmospheric-ocean interface fields.
+Although the plain LSTM has high performance in handling temporal sequences of point-like quantities it is not designed to recognize spatial features in sequences of two-dimensional maps as atmosphere-ocean interface fields.
 To address this limitation we employ a ConvLSTM architecture as described in the following.
@@
 
-In summary, the ConvLSTM excels at processing tasks that demand a combined understanding of spatial patterns and temporal sequences in data.
+In summary, the ConvLSTM is designed for processing tasks that demand a combined understanding of spatial patterns and temporal sequences in data.
 It merges the image-processing capabilities of Convolutional Neural Networks (CNNs) with the time-series modeling of Long Short-Term Memory (LSTM) networks.
@@
 
-As stated in Sec.
-@sec-LSTM, the final output $\vec{H}^t[N_\tau]$ and $\vec{C}^t[N_\tau]$ of the ConvLSTM encode information on the full input sequence.
+As stated in @sec-LSTM, the final output $\vec{H}^t[N_\tau]$ and $\vec{C}^t[N_\tau]$ of the ConvLSTM encode information on the full input sequence.
 In order to contract this information to obtain the runoff vector $\vec{R}^t$ representing the $N_r$ rivers, we propose to subject the final short-term memory (i.e.
@@
 In particular, the dimensionality of the vector field $\vec{H}^t[N_\tau]$ is sequentially reduced by three nested FC layers, each connected to the other by the Rectified Linear Unit (ReLU), see @fig-ConvLSTM_FC.
-Integrating over artificial degrees of freedom in a step-wise fashion has turned out to be benificial \[@XXX\].
+Integrating out artificial degrees of freedom in a step-wise fashion has turned out to be beneficial \[`missing`\].
 
@@
 
-In matrix-vector notation this can be compactified to $$
+In matrix-vector notation this can be compressed to $$
 \begin{aligned}
@@
 \end{aligned}
-$$
+$$ {#eq-FC_layer}
 
-Combining the last equation with @eq-L_nested provides finally an explicit formula for the initial assumption of modelling the runoff for time $t$ as a functional of a sequence of atmospheric fields, i.e.
+Combining @eq-FC_layer with @eq-L_nested provides finally an explicit formula for the initial assumption of modelling the runoff for time $t$ as a functional of a sequence of atmospheric fields, i.e.
 
@@
 
-![Schematic structure of the ConvLSTM implementation for river runoff forecasting.](ConvLSTM.png){#fig-baltNet}
-
-For the computation we use the following set of hyper parameters:
+During training, the following set of hyper parameters has turned out to yield sufficient quality of the results for the present study:
 
@@
 | Num. timesteps       | 30                        |
-| Conv. Kernelsize     | (7,7)                     |
+| Conv. kernel size    | (7,7)                     |
 | Num. ConvLSTM layers | 1                         |
@@
 
-As input the model receives $N_{\tau}$ = 30 days of atmospheric surface fields temperature $T$, precipitation $P$, specific humidity $Q$ and wind speed $W$, with a daily resolution to predict the river runoff $\vec{R}$.
-The choice of atmospheric fields was based on the implemented river runoff calculation in the atmospheric model COSMO-CLM which uses these atmospheric fields to calculate an river runoff estimate.
+As input the model receives $N_{\tau}$ = 30 days of atmospheric surface fields temperature, precipitation, specific humidity and wind speed, with a daily resolution to predict the river runoff $\vec{R}$ (@fig-baltNet).
+
+![Schematic structure of the ConvLSTM implementation for river runoff forecasting.](ConvLSTM.png){#fig-baltNet}
+
+The choice of atmospheric fields was based on the assumption that the runoff should be mainly given by the net precipitation (precipitation minus evaporation).
+The evaporation flux is often calculated as a function of wind speed, the air's humidity, the air's density and the involved turbulent exchange coefficients [@karsten2024], where the latter two are influenced by the air temperature.
 
@@
 
+### ConvLSTM model evaluation {#sec-evaluation}
+
 The model was trained with daily data for the period 1979 to 2011, as this period represents the only period of E-HYPE without further bias correction applied to the runoff to match observations.
-The data was divided into randomly chosen splits of 80$\%$ training data, 10$\%$ validation data to evaluate the performance of the model during training, and 10$\%$ training data which is finally used to evaluate the performance of the model after training.
+The data was divided into randomly chosen splits of 80$\%$ training data, 10$\%$ validation data to evaluate the performance of the model during training, and 10$\%$ test data which is finally used to evaluate the performance of the model after training.
 The model was trained for 400 epochs and the model weights with the lowest mean squared error during training have been stored.
 
-The accuracy of the model is displayed in Figure @fig-statistical-evaluationNN.
-As mention above for evaluation, the test dataset was utilized.
-The left panel panel (Figure @fig-statistical-evaluationNN a) illustrates the relative prediction error in relation to the original E-HYPE data, indicating that, on daily timescales, the model can predict river runoff with an accuracy of $\pm$ 5$\%$.
-The overall correlation is 0.997 with the resulting error metrics yielding a RMSE of 323.99 $m^3$/s and MAE of 249.51 $m^3$/s.
+The accuracy of the model is displayed in @fig-statistical-evaluationNN.
+As mention above, for this evaluation, the model's output is compared to the test data, which, importantly, the model has not seen yet during the training.
+`Which river is shown here?` The left panel a) illustrates the relative prediction error in relation to the original E-HYPE data, indicating that, on daily timescales, the model can predict river runoff with an accuracy of $\pm$ 5$\%$.
+The overall correlation is 0.997 with the resulting error metrics yielding a root mean square error (RMSE) of 323.99 $m^3$/s and mean absolute error (MAE) of 249.51 $m^3$/s.
 While the model's performance is satisfactory, the discrepancies between the actual values and the predictions could partly be attributed to the use of a different atmospheric dataset than the one originally used to drive the E-HYPE model.
@@
 
-![](../src/figures/paper_error_metrics.png)
+![](../src/figures/paper_error_metrics.png){#fig-statistical-evaluationNN}
 
-In the following we will now address the overall performance of the total river runoff while also zooming in on the four largest rivers entering the Baltic Sea.
-\@fig-PerformanceNeuralNetworkRunoff shows the predicted and the original river runoff using the test dataset.
-The predicted total river runoff for the Baltic Sea is closely matching the original data (\@fig-PerformanceNeuralNetworkRunoff a).
-Zooming in on the largest individual rivers (\@fig-PerformanceNeuralNetworkRunoff b-e) it can be seen that that also the prediction of the individual rivers is close to the original data.
+`The time axis should be improved here and in all other time series plots.`
 
-![](../src/figures/paper_total_river_runoff.png)
+In the following the model's performance to reproduce the total river runoff as well as the inflow of the four largest rivers into the Baltic Sea is addressed.
+@fig-PerformanceNeuralNetworkRunoff shows the predicted and the original river runoff in comparison to the test dataset.
+The predicted total river runoff for the Baltic Sea is closely matching the original data, see panel a) therin.
+Zooming in on the largest individual rivers (panels b-e), it can be seen that also the prediction of the individual rivers is close to the original data.
 
-Lastly, we evaluated the performance of the runoff model by incorporating the predicted river runoff as runoff forcing into the ocean model MOM5.
-This provides a robust validation of the runoff model against more complex real world conditions.
-This allows us to ensure that the predictions accurately reflect the impact of the river discharge on the ocean dynamics, validating the temporal and spatial variability of the the river discharge.
-@fig-by15 shows the salinity comparison between the original E-HYPE river runoff and the predicted river runoff at BY15 - a central stations in the Baltic Sea.
-It can be seen that the model simulation using the predicted river runoff by the ConvLSTM is closely mirroring the control simulation.
-The upper panel (@fig-by15 a) shows the surface salinity, representing the high-frequency variations in the salinity, which is heavily affected by river runoff.
-The lower panel (@fig-by15 b) shows the bottom salinity which can be viewed as a low-pass filter in the Baltic Sea, which is also closely mirrored by the ConvLSTM predictions.
+![](../src/figures/paper_total_river_runoff.png){#fig-PerformanceNeuralNetworkRunoff}
 
-![](../src/figures/by15_model.png)
+### Application of the ConvLSTM in combination with an ocean model {#sec-MOM-application}
 
-The final evaluation of the ConvLSTM model concentrates on the spatial accuracy of river runoff predictions.
-Figure xxx a) shows the vertically averaged salinity for the period 1981 to 2011 in the reference simulation.
+Lastly, we evaluate the performance of the ConvLSTM by incorporating the predicted river runoff as hydrological forcing into the ocean model MOM5.
+This provides a robust validation of the runoff model against more complex real world conditions and is thereby ensuring that the predictions accurately reflect the impact of the river discharge on the ocean dynamics.
+This in turn validates that the temporal and spatial variability of river runoff is well captured by ConvLSTM and that the residuals shown in @fig-statistical-evaluationNN are indeed insignificant when it comes to realistic applications.
+
+@fig-by15 shows the salinity comparison between the original E-HYPE river runoff and the predicted river runoff at BY15 - a central station in the Baltic Sea, east of the Gotland island.
+It can be seen that the model simulation using the predicted river runoff by the ConvLSTM is closely mirroring the control simulation, that is forced with the E-HYPE runoff.
+The upper panel a) shows the surface salinity, representing the high-frequency variations in the salinity, which is heavily affected by river runoff.
+The lower panel b) shows the bottom salinity which in turn is low-frequency feature in the Baltic Sea that is also well reproduced with the ConvLSTM predictions.
+
+![](../src/figures/by15_model.png){#fig-by15}
+
+The final evaluation of the ConvLSTM model concentrates on the spatial accuracy of river runoff predictions as visualized in @fig-evaluation_MOM. 
+Panel a) therein exhibits the vertically averaged salinity for the period 1981 to 2011 in the reference simulation.
 It highlights the strong horizontal gradients and complex topographic features in the Baltic Sea, as evidenced by salinity variations in deeper waters, captured by the vertical integration.
-Figure 4b compares these results by showing the percentage difference in vertically averaged salinity between the ConvLSTM simulation and the reference simulation.
-Overall, the differences remain below 1$\%$, except near a river mouth in the Gulf of Riga (22-24°E, 56.5-58.5°N for orientation), where the difference is approximately 1$\%$.
+In panel b), these reference results are compared to the ConvLSTM simulation by showing the percentage difference in vertically averaged salinity.
+Overall, the differences remain below 1$\%$, except in the Gulf of Riga (22-24°E, 56.5-58.5°N for orientation), where the runoff is dominated by the Daugava river.
+Here the difference is approximately 1$\%$.
 
-![](../src/figures/evaluation_MOM.png)
+![](../src/figures/evaluation_MOM.png){#fig-evaluation_MOM}
 
-## Discussion
+## Discussion {#sec-discussion}
 
 With the increasing demand of decision makers for regional climate projections, that allow to quantify regional climate change impacts, the availability of precise hydrological forecasting becomes invaluable.
-The quality of a projection for a coastal sea such as the Baltic Sea is too large parts based on the quality of the hyrdological conditions.
+The quality of a projection for a coastal sea such as the Baltic Sea is too large parts based on the quality of the hydrological conditions.
 In this work we analyze the implementation of a ConvLSTM networks for predicting river runoff, highlighting its potential to enhance river runoff forecasting across different coastal seas.
@@
 
-The transition from traditional hydrological models to machine learning approaches, such as the ConvLSTM model, offers significant advantages.
+The transition from traditional hydrological models to machine learning approaches, such as the ConvLSTM model, offers significant advantages with respect to the following aspects.
 The ConvLSTM can be seamlessly integrated into regional climate models, allowing for the real-time computation of river runoff while performing climate projections.
@@
 
-While the ConvLSTM represents an advancement for the climate community, given that running and tuning traditional hydrological models demands extensive expertise, models like E-HYPE maintain an essential role.
-They provide a comprehensive dataset that helps to train our ConvLSTM model effectively.
-This robust training enables the machine learning models to achieve highly accurate predictive weights.
+However, while the ConvLSTM represents an advancement for the climate community, given that running and tuning traditional hydrological models demands extensive expertise, models like E-HYPE maintain an essential role.
+They provide a comprehensive dataset that enables to train our ConvLSTM model effectively.
+This robust training helps the machine learning models to achieve highly accurate predictive weights.
 Thus, rather than rendering traditional methods obsolete, the integration of machine learning models builds upon and enhances the foundational data provided by them.
 
-The robust performance of the ConvLSTM model in simulating river runoff and its possible effective integration into regional climate models pave the way for a multitude of new storyline simulations.
-Hence, we can now explore various "what-if" scenarios more reliably, under the assumption that the model weights attained during training are robust.
+## Conclusions {#sec-conclusions}
 
 In summary, we showed that the ConvLSTM model demonstrated robust performance in forecasting river runoff across 97 rivers entering the Baltic Sea.
-Trained on data from 1979 to 2011, the model achieved an impressive daily prediction accuracy of ±5$\%$.
+Trained on data from 1979 to 2011, the model achieved an impressive daily prediction accuracy of $\pm 5 \%$.
+Importantly, this accuracy is achieved without any expert's domain knowledge on hydrological modelling.
 This capability to generate accurate and detailed simulations enables to examine the potential impacts of different climate scenarios.
@@
 
-Ultimately, the integration of ConvLSTM into regional climate models represents a significant step forward in our ability to understand and predict the complex dynamics of river systems and their impact on regional climate systems.
+The robust performance of the ConvLSTM model in simulating river runoff and its possible effective integration into coupled regional climate models, as for example in the IOW ESM [@karsten2024], pave the way for a multitude of new storyline simulations.
+Hence, we can now explore various "what-if" scenarios more reliably, under the assumption that the model weights attained during training are robust.
+This represents a significant step forward in our ability to understand and predict the complex dynamics of river systems and their impact on regional climate systems.
 

### 2024-06-12 paper/paper.tex
Worked on the overall text.
Open questions are embraced by `...`.
A lot of the citations are not yet in the bibliography.bib

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, enhanced, boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
-
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+\section{Introduction}\label{introduction}
 
@@
 models that integrate statistical correction, using land surface schemes
-of global or regional climate models {[}QUELLE{]}.
+of global or regional climate models {[}\texttt{missing}{]}.
 
@@
 LSTM networks, first introduced by (\textbf{hochreiter1997long?}) , are
-an evolution of the classical Recurrent Neural Networks (RNNs). Their
-structure enables them to learn long-term dependencies while avoiding
-the vanishing or exploding gradient problem. They have shown stability
-and efficacy in sequence-to-sequence predictions. However, a limitation
-of LSTMs is their inability to effectively capture two-dimensional
-structures, an area where Convolutional Neural Networks (CNNs) excel.
-Recognizing this limitation SHI et al. (2015) proposed a Convolutional
-LSTM (ConvLSTM) architecture, which combines the strengths of both LSTM
-and CNN. The ConvLSTM network has been proven useful for spatio-temporal
-applications such as precipitation nowcasting (SHI et al., 2015), flood
-forecasting (\textbf{moishin2021designing?}), and river runoff
-forecasting (\textbf{ha2021?}; \textbf{zhu2023spatiotemporal?}).
+an evolution of the classical Recurrent Neural Networks (RNNs). A
+significant advantage of this architecture is the memory cell's ability
+to retain gradients. This mechanism addresses the vanishing gradient
+problem, where, as input sequences elongate, the influence of initial
+stages becomes harder to capture, causing gradients of early input
+points to approach zero. LSTMs have shown stability and efficacy in
+sequence-to-sequence predictions. This performance in modeling
+long-range dependencies has been validated in various studies
+{[}\texttt{missing}{]}.
+
+However, a limitation of LSTMs is their inability to effectively capture
+two-dimensional structures, an area where Convolutional Neural Networks
+(CNNs) excel. Recognizing this limitation, SHI et al. (2015) proposed a
+Convolutional LSTM (ConvLSTM) architecture, which combines the strengths
+of both LSTM and CNN. The ConvLSTM network has been proven useful for
+spatio-temporal applications such as precipitation nowcasting (SHI et
+al., 2015), flood forecasting (\textbf{moishin2021designing?}), and
+river runoff forecasting (\textbf{ha2021?};
+\textbf{zhu2023spatiotemporal?}).
 
@@
 atmospheric forcing, even in the absence of a fully functional
-hydrological model with a complex parameterization. We use the Baltic
-Sea catchment as an example to illustrate our approach. Although the
-methodology we propose is universally applicable across various
+hydrological model with a complex parameterization. Still, data of a
+suitable HD model is used as reference to train the network. We use the
+Baltic Sea catchment as an example to illustrate our approach. Although
+the methodology we propose is universally applicable across various
 geographic regions, the Baltic Sea represents a challenging region due
@@
 
-In this work we will, we present a ConvLSTM architecture that is able to
-predict daily river runoff for 97 rivers across the Baltic Sea
-catchment. \textbf{\emph{mehr Fokus auf Neues?}}
+The main focus of this work is to present a ConvLSTM architecture that
+is able to predict daily river runoff for 97 rivers across the Baltic
+Sea catchment (Section~\ref{sec-methodology}). To train the network, we
+use data from the E-HYPE HD model {[}\texttt{missing}{]} as reference
+output data and data from the UERRA project (Uncertainties in Ensembles
+of Regional Re-Analyses,
+\href{http://www.uerra.eu/}{http://www.uerra.eu/)},) as atmospheric
+forcing (Section~\ref{sec-technical_details}). The quality of the model
+is evaluated in \textbf{?@sec-sec-evaluation} and subsequently the
+trained ConvLSTM is used in combination with the MOM5 ocean model for
+the Baltic Sea (Neumann et al., 2021)
+(Section~\ref{sec-MOM-application}). The obtained results are further
+discussed in Section~\ref{sec-discussion} and concluded in
+Section~\ref{sec-conclusions}.
 
-\hypertarget{implemented-model-architecture}{%
-\section{Implemented model
-architecture}\label{implemented-model-architecture}}
+\section{Implemented model architecture}\label{sec-methodology}
 
-\hypertarget{sec-main_idea}{%
-\subsection{The main idea}\label{sec-main_idea}}
+\subsection{The main idea}\label{sec-main_idea}
 
-We assume that the runoff at a specific point in time for all \(N_r\)
-considered rivers, collected in the vector
+We assume that the runoff at a specific point in time \(t\) for all
+\(N_r\) considered rivers, collected in the vector
 \(\vec{R}^t \in \mathbb{R}^{N_r}\), can be accurately approximated by a
@@
 represented by a combination of a convolutional Long-Short Term Memory
-(ConvLSTM) model with a subsequent fully connected neural network. This
-approach eliminates the need for detailed knowledge of the involved
+(ConvLSTM) model with a subsequent fully connected (FC) neural network.
+This approach eliminates the need for detailed knowledge of the involved
 processes and their modeling. Instead, these features can be ``learned''
-by the network in an automated manner. Our proposed network architecture
-is visualized in Figure~\ref{fig-ConvLSTM_FC} and described in detail in
-the following sections. To provide an overview, we will discuss the main
-components of this architecture one-by-one.
+by the network in an automated manner, i.e.~all free parameters are
+optimized such that the network's output reproduces best the HD model
+data with given atmospheric input fields. Our proposed network
+architecture is visualized in Figure~\ref{fig-ConvLSTM_FC} and described
+in detail in the following sections. To provide an overview, we will
+discuss the main components of this architecture one-by-one.
 
@@
 
-{\centering \includegraphics{./ConvLSTM_FC.png}
+\centering{
+
+\includegraphics{./ConvLSTM_FC.png}
 
@@
 \caption{\label{fig-ConvLSTM_FC}\textbf{Combined ConvLSTM and FC network
-architecture}. The starting point is the continuous timeseries of input
+architecture}. The starting point is the continuous time series of input
 data \(\{\vec{X}^t\}\), upper white block. From this series, a
@@
 feed a chain of \(N_\tau\) connected ConvLSTM cells (light blue block)
-building the ConvLSTM network (grey block). The input seuqence is mapped
+building the ConvLSTM network (grey block). The input sequence is mapped
 via weighting matrices \(\pmb{\mathcal{M}}^g\) (green blocks) onto gate
@@
 \(\vec{C}^t[\tau]\) and \(\vec{H}^t[\tau]\), respectively. The update is
-performed with LSTM core equations collectively described by the mapping
-L, see Equation~\ref{eq-LSTM}. The weighting matrices
+performed with the LSTM core equations collectively described by the
+mapping \(L\), see Equation~\ref{eq-LSTM}. The weighting matrices
 \(\pmb{\mathcal{N}}^g\) (green blocks) control how much of the last
@@
 \(\vec{H}^t[\tau]\) is then propagated to a FC network, which itself is
-a chain of three FC layers consisiting of weigthing matrices
+a chain of three FC layers consisting of weighting matrices
 \(\pmb{\mathcal{W}}\) and connected via ReLU functions, see
-Section~\ref{sec-FC_layer}. The final result is then the river runoff
+Section~\ref{sec-FC_layer}. The final result is the river runoff
 \(\vec{R}^t\) for all rivers considered at the current time instance
 \(t\) (white block on the lower left). Note that all bias vectors are
-ommitted for the sake of clarity. See text for more information.}
+omitted for the sake of clarity. See text for more information.}
 
-\end{figure}
+\end{figure}%
 
-\hypertarget{convlstm-network}{%
-\subsection{ConvLSTM network}\label{convlstm-network}}
+\subsection{The ConvLSTM network}\label{the-convlstm-network}
 
-\hypertarget{sec-LSTM}{%
-\subsubsection{The LSTM approach}\label{sec-LSTM}}
+\subsubsection{The LSTM approach}\label{sec-LSTM}
 
@@
 
-{\centering \includegraphics{./LSTM.png}
+\centering{
+
+\includegraphics[width=10.5cm,height=\textheight]{./LSTM.png}
 
@@
 
-\end{figure}
+\end{figure}%
 
-This performance in modeling long-range dependencies has been validated
-in various studies (\textbf{XXX?}). The key component of the LSTM's
-innovation is its cell state,
+The key component of the LSTM's innovation is its cell state,
 \(\vec{C}^t[\tau] = (C^t_h[\tau]) \in \mathbb{R}^{N_h}\), which stores
 state information, also referred to as \emph{long-term memory}. This
-state information complements the so-called hidden state
-\(\vec{H}^t[\tau] = (H^t_h[\tau]) \in \mathbb{R}^{N_h}\) vector, which
-is also known from simpler neural network architectures. In case of the
-LSTM, the hidden state vector plays the role of the \emph{short-term
-meory}.
-
-A significant advantage of this architecture is the memory cell's
-ability to retain gradients. This mechanism addresses the vanishing
-gradient problem, where, as input sequences elongate, the influence of
-initial stages becomes harder to capture, causing gradients of early
-input points to approach zero. The LSTM's activation function,
-inherently recurrent, mirrors the identity function with a consistent
-derivative of 1.0, ensuring the gradient remains stable throughout
-backpropagation.
-
+state information complements the so-called hidden state vector
+\(\vec{H}^t[\tau] = (H^t_h[\tau]) \in \mathbb{R}^{N_h}\), which is also
+known from simpler neural network architectures. In case of the LSTM,
+the hidden state vector plays the role of the \emph{short-term memory}.
 The cell state and the hidden state are vectors, where each element is
@@
 which are internal, \emph{artificial} degrees of freedom that enable the
-high adaptibility of neural networks. These two state vectors are
+high adaptability of neural networks. These two state vectors are
 determined through several self-parameterized gates, all in the same
@@
 optimized for the given problem during the training, which is at the
-heart of any machine learning approach. The matrix
+heart of the any machine learning approach. The matrix
 \(\pmb{\mathcal{M}}^{g} = (\mathcal{M}^{g}_{hk}) \in \mathbb{R}^{N_h \times N_k}\)
@@
 \(g_h^t[\tau]\) from the past cell state \(\vec{C}^t[\tau-1]\)
-(\textbf{XXX?}). Nevertheless, this mechanism called ``peeping'' is not
-further considered in this work.
+{[}\texttt{missing}{]}. Nevertheless, this mechanism called ``peeping''
+is not further considered in this work.
 
@@
 
-\begin{equation}\protect\hypertarget{eq-L_nested}{}{
+\begin{equation}\phantomsection\label{eq-L_nested}{
 \left(\vec{H}^t[N_\tau], \vec{C}^t[N_\tau] \right) = L \left( \vec{X}^t[N_\tau], L \left( \vec{X}^t[N_\tau-1], \ldots L \left( \vec{X}^t[1], (\vec{H}^t[0], \vec{C}^t[0]) \right) \ldots \right) \right) \ ,
-}\label{eq-L_nested}\end{equation}
+}\end{equation}
 
@@
 \(\vec{C}^t[N_\tau]\), encode information on the full input sequence
-ending at time \(t\). This information has to be decoded to obtain
-useful information via an appropriate subsequent network, as it is
-described in the Section~\ref{sec-FC_layer}.
+ending at time \(t\). This information has to be decoded via an
+appropriate subsequent network, as it is described in the
+Section~\ref{sec-FC_layer}.
 
-\hypertarget{combining-lstm-with-spatial-convolution}{%
-\subsubsection{Combining LSTM with spatial
-convolution}\label{combining-lstm-with-spatial-convolution}}
+\subsubsection{Combining the LSTM with spatial
+convolution}\label{combining-the-lstm-with-spatial-convolution}
 
@@
 sequences of point-like quantities it is not designed to recognize
-spatial features in sequences of, e.g., two-dimensional maps as
-atmospheric-ocean interface fields. To address this limitation we employ
+spatial features in sequences of two-dimensional maps as
+atmosphere-ocean interface fields. To address this limitation we employ
 a ConvLSTM architecture as described in the following.
@@
 
-In summary, the ConvLSTM excels at processing tasks that demand a
+In summary, the ConvLSTM is designed for processing tasks that demand a
 combined understanding of spatial patterns and temporal sequences in
@@
 
-\hypertarget{sec-FC_layer}{%
-\subsection{Fully connected layer}\label{sec-FC_layer}}
+\subsection{Fully connected layer}\label{sec-FC_layer}
 
-As stated in Sec. Section~\ref{sec-LSTM}, the final output
+As stated in Section~\ref{sec-LSTM}, the final output
 \(\vec{H}^t[N_\tau]\) and \(\vec{C}^t[N_\tau]\) of the ConvLSTM encode
@@
 each connected to the other by the Rectified Linear Unit (ReLU), see
-Figure~\ref{fig-ConvLSTM_FC}. Integrating over artificial degrees of
-freedom in a step-wise fashion has turned out to be benificial
-{[}(\textbf{XXX?}){]}.
+Figure~\ref{fig-ConvLSTM_FC}. Integrating out artificial degrees of
+freedom in a step-wise fashion has turned out to be beneficial
+{[}\texttt{missing}{]}.
 
@@
 
-In matrix-vector notation this can be compactified to \[
+In matrix-vector notation this can be compressed to
+\begin{equation}\phantomsection\label{eq-FC_layer}{
 \begin{aligned}
@@
 \end{aligned}
-\]
+}\end{equation}
 
-Combining the last equation with Equation~\ref{eq-L_nested} provides
-finally an explicit formula for the initial assumption of modelling the
-runoff for time \(t\) as a functional of a sequence of atmospheric
-fields, i.e.
+Combining Equation~\ref{eq-FC_layer} with Equation~\ref{eq-L_nested}
+provides finally an explicit formula for the initial assumption of
+modelling the runoff for time \(t\) as a functional of a sequence of
+atmospheric fields, i.e.
 
@@
 
-\hypertarget{sec-technical_details}{%
-\section{Technical details}\label{sec-technical_details}}
+\section{Technical details}\label{sec-technical_details}
 
-\hypertarget{runoff-data-used-for-training}{%
 \subsection{Runoff data used for
-training}\label{runoff-data-used-for-training}}
+training}\label{runoff-data-used-for-training}
 
@@
 
-\hypertarget{neural-network-hyper-parameters}{%
 \subsection{Neural network hyper
-parameters}\label{neural-network-hyper-parameters}}
+parameters}\label{neural-network-hyper-parameters}
+
+During training, the following set of hyper parameters has turned out to
+yield sufficient quality of the results for the present study:
+
+\begin{longtable}[]{@{}ll@{}}
+\caption{Hyperparameters}\label{tbl-letters}\tabularnewline
+\toprule\noalign{}
+Parameter name & Parameter size \\
+\midrule\noalign{}
+\endfirsthead
+\toprule\noalign{}
+Parameter name & Parameter size \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+Channel size & 4 \\
+Num. hidden layer & 9 \\
+Num. timesteps & 30 \\
+Conv. kernel size & (7,7) \\
+Num. ConvLSTM layers & 1 \\
+Batch size & 50 \\
+Learning Rate & 1e-3 with CosineAnnealing \\
+\end{longtable}
+
+As input the model receives \(N_{\tau}\) = 30 days of atmospheric
+surface fields temperature, precipitation, specific humidity and wind
+speed, with a daily resolution to predict the river runoff \(\vec{R}\)
+(Figure~\ref{fig-baltNet}).
 
@@
 
-{\centering \includegraphics{ConvLSTM.png}
+\centering{
+
+\includegraphics{ConvLSTM.png}
 
@@
 
-\end{figure}
+\end{figure}%
 
-For the computation we use the following set of hyper parameters:
+The choice of atmospheric fields was based on the assumption that the
+runoff should be mainly given by the net precipitation (precipitation
+minus evaporation). The evaporation flux is often calculated as a
+function of wind speed, the air's humidity, the air's density and the
+involved turbulent exchange coefficients (Karsten et al., 2024), where
+the latter two are influenced by the air temperature.
 
-\hypertarget{tbl-letters}{}
-\begin{longtable}[]{@{}ll@{}}
-\caption{\label{tbl-letters}Hyperparameters}\tabularnewline
-\toprule()
-Parameter name & Parameter size \\
-\midrule()
-\endfirsthead
-\toprule()
-Parameter name & Parameter size \\
-\midrule()
-\endhead
-Channel size & 4 \\
-Num. hidden layer & 9 \\
-Num. timesteps & 30 \\
-Conv. Kernelsize & (7,7) \\
-Num. ConvLSTM layers & 1 \\
-Batch size & 50 \\
-Learning Rate & 1e-3 with CosineAnnealing \\
-\bottomrule()
-\end{longtable}
+\section{Results}\label{results}
 
-As input the model receives \(N_{\tau}\) = 30 days of atmospheric
-surface fields temperature \(T\), precipitation \(P\), specific humidity
-\(Q\) and wind speed \(W\), with a daily resolution to predict the river
-runoff \(\vec{R}\). The choice of atmospheric fields was based on the
-implemented river runoff calculation in the atmospheric model COSMO-CLM
-which uses these atmospheric fields to calculate an river runoff
-estimate.
-
-\hypertarget{results}{%
-\section{Results}\label{results}}
+\subsection{ConvLSTM model evaluation}\label{sec-evaluation}
 
@@
 validation data to evaluate the performance of the model during
-training, and 10\(\%\) training data which is finally used to evaluate
-the performance of the model after training. The model was trained for
-400 epochs and the model weights with the lowest mean squared error
-during training have been stored.
+training, and 10\(\%\) test data which is finally used to evaluate the
+performance of the model after training. The model was trained for 400
+epochs and the model weights with the lowest mean squared error during
+training have been stored.
 
-The accuracy of the model is displayed in Figure
-\textbf{?@fig-statistical-evaluationNN}. As mention above for
-evaluation, the test dataset was utilized. The left panel panel (Figure
-\textbf{?@fig-statistical-evaluationNN} a) illustrates the relative
-prediction error in relation to the original E-HYPE data, indicating
-that, on daily timescales, the model can predict river runoff with an
-accuracy of \(\pm\) 5\(\%\). The overall correlation is 0.997 with the
-resulting error metrics yielding a RMSE of 323.99 \(m^3\)/s and MAE of
-249.51 \(m^3\)/s. While the model's performance is satisfactory, the
+The accuracy of the model is displayed in
+Figure~\ref{fig-statistical-evaluationNN}. As mention above, for this
+evaluation, the model's output is compared to the test data, which,
+importantly, the model has not seen yet during the training.
+\texttt{Which\ river\ is\ shown\ here?} The left panel a) illustrates
+the relative prediction error in relation to the original E-HYPE data,
+indicating that, on daily timescales, the model can predict river runoff
+with an accuracy of \(\pm\) 5\(\%\). The overall correlation is 0.997
+with the resulting error metrics yielding a root mean square error
+(RMSE) of 323.99 \(m^3\)/s and mean absolute error (MAE) of 249.51
+\(m^3\)/s. While the model's performance is satisfactory, the
 discrepancies between the actual values and the predictions could partly
@@
 
-\includegraphics{../src/figures/paper_error_metrics.png}
+\begin{figure}
 
-In the following we will now address the overall performance of the
-total river runoff while also zooming in on the four largest rivers
-entering the Baltic Sea. @fig-PerformanceNeuralNetworkRunoff shows the
-predicted and the original river runoff using the test dataset. The
-predicted total river runoff for the Baltic Sea is closely matching the
-original data (@fig-PerformanceNeuralNetworkRunoff a). Zooming in on the
-largest individual rivers (@fig-PerformanceNeuralNetworkRunoff b-e) it
-can be seen that that also the prediction of the individual rivers is
-close to the original data.
+\centering{
 
-\includegraphics{../src/figures/paper_total_river_runoff.png}
+\captionsetup{labelsep=none}\includegraphics{../src/figures/paper_error_metrics.png}
 
-Lastly, we evaluated the performance of the runoff model by
-incorporating the predicted river runoff as runoff forcing into the
-ocean model MOM5. This provides a robust validation of the runoff model
-against more complex real world conditions. This allows us to ensure
-that the predictions accurately reflect the impact of the river
-discharge on the ocean dynamics, validating the temporal and spatial
-variability of the the river discharge. \textbf{?@fig-by15} shows the
-salinity comparison between the original E-HYPE river runoff and the
-predicted river runoff at BY15 - a central stations in the Baltic Sea.
-It can be seen that the model simulation using the predicted river
-runoff by the ConvLSTM is closely mirroring the control simulation. The
-upper panel (\textbf{?@fig-by15} a) shows the surface salinity,
+}
+
+\caption{\label{fig-statistical-evaluationNN}}
+
+\end{figure}%
+
+\texttt{The\ time\ axis\ should\ be\ improved\ here\ and\ in\ all\ other\ time\ series\ plots.}
+
+In the following the model's performance to reproduce the total river
+runoff as well as the inflow of the four largest rivers into the Baltic
+Sea is addressed. Figure~\ref{fig-PerformanceNeuralNetworkRunoff} shows
+the predicted and the original river runoff in comparison to the test
+dataset. The predicted total river runoff for the Baltic Sea is closely
+matching the original data, see panel a) therin. Zooming in on the
+largest individual rivers (panels b-e), it can be seen that also the
+prediction of the individual rivers is close to the original data.
+
+\begin{figure}
+
+\centering{
+
+\captionsetup{labelsep=none}\includegraphics{../src/figures/paper_total_river_runoff.png}
+
+}
+
+\caption{\label{fig-PerformanceNeuralNetworkRunoff}}
+
+\end{figure}%
+
+\subsection{Application of the ConvLSTM in combination with an ocean
+model}\label{sec-MOM-application}
+
+Lastly, we evaluate the performance of the ConvLSTM by incorporating the
+predicted river runoff as hydrological forcing into the ocean model
+MOM5. This provides a robust validation of the runoff model against more
+complex real world conditions and is thereby ensuring that the
+predictions accurately reflect the impact of the river discharge on the
+ocean dynamics. This in turn validates that the temporal and spatial
+variability of river runoff is well captured by ConvLSTM and that the
+residuals shown in Figure~\ref{fig-statistical-evaluationNN} are indeed
+insignificant when it comes to realistic applications.
+
+Figure~\ref{fig-by15} shows the salinity comparison between the original
+E-HYPE river runoff and the predicted river runoff at BY15 - a central
+station in the Baltic Sea, east of the Gotland island. It can be seen
+that the model simulation using the predicted river runoff by the
+ConvLSTM is closely mirroring the control simulation, that is forced
+with the E-HYPE runoff. The upper panel a) shows the surface salinity,
 representing the high-frequency variations in the salinity, which is
-heavily affected by river runoff. The lower panel (\textbf{?@fig-by15}
-b) shows the bottom salinity which can be viewed as a low-pass filter in
-the Baltic Sea, which is also closely mirrored by the ConvLSTM
-predictions.
+heavily affected by river runoff. The lower panel b) shows the bottom
+salinity which in turn is low-frequency feature in the Baltic Sea that
+is also well reproduced with the ConvLSTM predictions.
 
-\includegraphics{../src/figures/by15_model.png}
+\begin{figure}
+
+\centering{
+
+\captionsetup{labelsep=none}\includegraphics{../src/figures/by15_model.png}
+
+}
+
+\caption{\label{fig-by15}}
+
+\end{figure}%
 
 The final evaluation of the ConvLSTM model concentrates on the spatial
-accuracy of river runoff predictions. Figure xxx a) shows the vertically
-averaged salinity for the period 1981 to 2011 in the reference
-simulation. It highlights the strong horizontal gradients and complex
-topographic features in the Baltic Sea, as evidenced by salinity
-variations in deeper waters, captured by the vertical integration.
-Figure 4b compares these results by showing the percentage difference in
-vertically averaged salinity between the ConvLSTM simulation and the
-reference simulation. Overall, the differences remain below 1\(\%\),
-except near a river mouth in the Gulf of Riga (22-24°E, 56.5-58.5°N for
-orientation), where the difference is approximately 1\(\%\).
+accuracy of river runoff predictions as visualized in
+Figure~\ref{fig-evaluation_MOM}. Panel a) therein exhibits the
+vertically averaged salinity for the period 1981 to 2011 in the
+reference simulation. It highlights the strong horizontal gradients and
+complex topographic features in the Baltic Sea, as evidenced by salinity
+variations in deeper waters, captured by the vertical integration. In
+panel b), these reference results are compared to the ConvLSTM
+simulation by showing the percentage difference in vertically averaged
+salinity. Overall, the differences remain below 1\(\%\), except in the
+Gulf of Riga (22-24°E, 56.5-58.5°N for orientation), where the runoff is
+dominated by the Daugava river. Here the difference is approximately
+1\(\%\).
 
-\includegraphics{../src/figures/evaluation_MOM.png}
+\begin{figure}
 
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
+\centering{
+
+\captionsetup{labelsep=none}\includegraphics{../src/figures/evaluation_MOM.png}
+
+}
+
+\caption{\label{fig-evaluation_MOM}}
+
+\end{figure}%
+
+\section{Discussion}\label{sec-discussion}
 
@@
 quality of a projection for a coastal sea such as the Baltic Sea is too
-large parts based on the quality of the hyrdological conditions. In this
+large parts based on the quality of the hydrological conditions. In this
 work we analyze the implementation of a ConvLSTM networks for predicting
@@
 The transition from traditional hydrological models to machine learning
-approaches, such as the ConvLSTM model, offers significant advantages.
-The ConvLSTM can be seamlessly integrated into regional climate models,
-allowing for the real-time computation of river runoff while performing
-climate projections. While the initial training of the model requires
-substantial computational resources, it remains less intensive than
-running comprehensive hydrological models. Furthermore, once trained,
-the ConvLSTM model is computationally efficient during inference,
-ensuring enhanced forecasting capabilities without significantly
-increasing computational demands.
+approaches, such as the ConvLSTM model, offers significant advantages
+with respect to the following aspects. The ConvLSTM can be seamlessly
+integrated into regional climate models, allowing for the real-time
+computation of river runoff while performing climate projections. While
+the initial training of the model requires substantial computational
+resources, it remains less intensive than running comprehensive
+hydrological models. Furthermore, once trained, the ConvLSTM model is
+computationally efficient during inference, ensuring enhanced
+forecasting capabilities without significantly increasing computational
+demands.
 
-While the ConvLSTM represents an advancement for the climate community,
-given that running and tuning traditional hydrological models demands
-extensive expertise, models like E-HYPE maintain an essential role. They
-provide a comprehensive dataset that helps to train our ConvLSTM model
-effectively. This robust training enables the machine learning models to
-achieve highly accurate predictive weights. Thus, rather than rendering
-traditional methods obsolete, the integration of machine learning models
-builds upon and enhances the foundational data provided by them.
+However, while the ConvLSTM represents an advancement for the climate
+community, given that running and tuning traditional hydrological models
+demands extensive expertise, models like E-HYPE maintain an essential
+role. They provide a comprehensive dataset that enables to train our
+ConvLSTM model effectively. This robust training helps the machine
+learning models to achieve highly accurate predictive weights. Thus,
+rather than rendering traditional methods obsolete, the integration of
+machine learning models builds upon and enhances the foundational data
+provided by them.
 
-The robust performance of the ConvLSTM model in simulating river runoff
-and its possible effective integration into regional climate models pave
-the way for a multitude of new storyline simulations. Hence, we can now
-explore various ``what-if'' scenarios more reliably, under the
-assumption that the model weights attained during training are robust.
+\section{Conclusions}\label{sec-conclusions}
 
@@
 Baltic Sea. Trained on data from 1979 to 2011, the model achieved an
-impressive daily prediction accuracy of ±5\(\%\). This capability to
-generate accurate and detailed simulations enables to examine the
-potential impacts of different climate scenarios. Such precision in
-forecasting and scenario testing is crucial for crafting effective water
-resource management strategies and adapting to the changing climate.
+impressive daily prediction accuracy of \(\pm 5 \%\). Importantly, this
+accuracy is achieved without any expert's domain knowledge on
+hydrological modelling. This capability to generate accurate and
+detailed simulations enables to examine the potential impacts of
+different climate scenarios. Such precision in forecasting and scenario
+testing is crucial for crafting effective water resource management
+strategies and adapting to the changing climate.
 
-Ultimately, the integration of ConvLSTM into regional climate models
-represents a significant step forward in our ability to understand and
-predict the complex dynamics of river systems and their impact on
+The robust performance of the ConvLSTM model in simulating river runoff
+and its possible effective integration into coupled regional climate
+models, as for example in the IOW ESM (Karsten et al., 2024), pave the
+way for a multitude of new storyline simulations. Hence, we can now
+explore various ``what-if'' scenarios more reliably, under the
+assumption that the model weights attained during training are robust.
+This represents a significant step forward in our ability to understand
+and predict the complex dynamics of river systems and their impact on
 regional climate systems.
 
-\hypertarget{acknowledgments}{%
-\section{Acknowledgments}\label{acknowledgments}}
+\section{Acknowledgments}\label{acknowledgments}
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-shiConvolutionalLSTMNetwork2015}{}}%
+\bibitem[\citeproctext]{ref-karsten2024}
+Karsten, S., Radtke, H., Gröger, M., Ho-Hagemann, H. T. M., Mashayekh,
+H., Neumann, T., \& Meier, H. E. M. (2024). Flux coupling approach on an
+exchange grid for the IOW earth system model (version 1.04.00) of the
+baltic sea region. \emph{Geoscientific Model Development}, \emph{17}(4),
+1689--1708. \url{https://doi.org/10.5194/gmd-17-1689-2024}
+
+\bibitem[\citeproctext]{ref-neumann2021}
+Neumann, T., Koponen, S., Attila, J., Brockmann, C., Kallio, K.,
+Kervinen, M., et al. (2021). Optical model for the baltic sea with an
+explicit CDOM state variable: A case study with model ERGOM (version
+1.2). \emph{Geoscientific Model Development}, \emph{14}(8), 5049--5062.
+\url{https://doi.org/10.5194/gmd-14-5049-2021}
+
+\bibitem[\citeproctext]{ref-shiConvolutionalLSTMNetwork2015}
 SHI, X., Chen, Z., Wang, H., Yeung, D.-Y., Wong, W., \& WOO, W. (2015).
